### 2022-01-04 chapters/conclusions/conclusions_david.tex
creation of thesis git

%\linespread{1.5}\selectfont

The astronomers' interest for Titan was raised by the observations of Voyager spacecrafts. Further explorations were conducted by means of Earth-based observation tools and the successful Cassini mission provided plentiful data whose analysis is an ongoing process. While this mission answered a variety of questions, crucial ones remain open. To this end, various exploration missions have been designed. One of them, Dragonfly \citep{lorenz2018dragonfly}, was recently selected by the NASA as part of its New Frontiers program while others are still under development and are proposed to several spatial agencies \citep{mitri2014exploration,hartwig2016exploring}. Despite the numerous projects, new \textit{in situ} data are not expected before at least 14 years. Indeed, Dragonfly, which is the most advanced project, is expected to be launched in 2026 and will land in 2034. Until then, Earth-based observations, data analysis by means of enhanced algorithms, and numerical modelling are the only tools able to provide new pieces of information about Titan.

\section*{Exploration of Titan's liquid bodies}

Titan is the only celestial body of the solar system, besides the Earth, housing significant surface liquid bodies. The latter played a key role in the apparition of life on our planet. It was thus instinctive to pay interest to the surface lakes and seas of Titan although the latter are filled with liquid hydrocarbons instead of water. The oceanography of Titan is still in its early stages due to the lack of data and observations, which leads to large uncertainties in the models predictions and prevents the scientists from validating their model(s) by comparison with \textit{in situ} measurements. Unfortunately, pieces of information only available through numerical simulations are requested to design an exploration mission focusing on the surface lakes and seas. For instance, an \textit{a priori} knowledge of the flow is a key feature in the design of an exploration mission. It is a chicken-and-egg problem. To circumvent this paradox, one has to proceed in stages. The first step consists in running numerical simulations of the flow in the surface lakes and seas using the scarce data available. Depending on the landing craft, different aspects of the liquid motion need to be known. For instance, the maximum flow speed is needed to design the engine of a motorised capsule while the orientation of the flow is useful to predict the path of a non-motorised capsule as a function of time and landing site and to select the latter such that the capsule does not beach within a short amount of time. The variation of the sea/lake surface elevation is a key parameter in order to ensure the capsule is able to be refloated from the beach, if it beaches on purpose at a given location. The influence of the poorly constrained parameters has to be assessed by means of sensitivity analyses. The results can then be used to design the selected operation mission taking into account the worst case scenario predicted by the sensitivity analyses. The data collected by this exploration mission could be used to validate more accurately the models and improve them as well as our understanding of the physical mechanisms at play. The results of the improved models could then be used to help analysing the data collected or to design more complex exploration missions.

While there are proposals for exploration missions focusing on the surface lakes and seas, such missions are not expected to take place before years. Indeed, the surface liquid bodies are located in the high latitudes, mostly in the northern hemisphere. The last spring equinox in this hemisphere occurred on August 11, 2009. A year on Titan lasting about 29.5 years on Earth, the autumn equinox in the northern hemisphere will occur before the next exploration mission could reach Titan (which is expected to be in 2034). Direct to Earth communication with a lander being impossible at high latitudes during the winter (without an orbiter acting as a relay antenna), it was decided that the next exploration mission, DragonFly, would land in the equatorial region, as it does not include an orbiter. The main drawback of a single lander is that it explores a single location, which can be circumvented if the lander is mobile. Among the concepts presented in the literature, \cite{lorenz2000post} proposed a helicopter but, a helicopter being mechanically complex, this idea was not very successful. Fortunately, technology developments in the last decades led to the advent of drones, which are much better suited for spatial exploration. The Dragonfly mission, which consists in sending a rotorcraft on Titan was recently selected by the NASA as part of its "New Frontiers" program. While it was first imagined granting the rotorcraft with amphibian capabilities, this idea was not pursued and it was decided to focus on operations on dry land \citep{lorenz2018dragonfly}. Dragonfly is expected to land at a location and season similar to the Huygens descent in 2005 to take advantage of the data collected by the probe. Once the drone is released from the entry system and parachute, it will take advantage of its mobility to scan the area and select the more favourable landing site. After Dragonfly safely landed, it will carry on small displacements and then flights of increasing range, duration, and height. The mission aims at improving our knowledge of the organic and methanogenic cycle on Titan and investigating the subsurface ocean and/or liquid reservoirs, particularly their evolution and interactions with the surface \citep{lorenz2018dragonfly}. To this end, Dragon fly will \citep{turtle2017dragonfly}
\begin{itemize}
\item Identify the chemical components and processes producing biologically relevant compounds by sampling surface material and conducting analyses with a mass spectrometer;
\item Analyse bulk elemental surface composition by means of a neutron-activated gamma-ray spectrometer;
\item Provide measurement of the atmospheric and surface conditions as well as their diurnal and spatial variations with meteorology sensors; 
\item Characterise geologic features by imaging;
\item Perform seismic studies to detect subsurface activity and structure;
\item Measure atmospheric vertical profile near the surface;
\item Provide aerial images of surface geology.
\end{itemize}

Several Jovian and Saturnian moons house a global subsurface ocean beneath their icy surface. Such subsurface oceans are interesting candidates as potential life-supporting areas. Unfortunately, the subsurface oceans are out of reach for common exploration missions although some aspects such as their density and depth can be inferred/modelled from the data collected. Different models are required to study the habitability of such oceans. Three main approaches exist: the "follow the energy", the "follow the water", and the "follow the nutriment" approaches \citep[see, e.g.,][]{lingam2018extraterrestrial}. Those three approaches require an \textit{a priori} knowledge of the liquid motion within the ocean, which can be provided by a hydrodynamic model.

While Dragonfly could provide data about Titan's subsurface ocean by performing seismic studies, the next advance regarding the subsurface oceans of icy moons will likely come from the JUpiter ICy moons Explorer (JUICE) mission which will be launched in 2022 and should start observing the Jovian moons in 2030. This mission will first orbit Jupiter, investigating Callisto and Europa and then orbit Ganymede. While the mission focuses on Jovian moons housing a subsurface ocean, it will also study other aspects of these natural satellites. Regarding Ganymede and, to a lesser extent, Callisto, JUICE will
\begin{itemize}
\item characterise the ocean layer and detect putative subsurface water reservoirs;
\item study the physical properties of the icy crust;
\item characterise the internal mass distribution, dynamics and evolution of the interiors.
\end{itemize}
JUICE will provide the first subsurface sounding of Europa and the minimal thickness of its icy crust over the most recently active regions.

The model used to study the liquid motion in Titan's global subsurface ocean could easily be modified to study the ocean of Callisto, Europa, and Ganymede. The data collected by JUICE could be used to better constrain the model's parameters regarding the Jovian moon and validate and/or improve our model. Our model is of little interest for the preparation of the mission as an orbiter does not need an \textit{a prioiri} knowledge of the domain studied. Nevertheless, the model could be used in a retro-engineering approach: once the deformations of the surface are well constrained and the number of poorly constrained parameters is reduced, the model could be used to infer the missing parameters from the results. Parameters such as the ocean depth could not be inferred as their influence on the elevation is insignificant but others could more easily be constrained. For instance, the surface heat flux generates a constant elevation which could be isolated from the temporally varying surface deformation measured. Conducting a great number of simulations for various boundary heat fluxes could provide a first approximation of the real boundary heat fluxes.

\section*{Numerical modelling}% of Titan's liquid bodies}
Numerical modelling allows virtually experiencing situations that cannot be observed and to study the impact of various parameters. It is therefore a powerful tool for scientists studying extra-terrestrial sciences for which data are scarce and observations occasional. Each model has its advantages and drawbacks. For instance, a full 3D hydrodynamic model coupled with a meso-scale atmospheric model is able to predict the surface flow which impacts the path of a capsule but it is very costly to use to conduct a sensitivity analysis with respect to the poorly constrained parameters. On the other hand, a 2D hydrodynamic model allows studying a wide range of parameters but only predicts the depth-averaged flow. The ideal tool to study a domain where the lack of data is significant is therefore a combination of these models. The less costly model should be used to study the impact of a wide range of parameters and build a database. Once the range of variation of each parameter is reduced by comparing the results with the observations, a more complex model should be used to predict more accurately the liquid motion. Depending on the domain and data, such models may not exist or the data may not be available to constrain the parameters based on the database.

\section*{My thesis in this context}
The objective of this thesis was as follows:\vspace{0.2cm}
\begin{adjustwidth}{25pt}{25pt}
\textit{Using the scarce available data and a numerical model to improve our knowledge of the motion of surface and subsurface liquid bodies of Titan, thereby helping design future exploration missions of this moon.}
\end{adjustwidth}
\vspace{0.1cm}
To this end, I adapted the Second generation Louvain-la-Neuve Ice ocean Model, SLIM, to extra-terrestrial conditions. The equations solved by the model were modified to take into account the impact of Titan's surface deformation on the liquid motion in the surface lakes and seas and in the subsurface ocean. Due to the lack of data, sensitivity analyses were conducted with respect to the poorly constrained parameters by means of the 2D version of SLIM. The 3D version was also modified to model the liquid motion on Titan. 3D simulations being significantly more costly, the model is used sparingly to study the interactions between the thermally driven flow and the tidal flow in the subsurface ocean (see Chapter~\ref{chap:ocean}) or is applied to a specific sub-domain such as Trevize Fretum, the strait linking Kraken and Ligeia Mare \citep[see][]{vincent2017study}.

The liquid motion predicted in the subsurface ocean could be used by astrobiologists studying the habitability of such an ocean or to constrain the internal heat fluxes. Such applications are beyond the scope of this thesis and are therefore not further developed. Regarding the surface lakes and seas, one goal of the thesis was to provide the elevation variations and the tidal flow, which are key data for the design and planning of an exploration mission focusing on the surface liquid bodies. The lakes/seas surface elevation variations allow highlighting area(s) where the capsule could beach. The tidal flow, on the other hand, can be used to predict the path of a drifted capsule, to provide design constraints for the engine of a motorised capsule/submarine, and to schedule some aspects of the mission such as the crossing of the straits linking the basins of the northern seas. Indeed, it would be convenient to go through these straits when the tidal current is in the right direction to save energy. The Lagrangian particle tracker module (LPT) implemented in SLIM was used to predict the path of a drifted capsule. It could be modified to study the path of a motorised capsule by adding the acceleration resulting from the thrust in the cinematic equations. This could be an asset to optimize the engine operation: using the latter by phase would save energy and reduce the fatigue stress. A modified version of the LPT could easily model the influence of the engine phases on the path, hence allowing conducting a sensitivity study in order to optimize the phases distribution such as the capsule follows a defined path.

Several aspects of SLIM were modified/improved as part of this thesis. The modified model used to study the surface lakes and seas could be used on Earth if the user provides the suitable parameters. There is no other application due to the absence of significant surface liquid bodies in the solar system. The model used to study the liquid motion in the subsurface ocean, on the other hand, could be easily modified to study the ocean of other icy moons housing a subsurface ocean (although particular attention should be paid to the aspect ratio for small moons such as Enceladus). Studying those oceans could be interesting in the context of the forthcoming JUICE mission. Numerical modifications of SLIM could also be useful for terrestrial applications. For instance, its ability to solve the equations on the sphere could be used to study domains where geographic projections are not well suited\footnote{In terrestrial domains, SLIM solves the equations on a flat domain which is obtained by geographic projection of the globe's surface.}. The post-processing tools developed have already been used by other members of the SLIM Team.

Various studies focusing on the surface lakes and seas were conducted in parallel with this thesis. My work differs by the high number of parameters taken into account as well as a better spatial representation of the nearshore areas by taking advantage of the geometrical flexibility of the unstructured grids. The internal structure of Titan and its ocean were also the main topic of studies conducted in parallel with this. My work focuses on the hydrodynamics of the ocean while most studies consider the ocean as a layer while studying the internal structure of Titan or pay interest to the dissipation in the ocean but not to the flow itself.

\subsection*{Ontario Lacus}

Ontario Lacus was the first lake studied as part of this thesis (see Chapters~\ref{chap:ontario} and~\ref{chap:eigenmode}). This lake was selected as the first application for two reasons: two bathymetry maps were available and unexplained shoreline variations were observed. The tides were ruled out as an explanation shortly after the beginning of this thesis, which was confirmed by our results. The elevation field is similar to that predicted by \cite{tokano2010simulation} while the velocity field significantly differs. This is due to the use of a complex bathymetry and the better spatial resolution of our simulations, which enhanced the representation of the coastline, resulting in a better capture of its effect on the flow. The normal modes of the lake were also studied, as a seiche could explain local but significant shoreline variations, and the lake response to arbitrary wind conditions was discussed.

The tides in Ontario Lacus cannot be observed by means of an orbiter. Indeed, the tidal range is much smaller than the vertical resolution of the state-of-the-art instruments. The shoreline variations induced by the tidal motion are also too small to be observed by a radar while Titan's atmosphere makes observation in the visible spectra harder. A spacecraft landing in the lake could easily provide temporal series of the surface elevation and velocity, an accurate bathymetry map as well as the liquid composition and atmospheric conditions. The temporal series of the surface elevation could be used to validate our numerical results (the diurnal signal should be isolated as other forcings could modify the elevation). The bathymetry map would decrease the uncertainties surrounding our results and the liquid composition and atmospheric conditions could be used to take into account additional forcings while studying the liquid motion in the lake. Unlike the tidal motion, the liquid motion associated with large scale normal modes could be observed by an orbiter, if they were resonantly forced. Indeed, the latter can result in high amplitude elevation variation, which could results in significant shoreline variations. The latter could be captured by an orbiter just as those observed by Cassini.

The numerical results can be used to constrain various aspects of a mission consisting in a capsule landing in the lake. For instance, in the light of the small tidal range, less than 0.06 m, beaching should be avoided as the tides could not refloat a beached capsule. The path of a drifted capsule as a function of time and sea landing site was studied by means of the LPT. As expected, the displacements induced by the tidal motion are small, less than 1 km over 3 Titan days (i.e. 1.5 months on Earth) except in the nearshore high speed areas where it can reach a few kilometers. A floating capsule should therefore be motorised to prevent beaching and to allow more significant displacements. In Ontario Lacus, the engine of the capsule should be designed while taking into account the wind forcing as the influence of the latter on the path could be more significant than that of the tidal motion.

The model in itself does not need to be improved to enhance our representation of the tidal motion in Ontario Lacus. Additional data allowing validating the results would be helpful but they are not expected before years. The major improvement that could be effective within a reasonable amount of time would be to couple SLIM3D with a meso-scale atmospheric model to study the air-sea interactions. Indeed, although the wind speed is weak regarding terrestrial standards (at most 10 m/s under stormy conditions), the larger atmosphere density and the smaller liquid density increase the influence of the wind, resulting in a surface stress of the same order of magnitude as on Earth. Two situations are of interest: the liquid motion under stormy conditions and under standard wind conditions. The former could be related to our study of the normal modes. If a mode with a recognizable pattern and a large amplitude develops, it could be observed by the next exploration mission. The liquid motion under standard wind conditions is useful as part of the creation and design of a lander. Due to the slow tidal motion, the wind could play a significant role in the drift of a capsule both by modifying the flow and by pushing the lander in a specific direction (this observation does not stand if the lander is a submarine). The driving effect of the wind on a capsule could be modelled by adding the acceleration resulting from this force in the cinematic equations of the LPT. Using the velocity field predicted by the 3D model would be more accurate as the surface velocity can locally significantly differ from the depth averaged velocity. The particle tracker module being 2D, the surface velocity would be used by the latter instead of the depth averaged velocity.

\subsection*{Northern seas}
The northern seas of Titan are another liquid area of high interest. The tidal motion and the liquid exchange between the two largest seas were studied by means of SLIM2D (see Chapter\ref{chap:northern}). The comparison of our results with those of \cite{tokano2014numerical} is deemed satisfactory both in term of magnitude and spatial patterns. Solar glints at the northern mouth of the strait linking Kraken and Ligeia Maria were observed by \cite{sotin2012observations}. Turbulent eddies or high velocity flows could be responsible for these glints. Therefore, SLIM3D was briefly used to study this area using the sea surface elevation and depth averaged flow as boundary conditions (see \cite{vincent2017study}). The surface speed can locally reach 0.12 m/s and small gyres were observed but their impact on the sea surface roughness was not discussed. Numerical simulations of the liquid motion due to the density gradient \citep{tokano2015sun} and wind \citep{tokano2015wind} based on the predictions of a global circulation model can also be found in the literature.

An orbiter would not be well suited to observe the tidal motion as the height variations are small (especially with respect to the deformation of Titan's surface), although the horizontal variations between low and high tides could induce shoreline variations significant enough to be observed by a radar, depending on the morphology of the surroundings. A lander would be better suited for the same reasons as in Ontario Lacus. Nevertheless, it would be more interesting to send a lander in the northern seas as the later could constrain spatial variations of the atmospheric fields (surface temperature, air moisture, wind velocity \dots) on a much larger scale than in Ontario Lacus and could provide a bathymetry for Kraken Mare and data about possible variations of the liquid composition between the basins.

The northern seas are a designated target for future exploration mission(s). Several proposals were published. At least two of them consist in sending a lander (a drifted capsule or a submarine) in the northern seas. An \textit{a priori} knowledge of the tidal motion in theses seas is an asset for the design and planning of such missions. Although the tidal range is larger than in Ontario Lacus, it is still small and unable to refloat a beached capsule. The path of a drifted capsule as a function of time and landing site was studied by means of the LPT. The capsules landing offshore drift by less than 2 km over 3 Titan days (TD) while those landing in the straits can drift up to 80 km over the same period. Their paths, however, mainly consist of back and forth motion: capsules landing during a specific period are the only one which can reach the straits outlets without being driven back. Therefore a motorised capsule would be better suited as it could travel longer distance and avoid being trapped in the straits. Other forcings such as the wind could impact the path of the capsule. The effect of the wind pushing on the superstructure of the capsule could be taken into account by adding the acceleration resulting from this force in the cinematic equations of the LPT but the influence of the wind, density spatial variations, surface heat fluxes \dots on the liquid motion should also be taken into account in the hydrodynamic model. Conducting a study of the liquid motion induced by the tides, atmospheric forcings, and density gradient is tricky as the timescale of these phenomena are significantly different while they are equally significant in terms of flow speed (at least in the middle of the seas according to \cite{tokano2015wind,tokano2015sun}). The best approach seems to consist in focusing on a specific area using an atmospheric mesoscale model to predict the wind and the surface boundary conditions for the temperature equation. The lateral boundary conditions could be given by SLIM2D and the results of \cite{tokano2015wind,tokano2015sun}. The depth averaged velocity field has already been used as boundary conditions for 3D modelling of sub-domains such as the straits linking the northern seas \citep[see][]{vincent2017study}. An advantage of this approach would be that the LPT could use the surface velocity field instead of the depth-averaged velocity field to predict the displacements of a capsule. While these velocity fields are not expected to differ significantly regarding the tidal motion, this is no longer the case when baroclinic phenomena and wind forcing are taken into account. The straits linking the basins are also a key point for the design of a motorised capsule/submarine as the speed there is one order of magnitude larger than elsewhere in the northern seas. It will therefore constrain the engine of the capsule/submarine. An \textit{a priori} knowledge of the velocity variation with respect to the depth in the straits would be an asset to design a submarine. Indeed, the velocity orientation and magnitude varies with the depth. Selecting the appropriate period and/or depth for a crossing would result in a significant energy saving and would decrease the maximum engine power needed.

It could also be interesting to take advantage of the developments of SLIM conducted in parallel with this thesis. Indeed, a member of the SLIM team, Thomas Dobbelaere, recently coupled SLIM with a wave model, SWAN (Simulating WAves Nearshore, \url{swanmodel.sourceforge.net/}, see \cite{booij1997swan}). Studying the waves of the northern seas could be interesting both as part of the design of a lander and to explain phenomena such as the solar glints and the "Magic Islands". Indeed, the capsule should be stable enough to not capsize due to the regular swell and its structure should be sized to be able to resist the small shocks with the waves. Therefore, the waves height and energy should be quantified as they are useful quantities for the design of a capsule. Regarding the "Magic Islands", four explanations were initially proposed. The tides being ruled out, the remaining suspects are sediment in suspension, rising bubbles, and small waves. Due to the lack of information, it is not possible to study the sediment transport in the area of the "Magic Islands" and rising bubbles request a different model to be studied. The third explanation could be discussed by means of a coupled SLIM-SWAN model. Such a model requires an \textit{a priori} knowledge of the wind distribution during the period studied. The wind predicted by a mesoscale atmospheric model (or a GCM, depending on the domain studied) could be used as input without coupling the atmospheric model with the SLIM-SWAN model.

Possible connection(s) with a third basin south of Kraken Mare and with Punga mare could be taken into account in order to improve our model. Although no connection is visible on the Cassini images, Punga Mare is likely connected to the northern seas as its surface is on the same equipotential as those of Kraken and Ligeia Maria. Regarding the basin south of Kraken Mare, a connection appears on ISS images but not on the radar images. This basin was therefore disregarded as our study is based on radar images. These basins could however be linked to Ligeia and/or Kraken Maria by canals too small to be observed by Cassini, by a subsurface flow, or by swamps. Taking into account the connection with these basins could modify the tidal motion as the liquid exchanges between basins have a significant impact on the tidal phase (see Chapter~\ref{chap:northern}). Nevertheless, conducting such a study without basic knowledge of the domain is pointless. Crucial parameters such as the horizontal dimensions of the canals/straits or an approximation of the volumetric flow rate are required to construct a realistic approximation of the domain.

\subsection*{Global ocean}
The penultimate chapter of this thesis focuses on the liquid motion in the subsurface ocean of Titan. Studying the liquid motion in this ocean could seem to be of little interest as models predicting the deformation of Titan or the tidal dissipation are available. Nevertheless, studying the liquid motion can help ruling out or validating assumptions about the heat fluxes at the top and bottom of the ocean and the results could be useful for astrobiologists paying attention to the ocean as a life-supporting area.

The ice shell has a significant impact on the surface elevation and the velocity field. Its effect should therefore be taken into account by scientists paying interest to the ocean hydrodynamics. Nevertheless, while a fully coupled model could be used to improve our knowledge of the fluid-structure interactions, the further insight gained would be very limited due to the poorly constrained parameters. Such a model could detect a resonance phenomenon induced by the interactions but such phenomenon seems to appear only in unrealistically shallow ocean \citep[see, e.g.,][]{hay2017numerically,tyler2014comparative}. Building such a model would be a giant with feet of clay as the accuracy gained would be rendered useless by the uncertainties in the parameters. The preliminary study also shows that the tidal and thermally driven flows strongly interacts. Taking into account disregarded phenomena such as the ice melting/water freezing (and the resulting bathymetry variations), a spatially varying bottom heat flux, and the heat produced by dissipation at the boundary and within the ocean would improve our understanding of the liquid motion within a subsurface ocean more significantly than coupling the model to study the interactions with the ice shell.

\section*{Conclusion}

Using a modified version of SLIM, an Earth-based model for geophysical and environmental flows, I studied the tides of Titan's liquid bodies. This version of SLIM is able to solve the depth-averaged shallow water equations as well as the 3D hydrostatic equations under the Boussinesq approximations on flat as well as spherical extra-terrestrial domains. The global ocean of Titan lying beneath an ice shell, the hydrodynamics equations were modified to take into account the impact of the shell deformation on the liquid motion.

The tidal motion in the surface liquid bodies of Titan was studied in order to provide useful data for the forthcoming dedicated exploration mission(s) and in the hope of explaining phenomena such as shoreline variations or "Magic Islands". The tidal motion in the surface lakes and seas of Titan is of little amplitude with respect to terrestrial standards. The surface elevation, although significantly varying depending on the lake/sea, is smaller than 0.4~m in the domains studied. The flow speed is $\mathcal{O}(10^{-2})$ m/s except in the straits linking the basins of the northern seas where it is one order of magnitude larger. The relatively weak tidal amplitude is due to the deformation of Titan's surface which damps the astronomical forcing acting on the surface lakes and seas. The same damping effect is observed on Earth, but at smaller scales. Indeed, the deformation of Titan's surface is much larger in terms of amplitude as the liquid layer is filled with water, instead of liquid metal on Earth, and the Young modulus of ice is one order of magnitude smaller than that of rocks. Therefore, the damping effect on Titan is much more significant: the attenuation factor is about 0.1 while it is about 0.7 on Earth. The various phenomena observed in the surface lakes and seas such as the shoreline variations of Ontario Lacus, the "Magic Islands" and the solar glints in the northern seas cannot be explained by the tidal motion. While excited normal mode(s) could be responsible for liquid motion whose amplitude is at least one order of magnitude larger than the tidal range in Ontario lacus, their natural period is such that only atmospheric forcings could resonantly force them. The numerical results were used to predict the path of drifted capsules. Based on the paths predicted for various landing sites and time, a motorised capsule seems better suited to avoid beaching or being trapped in small areas as well as for allowing significant displacements to take place. The results can also be used to constrain the engine of a capsule landing in the surface liquid bodies and to schedule the period at which a capsule should sail through  a strait in order to take advantage of the flow orientation. The model can be used to study sub-domains more accurately or to predict the influence of various parameters on the liquid motion and/or the path of a capsule.

 The simulations of the liquid motion within the global subsurface ocean show that the surface heat fluxes derived by \cite{kvorka2018does} do not result in unrealistic liquid motion. The resulting flow strongly interact with the tidal motion, modifying significantly the velocity orientation while the surface elevation is only slightly modified. Although the predicted surface elevation corresponds to the observed deformation of Titan's surface, the results cannot be validated by comparison with data. The ice shell lying at the top of the ocean significantly impacts the tidal motion: it slows down the tidal flow and decreases the amplitude of the surface elevation by a few percentage points. If viscous deformation significantly larger than the one expected took place, the elevation and speed would be increased by the ice shell as high frequency oscillations would appear due to the 90$^\circ$ out of phase term of the surface pressure representing the viscous deformation.

Modelling the tides of Titan liquid bodies is a complex task as the lack of data prevents from validating the results and leads to poorly constrained parameters. I took advantage of the current state-of-the-art knowledge of Titan and a proven Earth-based model to build an image of Titan liquid tides while taking into account the remaining uncertainties. Extra-terrestrial exploration is a scientific field suffering from many stop's and go's due to the elapsed time between the missions. Each mission is a giant leap forward which can contradict the predictions based on the observations of the previous mission(s). For instance, based on Voyager spacecrafts observations, a deep ocean of liquid hydrocarbons was assumed to take place on Titan surface while Cassini mission showed that the surface lakes and seas covered only $1.3\%$ of Titan's surface. Similarly, data from forthcoming exploration mission(s) could prove that Earth-based models are not suitable for modelling the liquid motion of Titan's liquid bodies or could infirm the values of the parameters used in this thesis.

This thesis is based on the data and observations collected by Cassini mission, which provided a clear outline of the surface lakes and seas shoreline, and bathymetry maps for some of them. Further insight were also gained in other aspects such as the composition of the liquid filling the lakes and seas, the internal structure of Titan, and the atmospheric conditions. While some observations such as the lakes/seas shoreline are hardly questionable, others are. For instance, the bathymetry map of Ontario Lacus significantly varies depending on the method used to derive it. To address this problem, sensitivity analysis with respect to the poorly constrained parameters were conducted. I am reasonably confident about the results presented in this thesis as they compare well with those available in the literature. While the quantified values do vary with the parameters, several aspects such as the global patterns of the elevation and velocity fields are rather insensitive to the parameters; except in the northern seas where the liquid exchanges between the basins have a significant influence on the phase of the main tidal component, hence modifying the spatial and temporal variations of the elevation field.

### 2022-01-04 chapters/intro/intro_david.tex
creation of thesis git

A common question in recent history is \textit{"Is there life anywhere else in the vastness of space?"}. There are countless books, films and songs about such an extraterrestrial life. Historically, the first planet suspected to shelter life was Mars (this is why the word "Martian" is often used to refer to extra-terrestrial life). Indeed, observations of canals on its surface in 1877 by the astronomer Giovanni Schiaparelli led to the outbreak of legends about an intelligent extraterrestrial life on Mars which would have dug those canals. Nowadays, the existence of life on Mars is almost completely ruled out -- at least on its surface -- and the focus for searching life switched from Mars to other celestial bodies such as icy moons and exoplanets\footnote{An exoplanet is a planet that is not part of our solar system.}. As far as we know, one (among other) essential condition to the emergence of life is the presence of liquid. In the solar system, one can find substantial amount of liquid on various planets and moons: Earth, Enceladus, Europa, Titan, Ganymede~\dots and, as shown by recent observations \citep{orosei2018radar}, on Mars. Nevertheless, most of these celestial bodies hide the liquid beneath their surface: only two of them have a significant amount of liquid bodies on their surface. The first one is well known as we live there: the Earth. Few have heard about the second one: Titan, the largest moon of Saturn. It has both surface lakes and seas filled with liquid hydrocarbons and a global subsurface water ocean.

This thesis aims at studying the motion of Titan's liquid bodies (i.e. both the surface lakes and seas and the global subsurface ocean). One purpose behind such studies is to suggest explanations for several observed phenomena such as shoreline variations, backscattering induced by lake surface roughness, appearance and disappearance of bright spots seemingly corresponding to solid ground on Cassini observations (the so-called "Magic Islands") and surface deformation -- the latter being strongly influenced by the liquid motion in the global subsurface ocean. On Earth, this can easily be done by means of numerical modelling -- whose results can be validated by measurements -- and \textit{in situ} or satellites observations. Nevertheless, on Titan, there is no way to get either \textit{in situ} measurements or new observations from any spacecraft as Cassini recently dived into Saturn at the end of its mission. Consequently, the only way to get new information about Titan in the next years will be by analysing "old" data with new method(s) or using numerical models. For instance, M. Mastrogiuseppe, A. G. Hayes and their team successfully derived a bathymetry map of some lakes and seas with such methods \citep{hayes2016bathymetry, mastrogiuseppe2014bathymetry,mastrogiuseppe2016radar, mastrogiuseppe2018bathymetry}. Modelling tools are expected to play an important part in order to gain further insight about Titan but also to help planning and designing future exploration mission(s) of Titan. Indeed, except for an orbiter, a lot of information are needed to be able to predict/select the landing site (depending on the starting point of the descent). Furthermore, depending on the locomotion, data such as the wind and/or liquid velocity encountered could be needed in order to either engineer the lander motorisation or to predict its position as a function of time.

\section{Titan} 
Titan was discovered in 1655 by the dutch astronomer Christiaan Huygens. The first spacecrafts to fly near Titan were Pioneer 11 and the Voyager I $\&$ II probes. Following their observations, a mission dedicated to the Saturnian system was launched in 1997: the Cassini mission. It was composed of a lander, Huygens, which landed on Titan on the 14$^{th}$ of January 2005 and an orbiter, Cassini, which orbited around Saturn from the 11$^{th}$ of June 2005 to the 15$^{th}$ of September 2017. At first, Cassini mission was supposed to last 4 years but, due to happy circumstances, it outlived its initial duration by 9 years (see Fig.~\ref{fig:intro-chrono}). Various projects were proposed to follow Cassini and study Titan in further details: Time~\citep{lorenz2012winds}, E2T~\citep{mitri2014exploration}, Titan sub~\citep{hartwig2016exploring}, Oceanus~\citep{sotin2017oceanus}, Dragonfly~\citep{turtle2017dragonfly}~\dots The latter was recently selected by the NASA as part of its New frontiers program. Its launch date is scheduled in 2026.

\begin{figure}
\begin{center}
\includegraphics[trim=5cm 1.15cm 5cm 1.2cm,clip=true,width=.7\textwidth]{chapters/intro/figures/titan.jpg}
\captionof{figure}{\label{fig:intro:titan}View of Titan from the Cassini spacecraft (in the visible spectrum). The orange aspect of Titan is due to a thick haze mostly made of methane, ethane and nitrogen (image credit: NASA/JPL-Caltech/Space Science Institute).}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{chapters/intro/figures/chronologie.png}
\captionof{figure}{\label{fig:intro-chrono}Chronology of Cassini mission. The original mission was first extended by about 2 years (the so-called Equinox mission) and then by 7 years (the Solstice mission) (image credit: \citet{Manor2013Cassini}).}
\end{center}
\end{figure}
Titan is the second largest moon in the solar system, the largest one being Ganymede, a Jovian moon. Its orbital period around Saturn is 15.945 Earth days, which is close to its rotational period. Consequently, one can consider Titan as tidally locked as a first approximation. Indeed, the mean non synchronous rotation is of about $-0.024^\circ\pm0.018^\circ$/year \citep{meriggiola2016rotational}. This results in length of day variations, polar motion and longitudinal libration due to the gravitational torque exerted by Saturn and the dynamic variation of the atmosphere \citep{coyette2018variations}. Nevertheless, the impact of such variations on the attraction forces is insignificant when looking at the tidal motion and one can hence consider Titan as tidally locked in this respect. Like the Earth, Titan is an ellispoid with semi-major and semi-minor axes given by $a =2575.15\pm0.02$~km, $b =2574.78\pm0.06$ km, $c =2574.47\pm0.06$~km \citep{zebker2009size}. A few characteristics of Titan are listed in Table~\ref{tab:intro:charac}. 
\begin{table}
\centering
\begin{threeparttable}
\caption{Main characteristics of Titan compared with those of the Earth. \label{tab:intro:charac}}
\begin{tabular}{p{6.cm}p{2.3cm}p{2.3cm}} \hline\noalign{\smallskip}

{}&{Titan}&{Earth} \\ \noalign{\smallskip}\hline\noalign{\smallskip}

{Orbital eccentricity}&{0.0288} &{0.0167}\\ 
{Obliquity [$^\circ$] $^a$}&{0.3} &{}\\ 
{Obliquity [$^\circ$] $^b$}&{26.73} &{23.44}\\ 
{Average orbital speed [km/s]}&{5.57} &{29.78}\\ 
{Mean radius [km]} &{2574.73}&{6371} \\ 
{Volume [km$^3$]}&{7.16$\times10^{10}$} &{$1.08\times10^{12}$}\\ 
{Surface [km$^2$]}&{8.3$\times10^7$} &{$5.1\times10^8$}\\ 
{Mass [kg]}&{$1.3452\times10^{23}$} &{$5.972\times10^{24}$}\\
{Mean surface temperature [K]}&{93.7} &{288}\\
{Surface pressure [kPa]}&{146.7} &{101.325}\\
{Mean gravitational acceleration [m/s$^{2}$]}&{1.352} &{9.807}\\
 \\\noalign{\smallskip}\hline
\end{tabular}
\begin{tablenotes}
\small
\item $^a$: Obliquity relative to Saturn equatorial plan. This obliquity is relevant for the study of the tidal motion as it contributes to the tidal potential.
\item $^b$: Obliquity relative to the Sun equatorial plan. This obliquity is relevant for a climate/solar illumination point of view.
\end{tablenotes}
\end{threeparttable}
\end{table}

\subsection{Atmospheric conditions}
\label{sec:intro:atm_cond}
Titan is the only natural satellite of the solar system which has a thick atmosphere\footnote{A thick atmosphere does not distinguish from a thin atmosphere by its total depth but by the gases that are present and their density. The larger the density is, the more thick the atmosphere.}. The presence of an atmosphere on Titan was first assumed by the astronomer Josep Comas i Sol\'a in 1903 and was confirmed in 1944 by Gerard Kuiper using a spectroscopic technique \citep{kuiper1944titan}. In 1980, Voyager I revealed that the surface pressure was about 1.48 times larger than on Earth (see Table~\ref{tab:intro:charac}). The major breakthrough regarding Titan's atmosphere came from the Cassini mission which was able to measure, among other things, the composition and the atmospheric conditions. The only \textit{in situ} measurements of Titan atmospheric conditions were provided by the Huygens probe during its descent and at its landing site \citep[see][]{fulchignoni2005situ} while Cassini instruments such as the CIRS\footnote{Composite InfraRed Spectrometer : it measures infrared energy from the surfaces, atmospheres, and rings of Saturn and its moons to study their temperature and compositions \citep{flasar2004exploring}.} and the VIMS\footnote{Visible and Infrared Mapping Spectrometer: it identifies the chemical composition of the surface, atmosphere, and rings of Saturn and its moons by measuring colors of visible light and infrared energy emitted or reflected \citep{brown2004cassini}.} were able to provide additional information through remote sensing.
% thick atmosphere:https://sciencing.com/difference-between-thick-thin-atmospheres-12302390.html
Unlike Mars, where surface water can be observed at large scales -- but only in a solid state --, methane was observed in both gaseous, liquid, and solid state on Titan as a methane cycle similar to the water cycle on Earth takes place on Titan \citep{atreya2006titan}. Most of the liquid methane is located in the surface lakes and seas (see Section~\ref{sec:intro:liquidBodies}). The atmosphere is mainly composed of nitrogen with a small amount of methane and a still smaller amount of other species (e.g. $^{40}$Ar, $^{36}$Ar, cyanogen, ethane) \citep{niemann2005abundances} \citep[for a detailed composition, see Table 1 in][]{cordier2009estimate}. The methane in the atmosphere can condense to form clouds \citep{toon1988methane, griffith2000detection, porco2005imaging} which can lead to methane precipitation \citep{toon1988methane}. In the high atmosphere, methane photolysis leads to the production of fast reacting radicals CH$_3$, CH$_2$, and CH. These radicals then react with atmospheric molecules such as nitrogen and hydrogen to form complex C$_x$H$_y$ (mostly ethane, C$_2$H$_6$) and nitrile species (C--N--H) \citep{atreya2006titan}.

\textit{In situ} measurements of Huygens probe gave the vertical pressure, temperature, and wind speed profile along its descent but, in order to have a data map of these fields, one has to rely on models and remote measurements from Cassini spacecraft. The surface temperature can be inferred from CIRS far infrared spectra while pressure field, precipitation and wind velocity have to be modelled. \citet{jennings2009titan} and \citet{cottini2012spatial} both derived the surface temperature from CIRS data respectively from 2004 (right after the northern winter solstice) to 2008 (late northern winter) and from 2004 to 2010 (in the early northern spring). Their results are consistent with each other and with Huygens measurements. Seasonal variations were observed by both authors. \citet{jennings2009titan} did not take into account the diurnal changes while \citet{cottini2012spatial} observed that the diurnal features disappeared at latitudes greater than $20-30^\circ$ in the north and $30-40^\circ$ in the south. Nevertheless, they took into account neither the data from latitudes greater than $60^\circ$ nor the local influence of liquid bodies. \citet{tan2015titan} established a formula describing the surface temperature as a function of the latitude and the time from January 2005 to December 2013 from \citet{jennings2011seasonal} and \citet{cottini2012spatial} temperature profiles. The values predicted by this formula have to be used with caution as no variation with respect to the longitude was taken into account.

Several global circulation models (GCMs) \citep[e.g.][]{tokano2008dune, tokano2009impact, friedson2009global, lebonnois2012titan,newman2011stratospheric,newman2016simulating, schneider2012polar} and mesoscale models \citep{soto2015mesoscale,charnay2015methane} were proposed in order to predict the atmospheric pressure, precipitation and winds. According to the GCMs, the atmospheric pressure variations are small at the surface and the winds are subject to significant spatial, seasonal and diurnal variations. Additionally, the wind speed may vary depending on the lake composition: if the seas are methane rich, the wind speed could be different than in the case of ethane-rich seas \citep{tokano2009impact,lorenz2012winds}. The tidal wind (i.e. the wind resulting from the diurnal astronomical forcing) has neither a preferential direction nor a specific time distribution while the wind due to the convergence of moist air over the lake area is directed offshore and can be stronger than the tidal wind \citep{tokano2009impact}. According to \cite{charnay2014Thesis}, the tidal winds are overestimated by \cite{tokano2009impact} as the tidal forces induce mostly change in pressure, resulting in slower wind. Except in stormy conditions where it can reach up to 10 m/s, the predicted wind speed ranges from 0 to 2~m/s depending on the location and the season \citep{lorenz2012winds, lorenz2013oceanography}. Each of these models uses simplifying assumptions and none of them is fully consistent with observations \citep{schneider2012polar}.

\subsection{Surface lakes and seas }
\label{sec:intro:liquidBodies}
Due to the atmospheric conditions at Titan's surface measured by Voyager spacecrafts, the presence of liquid methane was evoked as soon as 1981 by, among others, \citet{samuelson1981mean,hanel1981infrared}. It led \citet{sagan1982tide} and \citet{dermott1995tidal} to construct a pre-Cassini appearance of Titan's surface (which was unknown at that time) by conducting analyses of theoretical global surface ocean and disconnected seas and lakes on Titan's surface. Other features were also predicted on Titan: crater lakes \citep{lorenz1994crater}, submerged impact craters \citep{lorenz1997impacts} and cryovolvanic flows with pillow lava texture \citep{lorenz1996pillow}. Although such flows and crater lakes were not observed, the presence of surface lakes and seas was confirmed by Huygens and Cassini measurements \citep{mcewen2005mapping,stofan2007lakes}. Cassini first detected possible surface lakes in the southern polar region of Titan's surface in 2004 (only some years later confirmed to be liquid) by means of the imaging science subsystem\footnote{Imaging Science Subsystem: it takes pictures in visible, near-ultraviolet, and near-infrared light \citep[see][]{porco2004cassini}.} (ISS) \citep{mcewen2005mapping}. Its radar\footnote{Radar: it maps the surface of Titan using a radar imager to pierce the veil of haze. It is also used to measure heights of surface features. The synthetic aperture radar observed Titan in 13.78 GHz Ku-band with a resolution ranging from 0.35 to 1.7 km \citep[see][]{elachi2004radar}.} showed dark patches, interpreted as lakes, in the northern polar region on the 22$^{\text{nd}}$ of July 2006, during the flyby T16 \citep{stofan2007lakes}. These lakes and seas\footnote{The largest liquid bodies are referred to as seas (Mare) and the others as lakes (Lacus), according to the nomenclature of the International Astronomical Union. It only reflects the size of the body.} are asymmetrically distributed with respect to the equator \citep{aharonson2009asymmetric}: there are many more lakes in the northern high latitudes where they are larger and deeper \citep{hayes2008hydrocarbon, hayes2018post}. This asymmetry could be explained by the shorter and more intense southern summer induced by the obliquity of Saturn.

Various studies were carried out about Titan's seas and lakes: bathymetry \citep[e.g.][]{hayes2010bathymetry, ventura2012electromagnetic, mastrogiuseppe2014bathymetry,mastrogiuseppe2018bathymetry, mastrogiuseppe2018cassini,lorenz2014radar, hayes2016lakes, hayes2017topographic}, liquid composition \citep[e.g.][]{brown2008identification, cordier2009estimate, glein2013geochemical, tan2013titan, tan2015titan, luspay2015experimental}, dissipation due to friction in Titan's lakes and seas \citep[e.g.][]{sagan1982tide, dermott1995tidal, sears1995tidal, lorenz2014radar}, tidal response \citep{tokano2010simulation, tokano2014numerical,vincent2016numerical,vincent2018numerical}, normal modes \citep{dermott1995tidal,tokano2010simulation,vincent2019normal} interactions with the atmosphere \citep{tokano2015wind, tokano2015sun}, and phenomena such as the shoreline variations of Ontario Lacus \citep{turtle2011shoreline,hayes2011transient} and transient features in the northern seas \citep{hofgartner2014transient, hofgartner2016titan} were studied.

 The study of such lakes and seas belongs to a new scientific research field usually referred to as \textit{extra-terrestrial oceanography}. Indeed, although the governing equations are similar to those of terrestrial oceanography, there are significant differences that need to be taken into account: extra-terrestrial surface seas or lakes involve regimes of temperature, pressure, composition, and physical environment (gravity, tidal forces, rotational, and orbital periodicities, etc.) that are not observed in Earth's oceans. Despite all the observations and the results of previous studies, there are a lot of uncertainties about some essential parameters such as the liquid density, the bathymetry, the attenuation of the forcing induced by the lake/sea's bottom deformation \dots For instance, the bathymetry of a lake was derived following two different approaches, resulting in depth variations up to 28 m. Consequently, sensitivity analyses are conducted with respect to several parameters.

\subsubsection{Surface lakes and seas distribution}
The surface lakes and seas of Titan cover $1.04\times 10^6$~km$^2$, which represents about $1.3\%$ of the moon surface. Most of them are located at latitudes higher than $60^\circ$ \citep{aharonson2009asymmetric} (see Fig.~\ref{fig:intro:map_north}) although a few equatorial and mid latitudes dark spots are proposed as possible lakes \citep{griffith2012possible}. 295 of these lakes, covering a combined surface of $1.36\times 10^5$~km$^2$, are empty while there are 577 liquid-filled lakes, covering a combined surface of $2.33\times 10^5$~km$^2$ \citep{hayes2016lakes}. Most of them consist of small sharp-edged depressions filled with liquid but there are also larger lakes such as Jingpo and Bolsena Lacus in the northern hemisphere and Ontario Lacus in the southern one. Besides these lakes stand three large Maria representing $80\%$ of the liquid filled surface: Kraken Mare ($5\times 10^5$~km$^2$), Ligeia Mare ($1.3\times 10^5$~km$^2$) and Punga Mare ($6.1\times 10^4$~km$^2$) \citep{hayes2016lakes}. While Kraken Mare and Ligeia Mare are hydrologically connected by a strait, namely Trevize Fretum, \citep{sotin2012observations} (see Fig.~\ref{fig:intro:map_north}) one can only assume a connection between Kraken and Punga Mare: the maria's surface elevation are similar \citep{hayes2017topographic,hayes2016lakes} but no large canals were observed. Such a connection would therefore be made of small canals, shallow swamps, or underground flows. 
\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{chapters/intro/figures/map_bis.png}
\captionof{figure}{\label{fig:intro:map_north}Mosaic of Cassini radar images covering Titan northern hemisphere. (credit: NASA/JPL-Caltech/Agenzia Spaziale Italiana/USGS).}
\end{center}
\vspace{-0.9cm}
\end{figure}
 Kraken Mare is the largest sea identified on Titan, with an area of about $500,000$ km$^2$. It is centred at ($68^\circ$N, $50^\circ$E) \citep{tokano2010simulation} and it stretches from $55^\circ$N to $82^\circ$N \citep{lorenz2014radar}. Two basins can be highlighted, Kraken 1 (in the north) and Kraken 2, linked by a strait named Seldon Fretum (see Fig.~\ref{fig:intro:map_north}) and a set of small channels \citep{lorenz2014radar}. The horizontal dimensions of Seldon Fretum are similar to those of the Gibraltar strait on Earth, about $17$ km wide and $40$ km long \citep{lorenz2014radar}, but it is not the case of its depth. Ligeia Mare is located north-east of Kraken Mare: it is centred at ($79^\circ$N, $112^\circ$E). Its maximal dimensions are about $420$ by $350$ km \citep{stofan2012shorelines}. The Cassini radar altimeter also detected liquid filled canyons linked to Ligeia Mare in the northern region, near Xanthus Flumen, and south-eastern region, near Vid Flumina \citep{poggiali2016liquid} (see Fig.~\ref{fig:intro:map_north}). Transient features called "Magic Islands" were observed in Kraken Mare and Ligeia Mare \citep{hofgartner2014transient, hofgartner2016titan}. They consist of areas which appear bright on some observations and dark on others. Bright areas are consistent with solid ground while dark ones correspond to liquid. In Ligeia Mare, such phenomena were observed in regions centred at ($78^\circ$N, $123^\circ$E) and ($80^\circ$N, $111^\circ$E). They are best explained by ephemeral phenomena such as floating and/or suspended solids, bubbles, and waves while the diurnal phenomena, such as the tide, seem to be ruled out, the observations being all at nearly the same true orbital anomaly\footnote{The true orbital anomaly is the angle defining the position of Titan along its Keplerian orbit as seen from Saturn.} \citep{hofgartner2016titan}. In Kraken Mare, the phenomenon was observed at ($73^\circ$N, $55^\circ$E). In this sea, the same explanations stand but the tides cannot be ruled out. The third sea, namely Punga Mare, is also the smaller. It is located north of Kraken Mare, near the north Pole. It is centred at ($85.1^\circ$N, $339.7^\circ$W) and has a diameter of 380 km.

 \begin{figure}
 \centering
 \includegraphics[width=8.4cm]{chapters/intro/figures/ontario.pdf}
 \caption{Map of Ontario Lacus. Letters from A to M show the study areas of \citet{hayes2010bathymetry} which derived a nearshore bathymetry in these regions. The red letters, from N to R, are used in Chapter~\ref{chap:ontario} to refer to particular areas (adapted from www.titanexploration.com; original image credit: Cassini Radar Science team, NASA/JPL/Caltech) \label{fig:intro:ontario}}
 \vspace{-0.5cm}
 \end{figure}

Ontario Lacus is the largest lake in the southern hemisphere, covering approximately an area of 200 × 70 km \citep{wall2010active}. It is centred at ($72^\circ$S, $175^\circ$E). Ontario Lacus was flown over at least twice by each Cassini instrument (complete or partial flybys): ISS in 2004, 2005 $\&$ 2009, VIMS in 2007 $\&$ 2009, and radar in 2009 $\&$ 2010. It has the shape of a right footprint (see Fig.~\ref{fig:intro:ontario}) and is named after Lake Ontario (one of the Great Lakes of North America). It is connected to a hydrological network (for example, in area N in Fig.~\ref{fig:intro:ontario}) \citep{wall2010active, cornet2012geomorphological} which can provide liquid hydrocarbons during and after precipitations. Several, quite different, morphologies were observed along the shoreline of Ontario Lacus: deeply incised bays, mountainous region, beaches \dots \citep[see][for further details]{wall2010active, cornet2012geomorphological}. The two others significant lakes, Bolsena Lacus and Jingpo Lacus are located in the northern hemisphere. Bolsena Lacus is centred at ($76^\circ$N, $10^\circ$W) and has a diameter of 101 km while Jingpo Lacus has a diameter of 240 km and is centred at ($73^\circ$N, $336^\circ$W).

\subsubsection{Liquid composition}
\label{sec:intro:composition}
Studies showed that the surface lakes and seas are not filled with pure methane but with a mixture of methane, ethane, nitrogen, and other low-molecular mass hydrocarbons \citep{ mastrogiuseppe2014bathymetry,luspay2015experimental,mitchell2015laboratory, tan2013titan,tan2015titan}. The exact liquid composition varies between the lakes/seas. For instance, Ligeia Mare is much more methane rich than Ontario Lacus or Kraken Mare \citep{hayes2016bathymetry,le2016composition}. The VIMS identified liquid ethane in Ontario Lacus in 2008 while the identification of liquid methane on Titan's surface by means of this instrument is made difficult by the presence of methane in the atmosphere \citep{brown2008identification}. Using Cassini radar, \citep{mastrogiuseppe2014bathymetry,hayes2016bathymetry,le2016composition} were able to predict the composition of some lakes and seas such as Ontario Lacus and Ligeia Mare. This method relies on analysing the dielectric constant and the loss tangent\footnote{The loss tangent is the tangent of the loss angle. It is used to parametrise the electromagnetic energy dissipation inherent to a dielectric material.} of the lakes and seas: the proportion of methane, ethane and nitrogen is derived from the value of these parameters. \citet{luspay2015experimental} and \citet{mitchell2015laboratory} conducted laboratory experiments to study liquid properties such as evaporation rates and derived rough approximation of the lakes and seas composition. All these studies were able to predict the proportion of the major constituents such as ethane and methane but were not able to predict a detailed composition of the lakes and seas. To this end, models have to be resorted. Several models are available in the literature \citep[see, e.g.,][]{cordier2009estimate,tan2013titan,tan2015titan,glein2013geochemical}. \citet{cordier2009estimate} and \citet{tan2013titan, tan2015titan} distinguished two types of lakes: the near-equator lakes and the high latitudes lakes. This distinction is the result of the temperature difference between the equator and the high latitudes. \citet{cordier2009estimate} established the composition by considering the lakes as non ideal solutions in thermodynamic equilibrium with the atmosphere. Nevertheless, this model is quite sensitive to uncertainties about thermodynamic data and other parameters: the relative standard deviations remain between $10\%$ and $300\%$ depending on the species considered \citep{cordier2012titan}. \citet{tan2013titan} also considered this equilibrium and used an equation of state to model the chemical system of the atmosphere. \citet{tan2015titan} used the same model as \citet{tan2013titan} and included the effect of multicomponent mixtures. \citet{glein2013geochemical} developed a van Laar model parametrised by using experimental phase equilibrium data. \citet{cordier2009estimate} and \citet{luspay2015experimental} predicted an ethane-rich composition while \citet{tan2013titan, tan2015titan} predicted a methane-rich one. Nevertheless, \citet{tan2015titan} predicted that the composition varies seasonally: ethane and other heavy components would be present in larger quantities in winter than in summer. \citet{glein2013geochemical} also predicted a methane-rich composition but their results significantly vary with the mixing ratio of methane. If the mixing ratio of methane suggested by Voyager 1 was used, they would obtain an ethane-rich lake. The chemical models result in a liquid of similar density but there are some significant differences about the molecular viscosity: that predicted by \citet{cordier2009estimate} and \citet{luspay2015experimental} is larger than that predicted by \citet{tan2013titan} (respectively by a factor of 5 and up to 3). The lake also carries some solid particles like acetylene and high molecular weight organics produced by atmospheric photochemistry \citep[see e.g.][]{lorenz2010threshold, lorenz2011cyanide, lorenz2013oceanography} (\citet{tan2013titan} showed that tholins and acetylene could be present in Titan's lakes). These particles modify the lake properties. Therefore, the liquid density and viscosity cannot be accurately derived from the lake composition.

\subsection{Titan's internal structure}
\label{sec:intro:depth}
Titan is an icy moon. It is a class of natural satellites whose surface is composed mostly of water ice. The layers beneath the surface vary from one moon to another. As shown in Fig.~\ref{fig:intro:interior}, there is a liquid body on Titan other than the surface lakes and seas: a global subsurface ocean filled with liquid water lies beneath the icy surface of the moon. The latter is mostly composed of water ice and is therefore referred to as an icy shell.
\cite{lunine1987clathrate} were the first to predict the presence of such an ocean, although it had been discussed by \cite{consolmagno1978evolution}. Since then, the presence of a salt-rich or ammonia-rich water ocean has been supported by several observations and measurements such as Titan's obliquity, its surface motion, its gravity field and its topography \citep{rappaport2008can,nimmo2010shell, iess2012tides, hemingway2013rigid, lefevre2014structure, baland2011titan, baland2014titan} and has been widely studied \citep[see, e.g.,][]{beuthe2008thin, beuthe2015tides, beuthe2015tidal, fortes2012titan, sohl1995tidal, sohl2003interior, sohl2014structural}. Beneath this ocean lies a layer of high pressure water ice and, beyond this layer, the core. Although the layers forming Titan are clearly identified, uncertainties remain about their thickness and properties. For instance, there are discrepancies between the ice Young modulus inferred from Cassini measurements and that obtained in laboratory (i.e., on Earth) \citep{nimmo2004what}. The models used to infer the thickness and properties of the internal layers are validated by comparison with properties inferred from Cassini measurements such as the mean moment of inertia (which was questioned by \citet{hemingway2013rigid}, \citet{baland2014titan} and \citet{lefevre2014structure}), the second-degree gravity field coefficients, and Titan's mean density. The ocean's depth and density predicted can significantly vary depending on the model used. The commonly predicted depth ranges from several tens of kilometers to about 400 km. The ocean is dense: \citet{baland2014titan} predicted a density between 1275 and 1350 kg/m$^3$, \citet{sohl2014structural} studied two possible densities, 1150 and 1350 kg/m$^3$, while \citet{coyette2018variations} considered that the ocean density may vary from 950 to 1350 kg/m$^3$. This range can be further constrained by using properties inferred from Cassini measurements such as the second degree tidal Love number $k_2$. Indeed, low densities are not consistent with respect to the value of $k_2$ derived by \cite{iess2012tides}. Densities lower than 1150 kg/m$^3$ are consequently disregarded.

\begin{figure}
\begin{center}
\includegraphics[width=.7\textwidth]{chapters/intro/figures/interior.pdf}
\captionof{figure}{\label{fig:intro:interior}Internal structure of Titan as seen by the scientific community. The thickness of each layer is not to scale for the sake of clarity.}
\end{center}
\vspace{-0.5cm}
\end{figure}

\subsection{Tidal forcing}
\label{sec:intro:forcing}
There are numerous celestial bodies which can generate a tidal motion on Titan. Nevertheless, the forcing resulting from Titan's obliquity and its orbital eccentricity is much more significant. For instance, there is a ratio of $\mathcal{O}(10^{-6})$ between the solar tidal potential and those taken into account herein. Therefore, the solar gravitational tide and the tides due to other planets and moons are neglected \citep{sagan1982tide}. The period of the astronomical forcing is 1 Titan day (TD), resulting in a diurnal tide exhibiting no spring-neap tide cycle. The forcing is given by the horizontal gradient of the tidal potential. The latter is given by the sum of two contributions: the potential due to Titan's orbital eccentricity (see Eq.~\ref{eq:forcing_ecc}) and the potential due to Titan's obliquity (see Eq.~\ref{eq:forcing_ob}). The eccentricity potential is \citep{dermott1995tidal}
\begin{equation*}
\phi_{ecc} = -\frac{GM_s}{a} \left(\frac{R_T}{a}\right)^23e
\end{equation*}
\begin{equation}
\left(0.5(3\sin^2\theta\cos^2\lambda-1)\cos(nt)+\sin^2\theta\sin(2\lambda)\sin(nt)\right)
\label{eq:forcing_ecc}
\end{equation}
where $G=6.67259\times10^{-11}$ m$^3$/s$^{2}$/kg is the universal gravitational constant, $M_s=5.685\times10^{26}$ kg is Saturn's mass, $a=1.221865\times10^9$~m is the semi-major axis of Titan's orbit, $R_T=2,574,730$ m is Titan's mean radius, $e=0.0288$ is Titan's orbital eccentricity, $n=4.5601\times10^{-6}$ s$^{-1}$ is Titan's orbital angular velocity, $t$ is the time measured from perikron (point on Titan's orbit closest to Saturn), $\theta$ is the colatitude and $\lambda$ is the longitude. The potential due to the obliquity is \citep{tyler2008strong}
\begin{equation}
\phi_{ob} = \frac{3}{2}n^2R_T^2\theta_0\sin\theta\cos\theta\left(\cos(\lambda-nt)+\cos(\lambda+nt)\right)
\label{eq:forcing_ob}
\end{equation}
where $\theta_0=5.34\times10^{-3}$ is the obliquity of Titan expressed in radian. The forcings resulting from Titan non synchronous rotation (NSR) have a much smaller magnitude. Consequently, the NSR forcings are disregarded when studying the tides of Titan's liquid bodies.

Due to the global subsurface ocean, the deformation of the icy shell and, hence, of the lakes and seas' bottom is much larger than the shape variation of a rigid Titan. Such deformation modifies the tidal forcing to which surface liquids respond, since the ice crust itself partly follows the changing potential. The uplifting of Titan surface reduces the tidal potential acting on the surface lakes and seas by $h_2\left(\phi_{ecc}+\phi_{ob}\right)$, where $h_2$ is the second degree radial displacement Love number, while the additional gravitational potential arising from the mass redistribution increases the potential by by $k_2\left(\phi_{ecc}+\phi_{ob}\right)$, where $k_2$ is the second degree tidal Love number. These effects are modelled by an attenuation factor, $\gamma_2= 1+\Re(k_2)-\Re(h_2)$ ($\Re()$ denotes the real part), applied to the tidal forcing \citep{lorenz2014radar,tokano2014numerical,beuthe2015tides,beuthe2015tidal}, i.e. the forcing really applied to the surface lakes and seas is $\gamma_2\nabla_h\left(\phi_{ecc}+\phi_{ob}\right)$ where $\nabla_h$ is the horizontal del operator. The Love numbers depend on the internal structure of Titan and are particularly sensitive to the thickness, the rigidity and the rheological properties of the ice shell \citep[e.g.][]{sohl2003interior,lorenz2014radar,tokano2014numerical}. As a consequence, $\gamma_2$ should be considered as a free parameter \citep{lorenz2014radar}. \citet{iess2012tides} and \cite{durante2019titan} derived some values of $k_2$ from Cassini measurements by means of three analysis models while \citet{sohl2003interior, sohl2014structural, baland2014titan, lefevre2014structure,beuthe2015tides, beuthe2015tidal,mitri2014shape}, among others, used models to reconstruct Titan's internal structure and compute the second degree Love numbers. As a result, different values can be found depending on the working hypotheses and the model used (see Table~\ref{tab:intro:gamma}).

\begin{table}
\vspace{0.7cm}
\centering
\begin{threeparttable}
\caption{Real part of the Love numbers ($\Re(k_2)$ and $\Re(h_2)$) and attenuation factor ($\gamma_2$) corresponding to the presence of a subsurface ocean (as established by \citet{sohl2003interior, sohl2014structural, baland2014titan, lefevre2014structure, beuthe2015tidal} from observations and measurements) found in the literature. $k_2$ and $h_2$ respectively are the second degree tidal potential Love number and the second degree radial displacement Love number. \label{tab:intro:gamma}}
\begin{tabular}{p{4.3cm}p{2.3cm}ll} \hline\noalign{\smallskip}

{Authors}&{$\Re(k_2)$}& {$\Re(h_2)$}&{$\gamma_2$} \\ \noalign{\smallskip}\hline\noalign{\smallskip}
 {\citet{sohl1995tidal}} &{$0.36$}&{$1.19$}& {$0.17$} \\ 
{\citet{sohl2003interior}$^a$} &{$[0.39,0.32]$}&{$[1.25,1.05]$}&{$[0.13,0.27]$} \\ 
{} &{$[0.39,0.35]$}&{$[1.28,1.15]$}& {$[0.12,0.2]$}\\ 
{\citet{nimmo2010shell}} &{$-$}&{$1.28$}& {$-$}\\ 
{\citet{iess2012tides}$^b$} &{$0.589\pm0.075$}&{$-$}& {$-$}\\ 
{} &{$0.67\pm0.09 $}&{$-$}& {$-$} \\ 
{} &{$0.637\pm0.112 $}&{$-$}& {$-$} \\
{\citet{durante2019titan}} &{$0.62\pm0.067$}&{$-$}& {$-$}\\ 
{\citet{sohl2014structural}} &{$0.437$}&{$1.29$}& {$0.147$}\\
{\citet{beuthe2015tidal}$^c$} &{$0.7-0.5$}&{$1.7-1.3$}& {$0-0.2$}\\
{} &{$0.7-0.41$}&{$1.7-1.2$}& {$0-0.21$}
 \\\noalign{\smallskip}\hline
\end{tabular}
\begin{tablenotes}
\small
\item $^a$: Various ice thickness and two ammonia concentrations.
\item $^b$: Different analysis models.
\item $^c$: Two crust densities and various relative thickness of the ice shell.
\end{tablenotes}
\end{threeparttable}
\end{table}

 \citet{beuthe2015tidal} predicted values of $k_2$ matching with those computed by \citet{iess2012tides}. By setting the ice shell relative thickness in order to correspond with the value of $k_2$ obtained by \citet{iess2012tides,durante2019titan}, the corresponding attenuation factors range from 0 to 0.21 depending on the density of the crust and subsurface ocean \citep[See Fig.~5 in][]{beuthe2015tidal}. \citet{tokano2014numerical} computed $\gamma_2$ for different internal structures and only retained those which satisfy the moment of inertia and the Love number $k_2$ predicted by \citet{iess2012tides}, which results in $\gamma_2\in[0.1,0.2]$ depending on the ice shell thickness. However, the value retained from the moment of inertia was obtained by having recourse to the hydrostatic assumption, which is questionable due to the significant degree-three signal observed in the gravity field \citep{hemingway2013rigid, lefevre2014structure, baland2014titan}. All these models assume a thin and homogeneous crust. However, the crust could be non-homogeneous. There could be some clathrates outgassing methane in the atmosphere from time to time \citep{tobie2006episodic}. Such clathrates at the base of the ice shell might have a smaller shear modulus which would cause larger deformation of the shell and the Love number $k_2$ would be larger \citep{rappaport2008can}.

\section{Numerical modelling} 
Some phenomena/mechanisms cannot be observed because they occurred in the past, at unattainable places, or at a too small or too large scale. In order to study them, scientists often have to resort to numerical modelling. Nevertheless, such an approach is not perfect as models are limited by their inherent nature. There are three main limitations: the hypotheses associated to the equations solved, the knowledge of the domain (initial and boundary conditions, composition, dimensions\dots), and the scales resolved. Indeed, for computer cost reasons, it is not possible to model each and every phenomenon that impacting the domain. For instance, an oceanic model which aims at modelling the global ocean circulation is not able to take into account small scale phenomena and has to model their impact on the larger scales by means of a closure scheme. Consequently, as useful as the models can be, it should be remembered they are only approximations of reality. « Essentially, all models are wrong, but some are useful » \citep{box1987empirical}.

Although numerical methods are increasingly used nowadays thanks to the creation of more and more powerful computers, these methods are not recent. In fact, one could trace them back to more than 2000 years ago. For instance, Babylonian astronomers, mathematicians in Seleucid Mesopotamia, and Greek astronomer Hipparchus (2nd century BC) used linear interpolation to fill the gaps in tables. These methods were on a rise with the introduction of calculus. Nevertheless, the computation relying on tables, it was often slow and tedious. Another rise in these methods was induced by the invention of modern computers. The first numerical modelling of geophysical fluid dates back to the 1940s \citep{cushman2011introduction}. \citet{hansen1956theorie} was the first to create a model able to simulate regional storm surges and tide propagation in a two-dimensional model. It took ten additional years before the first general circulation model was developed \citep{bryan1967numerical}. For about 50 years, increasingly complex models emerged but most of them are based on structured grids using finite difference schemes. Recent improvements mainly concern the parametrisation of unresolved scales and limitations induced by the use of structured grids. Indeed, these grids suffer from significant drawbacks. A significant limitation is the difficulty to locally improve the spatial resolution. Although nesting techniques have been developed, they have their own drawbacks such as the numerical diffusion introduced \citep[see, e.g.,][]{brown2016impact}. Structured grids also result in a staircase representation of the coastline which impacts the boundary currents. Such grids also suffer from the pole problem: the spatial discretization is higher near the poles, decreasing the maximum explicit time step.

In order to circumvent these limitations, new models based on unstructured meshes were developed. Such meshes allow to locally increase the resolution and are therefore better suited for capturing the complex geometry of the domains. They are also particularly well suited for spherical domains as they do not suffer from the pole problem. These models are not based on finite different schemes but rather use finite volume (FVM) or finite element (FEM) methods. The assets of these methods also lead to their main drawback: they have a lower computational efficiency due to the complexity of their topology. FVM are well suited for advection-dominated problems and are based on a low-order discretisation, resulting in a smaller computational cost than the FEM which are based on a high-order discretisation. Nevertheless, the latter allows a better representation of the solution by element and suffer from a smaller discretisation error.

Most of the modern models simulating large scale circulation or both small and large scale hydrodynamics are based either on FVM or FEM. Among the most known, a majority opts for the finite volume methods. This is the case of, among other, MPAS \citep[Model for Precision Across Scales\footnote{\url{https://mpas-dev.github.io,}}][]{ringler2013multi}, the model of Alfred Wegener Institute \citep{danilov2012two}, and MITGCM \citep[MIT General Circulation Model\footnote{\url{http://mitgcm.org}},][]{marshall1997finite} which has the particularity of having already been applied to an extra terrestrial context as part of the study of Saturn atmosphere \citep{afanasyev2018cyclonic}. The finite element methods are used by, among other, TELEMAC \citep[open TELEMAC-MASCARET\footnote{\url{http://www.opentelemac.org}},][]{hervouet2007hydrodynamics}, FESOM \citep[Finite Element Sea Ice-Ocean Model\footnote{\url{https://fesom.de},}][]{wang2008finite}, and SELFE \citep[Semi-implicit Eulerian-Lagrangian Finite Element\footnote{\url{http://www.stccmop.org/knowledge_transfer/software/selfe}},][]{zhang2008selfe}, which is dedicated to coastal flows. A detailed review of the popular coastal models and concepts used in the United States is available in \cite{fringer2019future}.

There are several finite element methods. Among them the discontinuous Galerkin (DG) is one of the most interesting as it combines the advantages of both FEM and FVM. Indeed, they are well suited for advection dominated problems (which is one of the advantage of FVM) \citep{bernard2007high}, locally conservative, they allow a high order accuracy, they benefit from the functional flexibility of FEM, and they do not require any stabilization. The discontinuity between the elements allows using hybrid meshes (i.e. with various element types or different degree of the basis functions \citep{karna2012Thesis}) and simplifies the multi-core parallel implementation \citep{seny2013multirate,seny2014efficient}. The DG method also suffers from drawbacks. For instance, the larger number of discrete unknowns results in a high storage and high computational cost. This method is not widespread: one can cite as example Thetis\footnote{\url{http://thetisproject.org/}} model \citep{karna2018thetis} or the model developed by \citet{conroy2016hp}. In this thesis, 
use and partially develop another model, SLIM, the Second-generation Louvain-la-Neuve Ice-ocean Model\footnote{\url{http://www.climate.be/slim}}.

\section{The SLIM model}
SLIM, a project started in 1999, originally aimed at modelling coupled ice-ocean dynamics. It has been modified over the years and the ice component is presently frozen as it now focuses on modelling geophysical and environmental flows. One reason behind the project was to develop a model using the finite element method to take advantage of the unstructured meshes. Consequently, the first version of the model solved the depth-averaged shallow water equations using the continuous Galerkin method on various finite element pairs \citep{hanert2003comparison,leroux2005some}. The model then evolved to use the discontinuous Galerkin method \citep{lambrechts2008multiscale}. Such methods are well suited for domains characterised by a complex geometry as unstructured meshes allows high topological flexibility and locally increasing the resolution. SLIM solves the one- (section-averaged), two- (depth-integrated) or three-dimensional hydrostatic shallow water equations under the Boussinesq approximation. Depending on the domain constraints, the user either solve the 1D, 2D or 3D equations. Section-averaged equations can be used to model the flow of river upstream parts showing small lateral variability such as the Scheldt \citep{gourgue2009flux,debrye2010finite, debrauwere2011residence,debrye2012water} and the Mahakam \citep{debrye2011preliminary,van2016simulations} rivers. 2D equations are able to predict the hydrodynamics of tidally-dominated flows such as the Great Barrier Reef \citep{lambrechts2008multi,thomas2014numerical, delandmeter2017submesoscale}, the Scheldt River and the Mahakam River but is not reliable in domains where there is a strong vertical shear or stratification \citep{delandmeter2015transport} or in domains where there is a steep bathymetry, strong exchange flows, and vertical stratification such as the Congo estuary \citep{lebars2016unstructured,vallaeys2018thesis,vallaeys2018columbia, vallaeys2020congo}. In such domains, the third spatial dimension is needed to properly model the hydrodynamics as more complex phenomena influence the flow. Consequently, a baroclinic version of SLIM, SLIM3D, has been developed. The equations are solved on prismatic elements extruded from an unstructured horizontal mesh \citep{white2008three,white2008tracer}. The model was then enhanced to use the discontinuous Galerkin method \citep{comblen2010discontinuous,blaise2010discontinuous} and was further developed by \citet{karna2013baroclinic}, \citet{delandmeter2018fully} and \citet{vallaeys2018thesis,vallaeys2018columbia, vallaeys2020congo}.

 Besides these hydrodynamic equations, SLIM can also solve the reactive transport equation \citep{gourgue2013depth,deBrauwere2014integrated} and has a Lagrangian particle tracker (LPT) module \citep{thomas2014numerical, critchell2016modelling}. Several wetting and drying algorithms were also developed to deal with tidal flats as part of the study of Mahakam river delta, Scheldt river estuary and Mekong river delta \citep{gourgue2009flux,karna2011fully,le2018new,le2019implicit}. Other estuaries \citep{lebars2016unstructured,vallaeys2018columbia,vallaeys2020congo}, coral reefs areas \citep{legrand2006high,lambrechts2008multi,frys2018fine}, which all have open boundaries, and enclosed domains, such as lake Tanganyika \citep{gourgue2007toward,gourgue2011free,delandmeter2018fully}, were also studied as part of the SLIM project. %With very few modifications, the depth-averaged model also predicts the tidal movement and circulations in the lakes on Titan, a moon of Saturn \citep{vincent_numerical_2016,vincent_numerical_2017}.

Recently, SLIM has been applied to a domain covering the Florida Strait from the Gulf of Mexico to the Atlantic ocean in order to study the corral reefs connectivity \citep{frys2018fine}. The depth-averaged version of SLIM is used to predict the hydrodynamics at large scale (the area is characterised by strong flows induced by the Gulf Stream) as well as at small scales (near the reefs and islands). The particle tracker module is used to predict the motion of corral larvae at and after spawn.

\section{Scope of the thesis}
\label{sec:intro:scope}
The objective of this thesis is as follows:
\vspace{0.3cm}
\begin{adjustwidth}{25pt}{25pt}
\textit{Using the scarce available data and a numerical model to improve our knowledge of the motion of surface and subsurface liquid bodies of Titan, thereby helping design future exploration missions of this moon.}
\end{adjustwidth}
\vspace{0.3cm}
Gaining further insight in the motion of Titan liquid bodies can only be achieved by means of numerical modelling as direct observations are impossible since Cassini's Grand Finale\footnote{Cassini's Grand Finale is the final diving of Cassini into Saturn's atmosphere during which it disintegrated like a meteor, on the 15$^{th}$ of September 2017.}. The motion predicted can be used by numerous scientists from various fields of research. For instance, the motion in the subsurface ocean is a key information for astrobiologists studying its habitability but it could also be used to gain further insight in the internal structure of Titan by allowing to validate hypotheses about the heat flux within the ocean. One reason for studying the liquid motion in the surface lakes and seas is to search for explanations of observed phenomena such as the "Magic Islands", shoreline variations, and solar glints. However, the main purpose is to provide some of the information requested for the planification and design of the future exploration mission(s). For instance, studying the lakes/seas temporal evolution of the heights and flows is useful for predicting areas to be avoided to prevent beaching of a landed spacecraft or for providing constrains for the design of a motorised exploration device. Indeed, the motor must be powerful enough to propel the device through the flow and wind encountered. The tidal flow can also be used to predict the path of a drifted capsule as a function of time and landing point by means of the Lagrangian particle tracker module.

 %\red{The 2D results could also be used as boundary conditions for a more detailed 3D model taking into account additional phenomena such as the surface heat fluxes. The 3D model could be used to study the liquid motion in key areas such as the straits linking different basins. Indeed, the velocity likely varies with the depth (both in terms of magnitude and orientation) and a submarine could take advantage of this situation to cross more easily the straits while a floating capsule could schedule its crossing to take advantage of the velocity orientation.}

The results could also be used to select/design the instruments of the future mission(s). The next observation mission, Dragonfly, will not focus on the surface lakes and seas as the timing correspond to winter, and hence darkness, in the northern hemisphere and direct to Earth (DTE) communication are impossible in the high latitudes during winter \citep{lorenz2018dragonfly}. DTE communication with a lander being possible at lower latitudes, Dragonfly will land at the same location as Huygens so the wind profile measured during Huygens descent is relevant. The concept of a rotocraft was selected to facilitate the displacements while taking advantage of the dense atmosphere and low gravity. This mission will study prebiotic chemistry and extra terrestrial habitability at various locations of Titan but Dragonfly is unlikely to observe the surface lakes and seas. Therefore, the conclusions reached in this thesis regarding the instruments does not apply to this mission. Depending on the exploration mission, different data could be collected. An orbiter would not be well suited to provide data allowing a validation of the numerical results. Indeed, the vertical resolution of instruments such as a radar is too coarse to capture the predicted variations of the sea surface elevation. While an orbiter could measure the elevation of a specific point of Titan's surface, it would be complicated to isolate the liquid motion in the lake/sea from the deformation of Titan's surface as both phenomena are induced by the same forcing (and have, hence, the same period) and the amplitude of the surface deformation is several order of magnitude larger than the tidal range. Furthermore, the liquid motion in the surface lakes and seas does not result from the sole tides but is also induced by poorly constrained atmospheric forcings. An accurate knowledge of these forcings would be needed to model properly the resulting liquid motion, hence allowing comparison between model results and observations. An orbiter could also draw an accurate plan of the shoreline variations induced in a lake/sea over one Titan day. To this end, a high spatial resolution instrument able to detect accurately the presence of liquid (i.e., able to distinguish dry and wet areas for a small column of liquid) is required. While high resolution images of the Earth where provided by satellites, several complications arise when using the same instrument(s) on Titan. For instance, Titan's atmosphere is opaque to some wavelength and the presence of methane in the atmosphere makes it harder to detect a small amount of liquid methane on the surface, which would decrease the spatial resolution of the instrument(s). Therefore, while an orbiter could measure the deformation of Titan's surface (which was achieved by Cassini), it would not be able to provide any data allowing validating the numerical results concerning the surface lakes and seas. To this end, a lander would be better suited. For instance, a floating capsule as well as a lander would be able to provide time series of the ocean elevation by means of a sonar/depth sounder, for the former, or a rangefinder, for the latter. A capsule could also measure the velocity field, liquid composition and atmospheric data such as the wind speed, air moisture, surface temperature \dots While data about the velocity field are not useless, it is much more common to validate a model by comparing surface elevation time series at specific location. Indeed, the velocity field is much more sensible to poorly constrained parameters affecting a numerical model such as local variations of the bathymetry, bottom friction, atmospheric forcing (for instance, the wind orientation could be influenced by a cliff or a hummock)~\dots Measurements of the atmospheric data would allow to improve our model but would be useless for validation.

As part of this thesis, SLIM is modified and applied to the liquid bodies of Titan. This numerical model is based on the physical laws valid on Earth. It takes advantage of the state-of-the-art knowledge of Titan's liquid bodies while taking into account the uncertainties affecting the parameters. Predictions of the liquid motion in the surface lakes and seas can also be found in the literature as similar works were conducted in parallel with this thesis. The results presented in this thesis stand out by the number of parameters taken into account as well as by the wide range of values studied. While it allows us to constrain the impact of each of these parameters on the liquid motion, it did not result in a better range of their values. Therefore, the results presented are also subject to uncertainties and a safety factor should be used to design an exploration device while taking into account the worst case scenario prediction.

Besides the main goal, additional tasks were conducted to improve SLIM as a free and user-friendly model. An application programming interface (API) has been developed in collaboration with Dr. Jonathan Lambrechts, Dr. Valentin Vallaeys, and Dr. Philippe Delandmeter and post-processing scripts were developed to help analysing results.

\subsection{Tides of the surface lakes and seas}
In order to study the tidal motion in the surface lakes and seas of Titan, minor modifications are introduced (see Section~\ref{sec:num:modifsw2d_lake}). We first pay attention to Ontario Lacus, the largest lake in Titan's southern hemisphere. It had previously been studied by \citet{tokano2010simulation}. Our study takes advantage from a better representation of the shoreline and the recent bathymetry maps derived by \cite{ventura2012electromagnetic} and \cite{hayes2016lakes}. Due to the uncertainties, a sensitivity study is conducted with respect to the bathymetry and the bottom friction, among other parameters. \cite{tokano2010simulation} and our results, presented in Chapter~\ref{chap:ontario}, definitely rule out the tidal motion as the main source of the observed shoreline variations.

Our focus then switches toward two of the northern seas, Ligeia and Kraken Maria, as they are a designated target for oceanographic exploration on Titan. Indeed, they are the largest liquid bodies lying on Titan's surface, hence reducing the risk of missing the target and landing on dry ground. Another reason for paying attention to these seas is the unexplained phenomenon referred to as "Magic Islands". Once again, several parameters (bathymetry, bottom friction, depth of the straits, attenuation factor \dots) are poorly constrained and a sensitivity analysis is thus conducted. Other aspects such as the liquid exchanges between the basins, the implications for future missions and the liquid motion in several specific areas (bay, canyons \dots) are discussed (see Chapter~\ref{chap:northern}).

\subsection{Normal modes of Ontario Lacus}
The tides being ruled out as a possible explanation for the shoreline variations, we focus on another hypothesis: a seiche could explain a significant variation of the surface elevation. To this end, we numerically study the barotropic normal modes of the lake using a mathematical approach based on \cite{bernard2008dispersion}. The method is well suited for domain with a complex geometry as it takes advantage of the geometrical flexibility of finite element methods. Nevertheless, it is impossible, using this approach, to predict the amplitude of the motion associated to the modes. Therefore, numerical simulations are conducted for arbitrary configurations likely to resonantly force some of the modes (see Chapter~\ref{chap:eigenmode}).

%Although some of the modes predicted could affect the elevation on large scales, they cannot be excited by the astronomical forcings identified. Atmospheric forcings such as wind could resonantly force the modes but it would require stormy events to reach the wind speed needed to generate elevation larger than 1 meter. 
\subsection{Tides of the global subsurface ocean}
Finally, the liquid motion within the global subsurface ocean is studied. Although several studies about the tidal deformation of Titan exist, none of them focuses on the liquid motion within the ocean in itself, the latter being considered as a layer among others or the authors focusing on the energy dissipation without paying attention to the motion in itself. The main difficulties are raised by the fact that the shell also deforms due to the astronomical forcing. Depending on its properties, its deformation differs from the surface elevation of a free surface ocean, causing interactions between the ocean and the ice shell. In order to conduct simulations on such an ocean, significant modifications are applied to SLIM (see Section~\ref{sec:num:modif_subsurface}): the numerical implementation is modified to be able to solve the equation on a sphere and terms are added to the equations to take into account the impact of the ice shell. Due to the uncertainties about the depth of the ocean and the rheology of the ice shell, a sensitivity analysis with respect to these parameters is conducted. We also briefly study the interactions between the tidal flow and the flow generated by the heat fluxes at the top of the ocean for two configurations (see Chapter~\ref{chap:ocean}). To this end, the baroclinic version of SLIM is modified similarly to its depth-averaged version (see Section~\ref{sec:num:modifsw3d}).

\clearpage

\section*{Supporting publications}

%The research carried out in the framework of this doctoral thesis has led to the following articles:

\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{D. Vincent}, O. Karatekin, V. Vallaeys, A.G. Hayes, M. Mastrogiuseppe, C. Notarnicola, V. Dehant, E. Deleersnijder, 2016, Numerical study of tides in Ontario Lacus, a hydrocarbon lake on the surface of the Saturnian moon Titan, \textit{Ocean Dynamics}, 66: 461--482, doi: 10.1007/s10236-016-0926-2

\item \textbf{D. Vincent}, O. Karatekin, J. Lambrechts, R. D. Lorenz, V. Dehant, E. Deleersnijder, 2018, A numerical study of tides in Titan's northern seas, Kraken and Ligeia Maria. \textit{Icarus}, 310: 105-126, doi: 10.1016/j.icarus.2017.12.018

\item \textbf{D. Vincent}, J. Lambrechts, O. Karatekin, T. Van Hoolst, R. H. Tyler, V. Dehant, E. Deleersnijder, 2019, Normal modes and resonance in Ontario Lacus: a hydrocarbon lake of Titan. \textit{Ocean Dynamics}, 69: 1121-1132, doi: 10.1007/s10236-019-01290-2

\item \textbf{D. Vincent}, J. Lambrechts, O. Karatekin, T. Van Hoolst, R. H. Tyler, V. Dehant, E. Deleersnijder, Modelling of Titan's subsurface ocean liquid motion. \textit{Icarus}, (In preparation)

\item C. Frys, A. Saint-Amand, M. Le Hénaff, J. Figueiredo, A. Kuba, B. Walker, J. Lambrechts, V. Vallaeys, \textbf{D. Vincent}, E. Hanert, High-resolution marine connectivity modelling in the Florida Coral Reef Tract, \textit{Frontiers in Marine Science} (Submitted)
\end{list}

\section*{Public presentations}

%The research carried out in the framework of this doctoral thesis has led to the following articles:

\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, 2015 Numerical simulations of tides in Ontario Lacus. European geoscience Union - General Assembly, Vienna. \\http://hdl.handle.net/2078.1/158392

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, 2016, Modelling of tidal flows between Titan's seas Kraken Mare and Ligeia Mare. Titan surface meeting, Paris.

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, 2016, Modelisation of tidal flows between Titan's seas Kraken Mare and Ligeia Mare. European geoscience Union - General Assembly, Vienna.\\
http://hdl.handle.net/2078.1/173566

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, Eric, 2017, Study of liquid exchanges between Titan's seas Kraken Mare and Ligeia Mare. European geoscience Union - General Assembly, Vienna. \\
http://hdl.handle.net/2078.1/185118

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, Eric, 2017, Modelling the tides of Titan's liquid bodies. TECLIM Seminar, Louvain-la-Neuve.

\item \textbf{D. Vincent}, O. Karatekin, J. Lambrechts, V. Dehant, E. Deleersnijder, 2018, Tides and eigenmodes of an idealized subsurface global ocean on Titan. European geoscience Union - General Assembly, Vienna. \\
http://hdl.handle.net/2078.1/196902

\item H. A. Le,X. H. D. Vu, J. Lambrechts, S. Ortleb, V. Vallaeys, \textbf{D. Vincent}, N. Gratiot, S. Soares Frazao, E. Deleersnijder, 2018, A new wetting – drying algorithm integrated in SLIM, with an application to the Tonle Sap lake, Mekong. European geoscience Union - General Assembly, Vienna.\\ http://hdl.handle.net/2078.1/196904

\end{list}

### 2022-01-04 chapters/intro/intro_old.tex
creation of thesis git

A common question in recent history is \textit{"Is there life anywhere else in the vastness of space?"}. There are countless books, films and songs about such an extraterrestrial life. Historically, the first planet suspected to shelter life was Mars (this is why the word "Martian" is often used to refer to extra-terrestrial life). Indeed, observations of canals on its surface in 1877 by the astronomer Giovanni Schiaparelli led to the outbreak of legends about an intelligent extraterrestrial life on Mars which would have dug those canals. Nowadays, the existence of life on Mars is almost completely ruled out -- at least on its surface -- the focus for searching life switched from Mars to other celestial bodies such as icy moons and exoplanets\footnote{An exoplanet is a planet that is not part of our solar system.}. As far as we know, one (among other) essential condition to the emergence of life is the presence of liquid. In the solar system, one can find substantial amount of liquid on various planets and moons: Earth, Enceladus, Europa, Titan, Ganymede ... and, as shown by recent observations \citep{orosei2018radar}, on Mars. Nevertheless, most of these celestial bodies hide the liquid beneath their surface: only two of them have a significant amount of liquid bodies on their surface. The first one is well known as we live there: the Earth. Few have heard about the second one: Titan, the largest moon of Saturn. It has both surface lakes and seas filled with liquid hydrocarbons and a global subsurface water ocean.

This thesis aims at studying the motion of Titan's liquid bodies (i.e. both the surface lakes and seas and the global subsurface ocean). The main purpose behind such studies is to suggest explanations for several observed phenomena such as shoreline variations, backscattering induced by lake surface roughness, appearance and disappearance of bright spot corresponding to solid ground on Cassini observations (the so-called Magic Islands) and surface deformation -- the latter being strongly influenced by the motion of the global subsurface ocean. On Earth, this can easily be done by means of numerical modelling -- whose results can be validated by measurements -- and \textit{in situ} or satellites observations. Nevertheless, on Titan, there is no way to get either \textit{in situ} measurements or new observations from any spacecraft as Cassini recently dived into Saturn at the end of its mission. Consequently, the only way to get new information about Titan in the next years will be by analysing "old" data with new method(s). For instance, M. Mastrogiuseppe, A. G. Hayes and their team successfully derived a bathymetry map of some lakes and seas with such methods \citep{hayes2016lakes,hayes2016bathymetry, mastrogiuseppe2014bathymetry,mastrogiuseppe2016radar,mastrogiuseppe2018bathymetry}) or using numerical models. Modelling tools are therefore expected to play an important role in order to gain further insights about Titan but also in order to help planning the next discovery mission(s) of Titan. Indeed, except for an orbiter, a lot of information are needed to be able to predict the landing site (depending on the starting point of the descent). Furthermore, depending on the locomotion, data such as the wind and/or the liquid velocity encountered could be needed in order to either engineer the motorization of the lander or to predict its position as a function of time.

%
\section{Titan} 
\begin{figure}[H]
\begin{center}
\includegraphics[trim=5cm 1.15cm 5cm 1.2cm,clip=true,width=.7\textwidth]{chapters/intro/figures/titan.jpg}
\captionof{figure}{\label{fig:intro:titan}View of Titan from the Cassini spacecraft (in the visible spectrum). The orange aspect of Titan is due to a thick haze mostly made of methane, ethane and nitrogen (image credit: NASA/JPL-Caltech/Space Science Institute).}
~ \\
\end{center}
\end{figure}
Titan was discovered in 1655 by the dutch astronomer Christiaan Huygens. The first spacecrafts to fly near Titan were Pioneer 11 and the Voyager I $\&$ II probes. Following their observations, a mission dedicated to the Saturnian system was launched in 1997: the Cassini mission. It was composed of a lander, Huygens, which landed on Titan on the 14$^{th}$ of January 2005 and an orbiter, Cassini, which orbited around Saturn from 11$^th$ of June 2005 to 15$^{th}$ of September 2017. At first, Cassini mission was supposed to last 4 years but due to happy circumstances, it outlived its initial duration by 9 years (see Fig.~\ref{fig:intro-chrono}). Various projects were proposed to study Titan in further details, Time~\citep{lorenz2012winds}, E2T~\citep{mitri2014exploration}, Titan sub~\citep{hartwig2016exploring}, Oceanus~\citep{sotin2017oceanus} and Dragonfly~\citep{turtle2017dragonfly} ... but none was launched so far. Dragonfly was recently selected by the NASA as part of its New frontiers program but the launch date is scheduled in 2026 and it will not focus on the lakes and seas.

\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{chapters/intro/figures/chronologie.png}
\captionof{figure}{\label{fig:intro-chrono}Chronology of Cassini mission. The original mission was first extended by about 2 years (the so-called Equinox mission) and then again by 7 years (the Solstice mission) (image credit: \citet{Manor2013Cassini}).}
~ \\
\end{center}
\end{figure}

Titan is the second largest moon in the Solar System, the largest one being Ganymede, a Jovian moon. Its orbital period around Saturn is 15.945 Earth days, which is close to its rotational period. Consequently, one can consider Titan as tidally locked as a first approximation. Indeed, the mean non synchronous rotation is of about $-0.024^\circ\pm0.018^\circ$/year \citep{meriggiola2016rotational}. This results in length-of-day (LOD) variations, polar motion and longitudinal libration due to the gravitational torque exerted by Saturn and the dynamic variation of the atmosphere \citep{coyette2018variations}. Nevertheless, the impact of such variations on the attraction forces is insignificant when looking at the tidal motion and one can hence consider Titan as tidally locked in this respect. Like the Earth, Titan is an ellispoid with semi-major and semi-minor axis given by $a =2575.15\pm0.02$ km, $b =2574.78\pm0.06$ km, $c =2574.47\pm0.06$ km \citep{zebker2009size}. A few characteristics of Titan are listed in Table~\ref{tab:intro:charac}. 
\begin{table}
\centering
\begin{threeparttable}
\caption{Main characteristics of Titan and a comparison with respect to the Earth. \label{tab:intro:charac}}
\begin{tabular}{ccc} \hline\noalign{\smallskip}

{}&{Titan}&{Earth} \\ \noalign{\smallskip}\hline\noalign{\smallskip}

{Orbital eccentricity}&{0.0288} &{0.0167}\\ 
{Obliquity [$^\circ$]}&{0.3} &{23.44}\\ 
{Average orbital speed [km/s]}&{5.57} &{29.78}\\ 
{Mean radius [km]} &{2574.73}&{6371} \\ 
{Volume[km$^3$]}&{7.16$\times10^{10}$} &{$1.08\times10^{12}$}\\ 
{Surface [km$^2$]}&{8.3$\times10^7$} &{$5.1\times10^8$}\\ 
{Mass [kg]}&{$1.3452\times10^{23}$} &{$5.972\times10^{24}$}\\
{Mean gravitational acceleration [m/s$^{2}$]}&{1.352} &{9.807}\\
{Mean surface temperature [K]}&{93.7} &{288}\\
{Surface pressure [kPa]}&{146.7} &{101.325}\\
 \\\noalign{\smallskip}\hline
\end{tabular}
\end{threeparttable}
\end{table}

\subsection{Atmospheric conditions}
\label{sec:intro:atm_cond}
Titan is the only natural satellite of the solar system which has a thick atmosphere\footnote{A thick atmosphere does not distinguish from a thin atmosphere by its total depth but by the gases that are present and their density. The larger the density is, the more thick the atmosphere.}. The presence of an atmosphere on Titan was first assumed by the astronomer Josep Comas i Sol\'a in 1903 and was confirmed in 1944 by Gerard Kuiper using a spectroscopic technique \citep{kuiper1944titan}. In 1980, Voyager I revealed that the surface pressure was about 1.48 times larger than on Earth (see Table~\ref{tab:intro:charac}). The major breakthrough regarding Titan's atmosphere came from the Cassini-Huygens mission which was able to measure, among other things, the composition and the atmospheric conditions. The only \textit{in situ} measurements of Titan atmospheric conditions were provided by the Huygens probe during its descent and at its landing site \citep{fulchignoni2005situ} while Cassini instruments such as the CIRS\footnote{Composite Infrared Spectrometer : it measures infrared energy from the surfaces, atmospheres, and rings of Saturn and its moons to study their temperature and compositions \citep{flasar2004exploring}.} and the VIMS\footnote{Visible and Infrared Mapping Spectrometer: it identifies the chemical compositions of the surfaces, atmospheres, and rings of Saturn and its moons by measuring colors of visible light and infrared energy emitted or reflected \citep{brown2004cassini}.} were able to provide additional information through remote sensing.
% thick atmosphere:https://sciencing.com/difference-between-thick-thin-atmospheres-12302390.html
Unlike Mars where surface water can be observed at large scales -- but only in a solid state -- methane was observed in both gaseous, liquid, and solid state on Titan's surface. This is due to the fact that a methane cycle similar to the water cycle on Earth takes place on Titan \citep{atreya2006titan}. Most of the liquid methane is located in the surface lakes and seas (see Section~\ref{sec:intro:liquidBodies}). The atmosphere is mainly composed of nitrogen with a small amount of methane and a still smaller amount of other species (e.g. $^{40}$Ar, $^{36}$Ar, cyanogen, ethane) \citep{niemann2005abundances}\citep[for detailed composition, see Table 1 in][]{cordier2009estimate}. The methane in the atmosphere can condense to form clouds \citep{toon1988methane,griffith2000detection,porco2005imaging} which can lead to methane precipitation\citep{toon1988methane}. In the high atmosphere, methane photolysis leads to the production of fast reacting radicals CH$_3$, CH$_2$, and CH. These radicals then react with atmospheric molecules such as nitrogen and hydrogen to form complex C$_x$H$_y$ (mostly ethane, C$_2$H$_6$) and nitrile species (C--N--H) \citep{atreya2006titan}.

\textit{In situ} measurements of Huygens probe gave the vertical pressure, temperature, precipitation, and wind speed profile along its descent but, in order to have a data map of these fields, one has to rely on models and remote measurements from Cassini spacecraft. The surface temperature can be inferred from CIRS far infrared spectra while pressure field, precipitation and wind speed have to be modelled. \citet{jennings2009titan} and \citet{cottini2012spatial} both derived the surface temperature from CIRS data respectively from 2004 (right after the northern winter solstice) to 2008 (late northern winter) and from 2004 to 2010 (in the early northern spring). Their results are consistent with each other and with Huygens measurements. Seasonal variations were observed by both authors. \citet{jennings2009titan} did not take into account the diurnal changes while \citet{cottini2012spatial} observed that the diurnal features disappeared at latitudes greater than $20-30^\circ$ in the north and $30-40^\circ$ in the south. Nevertheless, they took into account neither the data from latitudes greater than $60^\circ$ nor the local influence of liquid bodies. \citet{tan2015titan} established a formula describing the surface temperature as a function of the latitude and the time from January 2005 to December 2013 from \citet{jennings2011seasonal} and \citet{cottini2012spatial} temperature profiles. The values predicted by this formula have to be used with caution as no variation with respect to the longitude was taken into account.

Several global circulation models (GCMs) \citep[e.g.][]{tokano2008dune, tokano2009impact, friedson2009global, lebonnois2012titan,newman2011stratospheric,newman2016simulating, schneider2012polar} and mesoscale models \citep{soto2015mesoscale,charnay2015methane} were proposed in order to predict the atmospheric pressure, precipitation and winds. According to the GCMs, the atmospheric pressure variations are small at the surface and the winds are subject to significant spatial, seasonal and diurnal variations. Additionally, the wind speed may vary depending on the lake composition: if the seas are methane rich, the wind speed could be different than for ethane-rich seas \citep{tokano2009impact,lorenz2012winds}. The tidal wind (i.e. the wind resulting from the diurnal astronomical forcing) has neither a preferential direction nor a specific time distribution while the wind due to the convergence of moist air over the lake area is directed offshore and can be stronger than the tidal wind \citep{tokano2009impact}. Except in stormy conditions where it can reach p to 10 m/s, the predicted wind speed ranges from 0 to 2 m/s depending on the location and the season \citep{lorenz2012winds, lorenz2013oceanography}. Each of these models uses simplifying assumptions and none of them is fully consistent with observations \citep{schneider2012polar}.

\subsection{Surface lakes and seas }
\label{sec:intro:liquidBodies}
Due to the atmospheric conditions at Titan's surface measured by Voyager spacecrafts, the presence of liquid methane was evoked as soon as 1981 by, among others, \citet{samuelson1981mean,hanel1981infrared}. It led \citet{sagan1982tide} and \citet{dermott1995tidal} to construct a pre-Cassini appearance of Titan's surface (which was unknown at that time) by conducting analyses of theoretical global surface ocean and disconnected seas and lakes on Titan's surface. Other features were also predicted on Titan: crater lakes \citep{lorenz1994crater}, submerged impact craters \citep{lorenz1997impacts} and cryovolvanic flows with pillow lava texture \citep{lorenz1996pillow}. Although such flows and crater lakes were not observed, the presence of surface lakes and seas was confirmed by Huygens and Cassini measurements \citep{mcewen2005mapping,stofan2007lakes}. \textit{Cassini} first detected possible surface lakes in the southern polar region of Titan's surface in 2004 (only some years later confirmed to be liquid) by means of the imaging science subsystem\footnote{Imaging Science Subsystem: it takes pictures in visible, near-ultraviolet, and near-infrared light. \citep[see][]{porco2004cassini}.} (ISS) \citep{mcewen2005mapping}. Its Radar\footnote{Radar: it maps the surface of Titan using a radar imager to pierce the veil of haze. It is also used to measure heights of surface features. The synthetic aperture radar observed Titan in 13.78 GHz Ku-band with a resolution ranging from 0.35 to 1.7 km \citep[see][]{elachi2004radar}.} showed dark patches, interpreted as lakes, in the northern polar region on 22 July 2006, during the flyby T16 \citep{stofan2007lakes}. These lakes and seas\footnote{The largest liquid bodies are referred to as seas (Mare) and the others as lakes (Lacus), according to the nomenclature of the International Astronomical Union. It only reflects the size of the body.} are asymmetrically distributed with respect to the equator \citep{aharonson2009asymmetric}: there are many more lakes in the northern high latitudes where they are larger and deeper \citep{hayes2008hydrocarbon}. This asymmetry could be explained by the shorter and more intense southern summer induced by the obliquity of Titan and its orbit eccentricity.

Various studies were carried out about Titan's seas and lakes: bathymetry \citep[e.g.][]{hayes2010bathymetry, ventura2012electromagnetic, mastrogiuseppe2014bathymetry,mastrogiuseppe2018cassini,mastrogiuseppe2018bathymetry,lorenz2014radar, hayes2016lakes,hayes2017topographic}, liquid composition \citep[e.g.][]{brown2008identification, cordier2009estimate, glein2013geochemical, tan2013titan, tan2015titan, luspay2015experimental}, dissipation due to friction in Titan's lakes and seas \citep[e.g.][]{sagan1982tide, dermott1995tidal, sears1995tidal, lorenz2014radar}, tidal response \citep{tokano2010simulation, tokano2014numerical,vincent2016numerical,vincent2018numerical}, normal modes \citep{dermott1995tidal,tokano2010simulation,vincent2019normal} interactions with the atmosphere \citep{tokano2015wind, tokano2015sun}, and phenomena such as the shoreline variations of Ontario Lacus \citep{turtle2011shoreline,hayes2011transient} and transient features in the northern seas \citep{hofgartner2014transient, hofgartner2016titan}.

 The study of such lakes and seas belongs to a new scientific research field usually referred to as \textit{extra -terrestrial oceanography}. Indeed, although the governing equations are similar to those of terrestrial oceanography, there are significant differences that need to be taken into account: extra-terrestrial surface seas or lakes involve regimes of temperature, pressure, composition, and physical environment (gravity, tidal forces, rotational, and orbital periodicities, etc.) that are not observed in Earth's oceans.

\subsubsection{Surface lakes and seas distribution}
The surface lakes and seas of Titan cover $1.04\times 10^6$ km$^2$, which represents about $1.3\%$ of the moon surface. Most of them are located at latitudes higher than $60^\circ$ \citep{aharonson2009asymmetric} (see Fig.~\ref{fig:intro:map_north}) although a few equatorial and mid latitudes dark spots are proposed as possible lakes \citep{griffith2012possible}. 295 of these lakes, covering a combined surface of $1.36\times 10^5$ km$^2$, are empty while there are 577 liquid-filled lakes, covering a combined surface of $2.33\times 10^5$ km$^2$ \citep{hayes2016lakes}. Most of them consist in small sharp-edged depressions filled with liquid but there are also larger lakes such as Jingpo and Bolsena Lacus in the northern hemisphere and Ontario Lacus in the southern one. Besides these lakes stand three large Maria representing $80\%$ of the liquid filled surface: Kraken Mare ($5\times 10^5$ km$^2$), Ligeia Mare ($1.3\times 10^5$ km$^2$) and Punga Mare ($6.1\times 10^4$ km$^2$) \citep{hayes2016lakes}. While Kraken Mare and Ligeia Mare are hydrologically connected by a strait, namely Trevize Fretum, \citep{sotin2012observations} (see Fig;~\ref{fig:intro:map_north}) one can only assume a connection between Kraken and Punga Mare: the maria's surface elevation are similar \citep{hayes2017topographic,hayes2016lakes} but no large canals were observed. Such a connection would therefore be made of small canals, shallow swamps, or underground flows.

\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{chapters/intro/figures/map_bis.png}
\captionof{figure}{\label{fig:intro:map_north}Mosaic of Cassini radar images covering Titan northern hemisphere. (credit: NASA/JPL-Caltech/Agenzia Spaziale Italiana/USGS).}
~ \\
\end{center}
\end{figure}

 Kraken Mare is the largest sea identified on Titan, with an area of at least $400,000$ km$^2$. It is centred at ($68^\circ$N, $50^\circ$E) \citep{tokano2010simulation} and it stretches from $55^\circ$N to $82^\circ$N \citep{lorenz2014radar}. It is formed of two basins, Kraken 1 (in the north) and Kraken 2, linked by a strait named Seldon Fretum (see Fig;~\ref{fig:intro:map_north}) and a set of small channels \citep{lorenz2014radar}. Its horizontal dimensions are similar to those of the Gibraltar strait on Earth, it is about $17$ km wide and $40$ km long \citep{lorenz2014radar}, which is not the case of its depth. Ligeia Mare is located north-east of Kraken Mare: it is centred at ($79^\circ$N, $112^\circ$E). Its maximal dimensions are about $420$ by $350$ km \citep{stofan2012shorelines}. The Cassini Radar altimeter also detected liquid filled canyons linked to Ligeia Mare in the northern region, near Xanthus Flumen, and south-eastern regions, near Vid Flumina \citep{poggiali2016liquid} (see Fig.~\ref{fig:intro:map_north}). Transient features called "Magic Islands" were observed in Kraken Mare and Ligeia Mare \citep{hofgartner2014transient, hofgartner2016titan}. They consist in areas which appear bright on some observations and dark on others. Bright areas are consistent with solid ground while dark ones correspond to liquid. In Ligeia Mare, such phenomena were observed in regions centred at ($78^\circ$N, $123^\circ$E) and ($80^\circ$N, $111^\circ$E). They are best explained by ephemeral phenomena such as floating and/or suspended solids, bubbles, and waves while the diurnal phenomena, such as the tides, seem to be ruled out, the observations being all at nearly the same true orbital anomaly\footnote{The true orbital anomaly is the angle defining the position of Titan along its Keplerian orbit as seen from Saturn.} \citep{hofgartner2016titan}. In Kraken Mare, the phenomenon was observed at ($73^\circ$N, $55^\circ$E). In this sea, the same explanations stand but the tides cannot be ruled out. The third sea, namely Punga Mare, is also the smaller. It is located north of Kraken Mare, near the north Pole. It is centred at ($85.1^\circ$ N, $339.7^\circ$W) and has a diameter of 380 km.

Ontario Lacus is the largest lake in the southern hemisphere, covering approximately an area of 200 × 70 km \citep{wall2010active}. It is centred at ($72^\circ$S, $175^\circ$E). Ontario Lacus was flown over at least twice by each Cassini instrument (complete or partial flybys): ISS in 2004, 2005 $\&$ 2009, VIMS\footnote{Visible and Infrared Mapping Spectrometer: it identifies the chemical compositions of the surfaces, atmospheres, and rings of Saturn and its moons by measuring colours of visible light and infrared energy emitted or reflected \citep{brown2004cassini}.} in 2007 $\&$ 2009, and Radar in 2009 $\&$ 2010. It has the shape of a right footprint (see Fig.~\ref{fig:intro:ontario}) and is named after Lake Ontario (one of the Great Lakes of North America). It is connected to a hydrological network (for example, in area N in Fig.~\ref{fig:intro:ontario}) \citep{wall2010active, cornet2012geomorphological} which can provide liquid hydrocarbons during and after precipitations. Several, quite different, morphologies were observed along the shoreline of Ontario Lacus: deeply incised bays, mountainous region, beaches\dots \citep[see][for further details]{wall2010active, cornet2012geomorphological}. The two others significant lakes, Bolsena Lacus and Jingpo Lacus are located in the northern hemisphere. Bolsena Lacus is centred at ($76^\circ$N, $10^\circ$W) and has a diameter of 101 km while Jingpo Lacus has a diameter of 240 km and is centred at ($73^\circ$N, $336^\circ$W).

 \begin{figure}[H]
 \centering
 \includegraphics[width=8.4cm]{chapters/intro/figures/ontario.pdf}
 \caption{Map of Ontario Lacus. Letters from A to M show the study areas of \citet{hayes2010bathymetry} which derived a nearshore bathymetry in these regions. The red letters, from N to R, are used in this article to refer to particular areas (adapted from www.titanexploration.com; original image credit: Cassini Radar Science team, NASA/JPL/Caltech) \label{fig:intro:ontario}}
 \end{figure}

\subsubsection{Liquid composition}
Studies showed that the surface lakes and seas are not filled only with pure methane but also with ethane, nitrogen, and other low-molecular mass hydrocarbons \citep{ mastrogiuseppe2014bathymetry,luspay2015experimental,mitchell2015laboratory,tan2013titan,tan2015titan}. The exact liquid composition varies between the lakes/seas. For instance, Ligeia Mare is much more methane rich than Ontario Lacus or Kraken Mare \citep{hayes2016bathymetry,le2016composition}. The VIMS identified liquid ethane in Ontario Lacus in 2008 while the identification of liquid methane on Titan's surface by means of this instrument is made difficult by the presence of methane in the atmosphere \citep{brown2008identification}. Using Cassini Radar, \citep{mastrogiuseppe2014bathymetry,hayes2016bathymetry,le2016composition} were able to predict the composition of some lakes and seas such as Ontario Lacus and Ligeia Mare. This method relies on analysing the dielectric constant and the loss tangent\footnote{The loss tangent is the tangent of the loss angle. It is used to parametrise the electromagnetic energy dissipation inherent to a dielectric material.} of the lake and seas: the proportion of methane, ethane and nitrogen are derived from the value of these parameters. \citet{luspay2015experimental} and \citet{mitchell2015laboratory} conducted laboratory experiments to study liquid properties such as evaporation rates and derived rough approximation of the lakes and seas composition. All these studies were able to predict the proportion of the major constituent such as ethane and methane but were not able to predict a detailed composition of the lakes and seas. To this end, models have to be resorted. Several model are available in the literature \citep[see, e.g.,][]{cordier2009estimate,tan2013titan,tan2015titan,glein2013geochemical}. \citet{cordier2009estimate} and \citet{tan2013titan, tan2015titan} distinguished two types of lakes: the near-equator lakes and the high latitudes lakes. This distinction is the result of the temperature difference between the equator and the high latitudes. \citet{cordier2009estimate} established the composition by considering the lakes as non ideal solutions in thermodynamic equilibrium with the atmosphere. Nevertheless, this model is quite sensitive to uncertainties about thermodynamic data and other parameters: the relative standard deviations remain between $10\%$ and $300\%$ according to the species considered \citep{cordier2012titan}. \citet{tan2013titan} also considered this equilibrium and used an equation of state to model the chemical system of the atmosphere. \citet{tan2015titan} used the same model as \citet{tan2013titan} and included the effect of multicomponent mixtures. \citet{glein2013geochemical} developed a van Laar model parametrized by using experimental phase equilibrium data. \citet{cordier2009estimate} and \citet{luspay2015experimental} predicted an ethane-rich composition while \citet{tan2013titan, tan2015titan} predicted a methane-rich one. Nevertheless, \citet{tan2015titan} predicted that the composition varies seasonally: ethane and other heavy components would be present in larger quantities in winter than in summer. \citet{glein2013geochemical} also predicted a methane-rich composition but their results significantly vary with the mixing ratio of methane. If the mixing ratio of methane suggested by Voyager 1 was used, they would obtain an ethane-rich lake. The chemical models result in a liquid of similar density but there are some significant differences about the molecular viscosity: that predicted by \citet{cordier2009estimate} and \citet{luspay2015experimental} is larger than that predicted by \citet{tan2013titan} (respectively by a factor of 5 and up to 3). The lake also carries some solid particles like acetylene and high molecular weight organics produced by atmospheric photochemistry \citep[see e.g.][]{lorenz2010threshold, lorenz2011cyanide, lorenz2013oceanography} (\citet{tan2013titan} showed that tholins and acetylene could be present in Titan's lakes). These particles modify the lake properties. Therefore, density and viscosity cannot be accurately derived from the lake composition.

\subsection{Titan internal structure}
\label{sec:intro:depth}
Titan is an icy moon. This is a class of natural satellites whose surface is composed mostly of water ice. The layers beneath the surface vary from one moon to another. As shown in Fig.~\ref{fig:intro:interior}, there is a liquid body on Titan other than the surface lakes and seas: a global subsurface ocean filled with liquid water lies beneath the surface of the moon. The latter is mostly composed of water ice and is therefore referred to as an icy shell. \citet{stevenson1992interior} was the first to suspect the presence of such an ocean. Since then, the presence of a salt-rich or ammonia-rich water ocean is supported by several observations and measurements such as Titan's obliquity, its surface motion, its gravity field and its topography \citep{rappaport2008can,nimmo2010shell,iess2012tides,hemingway2013rigid,lefevre2014structure,baland2011titan,baland2014titan}
and has been widely studied \citep{beuthe2008thin,beuthe2015tides,beuthe2015tidal,fortes2012titan,sohl1995tidal,sohl2003interior,sohl2014structural}
Beneath this ocean lies a layer of high pressure water ice and, beyond this layer, the core. Although the layers forming Titan are clearly identified, uncertainties remain about their thickness and properties. For instance, there are discrepancies between the ice Young modulus inferred from Cassini measurements and that obtained in laboratory (i.e., on Earth). The models used to infer the size and properties of the internal layers are validated by comparison of properties inferred from Cassini measurements such as the mean moment of inertia (which was questioned by \citet{hemingway2013rigid}, \citet{baland2014titan} and \citet{lefevre2014structure}), the second-degree gravity field coefficients, and Titan's mean density. The depth predicted can significantly vary depending on the model used. The commonly predicted depth for the ocean ranges from several tens of kilometers to about 400 km. The ocean is dense: \citet{baland2014titan} predicted a density between 1275 and 1350 kg/m$^3$, \citet{sohl2014structural} studied two possible densities, 1150 and 1350 kg/m$^3$, while \citet{coyette2018variations} considered that the ocean density may vary from 950 to 1350 kg/m$^3$.

\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{chapters/intro/figures/interior.pdf}
\captionof{figure}{\label{fig:intro:interior}Internal structure of Titan %(credit: A. D. Fortes/UCL/STFC)
.}
~ \\
\end{center}
\end{figure}

\subsection{Tidal forcing}
\label{sec:intro:forcing}
There are numerous celestial bodies which can generate a tidal motion on Titan. Nevertheless, the forcing resulting from Titan's obliquity and its orbital eccentricity is much more significant. For instance, there is a ratio of $\mathcal{O}(10^{-6})$ between the solar potential and those taken into account herein. Therefore, the solar gravitational tide and the tides due to other planets and moons are neglected \citep{sagan1982tide}. The period of the astronomical forcing is 1 Titan day (TD), resulting in a diurnal tide exhibiting no spring-neap tide cycle. The forcing is given by the horizontal gradient of the tidal potential. The latter is given by the sum of two contributions: the potential due to Titan's orbital eccentricity (see Equation~\ref{eq:forcing_ecc}) and the potential due to Titan's obliquity (see Equation~\ref{eq:forcing_ob}). The eccentricity potential is given by \citep{dermott1995tidal}
\begin{equation*}
\phi_{ecc} = -\frac{GM_s}{a} \left(\frac{R_T}{a}\right)^23e
\end{equation*}
\begin{equation}
\left(0.5(3\sin^2\theta\cos^2\lambda-1)\cos(nt)+\sin^2\theta\sin(2\lambda)\sin(nt)\right)
\label{eq:forcing_ecc}
\end{equation}
where $G=6.67259\times10^{-11}$ m$^3$/s$^{2}$/kg is the universal gravitational constant, $M_s=5.685\times10^{26}$ kg is Saturn's mass, $a=1.221865\times10^9$ m is the semi-major axis of Titan, $R_T=2,574,730$ m is Titan's mean radius, $e=0.0288$ is Titan's orbital eccentricity, $n=4.5601\times10^{-6}$ s$^{-1}$ is Titan's orbital angular velocity, $t$ is the time measured from perikron (point on Titan's orbit closest to Saturn), $\theta$ is the colatitude and $\lambda$ is the longitude. The potential due to the obliquity is \citep{tyler2008strong}
\begin{equation}
\phi_{ob} = \frac{3}{2}n^2R_T^2\theta_0\sin\theta\cos\theta\left(\cos(\lambda-nt)+\cos(\lambda+nt)\right)
\label{eq:forcing_ob}
\end{equation}
where $\theta_0=5.34\times10^{-3}$ is the obliquity of Titan expressed in radian. The forcings resulting from Titan non synchronous rotation (NSR) have much larger periods than the diurnal forcings above-mentionned. Furthermore, they have a much smaller magnitude. Consequently, the NSR forcings are disregarded when studying the tides of Titan's liquid bodies.\\

Due to the global subsurface ocean, the deformation of the icy shell and, hence, of the lakes and seas bottom is much larger than for a rigid Titan. Such deformation modifies the tidal forcing to which surface liquids respond, since the ice crust itself partly follows the changing potential. The lifting of Titan surface reduces the tidal potential acting on the surface lakes and seas by $h_2\left(\phi_{ecc}+\phi_{ob}\right)$ while the additional gravitational potential arising from the mass redistribution increases the potential by by $k_2\left(\phi_{ecc}+\phi_{ob}\right)$ where $h_2$ is the second degree radial displacement Love number and $k_2$ is the second degree tidal Love number. These effects are modelled by an attenuation factor, $\gamma_2= 1+\Re(k_2)-\Re(h_2)$, applied to the tidal forcing \citep{lorenz2014radar,tokano2014numerical,beuthe2015tides,beuthe2015tidal}, i.e. the forcing really applied to the surface lakes and seas is $\gamma_2\nabla_h\left(\phi_{ecc}+\phi_{ob}\right)$ where $\nabla_h$ is the horizontal del operator and $\Re()$ denotes the real part. The Love numbers depend on the internal structure of Titan and are particularly sensitive to the thickness, the rigidity and the rheological properties of the ice shell \citep[e.g.][]{sohl2003interior,lorenz2014radar,tokano2014numerical}. As a consequence, $\gamma_2$ should be considered as a free parameter \citep{lorenz2014radar}. \citet{iess2012tides} derived some values of $k_2$ from Cassini measurements by means of three analysis models while \citet{sohl2003interior, sohl2014structural, baland2014titan, lefevre2014structure, beuthe2015tidal,beuthe2015tides}, among others, used models to reconstruct Titan's internal structure and compute the second degree Love numbers. As a result, different values can be found depending on the working hypotheses and the model used (see Table~\ref{tab:intro:gamma}).

\begin{table}
\centering
\begin{threeparttable}
\caption{Real part of the Love numbers ($\Re(k_2)$ and $\Re(h_2)$) and attenuation factor ($\gamma_2$) corresponding to the presence of a subsurface ocean (as established by \citet{sohl2003interior, sohl2014structural, baland2014titan, lefevre2014structure, beuthe2015tidal} from observations and measurements) found in the literature. $k_2$ and $h_2$ respectively are the second degree tidal potential Love number and the second degree radial displacement Love number. \label{tab:intro:gamma}}
\begin{tabular}{lllllll} \hline\noalign{\smallskip}

{Authors}&{}&{$\Re(k_2)$}&{}& {$\Re(h_2)$}&{$\gamma_2$} \\ \noalign{\smallskip}\hline\noalign{\smallskip}
 {\citet{sohl1995tidal}}&{} &{$0.36$}&{}&{$1.19$}&{}& {$0.17$} \\ 
{\citet{sohl2003interior}$^a$}&{} &{$[0.39,0.32]$}&{}&{$[1.25,1.05]$}&{}& {$[0.13,0.27]$} \\ 
{}&{} &{$[0.39,0.35]$}&{}&{$[1.28,1.15]$}&{}& {$[0.12,0.2]$}\\ 
{\citet{nimmo2010shell}}&{} &{$-$}&{}&{$1.28$}&{}& {$-$}\\ 
{\citet{iess2012tides}$^b$}&{} &{$0.589\pm0.075$}&{}&{$-$}&{}& {$-$}\\ 
{}&{} &{$0.67\pm0.09 $}&{}&{$-$}&{}& {$-$} \\ 
{}&{} &{$0.637\pm0.112 $}&{}&{$-$}&{}& {$-$} \\
{\citet{sohl2014structural}}&{} &{$0.437$}&{}&{$1.29$}&{}& {$0.147$}\\
{\citet{beuthe2015tidal}$^c$}&{} &{$0.7-0.5$}&{}&{$1.7-1.3$}&{}& {$0-0.2$}\\
{}&{} &{$0.7-0.41$}&{}&{$1.7-1.2$}&{}& {$0-0.21$}
 \\\noalign{\smallskip}\hline
\end{tabular}
\begin{tablenotes}
\small
\item $^a$: Various ice thickness and two ammonia concentrations.
\item $^b$: Different analysis models.
\item $^c$: Two crust densities and various relative thickness of the ice shell.
\end{tablenotes}
\end{threeparttable}
\end{table}

 \citet{beuthe2015tidal} predicted values of $k_2$ matching with those computed by \citet{iess2012tides} from observations. By setting the ice shell relative thickness in order to correspond with the value of $k_2$ obtained by \citet{iess2012tides}, the corresponding attenuation factors range from 0 to 0.21 depending on the density of the crust and subsurface ocean \citep[See Fig.~5 in][]{beuthe2015tidal}. \citet{tokano2014numerical} computed $\gamma_2$ for different internal structures and only retained those which satisfy the moment of inertia and the Love number $k_2$ predicted by \citet{iess2012tides}, which results in $\gamma_2\in[0.1,0.2]$ depending on the ice shell thickness. However, the value retained from the moment of inertia was obtained by having recourse to the hydrostatic assumption, which is questionable due to the significant degree-three signal observed in the gravity field \citep{hemingway2013rigid, lefevre2014structure, baland2014titan}. All these models assume a thin and homogeneous crust. However, the crust could be non-homogeneous: there could be some clathrates outgassing methane in the atmosphere from time to time \citep{tobie2006episodic}. Such clathrates at the base of the ice shell might have a smaller shear modulus which would cause larger deformation of the shell and the Love number $k_2$ would be larger \citep{rappaport2008can}.

\section{Numerical modelling} 
Some phenomena/mechanisms cannot be observed because they occurred in the past, at unattainable places, at a too small or too large scale\dots. In order to study them,
scientists often have to resort to numerical modelling. Nevertheless, such an approach is not perfect as models are limited by their inherent nature. There are three main limitations: the assumptions associated to the equations solved, the knowledge of the domain (initial and boundary conditions, composition, dimension\dots), and the scales resolved. Indeed, for computer cost reasons, it is not possible to model each and every phenomenon that has an impact on the domain. For instance, an oceanic model which aims at modelling the global ocean circulation is not able to take into account small scale phenomena and has to model their impact on the larger scales by means of a closure scheme. Consequently, as useful as the models can be, it should be remembered they are only approximations of reality. « Essentially, all models are wrong, but some are useful » \citep{box1987empirical}.

Although numerical methods are increasingly used nowadays thanks to the creation of more and more powerful computers, these fields are not recent. In fact, one could trace them back to more than 2000 years ago (for instance, Babylonian astronomers, mathematicians in Seleucid Mesopotamia and Greek astronomer Hipparchus (2nd century BC) used linear interpolation to fill the gaps in tables. These methods were on a rise with the introduction of calculus. Nevertheless, the computation relying on tables, it was often slow and tedious. Another rise in these methods was induced by the invention of modern computers. The first numerical modelling of geophysical fluid dates back to the 1940s \citep{cushman2011introduction}. \citet{hansen1956theorie} was the first to create a model able to simulate regional storm surges and tide propagation in a two-dimensional model. It took ten additional years before the first general circulation model was developed \citep{bryan1967numerical}. For about 50 years, increasingly complex models emerged but all of them are based on structured grids using finite difference schemes. Recent improvements mainly concerns the parametrisation of unresolved scales and limitations induced by the use of structured grids. Indeed, these grids suffer from significant drawbacks. A significant limitation is the difficulty to locally improve the spatial resolution. Although nesting techniques have been developed, they have their own drawbacks \citep[see, e.g.,][]{hanert2004advection}. Structured grids also result in a staircase representation of the coastline which impacts the boundary currents. Such grids also suffer from the pole singularity problems: the north and south pole correspond to various nodes on the grids.

In order to circumvent these limitations, new models based on unstructured meshes were developed. Such meshes allow to locally increase the resolution and are therefore better suited for capturing complex geometry domains. They are also particularly well-suited to spherical domains as they do not suffer for the pole singularity problem. Such models are not based on finite different schemes but rather use finite volume (FVM) or finite element (FEM) methods. The assets of these methods also lead to their main drawback: they have a lower computational efficiency due to the complexity of their topology. FVM are well-suited for advection-dominated problems and are based on a low-order discretisation, resulting in a smaller computational cost than the FEM which are based on a high-order discretisation. Nevertheless, the latter allows for a better representation of the solution by element and suffer from a smaller discretisation error.

Most of the modern models simulating large scale circulation or both small and large scale hydrodynamics are based either on FVM or FEM. Among the most known, a majority opts for the finite volume methods. This is the case of, among other, MPAS \citep[Model for Precision Across Scales\footnote{\url{https://mpas-dev.github.io}}][]{ringler2013multi}, the model of Alfred Wegener Institute \citep{danilov2012two} and MITGCM \citep[MIT General Circulation Model\footnote{\url{http://mitgcm.org}}][]{marshall1997finite} which has the particularity to have already been applied to an extra terrestrial context as part of the study of Saturn atmosphere \citep{afanasyev2018cyclonic}. The finite element methods are used by, among other, TELEMAC \citep[open TELEMAC-MASCARET\footnote{\url{http://www.opentelemac.org}}][]{hervouet2007hydrodynamics}, FESOM \citep[Finite Element Sea Ice-Ocean Model\footnote{\url{https://fesom.de}}][]{wang2008finite}, and SELFE \citep[Semi-implicit Eulerian-Lagrangian Finite Element\footnote{\url{http://www.stccmop.org/knowledge_transfer/software/selfe}}][]{zhang2008selfe} which is dedicated to coastal flows.

There are several finite element methods. Among them the discontinuous Galerkin (DG) is one of the most interesting as it combines the advantages of both FEM and FVM. Indeed, they are well-suited for advection dominated problems (which is one of the advantage of FVM) \citep{bernard2007high}, locally conservative, they allow for a high order accuracy, they benefit from the functional flexibility of FEM, and they do not require any stabilization. The discontinuity between the elements allows for using hybrid meshes (i.e. with various element types or different degree of the basis functions \citep{karna2012Thesis}) and simplifies the multi-core parallel implementation \citep{seny2013multirate,seny2014efficient}. The DG method is not widespread: one can cite as example Thetis\footnote{\url{http://thetisproject.org/}} model \citep{karna2018thetis} or the model developed by \citet{conroy2016hp}. In this thesis, we use and partially developed another model, SLIM, the Second-generation Louvain-la-Neuve Ice-ocean Model\footnote{\url{http://www.climate.be/slim}}.

\section{The SLIM model}
SLIM, a project started in 1999, originally aimed at modelling coupled ice-ocean dynamics. It has been modified over the years and the ice component is presently frozen as it now focuses on modelling geophysical and environmental flows. One reason behind the project was to develop a model using the finite element method to take advantage of the unstructured meshes. Consequently, the first model solved the depth-averaged shallow water equations using the continuous Galerkin method on various finite element pairs \citep{hanert2003comparison,leroux2005some}. The model then evolved to use the discontinuous Galerkin method \citep{lambrechts2008multiscale}. Such methods are well-suited for domains characterised by a complex geometry as unstructured meshes allows for high topological flexibility and locally increasing the resolution. SLIM solves the one- (section-averaged), two- (depth-integrated) or three-dimensional hydrostatic shallow water equations under the Boussinesq approximation. Depending on the domain constraints, we either solve the 1D, 2D or 3D equations. Section-averaged equations can be used to model the flow of river upstream parts showing small lateral variability such as the Scheldt \citep{gourgue2009flux,debrye2010finite, debrauwere2011residence,debrye2012water} and the Mahakam \citep{debrye2011preliminary,van2016simulations} rivers. 2D equations are able to predict the hydrodynamics of tidally-dominated flows such as the Great Barrier Reef \citep{lambrechts2008multi,thomas2014numerical, delandmeter2017submesoscale}, the Scheldt River and the Mahakam River but is not reliable where there is a strong vertical shear or stratification \citep{delandmeter2015transport} or in domain where there is a steep bathymetry, strong exchange flows, and vertical stratification such as the Congo estuary \citep{lebars2016unstructured,vallaeys2018thesis,vallaeys2018discontinuous}
In such domains, the third spatial dimension is needed to properly model the hydrodynamics as more complex phenomena influence the flow. Consequently, a baroclinic version of SLIM, SLIM3D, has been developed. The equations are solved on prismatic elements extruded from an unstructured horizontal mesh \citep{white2008three,white2008tracer}. The model was then enhanced to use the discontinuous Galerkin method \citep{comblen2010discontinuous} and \citep{blaise2010discontinuous} and was further developed by \citet{karna2013baroclinic}, \citet{delandmeter2018fully} and \citet{vallaeys2018thesis}.

 Besides these hydrodynamic equations, SLIM can also solve the reactive transport equation \citep{gourgue2013depth,deBrauwere2014integrated} and has a Lagrangian particle tracker module \citep{thomas2014numerical, critchell2016modelling}. Several wetting and drying algorithms were also developed \citet{gourgue2009flux,karna2011fully,le2018new}
to deal with tidal flats as part of the study of Mahakam river delta, Scheldt river estuary and Mekong river delta \citep{le2018new,le2019implicit}. Besides these delta, estuaries, and coral reefs areas \citep{legrand2006high,lambrechts2008multi,frys2018fine}, which all have open boundaries, enclosed domains were also studied as part of the SLIM project. The first one was lake Tanganyika \citep{gourgue2007toward,gourgue2011free,delandmeter2018fully}. %With very few modifications, the depth-averaged model also predicts the tidal movement and circulations in the lakes on Titan, a moon of Saturn \citep{vincent_numerical_2016,vincent_numerical_2017}.

Recently, SLIM has been applied to a domain covering the Florida Strait from the Gulf of Mexico to the Atlantic ocean in order to study the corral reefs connectivity \citep{frys2018fine}. The depth-averaged version of SLIM is used to predict the hydrodynamics at large scale (the area is characterized by strong flows induced by the Gulf Stream) as well as at small scales (near the reefs and islands). The particle tracker module is used to predict the motion of corral larvae at and after spawn. %The mesh and the flow are shown in Fig.~\ref{fig:intro:florida}

%\begin{figure}
%\begin{center}
%\includegraphics[width=.9\textwidth]{chapters/intro/figures/florida.pdf}
%\captionof{figure}{\label{fig:intro:florida}Top row: Model computational domain with the bathymetry and location of the Lower, Middle and Upper Keys (left), and the unstructured mesh (right). The mesh contains about 10$^6$ elements and the resolution varies between about 100 m and 15 km. Close-up views of the mesh (middle) and snapshots of the currents (on September, 27 2010 at 22:00, bottom), on the Marquesas Keys (left) and the Lower Keys (right). This illustrates the benefits of unstructured meshes to represent the fine-scale details of the topography and hence simulate currents down to the scale of individual reefs (shown in grey) and islands (shown in black). Source: \citet{frys2018fine}}
%~ \\
%\end{center}
%\end{figure}

\section{Scope of the thesis}
In this thesis SLIM is applied to a new domain: the liquid bodies of Titan. The goals are to 
\BI
\item adapt SLIM to the specificity of an extra-terrestrial environment;
\item study the tides in the surface lakes and seas and their correlation to observed phenomena;
\item conduct sensitivity studies with respect to the poorly constrained parameters to improve our understanding of the mechanisms at work; 
\item study the normal modes of a lake using a mathematical approach adapted to DGFEM;
\item modify the model in order to solve the equations on a sphere;
\item modify the equations to take into account the impact of an ice shell lying at the top of the ocean;
\item study the liquid motion of such an ocean.
\EI
Besides these objectives, additional tasks were conducted to improve SLIM as a free and user-friendly model:
\BI
\item An application programming interface (API) has been developed in collaboration with Dr. Jonathan Lambrechts, Dr. Valentin Vallaeys, and Dr. Philippe Delandmeter;
\item Post-processing scripts were developed to help analysing results such as the decomposition of the sea surface elevation and flow velocity in periodic components which allows the user to get the tidal components or the tidal ellipses corresponding to a component.
\EI

\subsection{Tides of the surface lakes and seas}
In order to study the tidal motion in the surface lakes and seas of Titan, minor modifications are introduced: the mean gravity acceleration is modified and a momentum source term representing the astronomical forcing is added (see Section~\ref{sec:num:modifsw2d_lake}). We first pay attention to Ontario Lacus, the largest lake in Titan's southern hemisphere, covering approximately an area of 200 $\times$ 70 km \citep{wall2010active}. It was previously studied by \citet{tokano2010simulation} but this work was conducted in the ignorance of significant data such as the bathymetry (which was not available yet). Our study takes advantage from a better representation of the shoreline and the recent bathymetry maps derived by \cite{ventura2012electromagnetic} and \cite{hayes2016lakes}. This lake was brought to attention because possible shoreline variations were observed  \citep{lunine2009evidence,wall2010active,turtle2011shoreline,hayes2011transient} and, while several explanations were proposed, none could be asserted. Due to the uncertainties, a sensitivity study is conducted with respect to the bathymetry and the bottom friction, among other parameters. In order to facilitate the analysis of the results, several postprocessing tools are used and implemented in the API. The results, presented in Chapter~\ref{chap:ontario}, definitely rule out the tidal motion as the main source of these variations.

Our focus then switches towards two of the northern seas, Ligeia and Kraken Maria, as they are a designated target for oceanographic exploration on Titan. Indeed, they are the largest liquid bodies lying on Titan's surface, covering an area of at least 526,000 km$^2$ (hence reducing the risk of missing the target and landing on dry ground). Kraken Mare, Titan largest sea, is formed of two basins, Kraken 1 (in the north) and Kraken 2, linked by a strait named Seldon Fretum and a set of small canals \citep{lorenz2014radar}. Although Kraken and Ligeia Maria were first thought to be independent from each other, a strait linking them was latter discovered \citep{sotin2012observations}. Models are needed to predict the drift of a capsule as a function of time and landing site and/or the flow that could be encountered by a motorized vehicle, which is a key information for the design of the engine. Another reason for paying attention to these seas is the unexplained phenomenon referred to as Magic Islands. It was observed in both seas: transient features which seem to correspond to solid ground appeared in areas where liquid was previously observed \citep{hofgartner2014transient,hofgartner2016titan}. Once again, several parameters (bathymetry, bottom friction, depth of the straits, attenuation factor\dots) are poorly constrained and a sensitivity analysis is thus conducted. Other aspects such as the liquid exchanges between the basins, the implications for future missions and the liquid motion in several specific areas (bay, canyons\dots) are discussed (see Chapter~\ref{chap:northern}).

\subsection{Normal modes of Ontario Lacus}
The tide being ruled out as a possible explanation for the shoreline variations, we focus on another hypothesis: a seiche could explain a significant variation of the surface elevation. To this end, we numerically study the barotropic normal modes of the lake using a mathematical approach based on \cite{bernard2008dispersion}. The method is well-suited for domain with a complex geometry as it takes advantage of the geometrical flexibility of finite element methods. Nevertheless, it is impossible, using this approach, to predict the amplitude of the motion associated to the modes. Therefore, numerical simulations are conducted for arbitrary configurations likely to resonantly force some of the modes (see Chapter~\ref{chap:eigenmode}).

%Although some of the modes predicted could affect the elevation on large scales, they cannot be excited by the astronomical forcings identified. Atmospheric forcings such as wind could resonantly force the modes but it would require stormy events to reach the wind speed needed to generate elevation larger than 1 meter. 
\subsection{Tides of the global subsurface ocean}
Finally, the liquid motion within the global subsurface ocean is studied. Preliminary work consisted in predicting the tides of a free surface ocean. Although several studies about the tidal deformation of Titan exist, none of them focuses on the liquid motion within the ocean in itself, the latter being considered as a layer among others. The main difficulties is raised by the fact that the shell also deforms due to the astronomical forcing. Depending on its properties, its deformation differs from the surface elevation of the a free surface ocean, causing interactions between the ocean and the ice shell. In order to conduct simulations on such an ocean, significant modifications were applied to SLIM (see Section~\ref{sec:num:modif_surface}). Indeed, although the additional friction term can be modelled by doubling the friction coefficient, the surface pressure resulting from the shell has to be added to the equations and implemented. Our method is based on the membrane approach developed by \cite{beuthe2015tides,beuthe2015tidal} to study the deformation of Titan (see Section~\ref{sec:num:modif_eq_sphere}). Besides this additional term in the equations, the numerical implementation fo SLIM is also modified to solve the equations on a spherical domain (see Section~\ref{sec:num:mesh_sphere}). Although a method had been designed by Richard Comblen within the SLIM research team \citep[see][]{comblen2009finite}, it was not implemented in the current version of SLIM due to the successive rewritting of the code. Due to the uncertainties about the depth of the ocean and the rheology of the ice shell, a sensitivity analysis with respect to theses parameters is conducted. We also briefly study the interactions between the tidal flow and the flow generated by the heat fluxes between the bottom and the top of the ocean for two configurations (see Chapter~\ref{chap:ocean}). To this end, the baroclinic version of SLIM is modified similarly to its depth-averaged version (see Section~\ref{sec:num:modifsw3d}).

%\subsection{Additional developments in SLIM}
%In order to use SLIM, the model has to be adapted to Titan specific conditions. Concerning the surface lakes and seas, minors changes were needed: modifying the mean gravity acceleration and adding a momentum source term representing the astronomical forcing. More significant modifications were needed to study the liquid motion within the global subsurface ocean. Indeed, although a method was designed by Richard Comblen within the SLIM research team \citep[see][]{comblen2009finite}, the method was not implemented in the current version of SLIM. The equations also have to be modified to take into account the motion of the ice shell above the ocean. The model and the changes are detailed in Chapter~\ref{chap:numerics} along with the post-processing scripts implemented and the user friendly API.

%\subsection{Ontario Lacus}
% Ontario Lacus is the largest lake in Titan's southern hemisphere, covering approximately an area of 200 $\times$ 70 km \citep{wall2010active}. It was first studied by \citet{tokano2010simulation} but this work was conducted in the ignorance of significant data such as the bathymetry (which was not available yet). Our study takes advantage from the recent bathymetry maps derived \citep{ventura2012electromagnetic,hayes2016lakes} and a better representation of the shoreline. Besides the first article which focuses on the tidal motion, a second one was published where the normal modes of the lake are studied. Indeed, the first article showed that the observed shoreline variations \citep{lunine2009evidence,wall2010active,turtle2011shoreline,hayes2011transient} could not be explained by the tidal motion and we consequently study another hypothesis. The tidal motion and the normal modes of this lake are presented in Chapter~\ref{chap:ontario}and~\ref{chap:eigenmode}, respectively. Due to the scarcity of data, a sensitivity analysis with respect to parameters such as the liquid composition, the bottom friction and the bathymetry is conducted. ET GAMMA 2
% 
%\subsection{Ligeia and Kraken Maria} 
%
%In Chapter~\ref{chap:northern}, we apply the model to the two largest seas on Titan, Kraken and Ligeia Maria. Kraken Mare, Titan largest sea, is formed of two basins, Kraken 1 (in the north) and Kraken 2 linked by a strait named Seldon Fretum and a set of small canals \citep{lorenz2014radar}. Although they were first thought to be independent from each other, a strait linking them was latter discovered \citep{sotin2012observations}. Both seas cover an area of at least 526,000 km$^2$. An unexplained phenomena referred to as Magic Islands was observed in both seas: transient features which seem to correspond to solid ground appeared in areas where liquid was previously observed \citep{hofgartner2014transient,hofgartner2016titan}. These seas are also a designated target for Titan's surface lakes and seas exploration as they cover a wide area (and hence reduce the risk on landing on dry ground). Consequently, models are needed to predict the path of a capsule as a function of time and landing site and/or the flow that could be encountered by a lander, which is a key information in the design of the engine. Once again, a sensitivity analysis with respect to the bathymetry, the bottom friction, and the effect of Titan's surface deformation is conducted. The results are compared to those of \citet{tokano2014numerical} in order to validate our simulation.

%\subsection{Tides of the global subsurface ocean}
%Finally, the liquid motion within the global subsurface ocean is studied. Preliminary work consisted in predicting the tides of a surface free ocean, which had already been presented in the literature. Then, the model was modified in order to take the ice shell above the subsurface ocean into account (the implementation is detailed in Chapter~\ref{chap:numerics}). Indeed, the latter also deforms due to the astronomical forcing but, depending on the properties of the ice shell, this deformation could be different from the surface elevation of the ocean, causing interactions between the ocean and the ice shell. The results are discussed for various ocean depth and ice shell mechanical behaviour. Finally the interactions between the tidal motion and the liquid motion induced by a spatial varying boundary heat flux are studied.

\clearpage

\section*{Supporting publications}

%The research carried out in the framework of this doctoral thesis has led to the following articles:

\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{D. Vincent}, O. Karatekin, V. Vallaeys, A.G. Hayes, M. Mastrogiuseppe, C. Notarnicola, V. Dehant, E. Deleersnijder, 2016, Numerical study of tides in Ontario Lacus, a hydrocarbon lake on the surface of the Saturnian moon Titan, \textit{Ocean Dynamics}, 66: 461--482, doi: 10.1007/s10236-016-0926-2

\item \textbf{D. Vincent}, O. Karatekin, J. Lambrechts, R. D. Lorenz, V. Dehant, E. Deleersnijder, 2018, A numerical study of tides in Titan's northern seas, Kraken and Ligeia Maria. \textit{Icarus}, 310: 105-126, doi: 10.1016/j.icarus.2017.12.018

\item \textbf{D. Vincent}, J. Lambrechts, O. Karatekin, T. Van Hoolst, R. H. Tyler, V. Dehant, E. Deleersnijder, 2019, Normal modes and resonance in Ontario Lacus: a hydrocarbon lake of Titan. \textit{Ocean Dynamics}, 69: 1121-1132, doi: 10.1007/s10236-019-01290-2

\item \textbf{D. Vincent}, J. Lambrechts, O. Karatekin, T. Van Hoolst, R. H. Tyler, V. Dehant, E. Deleersnijder, Tides of Titan's subsurface ocean. \textit{Icarus}, (In preparation)

\item C. Frys, A. Saint-Amand, M. Le Hénaff, J. Figueiredo, A. Kuba, B. Walker, J. Lambrechts, V. Vallaeys, \textbf{D. Vincent}, E. Hanert, High-resolution marine connectivity modelling in the Florida Coral Reef Tract, \textit{Marine Ecology Progress Series} (In preparation)
\end{list}

\section*{Public presentations}

%The research carried out in the framework of this doctoral thesis has led to the following articles:

\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, 2015 Numerical simulations of tides in Ontario Lacus. General assembly 2015 of European geoscience union(EGU), Vienna. http://hdl.handle.net/2078.1/158392

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, 2016, Modelling of tidal flows between Titan's seas Kraken Mare and Ligeia Mare. Titan surface meeting, Paris.

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, 2016, Modelisation of tidal flows between Titan's seas Kraken Mare and Ligeia Mare. European geoscience Union - General Assembly 2016, Vienna.\\
http://hdl.handle.net/2078.1/173566

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, Eric, 2017, Study of liquid exchanges between Titan's seas Kraken Mare and Ligeia Mare. European geoscience Union General Assembly, Vienna. \\
http://hdl.handle.net/2078.1/185118

\item \textbf{D. Vincent}, O. Karatekin, V. Dehant, E. Deleersnijder, Eric, 2017, Modelling the tides of Titan's liquid bodies. TECLIM Seminar, Louvain-la-Neuve.

\item \textbf{D. Vincent}, O. Karatekin, J. Lambrechts, V. Dehant, E. Deleersnijder, 2018, Tides and eigenmodes of an idealized subsurface global ocean on Titan. European Geosciences Union General Assembly 2018, Vienna. \\
http://hdl.handle.net/2078.1/196902

\item H. A. Le,X. H. D. Vu, J. Lambrechts, S. Ortleb, V. Vallaeys, \textbf{D. Vincent}, N. Gratiot, S. Soares Frazao, E. Deleersnijder, 2018, A new wetting – drying algorithm integrated in SLIM, with an application to the Tonle Sap lake, Mekong. European Geosciences Union General Assembly 2018, Vienna. http://hdl.handle.net/2078.1/196904

\end{list}

### 2022-01-04 chapters/sctld/article.tex
creation of thesis git

\section{Introduction}

Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region  \citep{richardson1998florida, sutherland2004disease, aronson2001white, harvell2007coral, miller2009coral, brandt2009dynamics}. Indeed, the Florida Reef Tract (FRT), which was dominated by \textit{Acropora palmata} and \textit{Acropora cervicornis}, and often had ~30\% coral cover until the 1970s/80s \citep{dustan1987changes, porter1992quantification}, is now dominated by bare substrate, octocorals, and macroalgae with only approximately 5\% stony coral cover remaining \citep{ruzicka2013temporal}. The loss of the branching Acroporid species was attributed primarily to a disease outbreak, termed white band disease \citep{aronson2001white}, but several other threats such as habitat reduction, eutrophication, overfishing, hurricanes, and bleaching likely all contributed to these species decline \citep{acropora2005atlantic}. Subsequent losses of coral cover within the region were often linked to additional disease incidences and repeated regional coral bleaching events as a result of global climate change \citep{kuta1996abundance, richardson1998florida, sutherland2004disease, gardner2003long, aronson2006conservation, kuffner2015century, manzello2015rapid}. A novel coral disease outbreak, termed Stony Coral Tissue Loss Disease (SCTLD), is now threatening the last vestiges of coral throughout the Florida Reef Tract (FRT).

SCTLD was first documented off the coast of Miami-Dade County in the summer of 2014 by \cite{precht2016unprecedented} and has since spread throughout the entire FRT with the exception of the Dry Tortugas. To date, SCTLD has been observed affecting over 20 different stony corals species. A case definition of SCTLD has been compiled to describe the visual appearance and ecology of SCTLD \citep{noaa2018}. Briefly, the gross morphology of SCTLD is described as focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both. In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton. There is also a range in coral susceptibility to SCTLD, with species categorized as highly susceptible (e.g., \textit{Dendrogyra cylindrus}, \textit{Dichocoenia stokesii}, \textit{Meandrina meandrites}), moderately susceptible (e.g., \textit{Orbicella} spp., \textit{Montastraea cavernosa}, \textit{Siderastrea siderea}, \textit{Stephanocoenia intersepta}), or tolerant (e.g., \textit{Porites} spp., \textit{Acropora} spp.). Unfortunately, SCTLD has not remained isolated in the FRT and has now been recorded in Mexico \citep{alvarez2019rapid}, the US Virgin Islands \citep{blondeau2020coral} and several other locations around the Caribbean \citep{kramer2019map}. The continued persistence of the outbreak, the high number of species affected, and the large geographical range of reports consistent with the case definition suggests that SCTLD is the largest coral disease outbreak on record

Large-scale spatial epidemiologic analyses showed that the reefs in Florida with SCTLD are clustered, supporting a contagious mode of transmission \citep{muller2020spatial}. Similarly, aquaria-based experiments indicate SCTLD can be transmitted through direct contact or indirectly through the water column \citep{aeby2019pathogenesis} suggesting water can function as a SCTLD vector, at least within a controlled setting. The initial exponential increase in spread among reefs from the disease epicenter \citep{precht2016unprecedented} and the persistent subsequent linear rate of spread of SCTLD \citep{muller2020spatial}, north along South Florida reefs and south into the Florida Keys, indicates that water currents may play a role in disease transmission. Furthermore, the rate of spread, estimated at ~100 m per day, suggests surface currents are likely too fast to have spread SCTLD within the region. These results imply that either the middle layer or the bottom boundary layer, which are both significantly slower than surface currents, may be the vertical location in which transmission occurs \citep{muller2020spatial}. However, to date, there have been no attempt to link local hydrodynamic modeling efforts with the spatio-temporal dynamics of SCTLD in Florida.

Estimating the transport of the disease causative agent from reef to reef by currents cannot be performed empirically. However, experimentally-calibrated numerical models that simulate currents can provide a realistic picture of the dispersal of disease agents. Nonetheless, accurately modeling water circulation at the spatial scales that affect this dispersal remains a key challenge, as small-scale flow features such as recirculation eddies around reefs and islands strongly impact exchanges between reefs \citep{wolanski1994physical, burgess2007influence, figueiredo2013synthesizing}. In this context, models that can explicitly simulate flow features down to the reef scale are needed. This represents a spatial resolution of the order of 100-1,000 m in dense reef systems. As of today, most regional ocean models using traditional numerical methods cannot achieve such resolution because of the computational resources it requires. To our knowledge, the best resolution currently available among these models in the FRT is $\sim900$ m with the FKEYS-HYCOM model that has been developed for the Florida Keys region \citep{kourafalou2012florida, sponaugle2012observed, vaz2016perfect}. Unstructured-mesh ocean models offer a potential solution to this resolution issue by locally increasing the model resolution close to reefs and islands \citep{lambrechts2008multi, thomas2014numerical, thomas2015connectivity}, in order to focus the computational resources where they are most needed. High resolution bio-physical dispersal models can be used to build the potential connectivity between reefs and therefore approximate exchanges between colonies in the complex topography of the coral reef systems \citep{frys20}.

Marine diseases differ significantly from better studied terrestrial diseases, namely due to the potential for long environmental residence times, during which pathogens may survive and disperse through the water \citep{harvell2007coral, sokolow2009editor}. Several recent studies have attempted to adapt traditional epidemic models (Susceptible-Infectious-Recovered, or SIR models) to coral reef systems \citep{sokolow2009editor, bidegain2016microparasitic, bidegain2016marine}. Novel approaches have included developing pathogen pools \citep{bidegain2016microparasitic, bidegain2016marine}, and to model at the metapopulation scale, rather than at the scale of coral holobionts \citep{sokolow2009editor}. Both of these approaches are attempting to address the same issue: disease occurs between patches of entirely sessile animals, through the dispersal of pathogen(s). Thus, there are internal within-patch disease dynamics and metapopulation-scale between-patch dynamics occurring simultaneously. The epidemic model developed in the present study utilizes the same basic architecture of previous coral reef SIR models, but rather than assume pathogen pools (e.g. \cite{bidegain2016microparasitic, bidegain2016marine}) or ignore internal patch dynamics (e.g. \cite{sokolow2009editor}), we have modeled both within-patch disease dynamics and the dispersal of pathogenic agents explicitly using potential connectivity networks.

The objective of the present study is to deduce the probable propagation mechanism of the SCTLD throughout the FRT by developing an experimentally-calibrated epidemio-hydrodynamic model. With a resolution of about 100 m, this model can capture potential exchanges of disease-carrying material, further denominated as "infectious" in our modeling framework, between reefs that would be ignored by coarser models. By reproducing the observed spread of disease between 1st May 2018 and 1st April 2019, we provide insight on the characteristics of the disease agent and its vector. Ultimately, our model, coupled with lab and field studies, provide novel insight into the management of the epidemic, the identification of its causative agent, and mode of transmission.

% === METHODS === %
\section{Methods}

% --> Subsection 1\dan{SIR modeling background:}
\subsection{Modeling reef connectivity}
In the present study, we focused on the exchanges of infectious material between coral reefs driven by ocean currents, which therefore have to be accurately simulated. An ocean model should provide a realistic large-scale circulation while also resolving small-scale flow features down to the scale of individual reefs. In this context, we used the unstructured-mesh depth-integrated coastal ocean model \slim\footnote{\url{https://www.slim-ocean.be}} to simulate ocean currents over an area that includes the FRT but also the Florida Strait and part of the Gulf of Mexico (Fig. \ref{fig:setup}). By using an unstructured mesh, we increased the model resolution only over the FRT and hence concentrate computational resources where they were most needed. \slim has already been successfully applied in complex coastal systems such as the Great Barrier Reef \citep{lambrechts2008multi, thomas2014numerical} and is well suited to shallow-water flows. Details of the model formulation and validation are provided in \cite{frys20}.

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/sctld/figures/vel.jpg}
    \caption{Model computational domain and close-up views of the mesh with snapshots of the currents on May, 25 2018 at 00:00, for the Marquesas Keys (bottom) and the Lower Keys (top). This illustrates the benefits of unstructured meshes to represent the fine-scale details of the topography and hence simulate currents down to the scale of individual reefs (shown in light grey) and islands (shown in darker grey).}
    \label{fig:setup}
\end{figure}

The mesh resolution depended only on the distance to the coast, but we distinguished between the coastlines along the FRT where we imposed a maximum resolution of 100 m and the other coastlines along which the finest resolution was 2500 m. The mesh was generated with the open-source mesh generator GMSH \citep{Geuzaine2009} and is composed of approximately $7 \times 10^5$ elements. The coarsest elements, far away from the FRT, had a characteristic length size of about 10 km. Fig. \ref{fig:setup} depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as recirculation eddies and currents within the dense reef system in the Lower Keys that consist of many individual reefs with narrow passages in between.

The simulated currents were then used to model dispersal of disease agents throughout the FRT. In this study, 3 types of potential vectors carrying the disease causative agent were considered: positively buoyant (e.g. mucus and surfactant), neutrally buoyant (e.g. fines, pelagic organisms) and negatively buoyant (e.g. sediments, composites, demersal organisms). As SLIM is a depth-averaged model, the mean currents it generates are well suited to model the dispersal of neutrally buoyant material remaining within the water column. However, these currents must be modified to correctly represent the dynamics of material evolving in the surface and bottom boundary layers. Therefore, surface current response to winds was estimated by adding 1.5\% of the wind speed to SLIM currents with a stress-layer veering angle of 45$^\circ$ to the right for positively buoyant particles. Such parameterization is shown to be an accurate approximation of wave-induced Stokes drift and quasi-Eulerian surface currents by \cite{ardhuin2009observation}. For negatively buoyant material, bottom currents were obtained by taking 60\% of SLIM currents velocity with a veering angle of 15$^\circ$ to the left. This is an approximation based on observations of bottom currents and whole water column current profiles in the shallow waters ($<$15 m) of Hawk Channel in the middle Florida Keys by \cite{smith2009influence}, as well as observations obtained during the Atlantic Ocean Acidification Testbed project (Gramer, pers. comm.). This application is also consistent with the theory of current veering in the bottom Ekman layer, albeit that was previously observed in deeper (30-90 m) coastal waters, e.g., by \cite{perlin2007organization} and \cite{kundu1976ekman}.

Using these three velocity fields, virtual particles were then released on all the reefs composing the FRT to model the dispersal of material carrying the disease causative agent. The locations of the reefs of Florida were extracted from the "coral reefs and hardbottom" layer of the Unified Florida Reef Tract Map \citep{fwc2017unified}. The polygons of this reef map were then further divided into 500 m $\times$ 500 m squares in order to track the propagation of the disease with a finer geographical resolution, generating a total of 16,823 polygons. At the beginning of each simulated month and for each type of current, a total of about $1.5 \times 10^6$ particles were released over all the reef polygons. These particles had a state composed of their polygon of origin as well as their mass, that they lose at a constant rate $\gamma$ as they were moved by surface, mean or bottom currents. The value of $\gamma$ was chosen so that particles had a half-life of 30 days. When the particles were brought over a reef polygon by currents, the amount of disease mass that landed on the polygon was recorded in monthly potential connectivity matrices, whose entries are denoted $C_{ij}$. The matrix rows correspond to the source reefs and the columns correspond to the destination reefs. Hence $C_{ij}$ represents the mass of diseased material originating from sub-reef $i$ that had settled on sub-reef $j$. This matrix was then normalized by dividing each of its rows $i$ by the total mass of particles released on polygon $i$ in order to obtain the normalized potential connectivity matrix $\tilde{C}$, whose entry $\tilde{C}_{ij}$ gives the probability that disease agents produced on sub-reef $i$ settle on sub-reef $j$. Connectivity matrices were computed for each type of current and for each month of the simulated period.

These connectivity matrices are more easily handled by interpreting them as large graphs whose vertices are sub-reefs and whose edges represent connectivity pathways. They were analyzed using graph theory tools. In the present study, four potential connectivity measures were used to interpret the monthly computed graphs. These indicators are described in Table \ref{tab:indicator}. The first indicator is the weighted connectivity length (WCL), which gave the average dispersal distance from origin to destination for material produced on a given reef. The weighted connectivity of reef polygon $i$ is expressed as:
\begin{equation}
    \textrm{WCL}_i = \dfrac{\sum_j \tilde{C}_{ij} L_{ij}}{\sum_j \tilde{C}_{ij}}
\end{equation}
where $L_{ij}$ is the distance between origin reef $i$ and destination reef $j$. Another measure of the spreading potential of reef $j$ is its out-degree, \ie the product of the number of connections originating from reef $j$ by the quantity of disease agents it sent to the network. This indicator was obtained by computing the number of non-zero entries of row $i$ in the potential connectivity matrix $C$ and multiplying it with $\sum_j C_{ij}$. The information given by the out-degree was complemented by the fraction of disease agents produced on reef $i$ that successfully settled on a reef, called the fraction exchanged of reef $i$. This indicator is given by $\sum_{j} \tilde{C}_{ij}$. Finally, the isolation of reef $i$ in the network was given by its self-recruitment, \ie the fraction of disease agents settling on reef $i$ that originated from reef $i$, computed by $C_{ii}/\sum_jC_{ji}$. A large self-recruitment value indicates that little infectious material produced elsewhere settled on the reef and thus that it was isolated from the rest of the network. However, due to its relative nature, high self-recruitment could still be coupled with a large amount of exogenous diseased agents reaching the reef, although this quantity would be considered proportionally small in regard to the total mass of diseased material settling on the reef.

\begin{table}
    \centering
    \begin{tabular}{|p{3cm}|p{4cm}|p{3.5cm}|}
        \hline
        \textbf{Indicators} & \textbf{Description} & \textbf{What it shows} \\
        \hline
            Weighted connectivity length (WCL) & 
            Average dispersal distance from origin to destination reef for all disease agents released over a reef & 
            Average distance that disease agents from a reef travels successfully to another reef \\
        \hline
            Out-degree &
            Number of out-going connections originating from a given reef multiplied by the total mass exchanged  &
            Overall potential for a reef to spread the disease \\
        \hline
            Fraction exchanged &
            Fraction of disease agents produced on a given reef that settles on other reefs &
            Relative likelihood of successful spread of the disease to other reefs  \\
        \hline
            Self-recruitment &
            Fraction of disease agents settling on a given reef that has been released on the same reef &
            Relative likelihood of unsuccessful spread of the disease from other reefs \\
        \hline            
    \end{tabular}
    \caption{Indicators used to analyze the modeled exchanges of infected material for each considered type of currents and for each simulated month.}
    \label{tab:indicator}
\end{table}

% --> Subsection 2
\subsection{Epidemiological modeling}

\subsubsection{Model equations}

The spread of the SCTLD throughout the FRT was simulated using a connectivity-based \cite{kermack1927contribution} SIR model. SIR models are among the most standard epidemiological models. They divide individuals into three compartments: susceptible ($S$), infectious ($I$) and removed ($R$). When affected by the disease, susceptible individuals become infectious and infect other susceptible individuals until they are removed from the system, either through recovery or death. Such models usually rely on the hypothesis of an homogeneous, well-mixed population. To account for the spatial heterogeneity of the FRT, the basic SIR formulation was modified by considering the fractions of susceptible ($S_j$), infectious ($I_j$) and removed ($R_j$) corals of each sub-reef $j$. Although the pathogen of SCTLD has not been identified, studies suggest a contagious mode of transmission \citep{aeby2019pathogenesis,muller2020spatial}. Here, we use 'infectious' to denote disease agents that could be passed from individual to individual, which are responsible for causing disease signs. We note that the disease agents could be biological or non-biological in nature. In the present study's epidemiological model, individual reefs interact through the exchange of disease agents as represented by the different connectivity matrices. For each sub-reef $j$ and at any time, the following relations hold: $0\leq S_j,I_j,R_j\leq 1$ and $S_j+I_j+R_j=1$. The evolution of these fractions through time is governed by the following equations:
\begin{equation}
    \begin{aligned}
        \dfrac{dS_j}{dt} &= -\beta\sum_i\dfrac{A_i}{A_j}I_i\tilde{C}_{ij}S_j - \beta'(I_j)S_jI_j \\
        \dfrac{dI_j}{dt} &= \beta\sum_i\dfrac{A_i}{A_j}I_i\tilde{C}_{ij}S_j + \beta'(I_j)S_jI_j - \sigma I_j \\
        \dfrac{dR_j}{dt} &= \sigma I_j
    \end{aligned}\label{eq:epidemio}
\end{equation}
where $\tilde{C}_{ij}$ is the entry corresponding to reef pair $(i,j)$ of the normalized potential connectivity matrix [-], $A_i$ is the area of reef polygon $i$ [km$^2$], $\sigma$ is the mortality rate [day$^{-1}$], and $\beta$ and $\beta'(I_j)$ are the inter- and intra-reef disease transmission rates [day$^{-1}$], respectively. In this model, diseased corals of sub-reef $i$ can 'infect' corals of sub-reef $j$ if there is non-zero probability of disease agents exchange from sub-reef $i$ to sub-reef $j$, given by $\tilde{C}_{ij}$. Moreover, to account for coral resistance to the disease, the intra-reef transmission function $\beta'(I_j)$ has the shape of a smooth step function of the fraction of infectious corals $I_j$ and is expressed as follows:
\begin{equation}
    \beta'(I_j) = \dfrac{\beta_0'}{2}(1+\tanh[(I_j-I_0)/\tau]),\label{eq:beta}
\end{equation}
where $I_0$ is a threshold on the infection population above which intra-reef transmission becomes significant, and $\tau$ is a measure of the interval over which the transition from low to high transmission occurs. As long as the fraction of infectious corals on sub-reef $j$ is below $I_0$, the only infection mechanism taking place is connectivity-driven transmission at rate $\beta$. Once the threshold is approached, intra-reef transmission with rate $\beta'_0$ is activated. A larger value of threshold $I_0$ corresponds to a greater resistance of corals to the disease, and therefore a slower spread of the disease within reef $j$. Coral reproduction and natural (\ie non SCTLD-related) death rates are not taken into account in this model, which amounts to assume that they balance each other. For this study the same values were used for $\beta$ and $\beta'_0$.
%This assumption seems reasonable since coral populations of the FRT were declining prior to the emergence of the SCTLD outbreak.

\subsubsection{Calibration}

Transmission and removal parameters $\beta_0'$ and $\sigma$ were fitted to disease prevalence observations averaged over all colonies from 6 permanent monitoring sites in the Lower Keys to accurately simulate the temporal evolution of $S_j,I_j,R_j$ on each diseased reef polygon. Three focal reef sites were established in the lower Florida Keys, one offshore (Acer 17/18), one mid-channel (Wonderland), and one nearshore reef (N. Birthday). Sites were established in May 2018, when all colonies appeared healthy. Within each site, two 10 m $\times$ 10 m quadrats were established. Quadrats were generally set up from east to west although N. Birthday was established with one quadrat further north of the other two to better capture coral cover in the site. All coral colonies $>$ 10 cm in size were mapped using self-contained underwater breathing apparatus (SCUBA). Each coral was given an $(x,y)$ coordinate, identified to species, and maximum diameter was noted. After the initial data collection surveys, each site was visited every two to three weeks for rapid assessments to determine whether SCTLD was present. During these site visits, two divers conducted a visual assessment at each of the 6 quadrats. Disease was first observed in early October 2018. Detailed surveys were conducted every two to four weeks until December 2019. During the surveys, each individual coral was visually assessed for signs of SCTLD, including discoloration and tissue loss. Prevalence of diseased, apparently healthy, and dead were assessed for each time period. To relate our model framework to the compiled data, Eqs. $\ref{eq:epidemio}$ were simplified to a single-reef SIR model:
\begin{equation}
    \begin{aligned}
        \dfrac{dS}{dt} &= -\beta'_0SI \\
        \dfrac{dI}{dt} &= \beta'_0SI - \sigma I \\
        \dfrac{dR}{dt} &= \sigma I
    \end{aligned}\label{eq:simplified}
\end{equation}
Due to the low values of the entries in the normalized connectivity matrix $\tilde{C}_{ij}$, intra-reef transmission, when activated, is the dominant infection mechanism of Eqs. \ref{eq:epidemio}. Consequently, Eqs. \ref{eq:simplified} gave a reasonable approximation of the evolution of the disease on sub-reefs for which $I_j \geq I_0$. Using this approximation, the ratio $\beta_0'/\sigma$ was estimated by matching the modeled fraction of susceptible corals remaining after the disease activity had stopped ($S_\infty$) with observations. A standard property of a SIR model solution is such that
\begin{equation}
    S_\infty - \frac{\sigma}{\beta_0'}\log(S_{\infty}/S(t=0)) = 1\label{eq:ratio}
\end{equation}
where the initial fraction of susceptible corals ($S(t=0)$) was taken equal to $1-I_0$ (see for instance \cite{Murray07}). In the framework of Eqs. \ref{eq:simplified}, the ratio $\beta_0'/\sigma$ gave the value of the basic reproduction number $R_0$, defined as the average number of secondary cases produced by one infected individual introduced into a population of susceptible individuals \citep{keeling2007stochastic}. This number is used in epidemiological models to determine whether an emerging infectious disease can spread in a population ($R_0 > 1$) or not ($R_0 < 1$). The basic reproduction number that was obtained was then used to express $\sigma$ in terms of $\beta_0'$ and calibrate its value to reproduce, as well as possible, the temporal evolution of the colonies-averaged susceptible population shown in Fig. \ref{fig:calibration}.

\subsubsection{Initialization} \label{sec:init}

In order to solve Eqs. \ref{eq:epidemio}, initial conditions were needed, \ie fractions of susceptible, infectious and recovered corals at the beginning of the simulated period. This information was constructed from 9 different field-collected datasets: (i) Coral Reef Evaluation and Monitoring Project (CREMP; 2014–2017), (ii) CREMP Presence/Absence Data (CREMP P\_A; 2016–2017), (iii) Southeast Florida Coral Reef Evaluation and Monitoring Project (SECREMP; 2014–2017), (iv) Florida Reef Resilience Program Disturbance Response Monitoring (FRRP; 2014–2017), (v) Hurricane Irma Rapid Reef Assessment (IRMA; 2017, \cite{viehman2018}), (vi) the Southeast Florida Action Network citizen science program (SEAFAN; 2014–2017), and (vii) the Southefrn Coral Disease Margin field effort (2017 and 2018; \cite{neely2018surveying}), (viii) Mote Marine Laboratory’s Field operations data (2018-2019) and (ix) data compiled through the citizen science BleachWatch program (2018). Every dataset provided data on the presence or absence of the SCTLD (or tissue loss consistent with the SCTLD case definition) within each survey. Some also provided detailed disease metrics such as the species affected and the disease prevalence, which was subsequently compiled into presence/absence of SCTLD data by surveyed site. The locations of these observations are shown in Fig. \ref{fig:stns}. Using this information, we first delineated the diseased zone by constructing the concave hull of the points where the disease was observed before May 2018. The reefs diseased prior to the beginning of our simulated period were then defined as the reefs located inside the constructed zone. The time of disease observation was then spatially interpolated on each reef of the diseased zone by kriging with a Gaussian semivariogram using Python \texttt{pyKrige} module \citep{murphy2014pykrige}. Assuming an initial state $(S,I,R)=(1-I_0, I_0, 0)$ when the disease was observed, the fractions of susceptible, infectious and removed corals on each reef of the diseased zone on the 1st May 2018 was finally approximated using Eqs. \ref{eq:simplified}. Reefs outside of the diseased zone were initialized with an entirely susceptible population.

\begin{figure}
    \center
    \includegraphics[width=.8\textwidth]{chapters/sctld/figures/monitoring.jpg}
    \caption{Locations of the disease observations between 2014 and 2019 recorded in the data sets used in this study.}
    \label{fig:stns}
\end{figure}

\subsubsection{Computation of front speed}

\begin{figure}
    \centering
    \includegraphics[width=.99\linewidth]{chapters/sctld/figures/hull_example_2.jpg}
    \caption{Method used to compute the disease front displacement during a simulated time interval.\textbf{a.} Concave hull $H_0$ of the diseased sub-reefs at the beginning of the simulated period. \textbf{b.} Concave hull $H_1$ of the sub-reefs diseased during the simulated time interval. \textbf{c.} Arrow showing the computed front displacement during simulated time interval between $H_0$ and $H_2$, the union of $H_0$ and $H_1$.}
    \label{fig:hull}
\end{figure}

\cite{muller2020spatial} estimated the speed of the spreading STCLD epidemics at around 92 m/day in the southern section of the FRT. In order to assess our simulation results in regard to this value, we developed a methodology to compute the displacement of the disease front during a given time interval within our simulated period. First, the concave hull $H_0$ of the diseased polygons at the beginning of the time interval was delineated. Then the concave hull $H_1$ of the polygons diseased during the time interval was computed while the concave hull $H_2$ was defined as the union of $H_0$ and $H_1$. This methodology is illustrated in Fig. \ref{fig:hull}. The distance traveled by the disease front was then obtained by computing the maximum distance between all pairs of points of $H_0$ and $H_2$. The epidemic's front speed was finally obtained by dividing the resulting distance by the number of days in the simulated time interval.

\subsection{Sensitivity analysis}
Disease agents and their vector were assumed to have a half-life of 30 days in the water column, governed by parameter $\gamma$ while inter- and intra-reef infection were assumed to occur at the same rate $\beta$. Although the sensitivity of our model to these two parameters was not extensively investigated in the present study, some additional simulations were performed to estimate how variations of their value might impact our results. To assess the influence of mass loss rate $\gamma$, additional connectivity matrices were computed using particles with a half-life of 15 days ($\gamma$ increased with a factor 2) in our dispersal model. Connectivity indicators were then computed for these matrices and their value was compared with the ones derived from 30-day-half-life connectivity matrices. Moreover, to evaluate whether the 'infectiousness' of disease agent might be reduced during its journey from reef to reef, epidemiological model simulations were performed with reduced inter-reef 'infection' rate $\beta=\frac{1}{2}\beta_0'$.

\subsection{Transmission experiments}\label{sec:transmission}
In parallel to this modeling study, laboratory-based transmission experiments of SCTLD were conducted by several independent groups for various end points including transmission dynamics and samples for molecular and histological analysis. Requests for transmission data were sent to members of the ‘Transmission’ sub-group of the Florida Disease Advisory Committee’s ‘Research’ working group as well as any other additional researchers that may have been conducting transmission studies on SCTLD. Data that was requested and subsequently provided included the location, dates, and duration of the experiment, the species used as the diseased colony (donor of disease agents) and apparently healthy colony (exposed to disease agents), the number of successful transmissions as well as the 'incubation' period following a contact with disease agents prior to disease signs. Additional information included the size of the colonies used in the experiment, the percent tissue loss of the diseased (donor) colony at the beginning of the experiment, and whether the apparently healthy (exposed) fragment was touching the diseased colony or not.

The average probability of successful disease transmission was determined by taking the mean of the number of colonies exposed to the disease in each study divided by the total number of coral colonies exposed to diseased colonies. The ‘incubation’ period was identified as the average number of days after an apparently healthy coral colony was exposed to a diseased colony before visual disease signs occurred (i.e., active tissue loss). Only corals that eventually showed disease signs were integrated within the 'incubation' period calculation.

Data were provided from 8 different research groups representing 15 institutions and 19 total collaborators providing a total of 109 data points. After amalgamating the contributed data, the mean probability of transmission of SCTLD to an apparently healthy coral had a likelihood of approximately $44.8 \pm 3.6$ \%. The probability of transmission ranged from 0 to 100\% depending on the experiment. Additionally, the time between exposure of an apparently healthy coral to a diseased coral and subsequently showing initial signs of tissue loss (i.e., 'incubation' period) was $9.7 \pm 1$ days.

%%%%%%%%%%%%%%%%%%%
% --- RESULTS --- %
%%%%%%%%%%%%%%%%%%%
\section{Results}

\subsection{Exchanges of infected material}

Among the three modes of transport, bottom currents exhibited the lowest propagation range as they generated the networks with the smallest weighted connectivity length (Fig. \ref{fig:connect}). However, disease agents transported by bottom currents had the largest settlement success rate as these currents produced the graphs with the largest fraction exchanged. Therefore, bottom currents tend to transport more disease agents ro closer reefs compared to the two other modes of transport. Mean and surface currents, on the other hand showed similar spreading ranges with mean WCL of 20.63 km and 21.39 km respectively. However, the disease agents that surface currents transported had the lowest probability of settling successfully on reefs. Consequently, surface currents and bottom currents produced networks with similar mean out-degree, although surface currents have the potential to transport disease agents accross larger distances. Nonetheless, networks had larger median out-degree with bottom currents than with surface currents, which suggests that surface currents have a lower spreading potential than bottom currents. As a result of their large WCL and fraction exchanged, barotropic currents on the other hand exhibited the largest mean out-degree, which indicates that they have the strongest dispersal potential.

Self-recruitment gives the fraction of disease agents settling on a reef that was produced on the same reef. The greater the self-recruitment value, the more the reef was isolated from the rest of the network. Since diseased material is less likely to settle on isolated reefs, self-recruitment informs on the potential for the disease to reach a given reef, whereas all three other indicators inform on the reef spreading potential. As mentioned before, self-recruitment is a relative measure and it might thus not give a complete picture of the amount of diseased material coming from upstream reefs that settles on a given reef. This is particularly true when considering reefs with different areas and coral cover. However, in the present study, we assumed a uniform coral cover and divided all the reefs into square sub-reefs with the same area. It is therefore reasonable to assume that sub-reefs with high self-recruitment act as weak sinks for diseased material. Fig. \ref{fig:connect} shows that the disease was more likely to settle on the reefs of networks generated by mean currents. This result is consistent with the values of the other connectivity measures, as reefs tended to be more strongly connected with mean currents. On the other hand, reefs were more isolated with bottom currents, as they produced the graphs with lowest WCL and out-degree. Finally, surface currents generated larger self-recruitment values than mean currents as they exhibited the lowest fraction exchanged. Therefore, although bottom currents exhibited stronger spreading potential than surface currents, reefs were more likely to become diseased with surface currents.

% \edit{Although the SCTLD has not reached the Marquesas yet in 2018, a preliminary analysis of the potential for disease transmission from the Florida Keys to the Dry Tortugas (where no sign of the SCTLD has been observed to date) can be performed based on the connectivity results of the present study (see appendix \ref{section:appendix}).}

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/sctld/figures/connect_paper.jpg}
    \caption{Distribution of the indicators derived from the monthly connectivity matrices computed for each type of current during our simulated period. Mean values are indicated by white squares.}
    \label{fig:connect}
\end{figure}

\subsection{Epidemiological model results}

As aggregated observations showed a fraction of susceptible individuals of about $85\%$ after a year, a basic reproduction number $R_0=\beta'_0/\sigma=1.0345$ was found with Eq. \ref{eq:ratio}. Using this ratio, best fit to averaged disease prevalence observations was obtained with transmission rate $\beta_0'=\frac{1}{6.45}$ days$^{-1}$ and mortality rate $\sigma=\frac{1}{6.99}$ days$^{-1}$. Comparison of the evolution of the state described by Eqs \ref{eq:simplified} with observations is shown in Fig. \ref{fig:calibration}. Our model results accurately reproduced the observed fraction of susceptible individuals on colonies through time. However, the modeled fraction of removed individuals overestimated observations by about $5\%$. These discrepancies might be explained by the presence of "Unknown" values in our data sets as well as the simplifying assumptions of SIR models. Since 'infection' and removal occur at very close rates, the instantaneous fraction of diseased individuals on the reefs remained low through the outbreak, with a maximum value of about $0.4\%$.

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/sctld/figures/sir_obs.jpg}
    \caption{Disease prevalence averaged over all monitored sites over time as modeled by Eqs. \ref{eq:simplified} using calibrated transmission and removal parameters $\beta_0'=\frac{1}{6.45}~\text{days}^{-1}$ and $\sigma=\frac{1}{6.99}$ days$^{-1}$.}
    \label{fig:calibration}
\end{figure}

Once the model was calibrated, epidemio-hydrodynamic model simulations were performed from 1st May 2018 to 1st April 2019 for each type of current and different values of the 'infection' threshold $I_0$. Two metrics were used to assess the accuracy of the model. First, the modeled front speed was compared to the reference rate of 92 m/day derived by \cite{muller2020spatial}. Furthermore, we computed the mean of the distances between each point where SCTLD had been observed during our simulated period (extracted from the 2018-2019 data sets described in section \ref{sec:init}) and the centroid of the closest reef polygon predicted to be diseased by our model during the same period (Fig. \ref{fig:results}). Bottom currents produced the slowest modeled disease propagation with a maximum front speed of $\sim 20$ m/day, while simulations performed with surface currents spread the disease at a maximum speed of about 60 m/day. However, surface currents tended to propagate the disease to the north, rather than westward, along the Florida Keys. This explains why bottom currents predict disease occurrence closer to field observations despite exhibiting slower front speed. Finally, mean barotropic currents outperform other types of current predictions regarding both criteria with a front speed of 107 m/day and a mean geographical accuracy of $\sim1.2$ km. This suggests that the disease agents of SCTLD may be transported within neutrally buoyant material driven from reef to reef inside the water column by mean currents.

Moreover, Fig. \ref{fig:results} shows a strong dependence of the model results to 'infection' threshold $I_0$, that gives the fraction of diseased individual that colonies can withstand before exponential disease growth is triggered on the reef. Front speeds of both mean and bottom currents reached a plateau for values of the threshold between $I_0=0.05\%$ and $I_0=0.1\%$, while the minimal prediction error was reached around $I_0 \approx 0.078\%$ with mean currents. For $I_0 > 0.1\%$, intra-reef 'infection' was strongly impeded and populations of 'infectious' individuals on diseased reefs were not able to become sufficiently large to infect other colonies on connected reefs. For values of $I_0$ lower than $0.05\%$ on the other hand, intra-reef 'infection' dominated and coral populations on diseased reefs were removed too fast to efficiently spread the disease through the network. Since disease propagation throughout the FRT only occurred for fairly small values of $I_0$ in our model, corals are expected to have low resistance to the causative agent of the SCTLD.

\begin{figure}
    \centering
    \includegraphics[width=.8\textwidth]{chapters/sctld/figures/sctld_validation_corrected.jpg}
    \caption{Summary of epidemiological model simulations with calibrated transmission parameters. \textbf{Top:} Modeled disease front speed for each type of current with respect to intra-reef 'infection' threshold $I_0$. \textbf{Bottom:} Mean distance between predicted diseased reefs and observed disease points. These results show that mean barotropic currents outperformed other modes of transport at reproducing the observed spread of the disease. The appearance of a plateau suggests that the modeled spread of the disease occurs for a well-defined range of values of $I_0$.}
    \label{fig:results}
\end{figure}

The results shown in Fig. \ref{fig:results} were obtained by removing the large reef located North to Vaca key, denoted Vaca reef in \cite{frys20}, from our reef polygons. Preliminary simulations showed that this reef had close to no impact on the modeled spread of the disease to the rest of the FRT, as it sent very few disease agents to southerly and easterly neighboring reefs. Moreover, Vaca reef has a low coral coverage ($0-10\%$), which strongly impedes disease spread on the reef. However, as coral coverage was not taken into account in our epidemiological model, propagation of the disease on the reef was overestimated. This led to unrealistically strong modeled front speed variations due to the large size of the reef. Consequently, and in the absence of SCTLD observations on Vaca reef, this area was removed from our reef polygons in order to avoid overestimating the front speed.

% \begin{figure}
%     \centeringSubsection 1
%     \includegraphics[width=.9\textwidth]{figures/traj.png}
%     \caption{}
%     \label{fig:traj}
% \end{figure}

%\begin{figure}[h]
%    \centering
%    \includegraphics[width=.9\textwidth]{figures/hull_result.png}
%    \caption{Take home messages:\begin{itemize}
%        \item Our methods accurately captures disease front evolution through time (mtch between points and concave hulls)
%    \end{itemize}}
%    \label{fig:front}
%\end{figure}

\subsection{Sensitivity analysis}
Connectivity matrices computed using particles with a half-life of 15 days exhibited shorter range connectivity. However, the impact of increasing mass loss rate $\gamma$ in our dispersal model remained limited with variations in connectivity indicator values weaker than $10\%$. However, reducing inter-reef 'infection' rate $\beta$ by a factor 2 had a significant impact on epidemiological model outputs as predicted disease front speed did not exceed 20 m/day, far less than the observed speed of the disease front. This strong decrease can be explained by the interplay between inter- and intra-reef disease activity. Reducing inter-reef transmission rates decreases the fraction of diseased corals on reefs attained by disease agents, which in turn reduces the amount of disease agents sent to the rest of the network.

% === DISCUSSION === %
\section{Discussion and conclusions}

%% --- SUMMARY OF TAKE HOME MESSAGES --- %%

We developed an epidemio-hydrodynamic model to simulate the spread of the SCTLD throughout the entire FRT. Calibrating our model with colony-averaged prevalence observations, we estimated the species-averaged reproduction number $R_0$ to be slightly larger than one. Our model simulations suggest that barotropic currents best reproduce the observed spread of the disease among reefs in the lower Florida Keys. Bottom current did not spread disease agents far enough while surface currents did not allow disease agents to reside within the reefs long enough to ensure disease transmission that matched field observations. The results, therefore, suggest the causative agent of SCTLD is likely to be transported within neutrally buoyant particles in the water column. With this mode of transport, the propagation of the disease from reef to reef only occurred for a well-defined range of values of the 'infection' threshold $I_0$. This threshold is defined as the fraction of all sub-reef colonies that were diseased which triggered a rapid spread of the disease over the entire sub-reef area. Our results suggest that this occurred once $0.05-0.1\%$ of the colonies were diseased. These results suggest that, on average, corals appear to have a low resistance to SCTLD.

%% --- DISCUSSION OF TAKE HOME MESSAGES/MAIN RESULTS --- %%

% 1. MODEL PARAMETERS
After calibration, we estimated the species-averaged basic reproduction number $\beta_0'/\sigma$ to be equal to $1.0835$. This value, being close to 1, suggests modeled diseased individuals were removed from the system almost as fast as susceptible individuals became diseased. This situation caused the fraction of infectious corals on the reefs to remain fairly low (\ie $\leq 0.4\%$) through time. These results suggest that only a small fraction of the colonies caused the disease to spread on the reef during the outbreak. The observation-based species-averaged transmission period of 6.45 days used in the model was a reasonable estimation of the disease transmission dynamics, being of the same order of magnitude as the experimentally-derived mean incubation period of 9.7 days. The difference between the two values can be explained by field measurement uncertainties as well as the inability to perfectly mimic field conditions in laboratory. In the present study, the same values were used for inter- and intra-reef 'infection' rates $\beta$ and $\beta_0'$, \ie disease agents 'infectiousness' was assumed to remain unchanged during their journey from reef to reef. This assumption is consistent with sensitivity analysis results, which showed that reducing the value of $\beta$ strongly impeded disease propagation through the network. This outcome suggests that, to reproduce the observed spread, inter- and intra-reef transmission rates must have similar magnitude, \ie that the causative agent is not degraded while traveling from reef to reef. Given that the etiology of SCTLD is still unknown, the lack of degradation of the SCTLD causative agent over distances of tens of kilometers between reefs may help to narrow the search.

% 2. MEAN BAROTROPIC CURRENTS => NEUTRALLY BUOYANT VECTOR

The fact that mean barotropic currents outperformed the other modes of transport can be explained by considering the trajectories of the particles used to model the transport of the disease causative agent. Due to the impact of winds on positively buoyant material, particles driven by surface currents are likely to be blown away from the reefs. Moreover, even when winds are pushing particles along the reef line, these particles spend less time over the same region than particles driven by mean and bottom currents. Smaller amounts of particle mass will therefore settle on reef polygons, leading to lower connection strengths in the potential connectivity matrix, \ie lower exchange of infectious material between reefs. Hence, despite being able to transport the disease over greater distances, surface currents are less efficient to drive the propagation of the disease. Particles driven by bottom currents, on the other hand, remain over the same region longer, producing larger entries of the potential connectivity matrix. Due to these large exchange probabilities between reefs, bottom currents are better at propagating the disease (Fig. \ref{fig:results}). Nevertheless, because bottom currents are slower, exchanges of disease agents occur within a more limited geographical range. Mean barotropic currents that carry particles greater distances while allowing for sufficiently large amounts of diseased material to settle on reef polygons, were thus best suited to propagate the disease at a speed similar to field-collected data (Fig. \ref{fig:results}).

Since mean currents were the only mode of transport that successfully reproduced the observed propagation speed of the disease in our model, the disease causative agent is expected to be transported within neutrally buoyant material inside the water column. Current-driven propagation seems reasonable as water-borne transmission is an important spreading mechanism for multiple coral diseases, including white band disease, white plague disease, white pox disease, white syndrome disease, \textit{Porites} ulcerative white spots diseases, and skeletal eroding band disease \citep{shore2019modes}. These results suggest that coral mucus or sloughed off diseased coral tissue may directly transmit SCTLD if they remain suspended within mean currents. \cite{aeby2019pathogenesis} showed that corals held within aquaria can transmit SCTLD to healthy colonies suggesting that these mechanisms may be occurring within reefs. Additionally, vectors such as zooplankton, phytoplankton, or even marine snow, may be transporting disease agents from reef to reef within the water column. \cite{landsberg2018} used histology to show an association between the location of the in hospite symbiotic algae that reside within the coral gastrovascular cavity and the area of tissue necrosis occurring within SCTLD-affected corals. These results have led to the working hypothesis that SCTLD could be associated with the symbiotic algae or the process of heterotrophic feeding and ingestion. The causative agent could also be transported within fine sediments such as silt, as suggested by \cite{rosales2020rhodobacterales}. Such sediments are easily eroded in shallow areas around coral reefs and would therefore be mostly transported inside the water column by mean barotropic currents. This hypothesis might be tested by adapting the deposition rate $\gamma$ used in our experiments to be consistent with the sedimentation rate of silt. However, such modification of $\gamma$ would alter the entries of our potential connectivity matrices. Nonetheless, sensitivity analysis of connectivity indicators suggest that such modifications would have a minor impact on model behavior and that the main results of this study would remain valid for larger deposition rates.

% 3. IMPACT OF I0/APPEARANCE OF PLATEAU
Coral resistance to SCTLD was represented by parameter $I_0$, defined as the maximum fraction of the colony that can become diseased without causing the disease to spread to the rest of the sub-reef site. The plateau shown in Fig. \ref{fig:results} highlights the impact of this parameter on the modeled propagation of the disease. On the one hand, when corals were strongly susceptible to the disease, diseased individuals were removed from the system too fast to become sustainable sources of disease agents in the network. On the other hand, if corals were weakly susceptible to the disease, very few corals became diseased and the disease barely propagated. Our simulations suggest that this value of resistance must be fairly low (around 0.07\%) in order to successfully spread the disease throughout the FRT. This seems to imply that susceptible coral species have very weak defense mechanisms against the causative agent of the disease.

%% --- LIMITATION AND POTENTIAL IMPROVEMENTS OF THE MODEL --- %%
% 1. Limitations of hydro
As with any modeling study, it is important to understand the assumptions on which the model is based. Here, we have used a 2D barotropic ocean model forced by the 3D model HYCOM \citep{Chassignet2007} in order to indirectly represent baroclinic phenomena. Such a model is well-suited to simulate the fate of neutrally-buoyant material in shallow regions. However, as depth-averaged currents do not accurately approximate the motion of particles in the bottom and surface layers, they have been modified to simulate the exchanges of negatively and positively buoyant material. In order to derive bottom and surface currents from mean barotropic currents, we used parameterizations consistent with both observations and theory \citep{ardhuin2009observation,perlin2007organization, kundu1976ekman, smith2009influence}. Although such parametric estimation of surface and bottom currents adds additional uncertainty to those simulations, using a 2D model as the basis for the study allowed for reef-scale resolution throughout the whole FRT with available computational resources. Such high-resolution allowed the simulation to explicitly represent recirculation eddies around islands and reefs, that significantly impact the weighted connectivity length as well as the local retention of pathogenic material on the reefs \citep{frys20}.

% 2. Limitations of epidemiological model
The appearance of an interval of optimal values of threshold $I_0$ for the propagation of the disease in our results highlights the impact of coral resistance on the spread of SCTLD through the FRT. Therefore, a next step in our modeling approach would be further dividing coral populations of our polygons into highly susceptible (\eg \textit{Dichocoenia stokesii}, \textit{Meandrina meandrites}), intermediately susceptible (\eg \textit{Orbicella faveolata}, \textit{Montastrea cavernosa}), and weakly susceptible (\eg \textit{Acropora Palmata}, \textit{Acropora cervicornis}) sub-populations. The fractions of susceptible, infectious and removed individuals within these sub-populations would then be modeled with specific transmission ($\beta$, $\beta_0'$) and removal ($\sigma$) rates as well as specific infection thresholds $I_0$. Such approach would however require a detailed knowledge of the fine-scale distribution of the different coral species throughout the FRT. This knowledge about coral coverage could also be used to avoid overestimation of the front propagation, as in the case of Vaca reef.

%% --- NICE FINAL PARAGRAPH --- %%
Despite the limitations of its current formulation, our model brings unprecedented perspectives on the propagation mechanism of SCTLD throughout the FRT. Using a reef-scale spatial resolution, we determined the most probable mode of transport for the vector of the disease agent and deduced its host-species-averaged reproduction number based on prevalence observations. In addition, our model formulation provides a framework to quantify coral resistance to the disease.  This framework is a novel contribution to the study and modeling of marine diseases, as both inter- and intra-patch disease dynamics are modeled explicitly and realistically in time and space. As our model results are continuous through time, they can exhibit the variability of the propagation of SCTLD temporally and therefore bring additional insight into observation data. This study provides much-needed complementary insight into the identification of the causative agent of the SCTLD and the management of the crisis it generates. Furthermore, our modeling approach could be applied to other affected areas of the Caribbean, where there is still time to perform active management as the disease spreads throughout the region.

% === APPENDIX === %
% \appendix
% \section*{Appendix}
% 
% % --- MODEL FORMULATION --- %   
% \subsection*{Transmission data contributors}

% % --- MODEL EVALUATION --- %
% \subsection*{Subsection 2}

% \section*{Conflict of Interest Statement}
% The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.

% \section*{Author Contributions}
% TD developed the model, run the simulations and analyzed the results. EM, LG and EH conceptualized the study and designed the modeling experiments. EM collected the biological data. DH designed the epidemiological model. All authors contributed to the writing of the manuscript.

% \section*{Funding}
% This paper is a result of research funded by the Florida Department of Environmental Protection under award PO: B6A24 to Mote Marine Laboratory.

% \section*{Acknowledgments}
% Computational resources were provided by the Consortium des \'Equipements de Calcul Intensif (\textsc{c\'eci}), funded by the \textsc{f.r.s.-fnrs} under Grant No. 2.5020.11. Thomas Dobbelaere is a PhD student supported by the Fund for Research training in Industry and Agriculture (\textsc{FRIA}/\textsc{FNRS}).

% Field data was collected at the permanent SWG sites were collected by Sara Williams, Cory Walter, Joe Kuehl, Ray Bannister, and Erich Bartels. These sites were funded from "Florida's State Wildlife Grant, FWC Agreement no. 16007".

% Contributors listed in Tab. \ref{tab:contributors} provided data to the transmission study.
% \begin{table}[h!]
%     \center
%     \begin{tabular}{|l|l|}
%         \hline
%         \textbf{Contributors: Transmission Data} & \textbf{Institutions} \\ \hline
%         Erinn Muller$^\ast$     & Mote Marine Laboratory                 \\ \hline
%         Katie Eaton$^\ast$      & Mote Marine Laboratory                 \\ \hline
%         Jan Landsburg           & Florida Fish and Wildlife              \\ \hline
%         Yasu Kiryu              & Florida Fish and Wildlife              \\ \hline
%         Esther Peters           & George Mason University                \\ \hline
%         Ray Banister            & Mote Marine Laboratory/Florida Tech    \\ \hline
%         Valerie Paul            & Smithsonian Marine Station             \\ \hline
%         Blake Ushijima$^\ast$   & Smithsonian Marine Station             \\ \hline
%         Nikki Traylor Knowles   & University of Miami                    \\ \hline
%         Michael Studivan$^\ast$ & University of Miami/NOAA AOML          \\ \hline
%         Joshua Voss             & Harbor Branch Oceanographic Institute  \\ \hline
%         Greta Aeby$^\ast$       & Qatar University                       \\ \hline
%         Marilyn Brandt$^\ast$   & University of the Virgin Islands       \\ \hline
%         Adrienne Corea          & Rice University                        \\ \hline
%         Laura Mydlarz           & University of Texas - Arlington        \\ \hline
%         Dan Holstein            & Louisiana State University             \\ \hline
%         Amy Apprill             & Woods Hole Oceanographic Institute     \\ \hline
%         Tyler Smith             & University of the Virgin Islands       \\ \hline
%         Sonora Meiling$^\ast$   & University of the Virgin Islands       \\ \hline
%     \end{tabular}
%     \caption{Data contributors to the transmission experiments described in section \ref{sec:transmission}, to which the calibrated model parameters were compared. Contributors highlighted with "$^\ast$" conducted the Data Sharing.}
%     \label{tab:contributors}
% \end{table}

% \section*{Supplementary Material}
% The Supplementary Material for this article is attached to the submitted document.

% === BIBLIOGRAPHY === %

%%% Make sure to upload the bib file along with the tex file and PDF
%%% Please see the test.bib file for some examples of references

% \appendix
% \section*{Appendix}
% \renewcommand{\thesection}{A}
% \setcounter{figure}{0}
% \renewcommand{\thefigure}{\thesection\arabic{figure}}
% \subsection{Connectivity between the Marquesas and the Dry Tortugas}\label{section:appendix}
% To assess the potential for disease transmission from the Marquesas to the Dry Tortugas, we computed the paths with the highest transmission probability between each possible pairing of a source reef in the Marquesas and a sink reef in the Dry Tortugas in our connectivity networks. To do so, a weight $w_{ij} = 1-C_{ij}$ was attributed to the edge between sub-reefs i and j, so that paths with lowest weight (i.e. shortest paths) in the network would correspond to connections with highest probability of exchange of disease agents. The shortest paths were then computed using the Python package python-igraph \citep{csardi2006igraph} for each monthly matrix obtained with barotropic currents, as they drive the modeled disease propagation that best matches observations (Fig. \ref{fig:results}).

% The most likely transmission pathways from the Marquesas to the Dry Tortugas are computed by taking the yearly average of the shortest paths obtained for each monthly connectivity (Fig. \ref{fig:appendix1}). Although the SCTLD has not reached the Marquesas yet in 2018, these preliminary results bring some insight into the exchange mechanisms between the Marquesas and the Dry Tortugas. First, the existence of multiple paths from the Marquesas to the Dry Tortugas suggests a possibility for disease agents produced in the Marquesas to reach the Dry Tortugas. Moreover, disease transmission seems more likely to originate from inshore reefs as paths starting from these reefs exhibit the largest exchange probability. Furthermore, the presence of indirect transmission pathways from the Marquesas to the Dry Tortugas suggests that reefs located between these two zones act as stepping stones that facilitate the propagation of the disease. This is particularly the case for disease agents produced in reefs located more offshore. Additionally, our results suggest a seasonal variation of these exchanges as no exchange from the Marquesas to the Dry Tortugas was modeled between September and November 2018 (Fig. \ref{fig:appendix2}). This absence of connection might be explained by the seasonal variation of the wind regime on the Western Florida Shelf. A non-seasonal influence which may have played a role is the offshore circulation associated with the Florida Current. The presence or absence of an oceanic mesoscale Tortugas gyre and associated frontal eddies has been shown \citep{lee1994evolution,lane2003mesoscale,sponaugle2005florida} to play an important role in connectivity patterns within the lower Keys, and between this region and the Dry Tortugas. Analysis based on operational ocean models from this period in 2018 (e.g., RTOFS, \cite{mehra2010real}) shows that the Tortugas Gyre was absent, with the Loop Current in a strongly retracted position - essentially, flowing straight from the Yucatan Channel into the Straits of Florida in November 2018; this may have significantly reduced connectivity during this period. The most likely period during which the infection could occur seems to be winter-early spring, so the presence or absence of this larger scale setup for later years may be important in understanding both disease and larval connectivity with the Dry Tortugas.

% \begin{figure}
%     \centering
%     \includegraphics[width=.95\textwidth]{figures/paths_average.jpg}
%     \caption{Shortest paths to each sub-reef of the Dry Tortugas from the Marquesas computed from the yearly-averaged connectivity matrix obtained with mean barotropic currents. Reefs are  highlighted in light grey while islands are shown in dark grey.}
%     \label{fig:appendix1}
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=.7\textwidth]{figures/time_series_MQ_DT_2.jpg}
%     \caption{Number of shortest paths from the Marquesas to the Dry Tortugas multiplied by their mean probability for each simulated month with barotropic currents.}
%     \label{fig:appendix2}
% \end{figure}

### 2022-01-04 chapters/sctld/sctld.tex
creation of thesis git

This chapter is based on the following article:
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{T. Dobbelaere}, E. M. Muller, L. J. Gramer, D. M. Holstein, $\&$ E. Hanert, 2020, Coupled Epidemio-Hydrodynamic Modeling to Understand the Spread of a Deadly Coral Disease in Florida, \textit{Frontiers in Marine Science},69: 1121-1132. \href{https://www.frontiersin.org/article/10.3389/fmars.2020.591881 }{doi: 10.3389/fmars.2020.591881}
\end{list}

\begin{abstract}
    For the last six years, the Florida Reef Tract (FRT) has been experiencing an outbreak of the Stony Coral Tissue Loss Disease (SCTLD). First reported off the coast of Miami-Dade County in 2014, the SCTLD has since spread throughout the entire FRT with the exception of the Dry Tortugas. However, the causative agent for this outbreak is currently unknown. Here we show how a high-resolution bio-physical model coupled with a modified patch Susceptible-Infectious-Removed epidemic model can characterize the potential causative agent(s) of the disease and its vector. In the present study, the agent is assumed to be transported within composite material (e.g., coral mucus, dying tissues, and/or resuspended sediments) driven by currents and potentially persisting in the water column for extended periods of time. In this framework, our simulations suggest that the SCTLD is likely to be propagated within neutrally buoyant material driven by mean barotropic currents. Calibration of our model parameters with field data shows that corals are diseased within a mean transmission time of 6.45 days, with a basic reproduction number slightly above 1. Furthermore, the propagation speed of the disease through the FRT is shown to occur for a well-defined range of values of a disease threshold, defined as the fraction of diseased corals that causes an exponential growth of the disease in the reef site. Our results present a new connectivity-based approach to understand the spread of the SCTLD through the FRT. Such a method can provide a valuable complement to field observations and lab experiments to support the management of the epidemic as well as the identification of its causative agent.
\end{abstract}

%\newpage

\input{chapters/sctld/article}

### 2022-01-04 include/title.tex
creation of thesis git

\thispagestyle{empty}
\begin{minipage}[b]{0.3\textwidth}
\hspace*{-0.8cm}\includegraphics[width=5.5cm]{./figures/logo_ucl.jpg}\vspace{0.85cm}
%\hspace*{3.4cm}\includegraphics[width=4.5cm]{./figures/orb.png}\vspace*{.1cm}
\end{minipage}
\begin{minipage}[b]{0.69\textwidth}

\begin{center}
\hspace*{0.5cm}Université Catholique de Louvain\\
\hspace*{0.5cm}\'Ecole Polytechnique de Louvain\\
\hspace*{0.5cm} Earth and Life Institute\\
\end{center}
\end{minipage}

\vspace*{1.2cm}
\hspace{-0.8cm}
\begin{minipage}{1.02\linewidth}
\centering
\LARGE\bfseries How modeling tools can inform coral reef management in the Florida Reef Tract
\end{minipage}

\vspace*{0.5cm}
\hspace{-0.5cm}\begin{minipage}{\textwidth}
\centering
\rule{50pt}{.5pt}\\[1.2ex]
Doctoral dissertation presented by \\ [2.ex]
{\Large Thomas Dobbelaere\\ [2.ex]
}
in partial fulfillment of the requirements for the degree of\\ [2.ex]
\Large Docteur en Sciences de l'ingénieur

\rule{50pt}{.5pt}\\
\end{minipage}

\vspace*{1.5cm}
\begin{minipage}{\textwidth}
\begin{center}

\large Thesis committee:\\

\vspace*{0.5cm}
\footnotesize{
\hspace{-1.2cm}
\begin{tabular}{ll}
Prof. Emmanuel Hanert (Supervisor) & Université Catholique de Louvain \\
Prof. Eric Deleersnijder           & Université Catholique de Louvain \\
Dr. Jonathan Lambrechts            & Université Catholique de Louvain\\
???                                & ??? \\
???                                & ??? \\ 
\end{tabular}
}
\end{center}
\end{minipage}

\vspace*{.5cm}
\vspace*{\stretch{1}}
\begin{minipage}{\textwidth}
\centering
\large Louvain-la-Neuve, March 2022
\end{minipage}

### 2022-01-04 thesis.tex
creation of thesis git

\documentclass[twoside,12pt,a4paper,openright]{memoir}
\maxsecnumdepth{subsubsection}			%Automatically number sections down to ``section'' level
\setsecnumdepth{subsubsection}
% \hyphenation{de-leer-snij-der}
\usepackage[utf8]{inputenc}			%Optimal font encoding for most latin languages
\usepackage[T1]{fontenc}
\usepackage{changepage}
\usepackage{pgfplots}
\usepackage{bm}
\usepackage{lipsum}
\input{include/specific_headers.tex}
\newcommand{\GG}[1]{}
\pgfplotsset{compat=1.15}
\usepackage[pdftex,linktocpage]{hyperref}		%For hyperlinks within doc, and setting doc info
\hypersetup{pdftitle={How modeling tools can inform coral reef management in Florida}}
\hypersetup{pdfauthor={Thomas DOBBELAERE}}
\hypersetup{pdfsubject={How modeling tools can inform coral reef management in Florida}}
\hypersetup{pdfkeywords={Biophysical modeling, Florida Coral Reef, SCTLD, SLIM}}
% % Comment out next hypersetup section for print version:
%\hypersetup{
%  colorlinks=true,
%  citecolor=blue,
%  linkcolor=blue,
%  urlcolor=blue
%}
\hypersetup{
  colorlinks=true,
  citecolor=black,
  linkcolor=black,
  urlcolor=black
}
% % % %   citebordercolor={0.8 0.8 0.8},
% % % %   filebordercolor={0.8 0.8 0.8},
% % % %   linkbordercolor={0.8 0.8 0.8},
% % % %   menubordercolor={0.8 0.8 0.8},
% % % %   urlbordercolor={0.8 0.8 0.8},
% % % %   runbordercolor={0.8 0.8 0.8}

% \usepackage{cmbright}				%TODO: this doesn't compile when I add it *** Font: Computer modern bright
% \usepackage{sfmath}				%Display all maths with sans serif font

\newcommand{\hycom}{\textsc{hycom} }
\newcommand{\ie}{\textit{i.e}} 
\newcommand{\eg}{\textit{e.g.}}

\begin{document}
\frontmatter
  \input{include/title}
  \clearemptydoublepage
  \input{include/abstract}
  %\clearemptydoublepage
  \input{include/thanks}
  \clearemptydoublepage
  \tableofcontents
  \clearemptydoublepage
  % \input{include/glossary}
  % \clearemptydoublepage
  \mainmatter 
%%%%
\mychapter{Introduction} \label{chap:intro}
 \input{chapters/intro/intro}    
%%
%%%%%% 
  \mychapter{Modeling the spread of SCTLD in Florida} \label{chap:sctld}
  \chaptermark{Modeling the spread of SCTLD in Florida}  
  \input{chapters/sctld/sctld}
%%
% \mychapter{Paper Dry Tortugas} \label{chap:ontario}
% \chaptermark{Paper DRTO}  
%   \input{chapters/ontario/ontario}
%
% \mychapter{Paper onset} \label{chap:eigenmode}
% \chaptermark{Paper onset}  
%   \input{chapters/eigenmode/eigenmode}
%%%   
%%%%%%%%%%%
%
%%%%%  
%%%
  \mychapter{Conclusions and perspectives} \label{chap:conclusion}
  \input{chapters/conclusions/conclusions}    
%%
%
  \clearemptydoublepage
  \bibliography{include/bibliography.bib}

\end{document}

### 2022-01-04 include/title.tex
'c' minuscule à 'catholique'

@@
 \begin{center}
-\hspace*{0.5cm}Université Catholique de Louvain\\
+\hspace*{0.5cm}Université catholique de Louvain\\
 \hspace*{0.5cm}\'Ecole Polytechnique de Louvain\\
@@
 \begin{tabular}{ll}
-Prof. Emmanuel Hanert (Supervisor) & Université Catholique de Louvain \\
-Prof. Eric Deleersnijder           & Université Catholique de Louvain \\
-Dr. Jonathan Lambrechts            & Université Catholique de Louvain\\
+Prof. Emmanuel Hanert (Supervisor) & Université catholique de Louvain \\
+Prof. Eric Deleersnijder           & Université catholique de Louvain \\
+Dr. Jonathan Lambrechts            & Université catholique de Louvain\\
 ???                                & ??? \\

### 2022-01-05 chapters/irma/article.tex
slim+swan paper added

\section{Introduction}

Major hurricanes are becoming more intense under the effect of global warming \citep{bhatia2019recent, knutson2020tropical}. Better understanding their repercussions on coastal areas becomes therefore critical. However, estimating the impact of hurricanes on the coastal ocean circulation remains a challenge. Understanding wave-current interactions and representing their impact on coastal ocean transport processes is central to many coastal activities such as dredging, erosion management, oil and gas activities, search and rescue, and insurance \citep{bever2013simulating,li1998three, breivik2013advances}. All these activities require wave-current models to predict the impact of tropical cyclones on the coastal circulation and on the sea surface elevation.

Wave-current interactions during a cyclone are highly nonlinear and vary significantly in space and time \citep{wu2011fvcom}. Wave-induced currents are generated by wave radiation stress gradients \citep{longuet1970longshore}, affecting water levels near shorelines and wave breaking points \citep{longuet1964radiation}. Changes in water levels and currents, in turn, affect the motion and evolution of the waves \citep{sikiric2013coupling}. Coupled wave-current models hence require the calculation of the full directional wave spectrum in order to correctly reproduce the dynamics of wind-driven surface waves. This is usually achieved by spectral wave models, which describe the evolution of the wave energy spectrum. As of today, the most popular spectral wave models are the WAve Model (WAM) \citep{group1988wam}, %TOMAWAC \citep{benoit1997tomawac},
Simulating WAve Nearshore (SWAN) \citep{booij1999third}, and WAVEWATCH III \citep{tolman2009user}. Among these models, SWAN has been specifically developed for coastal applications, as it represents depth-induced wave breaking and triad wave-wave interactions using numerical techniques adapted to small-scale, shallow water regions \citep{booij1999third}. WAVEWATCH III has also recently been equipped with new parallelization algorithm, domain decomposition and numerical schemes for high resolution coastal applications \citep{ww3dg,abdolali2020large}.

Coastal oceans are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures. Traditional structured-grid models lack the flexibility to simulate near-shore processes at a sufficiently small scale. Although the use of nested structured grids allows local grid refinement \citep{warner2010development}, staircase-like representation of complex coastal topographies cannot be avoided. Instead, unstructured-mesh models easily adapt to the topography and are hence better suited to coastal processes \citep{fringer2019future}. Capturing the impact of the topography on wave interactions becomes even more important in the case of tropical cyclones. Heavy winds generate large wind-waves and disturb ocean conditions \citep{liu2020impacts} by causing coastal upwellings on continental shelves \citep{smith1982response} and inducing strong currents, waves and storm surges in nearshore and coastal regions \citep{dietrich2010high, weisberg2006hurricane}.

Ocean waves act as the dynamical interface between the atmosphere and the ocean. Through this interface, tropical cyclones cause a cooling of the upper ocean layer by vertical mixing and heat transfer \citep{aijaz2017nonbreaking,varlas2020investigating}. By altering the structure of the upper-ocean, hurricane can cause the disruption of major ocean currents such as the Florida Current and Gulf Stream \citep{oey2007hurricane}. Interaction with hurricanes alters the thermal structure of these currents and can cause a significant decline of their flow, resulting in delayed increased coastal levels along their path, even in locations out of reach of the hurricane itself \citep{ezer2017observations, ezer2018interaction,ezer2020long}.

Near the storm, heavy wind conditions also affect material transport at the ocean surface. The transport of drifting objects or substances that are locally released is often best represented by a Lagrangian individual-based model. Such an approach is routinely used to model the dispersal of larvae, pollutants, sediments and many other tracers (e.g. \citealp{le2012surface,liubartseva2018tracking, figueiredo2013synthesizing,frys2020fine}). Although some transport model might take wave-induced currents into account, most of them neglect wave-current interactions, which can lead to significant errors in storm conditions \citep{rohrs2012observation,curcic2016hurricane}. \cite{niu2017role} and \cite{mao2018wave,mao2020particle} investigated the impact of wave-current interactions during storm event in lakes and inlets. However, to our knowledge, there have been no similar studies on the impact of hurricane-induced wave-current interactions in coastal environments such as the Florida Reef Tract (FRT), where changes in transport processes might significantly impact the biological connectivity.

The main questions we want to answer are the following: (1) How important are wave-current interactions during a tropical cyclone? (2) What effect do they have on the transport of drifting material? We tackle these issues by investigating the transport of drifting particles on the Florida shelf during Hurricane Irma, one of the strongest and costliest tropical cyclones on record in the Atlantic Basin \citep{xian2018brief}, which made landfall in Florida in September 2017. To do that, we developed an unstructured-mesh coupled wave-current model of the whole FRT to simulate the ocean circulation under hurricane conditions. Both modeled currents and waves were validated against field measurements and then used to simulate the transport of drifting material in the Florida Keys and over the Florida inner shelf. Model outputs were then compared with uncoupled simulation results in order to assess the impact of the radiation stress gradient and Stokes drift on the modeled currents and transport.

% === METHODS === %
\section{Methods}
\subsection{Study area and observational data}
We study the ocean circulation in an area that covers the whole FRT and includes the nortwestern end of the Gulf of Mexico and the Straits of Florida (Fig. \ref{fig:mesh}). The large-scale ocean circulation around South Florida is dominated by the Florida Current (FC), which originates from the Loop Current (LC) where it enters the Florida Straits from the Gulf of Mexico, and, downstream, forms the Gulf Stream. The FC is a major western boundary current characterized by spatial variability and meandering, associated with the presence of cyclonic eddies between the core of the current and the complex reef topography of the FRT \citep{lee1995florida,kourafalou2012florida}. The variability of the FC extends over a large range of spatial and temporal scales, with periods of 30-70 days in the Lower Keys \citep{lee1995florida} and shorter periods of 2-21 days in the Upper Keys \citep{lee1977low}, and exhibits significant seasonal and interannual cycles \citep{johns1987meandering, lee1988wind,schott1988variability}. Circulation on the West Florida Shelf (WFS), on the other hand, is forced by local winds and tidal fluctuations \citep{lee2002volume,liu2012seasonal}. Furthermore, due to its location relative to the warm waters of the North Atlantic, Florida is particularly vulnerable to tropical cyclones. On average, the state is hit by a hurricane every two years and strong hurricanes, some of which are among the most destructive on record, strike Florida on average once every four years \citep{malmstadt2009florida}.

The state of the ocean around Florida is monitored by an extensive array of tide gauges, current meters and buoys. In this study, we used sea surface elevation measurements from the National Oceanic and Atmospheric Administration’s (NOAA) Tides and Currents dataset. These measurements were taken at four locations: two in the Florida Keys (Key West and Vaca Key); one on the East coast of Florida (Virginia Key); and one on the West coast (Naples). For the currents, we used ADCP measurements from the University of South Florida's College of Marine Science's (USF/CMS) Coastal Ocean Monitoring and Prediction System (COMPS) for the WFS \citep{weisberg2009mean}. More specifically, we used measurements from moorings C10, C12 and C13, respectively located at the 25, 50, and 50 m isobaths of the WFS \citep{liu2020impacts}. Finally, for the waves, we used measurements from four buoys of the NOAA's National Data Buoy Center (NDBC); two on Florida's eastern shelf and two on the WFS. The locations of all measurement stations are shown in Fig. \ref{fig:mesh}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/irma/figures/fig_mesh_hurr.png}
    \caption{(\textbf{A}) Mesh of the computational domain with the trajectory of Irma. The category of the hurricane is given by the Saffir-Simpson color scale. (\textbf{B}) Bathymetry of the domain with the location of stations used for the validation of the model outputs. (\textbf{C}) Close up view of the Lower Keys area (red box in (A)), where the mesh resolution reaches 100 m near reefs (shown in dark orange) and islands (shown in dark grey)}
    \label{fig:mesh}
\end{figure}

\subsection{Wind and atmospheric pressure during Hurricane Irma}

Hurricane Irma made landfall in Florida on 10 September 2017 as a category 4 hurricane at Cudjoe Key (Florida Keys) and later as a category 3 hurricane on Marco Island, south of Naples (see hurricane track in Fig. \ref{fig:mesh}). It then weakened to a category 2 hurricane as it moved further inland \citep{irmaNOAA}. The storm damaged up to 75\% of the buildings at its landfall point in the Florida Keys, making it one of the strongest and costliest hurricanes on record in the Atlantic basin \citep{xian2018brief,zhang2019modeling}. The strongest reported sustained winds on Marco Island were 50 m/s while the highest recorded storm surge was 2.3 m, although larger wind speed likely occurred in the Florida Keys \citep{pinelli2018overview}. To reproduce the wind profile of Irma in our model, we used high-resolution H$^\ast$Wind wind fields \citep{powell1998hrd}. As these data represent 1-min averaged wind speeds, we multiplied them by a factor 0.93 to obtain 10-min averaged wind speeds \citep{harper2010guidelines}. This operation reduces the erratic values caused by the greater variance of mean winds measured over periods shorter than 10 minutes. Furthermore, H*Wind wind profiles did not cover the whole model extent during the passage of the hurricane and were thus blended within a coarser wind field extracted from the European Centre for Medium-Range Weather Forecasts (ECMWF) ERA-5 dataset (Fig. \ref{fig:atm}A). The pressure field during the passage of Hurricane Irma was also reconstructed using ERA-5 data. However, the coarse resolution of the dataset smoothes out the depression at the center of the hurricane, leading to an underestimation of the pressure gradient (Fig. \ref{fig:atm}B). To better capture the central depression of Irma, we therefore built a hybrid pressure field using the position and the minimal pressure of the core of the hurricane based on its track, as recorded in the HURricane DATabases (HURDAT) 2 \citep{landsea2013atlantic}. Based on this information, the hybrid pressure field was constructed by combining an idealized Holland pressure profile \citep{lin2012hurricane} within the radius of maximum wind speed of Irma \citep{knaff2018statistical} with ERA-5 pressure field. The transition from the Holland profile to ERA-5 data outside the radius of maximum wind speed data was performed using a smooth step function (Fig. \ref{fig:atm}).

\begin{figure}
    \centering
    \includegraphics[width=.99\textwidth]{chapters/irma/figures/hwind+holland_vs_era.png}
    \caption{Snapshot of the hybrid wind (\textbf{A}) and pressure (\textbf{B}) profiles constructed to capture the passage of Hurricane Irma at 1800 UTC on 9 September 2017. Wind profiles were obtained by combining high resolution H*Wind with coarser ERA-5 wind fields. The pressure field was built by combining the ERA-5 pressure field with an idealized Holland pressure profile based on the track of Irma in the HURDAT 2 database. Holland field was only used within the radius of maximum wind speed (dashed grey line) of the hurricane to capture its central depression. }
    \label{fig:atm}
\end{figure}

\subsection{Hydrodynamic model}

Ocean currents generated during Hurricane Irma around South Florida were modeled using the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-neuve Ice-ocean Model\footnote{\url{https://www.slim-ocean.be}} (SLIM) \citep{lambrechts2008multi}. The model mesh covers an area similar to the model extent of \cite{dobbelaere2020coupled}, that includes the FRT but also the Florida Straits and part of the Gulf of Mexico (Figure \ref{fig:mesh}). However, this area has been slightly extended northeastward and westward in order to include the NOAA-NDBC buoys. Furthermore, to withstand potential cell drying during the hurricane, we solved the conservative shallow water equations with wetting-drying:
\begin{equation}
    \begin{split}
        \dfrac{\partial H}{\partial t} +\nabla\cdot(\UV) =& ~0~, \\
        \dfrac{\partial \UV}{\partial t}  + \nabla\cdot\left(\dfrac{\UV\UV}{H}\right) + f\mathbf{e}_z \times \UV =& ~\alpha gH\nabla(H-h) - \dfrac{1}{\rho}\nabla p_\text{atm} + \dfrac{1}{\rho}\text{\boldmath$\tau$}_s \\
         &+\nabla\cdot(\nu\nabla\UV) - \dfrac{C_b}{H^2}|\UV|\UV + \gamma(\UV_\text{ref}-\UV)~,
    \end{split} \label{eq:slim}
\end{equation}
where $H$ is the water column height and $\UV$ is the depth-averaged transport; $f$ is the Coriolis coefficient; $g$ is the gravitational acceleration; $h$ is the bathymetry; $\alpha$ is a coefficient indicating whether the mesh element is wet ($\alpha=1$) or dry ($\alpha=0$) \citep{le2020implicit}; $\nu$  is the Smagorinsky viscosity; $C_b$ is the bulk bottom drag coefficient; $p_\text{atm}$ is the atmospheric pressure; {\boldmath$\tau$}$_s$ is the surface stress, usually due to wind; and $\gamma$ is a relaxation coefficient towards a reference transport $\UV_\text{ref}$. As this study focuses on transport processes and not coastal flooding, wetting-drying is only applied on wet grid cells that may become dry under the influence of the hurricane. As in \cite{frys2020fine} and \cite{dobbelaere2020coupled}, SLIM currents were gradually relaxed towards the operational Navy HYbrid Coordinate Ocean Model (HYCOM) product (GOMl0.04\footnote{\url{https://www.hycom.org/data/goml0pt04}}, \cite{chassignet2007hycom}) in regions where the water depth exceeds 50 m. HYCOM's 3D currents were depth-integrated into 2D transports to be used as forcing in the model. Moreover, these transports as well as HYCOM's sea surface elevation were used as boundary condition in the model.

We adapted the parameterization of the wind-induced surface stress to storm conditions. At very high wind speeds, the white cap is blown off the crest of the waves. This phenomenon, also known as spume, has been hypothesized to generate a layer of droplets that acts as a slip layer for the winds at the ocean-atmosphere interface \citep{holthuijsen2012wind}. It causes a saturation of the wind drag coefficient for strong winds \citep{powell2003reduced,donelan2004limiting,curcic2020revised}. We take this saturation effect into account by using the wind  drag parameterization of \cite{moon2007physics}. In this parameterization, the drag coefficient $C_d$ depends on the wind speed at 10-m height $U_{10}$ according to:
\begin{equation}
    C_d = \kappa^2 \log\left(\dfrac{10}{z_0}\right)^{-2}\label{eq:drag}
\end{equation}
where $\kappa$ is the von Karman constant and $z_0$ is the roughness length expressed as: 
\begin{equation}
    z_0 =\left\{\begin{array}[]{ll}
        \dfrac{0.0185}{g}u_{\ast}^2 & \text{if }U_{10} \leq 12.5\text{ m/s}~, \\
        \lbrack 0.085(-0.56u_{\ast}^2+20.255u_{\ast} & \text{if }U_{10} > 12.5\text{ m/s}~, \\
        \quad +2.458) - 0.58\rbrack\times 10^{-3} 
    \end{array}\right.
\end{equation}
with $u_\ast$ the friction velocity. The relation between $U_{10}$ and $u_{\ast}$ is given by:
\begin{equation}
    U_{10}=-0.56u_{\ast}^2+20.255u_{\ast}+2.458~.
\end{equation}

The mesh resolution depends on the distance to the coastlines and reefs following the approach of \cite{dobbelaere2020coupled}. The mesh is then further refined according to bathymetry value and gradient, as suggested in the SWAN user-guide\footnote{\url{http://swanmodel.sourceforge.net/unswan/unswan.htm}}. Such an approach improves the model efficiency as the mesh resolution is only increased where required by the currents and waves dynamics. The mesh was generated with the seamsh\footnote{\url{https://pypi.org/project/seamsh/}} Python library, which is based on the the open-source mesh generator GMSH \citep{geuzaine2009gmsh}. It is composed of approximately 7.7 $\times$ 10$^5$ elements. The coarsest elements, far away from the FRT, have a characteristic length of about 5 km whereas the finest elements have a characteristic length of about 100 m along the coastlines and over the reefs (Fig \ref{fig:mesh}).

\subsection{Wave model}
Waves were modeled using the parallel unstructured-mesh version of the Simulating WAves Nearshore (SWAN) model \citep{booij1999third}, one of the most popular wave models for coastal areas and inland waters. It solves the action balance equation \citep{mei1989applied}:
\begin{equation}
    \dfrac{\partial N}{\partial t} + \nabla_\mathbf{x}\cdot[(\mathbf{c}_g+\mathbf{u})N] + \dfrac{\partial }{\partial \theta}[c_\theta N] + \dfrac{\partial}{\partial \sigma}[c_\sigma N] = \dfrac{S_{in}+S_{ds}+S_{nl}}{\sigma}~, \label{eq:swan}
\end{equation}
where $N=E/\sigma$ is the wave action density and $E$ is the wave energy spectrum; $\theta$ is the wave propagation direction; $\sigma$ is the intrinsic wave frequency; $\mathbf{c}_g$ is the wave group velocity, $\mathbf{u}=\mathbf{U}/H$ is SLIM depth-averaged current velocity; $c_\theta$ and $c_\sigma$ are the propagation velocities in spectral space due to refraction and shifting in frequency due to variations in depth and currents; and $S_{in}$, $S_{ds}$, and $S_{nl}$ respectively represent wave growth by wind, wave decay and nonlinear transfers of wave energy through four and three-wave interactions, \ie quadruplets and triplets. The wave spectra were discretized with 48 direction bins and 50 frequency bins logarithmically distributed from 0.03 to 2 Hz. Exponential wind growth was parameterized using the formulation of \cite{janssen1991quasi}, while dissipations by whitecapping and bottom dissipation followed the formulations of \cite{komen1984existence} and \cite{madsen1989spectral}, respectively.

Coefficients for exponential wind growth and whitecapping parameterizations were based on the results of \cite{siadatmousavi2011evaluation}, and significantly differ from SWAN's default settings. By default, SWAN implements the wind input formulation of \cite{komen1984existence} and the steepness-dependent coefficient governing dissipation by whitecapping is a linear function of the wavenumber. In this study, this steepness-dependent coefficient is a quadratic function of the wavenumber, as it showed better predictions of the significant wave height in the study of \cite{siadatmousavi2011evaluation}. The choice of these formulations was motivated by the appearance of numerical instabilities in the region of the Gulf Stream when using SWAN's default parameter values. Finally, ERA-5 wave spectra was used as boundary condition for SWAN. Wave spectra is obtained from the ocean wave model WAM and is given on a $1^\circ \times 1^\circ$ grid with 24 directions and 36 frequencies.

Surface waves induce a net drift in the direction of the wave propagation, known as the Stokes drift \citep{van2018stokes,stokes1880theory}. This net drift has a significant impact on sediment transport in nearshore regions \citep{hoefel2003wave}, on the formation of Langmuir cells \citep{langmuir1938surface, craik1976rational} as well as on the transport of heat, salt or pollutants such as oil or micro-plastic in the upper ocean layer \citep{mcwilliams2000vertical,rohrs2012observation,drivdal2014wave}. To correctly model the Stokes drift profile in mixed wind-driven sea and swell conditions, the full two-dimensional wave spectrum must be represented by a spectral wave model within a wave-current coupling \citep{van2018stokes}. We therefore used SWAN modeled spectra to compute the Stokes drift as follows:
\begin{equation}
    \mathbf{u}_{s} = \int_0^{2\pi}\int_0^{+\infty} \dfrac{\sigma^3}{h\tanh(2kh)}E(\sigma,\theta)(\cos\theta, \sin\theta)d\sigma d\theta~, \label{eq:stokes}
\end{equation}
where $k$ is the norm of the wave vector; $h$ is the water depth; and $E(\sigma,\theta)$ is the wave energy density. The computed Stokes drift velocity is then added to SLIM depth-averaged current velocity to transport drifting particles in the experiments described in section \ref{sec:traj}.

\subsection{Coupled model}

SLIM and SWAN are coupled so that they run on the same computational core and the same unstructured mesh. SLIM is run first and passes the wind velocity ($\mathbf{U}_{10}$), water level ($\eta=H-h$) and depth-averaged current ($\mathbf{u}=\mathbf{U}/H$) fields to SWAN, as well as a roughness length ($z_0$) for the bottom dissipation formulation of \cite{madsen1989spectral}. This roughness length is computed from SLIM's bulk drag coefficient $C_b$ following the approach of \cite{dietrich2011hurricane} so that both models have consistent bottom dissipation parameterizations. SWAN then uses these quantities to compute the wave radiation stress gradient, that is then passed to SLIM as the force exerted by waves on currents {\boldmath$\tau$}$_\text{wave}$ (Fig. \ref{fig:coupling}). SLIM then uses this quantity to update the value of the surface stress {\boldmath$\tau$}$_s$ in Eq. (\ref{eq:slim}), that now becomes the sum of wind and wave-induced stresses $\text{\boldmath$\tau$}_s = \text{\boldmath$\tau$}_\text{wind}+\text{\boldmath$\tau$}_\text{wave}$. Here, the momentum flux from the atmosphere to the ocean is taken as the usual full wind stress {\boldmath$\tau$}$_\text{wind}$. Doing so, we neglect the momentum advected away from the storm by the waves, leading to a 10-15\% overestimation of the momentum flux in hurricane winds \citep{curcic2015explicit}.

We followed the approach of \cite{dietrich2012performance} by characterizing the wave-induced forces on currents using the radiation-stress (RS) gradient formalism, which has been successfully applied in both 2D and 3D coupled wave-current models under storm conditions \citep{hope2013hindcast, sebastian2014characterizing, brown2013depth}. An alternative formalism is the vortex-force (VF) representation \citep{mcwilliams2004asymptotic}, that provides a clearer decomposition of the wave effect \citep{lane2007wave}. Although both approaches were adopted by coastal modeling communities, there is an ongoing scientific debate over the correctness and applicability of the two concepts \citep{ardhuin2008comments, mellor2013waves, mellor2015combined, ardhuin2017comments}. \cite{xia2020implementation} recently implemented the two formalisms in a 3D unstructured-grid model and compared them in three typical coastal systems, showing that the 3D RS algorithm could generate unrealistic offshore currents near shorelines. Despite these shortcomings, the 3D RS method reproduced most wave-induced currents and the 2D RS formalism remains a well-validated modeling approach. Furthermore, \cite{mellor2013waves} showed that the RS approach was valid when $[\frac{\partial H}{\partial\mathbf{x}}/\sinh(kh)]^2$ is smaller or of the same order as $(ka)^2$, where $a$ is the wave amplitude. We evaluated these quantities and verified that the validity criterion was met in our model domain. Additionally, since the VF and RS approached are formally equivalent \citep{lane2007wave}, we selected the 2D RS formalism, as it has the advantage of summarizing the impact of waves on the currents in a single additional stress term in the hydrodynamic model equations.

SLIM's governing equations are integrated using an implicit time integration scheme while SWAN is unconditionally stable \citep{dietrich2012performance}, allowing both models to be run with relatively large time steps. In this study, the stationary version of SWAN was used, \ie the first term of Eq. (\ref{eq:swan}) was set to zero. This resulted in reduced scaling and convergence rates than with the nonstationary version of SWAN but increased the model stability. The wave spectra at each node of the mesh was saved at the end of each iteration to serve as initial conditions for the next one. Both models were run sequentially using a time step of 600 s, so that each computational core was alternatively running either SLIM or SWAN. As in the coupling between SWAN and the ADvanced CIRCulation model (ADCIRC) \citep{dietrich2012performance}, both models use the same local sub-mesh, allowing for a one-to-one correspondence between the geographic locations of the mesh vertices. No interpolation is therefore needed when passing the discretised variables from one model to the other, which allows an efficient inter-model communication. However, as SLIM is based on  a discontinuous Galerkin finite element method, an additional conversion step to a continuous framework was required to transfer SLIM nodal quantities to SWAN. The coupling increases the computation time by 3\% as compared to the sum of the uncoupled SLIM and SWAN simulations wall-clock times for the same number of CPUs and the same simulation period.
%The coupling increased the computation time by 400\% and 25\% respectively compared to uncoupled SLIM and SWAN model runs with the same number of CPUs over the same simulated period.

% Wall time simulations
%   - slim     : 00:56:56 ->  3416 s
%   - swan     : 03:56:57 -> 14217 s
%   - coupled  : 05:02:51 -> 18171 s
% => overshoot of 538 sec -> increase computation time by 3%

\begin{figure}
    \centering
    \includegraphics[width=.99\textwidth]{chapters/irma/figures/coupling_v2.png}
    \caption{Schematic illustration of the coupled SLIM+SWAN model.}
    \label{fig:coupling}
\end{figure}

\subsection{Quantifying the effect of wave-current interactions on transport}\label{sec:traj}

To quantify the impact of wave-current interactions on transport processes, we compared the trajectories of passive particles advected by the uncoupled SLIM and coupled SLIM+SWAN currents during the passage of Irma in the Lower Keys. Furthermore, the depth-averaged Stokes drift was computed using the wave spectra of the coupled model SLIM+SWAN run as well as those of an uncoupled SWAN run. Particles were released on the inner and outer shelves at the points highlighted by red and blue dots in Fig. \ref{fig:init} on Sept. 7 at 0000 UTC and then tracked until Sept. 15. These initial particle positions were found using backtracking methods \citep{spivakovskaya2005simulation} to ensure that the release particles would intersect the path of Irma during its passage through the Florida Keys. We first defined two 25 km$^\text{2}$ circular regions on the trajectory of the hurricane (see red and blue circles in Fig. \ref{fig:init}). Particles within these two regions were then tracked backward in time using uncoupled SLIM currents from the exact time of the passage of the hurricane until Sept. 7 at 0000 UTC. Their positions at the end of the backward simulation (see red and blue particle clouds in Fig. \ref{fig:init}) corresponds to the initial condition of the forward transport simulations described below. We then compared the trajectories of particles originating from these regions and advected forward in time by different sets of currents: (i) uncoupled SLIM currents alone; (ii) coupled SLIM+SWAN currents; (iii) SLIM currents with the addition of the depth-averaged Stokes drift computed with the coupled wave-current model (Stokes-C); (iv) SLIM+SWAN currents with Stokes-C; and (v) SLIM currents with the depth-averaged Stokes drift computed with the uncoupled wave model (Stokes-U). The different combinations of Eulerian currents and Stokes drifts used to model the transport of passive drifters in the Lower Keys during the passage of Irma are summarized in Table \ref{tab:summary}. Particle trajectories are compared by computing the distances between the centers of mass of the particle clouds through time.

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/irma/figures/figure_initial_positions.png}
    \caption{Release regions of the passive particles on the inner and outer shelves on Sept 7 at 0000 UTC (red and blue clouds) obtained by backtracking particles released in the red and blue circular areas during the passage of Irma.}
    \label{fig:init}
\end{figure}

\begin{table}
    \centering
%    \begin{tabular}{|p{4.5cm}|p{8cm}|}
%        \hline
%        \textbf{Experiment name} & \textbf{Model description} \\ \hline
%        SLIM               & Eulerian currents from the uncoupled SLIM simulation \\ \hline
%        SLIM+SWAN          & Eulerian currents impacted by the RS gradient from the coupled SLIM+SWAN simulation \\ \hline
%        SLIM+Stokes-U      & Eulerian currents from the uncoupled SLIM run with the Stokes drift from the uncoupled SWAN simulation \\ \hline
%        SLIM+Stokes-C      & Eulerian currents from the uncoupled SLIM run with the Stokes drift from the coupled SLIM+SWAN simulation \\ \hline
%        SLIM+SWAN+Stokes-C & Eulerian currents impacted by the RS gradient and the Stokes drift from the coupled SLIM+SWAN simulation \\ \hline
%    \end{tabular}
    \begin{tabular}{|p{4.5cm}|p{4cm}|p{4cm}|}
        \hline
        \textbf{Experiment name} & \textbf{Eulerian currents from} &  \textbf{Stokes drift from}\\ \hline
        SLIM               & uncoupled SLIM simulation & None \\ \hline
        SLIM+SWAN          & coupled SLIM+SWAN simulation (impacted by RS gradient) & None \\ \hline
        SLIM+Stokes-U      & uncoupled SLIM simulation & uncoupled SWAN simulation \\ \hline
        SLIM+Stokes-C      & uncoupled SLIM simulation & coupled SLIM+SWAN simulation \\ \hline
        SLIM+SWAN+Stokes-C & coupled SLIM+SWAN simulation (impacted by RS gradient) & coupled SLIM+SWAN simulation \\ \hline
    \end{tabular}
    \caption{Summary of the different combinations of Eulerian currents and Stokes drifts used to model the transport of passive drifters on the passage of Hurricane Irma in the Lower Keys}
    \label{tab:summary}
\end{table}

%%%%%%%%%%%%%%%%%%%
% --- RESULTS --- %
%%%%%%%%%%%%%%%%%%%
\section{Results}

We first validated the reconstructed atmospheric fields of Hurricane Irma as well as the outputs of our coupled wave-current model against field measurements. We then used the validated model outputs to simulate the transport of passive particles in the Lower Keys during the passage of Hurricane Irma. These particles were advected by the sets of currents described in Table \ref{tab:summary} and their trajectories were compared to evaluate the impact of the wave-current interactions and the Stokes drift on the transport processes during the passage of Irma.

\subsection{Model validation}

H*Wind winds and hybrid pressure field agree well with station measurements at Vaca Key station (Fig. \ref{fig:forcings}). The hybrid pressure field shows a better agreement with observations than ERA-5 pressure as it successfully reproduces the storm depression. ERA-5 fields, on the other hand, fail to reproduce the low pressure at the core of the hurricane due to their coarser grid, leading to an overestimation of 8 mbar of the storm depression. Both H*Wind and ERA-5 agree well with observed wind speeds although both data sets tend to slightly overestimate the width and intensity of the wind peak. However, H*Wind profiles better reproduce the timing of the observed peak, as ERA-5 winds tend to anticipate it. H*wind also exhibits a slightly narrower peak in wind speed, which better agrees with observations.

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/irma/figures/validation_met_2.png}
    \caption{Comparison of reconstructed atmospheric pressure (top) and wind speed (bottom)} with field measurements and coarser ECMWF ERA-5 profiles at Vaca Key station. The generated hybrid atmospheric forcings better reproduce the observed storm depression while H*wind winds better reproduce the measured peak in wind speed.
    \label{fig:forcings}
\end{figure}

Hydrodynamic outputs of the coupled wave-current model agree well with tide gauge (Fig. \ref{fig:sse}) and ADCP measurements (Fig. \ref{fig:uv}). The coupled model reproduces well the timing of the positive and negative storm surges at all tide gauge locations. The amplitude of the positive surges are especially well captured at Naples and Vaca Key, with errors of 2 and 6 centimeters respectively. However, the model underestimates the positive surges at Virginia Key and Key West by 24\% and 15\% at the peak respectively. The amplitude of the negative surge at Naples is also underestimated by about 16\% at the peak. Nonetheless, on average, the absolute error between the model and observations does not exceed 10 cm (Table \ref{tab:stat}). Modeled 2D currents were validated against depth-averaged ADCP measurements at mooring stations C10, C12 and C13 (Fig. \ref{fig:uv}). As in \cite{liu2020impacts}, we performed the vector correlation analysis of \cite{kundu1976ekman} to compare modeled and observed current velocity vectors. Correlation coefficients ($\rho$) between simulated and observed depth-averaged currents are 0.87, 0.84 and 0.81 at stations C10, C12 and C13, respectively. The average veering angles are below 12$^\circ$, as in \citep{liu2020impacts}. Furthermore, the positive bias in Table \ref{tab:stat} indicates that our model tends to underestimate the southward component of the currents at the different stations. As expected from a depth-averaged model, the best fit with observations is obtained at the shallowest mooring C10, located on the 25 m isobath.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/irma/figures/validation_eta.png}
    \caption{Comparison of modeled sea surface elevation at the 4 tide gauges shown in Fig. \ref{fig:mesh}B. The timing and amplitude of the storm surges are well reproduced by the model}
    \label{fig:sse}
\end{figure}
\begin{figure}
    \includegraphics[width=\textwidth]{chapters/irma/figures/figure_currents_all.png}
    \caption{Comparison of modeled current velocity with observed velocity at the moorings (see Fig. \ref{fig:mesh}B for their location). Modeled current velocities agree well with observations, with a correlation coefficient of 0.87, 0.83 and 0.81 at moorings C10, C12 and C13, respectively. The corresponding veering angles are $5.4^\circ$, $0.07^\circ$ and $10.5^\circ$, respectively.}
    \label{fig:uv}
\end{figure}

The simulated significant wave height agrees well with observations at all buoy locations (Fig. \ref{fig:waves}). The timing of the peak in wave height is well captured at all buoys, while the amplitude is better reproduced on the WFS (buoys 42036 and 42097) with errors below 10\%. The error on the peak amplitude on Florida's eastern shelf is of 13\% and 21\% at buoys 4114 and 4113, respectively. On average, observed significant wave height and wave period are better reproduced on the WFS while wave direction is better captured by the model on Florida's eastern shelf (Table \ref{tab:stat}). The fit is especially good at buoy 41113, where the mostly westward-northwestward wave propagation is less perturbed by Irma's wind field.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/irma/figures/hsig_validation.png}
    \includegraphics[width=\textwidth]{chapters/irma/figures/wave_validation_42036.png}
    \caption{Comparison of modeled wave parameters with observation at the 4 buoys location (locations shown in Fig. \ref{fig:mesh}B). Overall, the modeled significant wave heights agree well with field measurement (mean error $<25$\%). }
    % As model parameters were calibrated for the Northern Gulf of Mexico, observations are better reproduced at buoys located on the WFS, as illustrated for buoy 42036
    \label{fig:waves}
\end{figure}

\begin{table}
    \centering
    \begin{tabular}{|p{2.8cm}p{3cm}p{1.8cm}p{1.8cm}p{1.8cm}|}
        \hline
        \textbf{Station} & \textbf{Variable} & \textbf{Bias} & \textbf{MAE} & \textbf{RMSE} \\
        \hline
        Vaca Key     & sse (m)              & 0     & 0.112 & 0.142 \\
                     & $U_{10}$ (m/s)       & 1.51  & 1.85  & 2.61 \\
                     & $p_\text{atm}$ (hPa) & -0.21 & 0.59  & 1.03 \\ 
        Key West     & sse (m)              & 0     & 0.066 & 0.085 \\
        Virginia Key & sse (m)              & 0     & 0.087 & 0.120 \\
        Naples       & sse (m)              & 0     & 0.099 & 0.180 \\
        \hline
        C10   & $u$ (m/s)           &  0.002 &  0.045 &  0.056 \\
              & $v$ (m/s)           &  0.039 &  0.102 &  0.121 \\
        C12   & $u$ (m/s)           &  0.002 &  0.059 &  0.074 \\
              & $v$ (m/s)           &  0.047 &  0.073 &  0.094 \\
        C13   & $u$ (m/s)           & -0.009 &  0.065 &  0.077 \\
              & $v$ (m/s)           &  0.039 &  0.086 &  0.102 \\
        \hline
        41113 & $H_s$ (m)           &  0.150 &  0.357 &  0.430 \\
              & $T_m$ (s)           &  1.608 &  1.671 &  1.878 \\
              & $\theta_m$ (degree) & -1.555 &  7.036 &  9.250 \\
        41114 & $H_s$ (m)           &  0.361 &  0.424 &  0.560 \\
              & $T_m$ (s)           &  0.899 &  1.506 &  1.594 \\
              & $\theta_m$ (degree) & -8.236 & 14.616 & 22.560 \\
        42036 & $H_s$ (m)           &  0.082 &  0.312 &  0.398 \\
              & $T_m$ (s)           &  0.430 &  0.528 &  0.645 \\
              & $\theta_m$ (degree) & -2.307 & 17.144 & 22.734 \\
        42097 & $H_s$ (m)           &  0.048 &  0.326 &  0.432 \\
              & $T_m$ (s)           &  0.476 &  0.755 &  0.892 \\
              & $\theta_m$ (degree) &  2.538 & 34.760 & 55.892 \\
        \hline
    \end{tabular}
    \caption{Error statistics on the wave-current model outputs as compared to the measured sea surface elevation (sse), eastward and northward depth-average current velocities ($u$,$v$), significant wave height ($H_s$), zero-crossing mean wave period ($T_m$) and mean wave direction ($\theta_m$). Model bias, mean absolute error (MAE), and root mean squared error (RMSE) are listed by variable (unit) and value.}
    \label{tab:stat}
\end{table}

\subsection{Impact of waves on currents and transport}

We evaluated the impact of the RS gradient on the modeled currents during the passage of Irma in the Lower Keys, between Sept. 7 and 13, 2017. First, we computed the maximum difference between currents modeled by SLIM and SLIM+SWAN during this period (Fig. \ref{fig:diff}A). The largest differences in current speed were observed over the reefs, on the shelf break and around islands. They locally reach 1 m/s, with the coupled SLIM+SWAN model yielding the largest amplitudes. The regions where the differences were the largest correspond to areas that experienced large maximum values of the RS gradient {\boldmath$\tau$}$_\text{wave}$ (Fig. \ref{fig:diff}B). These areas of large RS gradient are located on the shelf break and over coral reefs, where important wave energy dissipation occurred through depth-induced wave breaking and bottom dissipation \citep{longuet1964radiation}. This highlights the important protective role of the barrier formed by the offshore reefs, that require a fine spatial resolution to be accurately represented by the model. RS-induced differences in current speed were amplified by the action of the wind stress {\boldmath$\tau$}$_\text{wind}$ (Fig. \ref{fig:diff}C). Wind speeds were larger in the front right quadrant of the hurricane \citep{zedler2009ocean}, yielding larger differences on the right-hand side of the storm trajectory. This is especially clear in the area between the Florida Keys and the Everglades, where relatively small values of {\boldmath$\tau$}$_\text{wave}$ produce current speed differences larger than 0.5 m/s because of the wind stress.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/irma/figures/max_diff_new.png}
    \caption{(\textbf{A}). Maximum difference between SLIM and SLIM+SWAN currents during the passage of Irma in the Lower Florida Keys; (\textbf{B}) Maximum wave radiation stress gradient {\boldmath$\tau$}$_\text{wave}$ and (\textbf{C}) maximum wind stress {\boldmath$\tau$}$_\text{wind}$ (\textbf{C}) generated by the hurricane. Radiation stress gradient yields currents speed differences reaching 1 m/s, especially over offshore reefs.}
    \label{fig:diff}
\end{figure}

Our results suggest that the RS gradient alone can deflect particle trajectories by up to 1 km on the inner shelf and 5 km on the outer shelf (Fig. \ref{fig:traj}A,B). The RS mainly affects transport processes during the passage of the hurricane, as the distance between particle cloud advected by SLIM and SLIM+SWAN currents remains roughly constant afterwards. The Stokes drift, however, has a longer-lasting effect on the particle trajectories on the outer shelf. When adding a Stokes drift component to the Eulerian currents, the distances between the particle cloud centers keeps increasing during 2 days after the passage of Irma (Fig. \ref{fig:traj}B). Under the effect of the Stokes drift, particles from the outer shelf can be moved inshore on the passage of the hurricane. This motion is less pronounced for particles that are advected by Eulerian currents only. The particle cloud thus moves quickly northeastward under the action of the FC. After 2 days, the particles advected inshore under the action of the Stokes drift are in turn entrained by the FC and the distance between the clouds of particles starts decreasing. The impact of the Stokes drift on particle motion appears to be five times larger than the one of the RS on the inner shelf (Fig. \ref{fig:traj}A). However, both processes yield a similar impact on the particle trajectories at the moment of the passage of Hurricane Irma on the outer shelf (Fig. \ref{fig:traj}B).

Taking wave-currents interactions into account appears to significantly impact the modeled Stokes drift (Fig. \ref{fig:traj}C,D). Our results suggest that neglecting the wave-current coupling when computing the Stokes drift in storm conditions can yield deflections of the particle trajectories by up to 5 km on both the inner and outer shelves. On the outer shelf, differences in particle trajectories mostly appear during the two days following the passage of the hurricane. This is explained by the stronger shoreward component of the coupled SLIM+SWAN Stokes drift compared to the uncoupled one. On the inner shelf, however, differences in particle trajectories of up to 5 km occur at the moment of the passage of Hurricane Irma. The distance between the particle clouds then stabilizes directly after the passage of the hurricane.

% Milan: Table in complement of graph for comparison of trajectories ?

\begin{figure}
    \centering
    \includegraphics[width=.99\textwidth]{chapters/irma/figures/trajectory_comparisons.png}
    \caption{Difference between the centers of mass of the particle clouds released from the regions highlighted in Fig. \ref{fig:init} and advected by the different combinations of Eulerian currents and Stokes drifts described in Table \ref{tab:summary}}.
    \label{fig:traj}
\end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=.98\textwidth]{fig/dispersion.png}
%     \caption{Dispersion of particles with different velocity fields $\rightarrow$ not sure to know what to say about that...}   
%     \label{fig:dispersion}
% \end{figure}

% === DISCUSSION === %
\section{Discussion}
% SUMMARY
The coupled SLIM+SWAN model correctly reproduced the hydrodynamics and wave dynamics during Hurricane Irma. Such good agreement with field measurements could only be achieved using accurate forcings and adequate wave parameterizations. By comparing coupled and uncoupled model runs, we showed that neglecting wave radiation stress gradient can induce differences of up to 1 m/s in modeled current velocities. The radiation stress gradient during the hurricane was especially large over the shelf break, where waves are strongly dissipated by the offshore coral reef barrier. The radiation stress gradient alone can deflect drifting particles by up to 5 km during the passage of the hurricane. The impact of the Stokes drift dominates the effects of the radiation stress gradient on transport processes, except during the passage of the hurricane, when both contributions are similar on the outer shelf. The Stokes drift induces a shoreward transport during the passage of the hurricane that moves particles towards  the inner shelf, and hence away from the FC. Finally, neglecting wave-current interactions when computing Stokes drift leads to variations of up to 5 km in modeled trajectories on the passage of the hurricane.

The coupled wave-current model correctly reproduced the timing of the observed storm surges and captured the elevation peaks with a 15\% accuracy at every tide gauge except Virginia Key. Such accuracy is key to predict the damages caused by the hurricane, as they were mostly due to the storm surge and  high waves \citep{xian2018brief}. Furthermore, by using a high-resolution model, we can explicitly reproduce the circulation between all the reefs and islands of the Florida Keys. The fine-scale details of the storm surge, and hence the associated risk, are thus accurately represented. In addition to accurately capturing positive surges, the model also reproduced the observed negative surge in Naples. This result is of interest from a biological point of view as negative surges, although less studied, affect water exchanges between the estuaries and the coastal ocean and disturb the benthic ecosystems \citep{liu2020impacts}. Such rapid decrease in water level followed by a positive surge cause massive freshwater inflows, causing a significant decrease in water salinity \citep{wachnicka2019hurricane}. Surface waters are also significantly impacted by storms and hurricanes through induced cooling, upwelling and mixing \citep{varlas2020investigating}, but these processes were not accounted for in our model.

Strong currents such as the Gulf Stream affect waves through refraction over gradients in current velocity, shoaling and breaking of opposing waves or lengthening of following waves \citep{hegermiller2019wave}. Under hurricane conditions, these interactions can cause numerical instabilities in the wave model if the parameterizations are not appropriate and the model resolution not sufficient. \cite{hegermiller2019wave}, for instance, used a 5-km model grid and 48 directional bins to capture spatial gradients in wave height induced by wave-currents interactions in the Gulf Stream during Hurricane Matthew (2016). We followed these guidelines when defining the coarsest mesh resolution as the wave model spectral discretization. Boundary conditions and directional spreading of the incident waves also play a significant role when modeling wave-current interactions at meso- and submesoscales \citep{villas2020wave}, which motivated our choice of imposing full spectra on the boundary of the wave model instead of bulk parameters.

Tropical cyclones interact with the Gulf Stream and the FC through cooling and mixing of the upper ocean. These interactions can momentarily disrupt these currents and cause a significant reduction of their transport \citep{oey2007hurricane,ezer2017observations,ezer2020long}. As a 2D barotropic model, SLIM does not explicitly capture the vertical structure of the FC and Gulf Stream. Furthermore, a coupling with an atmospheric model would be required to represent heat fluxes between the upper ocean and the hurricane. However, we argue that a 2D model is sufficient for the scope of this study, that focuses on nearshore processes on the shelf and the shelf slope. Furthermore, by coupling the model with HYCOM, SLIM is able to represent indirectly the baroclinic features such as the meandering of the FC and eddie formation \citep{frys2020fine}. Furthermore, using a 2D model allows us to capture the impact of wave-current interactions on transport processes at the reef scale in the topologically complex coastal system of the FRT. Such a fine resolution is key to estimate the amount of wave energy dissipated over offshore reefs and accurately capture the generated RS gradient.

%Finally, SWAN default parameterizations for wind energy input and whitecapping caused numerical instabilities by overestimating wave growth and steepness on the boundary of the Gulf Stream on the passage of Irma. This overestimation was solved by using the parameterization of \cite{siadatmousavi2011evaluation}. The parameters used in this study were calibrated on the Northern Gulf of Mexico conditions, which might explain that our model better reproduces wave parameters at buoys located on the WFS. However, these calibrated parameters might underestimate wind-induced wave growth on Florida's eastern shelf. Consequently, incident wave do not receive enough energy to grow after breaking on the bank boundary, leading to an underestimation of the significant wave height at buoys located on Florida's eastern shelf. A more extensive calibration study might therefore be necessary to further improve the agreement with field measurements on both sides of Florida. Nonetheless, as this study focuses on the wave produced by Irma, which made landfall on the West coast of Florida, the use of parameterizations calibrated for the Gulf of Mexico seems reasonable.

The RS gradient significantly impacts currents during the passage of the hurricane. It can induce differences of up to 1 m/s in the current speed on the shelf break. In this region, waves are strongly dissipated due to action of depth-induced breaking and bottom dissipation on coral reefs. This link between wave breaking, RS gradient and wave-induced nearshore currents is consistent with previous studies on wave-current dynamics during storms \citep{mao2017dynamics,mao2018wave,mao2020particle}. Furthermore, our results highlight the protective role of coral reefs against strong incoming waves \citep{lowe2005spectral}, which requires a sufficiently fine spatial resolution to be explicitly represented in the model. As wave energy mostly dissipates on the shelf break, the impact of the RS gradient on transport processes is 5 times larger on the outer shelf. In the sheltered area of the inner shelf, wave impact on transport processes is dominated by the Stokes drift. Trajectory deflection under the influence of wave-induced motions mostly occurs at the moment of the passage of the hurricane on the inner shelf. After that moment, the distance between the clouds of particles remained roughly constant through time. On the outer shelf, RS and Stokes drift have a similar impact on transport processes at the moment of the passage of the cyclone and deflect particle trajectories by up to 5 km. However, by inducing shoreward transport, the Stokes drift delayed the advection of particles by the FC, therefore causing differences in trajectories of up to 20 km during the days following the passage of the hurricane. The dominance of the Stokes drift on particle transport in storm conditions was also observed in Lake Michigan by \cite{mao2020particle}. Finally, neglecting wave-current coupling in Stokes drift computation leads to differences in modeled trajectories of the order of 5 km on both the inner and the outer shelves. This fact, coupled with the impact of RS-induced currents strongly advocates for the use of coupled wave-current models when studying transport processes in storm conditions.

%Due to the dissipation of incoming waves on the reefs, wave impact during Irma is different on the inner and outer shelves. It is less important on the inner shelf because of the sheltering of the inner shelf due to reefs and islands as well as wave breaking on the shelf break. The inner shelf hence experiences weaker waves and currents, inducing weaker and more localized transport. Furthermore, the impact of winds on waves is reduced in shallower areas under the action of depth-induced breaking. This might explain why differences between particle trajectories stabilize on the inner shelf just after the passage of Hurricane Irma. However, the Florida Keys still experienced strong winds after the passage of the core of the hurricane, which generated high waves in the deeper areas. This might explain why the differences between the modeled trajectories kept increasing on the outer shelf under the action of strong Stokes drift up to two days after the passage of the hurricane.

\section{Conclusion}

We developed a coupled wave-current model to study the impact of waves on transport processes during Hurricane Irma. In order to accurately represent the wind and pressure profiles of the hurricane, we built hybrid fields by combining coarser ERA-5 data with high-resolution H*Wind data for the wind speed and idealized Holland profiles for the pressure. Comparing these hybrid profiles with field observations showed that they were better at reproducing the observed central depression of the hurricane as well as the peak in wind speed than ERA-5 data. Using these hybrid fields as forcings, our coupled model accurately reproduced the storm surge at tide gauge locations and produced currents and wave parameters in good agreement with field observations. The modeled currents and Stokes drift were then used to evaluate the impact of wave-current coupling on the modeled trajectories of passive drifters on the passage of the hurricane through the Florida Keys. Our results show that waves had a significant impact on heavy-wind transport processes and caused deflections of the drifters trajectories by more than 20 km on the outer shelf.

Despite its good agreement with observations, our model could be further refined by improving the representation of wind-wave interactions. In particular, we did not consider the momentum loss due to the action of surface waves, which can lead to overestimations of the momentum flux from the atmosphere to the ocean under hurricane conditions. Our model could therefore be further improved by using a wave-dissipative stress instead of the full wind stress as the momentum flux from the atmosphere to the ocean. As a 2D barotropic model, SLIM does not explicitly represent heat fluxes between the ocean and the atmosphere and the vertical structure of the ocean. However, our study focused on relatively shallow and vertically homogeneous coastal waters using a reef-scale resolution throughout the whole FRT. Such fine resolution allows to explicitly represent wave dissipation over coral reefs and is only achievable using a 2D model due to computational resource limitations.

Wave coupling needs to be taken into account during heavy-wind events but not necessarily in milder conditions. While the RS gradient plays an important role and can lead to differences of up to 5 km, the Stokes drift is about 4 times more intense and is thus the most important wave-induced transport process. Nonetheless, neglecting wave-current coupling through RS when modeling Stokes drift leads to differences of up to 5 km in predicted drifter trajectories. Such discrepancies reveal the strong influence of wave-current interactions on transport under storm conditions. This study brings new insight on the impact of waves on the transport processes nearshore during a tropical cyclone. Due to its fine spatial resolution, our coupled wave-current model can be used to accurately represent the dispersal of pollutants, sediments or larvae in topologically complex coastal areas in heavy-wind conditions.

### 2022-01-05 chapters/irma/irma.tex
slim+swan paper added

This chapter is based on the following article:
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{T. Dobbelaere}, M. Curcic, M. Le Hénaff, $\&$ E. Hanert, 2022, Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes, \textit{Ocean Modelling},69: 1121-1132. \href{https://www.frontiersin.org/article/10.3389/fmars.2020.591881 }{doi: 10.3389/fmars.2020.591881}
\end{list}

\begin{abstract}
    The intensity of major tropical cyclones has increased during the past decade. Their effect is particularly acute in coastal areas where they cause extensive damage leading to an influx of debris, sediments and waste to the sea. However, most operational coastal ocean models do not represent heavy-wind transport processes correctly \modif{if} the hydrodynamics is not coupled with the wind-generated waves. This may lead to significant errors in ocean simulations under tropical cyclone conditions. Here, we investigate current-wave interactions during a major hurricane and assess their impact on transport processes. We do that by coupling the unstructured-mesh coastal ocean model SLIM with the spectral wave model SWAN, and applying it to the Florida Reef Tract during Hurricane Irma (September 2017). We show that the coupled model successfully reproduces the wave behavior, the storm surge and the ocean currents during the passage of the hurricane. We then use the coupled and uncoupled wave-current model to simulate the transport of passive drifters. We show that the wave radiation stress gradient alone can lead to changes of up to 1 m/s in the modeled currents, which in turn leads to differences of up to 5 km in the position of drifting material over the duration of the hurricane. The Stokes drift however appears to cause deflections up to 4 times larger and hence dominates wave-induced transport. Wave-current interactions therefore strongly impact the transport of drifting material such as sediments and debris in the aftermath of a hurricane. They should thus be taken into account in order to correctly assess its overall impact.
\end{abstract}

%\newpage

\input{chapters/irma/article}

### 2022-01-06 chapters/drto/article.tex
chapter on drto added to thesis

\section{Introduction}
Over the last seven years, Florida's Coral Reef (FCR) has suffered from a widespread decline in living, reef-building corals due to a multi-year coral-disease outbreak \citep{precht2016unprecedented,walton2018impacts}, called stony coral tissue loss disease (SCTLD). First reported off the coast of Miami-Dade county in 2014, the disease has since spread through the entirety of FCR as well as other sites within the Caribbean \citep{alvarez2019rapid,kramer2019map,estrada2020reef} and affected more than 20 scleractinian coral species \citep{noaa2018}. The range of species affected coupled with the large spatial and temporal extents of the outbreak make SCTLD the most severe disease outbreak to have affected coral reefs on record.

Previous study of the spatio-temporal dynamics of SCTLD between West Palm Beach and Marathon showed that the epizootic propagated at a rate of $\sim$100 m/day to the north and $\sim$92 m/day to the south, hitting the Middle Florida Keys in 2017 \citep{muller2020spatial}. The disease was then reported in the Lower Keys in summer 2018 \citep{frrp2018}; near Key West in winter 2018 \citep{frrp2018}; on the easternmost boundary of the Marquesas subregion in October 2019 \citep{frrp2019}; and at its western end, approximately 11 km. west of halfmoon shoal, in August 2020 \citep{frrp2020}. SCTLD was observed propagating faster than 100 m/day within the Marquesas between October 2019 and August 2020 and would have reached the reefs of the Dry Tortugas (DRTO) by February 2021 if it had spread at a similarly high propagation rate \citep{kramer2019map}. These propagation rates were observed in a relatively dense shallow-water reef system and mostly occurred through transmission from neighbor to neighbor. Disease agents then had to cross 30 km ($\sim$20 miles) of open waters to reach the reefs of the DRTO. The disease was ultimately observed in May 2021 \citep{kramer2019map} in the southeast of the DRTO. To date, questions remain about the timing of the spread of the disease to the DRTO as well as the role played by hydrodynamics in the process.

The DRTO coral reefs are among the most preserved, but also vulnerable, marine ecosystems of the continental United States \citep{kourafalou2018physical}. They are part of the Florida Keys National Marine Sanctuary (FKNMS), under the jurisdiction of the Tortugas Ecological Reserve that covers around 566 km$^2$ of no-take marine reserves \citep{ault2006building}. Since the establishment of DRTO in July 2001, fish populations within the managed areas have significantly increased in size, adult abundance, and occupancy rates, therefore favoring the establishment of sustainable fisheries \citep{ault2013assessing}. Furthermore, because of their location and hydrodynamics, DRTO reefs are believed to be important sources of recruitment for coral-reef fishes downstream in the Florida Keys \citep{domeier2004potential} and thus significantly contribute to the overall resilience of the Florida Keys coral-reef ecosystem.

Although the pathogen responsible for the disease remains unknown to date, hydrodynamics are likely to play an important role in its propagation. Modeling studies suggest that the observed spread of the disease can be reproduced by simulating the dispersal of neutrally buoyant particles \citep{dobbelaere2020coupled}, while ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis, eaton2021measuring,meiling2021variable}. Accounting for the underlying ocean circulation is therefore important to understand the evolution of the propagation of the disease. The observed apparent stalling of the spread of SCTLD from the Marquesas to DRTO reefs may be explained by ephemeral or persistent oceanographic features in the area that initially prevented the transmission of disease agents further west.

The ocean circulation between the Marquesas and the DRTO is strongly driven by the Loop Current (LC). This ocean current brings warm waters from the Caribbean Sea through the Yucatan Channel, loops anti-cyclonically through the northeastern Gulf of Mexico (GoM) and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC). The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual,vukovich1988loop}. The dominant mesoscale feature near the DRTO is the formation of quasi-stationary eddies with a diameter of 100-200 km \citep{lee1994evolution,fratantoni1998influence}, usually called Tortugas gyres or eddies. \cite{fratantoni1998influence} hypothesized that the formation of these eddies as well as their lifespan was linked to the generation of anticyclonic rings from the LC in the GoM, while \cite{kourafalou2012florida} showed the relation between eddy formation dynamics near the DRTO and meandering of the FC. These eddies can remain up to 120 days near the DRTO before moving eastward at a speed of 5 km/day under the action of LC frontal eddies \citep{maul1977annual}. Tortugas gyres have been identified as an important mechanism for the potential retention of larvae of invertebrates and fishes on the southwest Florida shelf \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida} and might therefore have impacted the dispersal of disease agents from the Marquesas to the DRTO.

The objective of the present study was to apply the same modeling tools as \cite{dobbelaere2020coupled} to determine the potential impact of the circulation in the GoM on the delayed propagation of SCTLD between the Marquesas and the DRTO. This was performed by evaluating the evolution of the hydrodynamic-predicted connectivity between these two regions between May 2018 and May 2021. We then assessed how large-scale ocean circulation features, such as Tortugas gyres, could have prevented the transport of potential disease agents from the Marquesas to the DRTO. Ultimately, this study aims to provide novel insight on the role hydrodynamics have played in the spread of SCTLD throughout the western-most area of FCR, which further informs how it has, and will continue to, spread throughout the tropical western Atlantic

% === METHODS === %
\section{Methods}
Stony Coral Tissue Loss Disease data was obtained by using the Florida Reef Resilience Program's Disturbance Response Monitoring Report for 2019 and 2020 \citep{frrp2019,frrp2020} as well as the Atlantic and Gulf Rapid Reef Assessment (AGGRA) resources for SCTLD reported throughout the Caribbean, an open access spatial and temporal database curated by experts in Caribbean coral reef ecology \citep{kramer2019map}. Assumptions associated with SCTLD included the following: $(i)$ SCTLD is different from other tissue loss diseases previously reported on corals (i.e. white plague, white syndrome) based on its ecology and case definition \citep{noaa2018}, $(ii)$ SCTLD is transmitted, in part, by ocean currents \citep{aeby2019pathogenesis, muller2020spatial,eaton2021measuring} and $(iii)$ SCTLD was along the western most reef area of the Lower Florida Keys in October 2019, reached the western most part of the Marquesas in August 2020, and did not occur in DRTO until May 2021 based on curated data provided by the FRRP reports and the AGGRA open access database.

We studied the transmission patterns of SCTLD between the Marquesas and the DRTO by using the hydro-epidemiological model of \cite{dobbelaere2020coupled}. This model relies on the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}} to simulate the hydrodynamics over the entirety of FCR. The computational domain includes the Florida Straits as well as part of the GoM (Fig. \ref{fig:fig1_drto}a). SLIM is coupled with the large-scale ocean model HYCOM in order to correctly represent the dynamics of the FC which is a major western boundary current that flows out of the GoM, as a continuation of the Loop Current, before joining the Gulf Stream. SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features. Here we used the same mesh as \cite{dobbelaere2020coupled}, composed of $7\times 10^5$ elements, and with resolution that reaches 100 m over reefs and along the coastline (Fig \ref{fig:fig1_drto}a). By using such a fine resolution, we can explicitly represent circulation and resolve tides at the reef scale, such as around the Marquesas island group (Fig. \ref{fig:fig1_drto}a, inset), while also representing the large-scale flow features within the FC (Fig 1b and c). Further details on the model formulation and validation are provided by \cite{frys2020fine}.

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/drto/figures/fig1.jpg}
    \caption{(a) Mesh of the computational domain and close-up view of the instantaneous currents on November 15, 2020 at 12:00 in the Marquesas (inset), (b) close-up view of the monthly-averaged circulation between the Marquesas and the DRTO in October 2020, and (c) November 2020. (d) Reefs of the areas of the DRTO, the Western Marquesas and the Marquesas, as defined in this study. Land is shown in dark gray and reefs in light gray, except in panel (d) where reefs from the Marquesas, the Western Marquesas and the DRTO are respectively shown in green, yellow and red. Streamlines indicate the presence of the Tortugas gyre in October 2020, which leads to northward currents between the Marquesas and the DRTO. In November, the gyre left its static location to transit through the Straits of Florida and currents between the Marquesas and DRTO became westward.}
    \label{fig:fig1_drto}
\end{figure}

We simulated the dispersal of disease agents throughout FCR between May 2018 and May 2021. As in \cite{dobbelaere2020coupled}, disease agents were modeled as passive particles transported by depth-averaged currents which constantly lost mass with a half-life of 30 days. We ran the particle transport model for every month between May 2018 and May 2021 and hence obtained monthly normalized connectivity matrices. An entry $C_{ij}$ of the connectivity matrix represents the probability that disease agents produced on sub-reef $i$ settle on sub-reef $j$. These sub-reefs were defined by extracting polygons from the “coral reefs and hard bottom” layer of the Unified Florida Reef Tract Map \citep{fwc2017unified} and further dividing them into 500 m $\times$ 500 m squares, generating a total of $16,823$ sub-reef polygons. Since the connectivity matrices had more than 200 million entries, they were not directly interpreted. Instead, they were represented as large networks where nodes corresponded to sub-reefs of FCR and edges corresponded to nonzero entries in the connectivity matrix. Graph-theory algorithms were then used to analyze the properties of the connectivity networks.

To quantify the connectivity between the Marquesas and the DRTO, we defined 3 sub-regions within the sub-reefs of the western part of FCR: $(i)$ the Marquesas, $(ii)$ the Western Marquesas and $(iii)$ the DRTO (Fig. \ref{fig:fig1_drto}d). The sub-region of the Marquesas was defined as all the sub-reefs whose centroid is located between $-82.21^\circ$ and $-82.05^\circ$ longitude; the Western Marquesas included all sub-reefs between $-82.73^\circ$ and $-82.21^\circ$; and the DRTO were composed of  all the sub-reefs in our domain located west of $-82.73^\circ$ longitude. We evaluated the exchanges between two given sub-regions by identifying all edges connecting any sub-reef from one sub-region to any sub-reef from the other sub-region in our monthly connectivity networks. We then defined a connectivity indicator by multiplying the number of these direct edges by their mean connection probability. This indicator was computed monthly between May 2018 and May 2021 to evaluate the connectivity $(i)$ from the Marquesas to the DRTO, $(ii)$ from the Western Marquesas to the DRTO and $(iii)$ from the Marquesas to the Western Marquesas.

The spatial information of the network can be used to predict which part of the DRTO was likely to be first exposed to potential disease agents via hydrodynamic connections. We built a monthly spatial vulnerability indicator by summing the probabilities of all edges connecting sub-reefs of the Marquesas and the Western Marquesas to a given sub-reef of the DRTO. Sub-reefs of the DRTO with larger values of this indicator would have a larger cumulative likelihood of being infected during the simulated month. Such areas would thus be more likely to be the "entry points" of SCTLD in the DRTO during the month, as they were the destination of more pathways with larger exchange probabilities. A similar indicator can be computed for the Marquesas and the Western Marquesas by summing the probabilities of the edges connecting a given sub-reef of these sub-regions to the DRTO. Sub-reefs with higher value of this source indicator would be more likely to initiate the propagation of the disease to the DRTO, as they were the origin of more connectivity pathways with larger exchange probabilities. Model predictions were then compared to actual observations of the first cases of SCTLD in the DRTO collected on the interactive map of \cite{kramer2019map}.

To understand the effect of the ocean circulation on disease transmission, we computed the monthly-averaged circulation for each simulated month. These residual currents were used to visually assess the presence of a Tortugas gyre during that month, i.e. the presence of eastward moving eddies persisting on temporal and spatial scales consistent with the description of \cite{lee1994evolution} in the vicinity of the DRTO. Moreover, we defined an axis from the Marquesas to the DRTO over which we averaged the monthly residual currents. These temporally and spatially averaged currents were used as a proxy of the circulation between the Marquesas and the DRTO during each simulated month.

\section{Results}
The evolution of the monthly connectivity indicator suggests that there were no exchanges from the Marquesas and the Western Marquesas to the DRTO between February and October 2020, and between January and May 2021 (Fig. \ref{fig:fig2_drto}a). This is quite remarkable as outside of these periods of low connectivity, there were no periods greater than four months without exchange from both the Marquesas and the Western Marquesas between May 2018 and May 2021. Furthermore, February-October 2020 and March 2021 were the only months characterized by mean currents with an eastward component between the Marquesas and the DRTO (Fig. \ref{fig:fig2_drto}b). These periods of low connectivity and circulation inversion appeared correlated with the presence of eddies near the DRTO, as Tortugas gyres were observed during most months with no simulated connections to the DRTO. Additionally, no exchange occurred in September 2019 or from January to February 2021, during which a strong southward mean circulation occurred between the Marquesas and the DRTO (Fig. \ref{fig:fig2_drto}b). This suggests that southward currents might also act as a barrier to disease propagation by flushing away disease agents south to the GoM. Connectivity from the Marquesas to the Western Marquesas was never completely interrupted during our simulated period (Fig 2a). Nonetheless, periods of low connectivity to the DRTO occurred in the presence of Tortugas eddies (April-June 2020) and southward circulation (February 2021). This suggests that propagation of disease agents from the Marquesas to the Western Marquesas was routinely possible, even though exchanges may have been hindered by the local circulation. This latter result is consistent with the observed westward propagation of the disease in the Marquesas between October 2019 and August 2020.

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig2.jpg}
    \caption{(a) Monthly connectivity indicator between the Marquesas and the DRTO (blue), the Western Marquesas and the DRTO (orange) and the Marquesas and the Western Marquesas (green) between May 2018 and May 2021. Months during which Tortugas gyres were observed in the ocean model outputs are highlighted in grey. Milestones of the observed propagation of the SCTLD through the Marquesas are indicated by dark red dotted lines. (b) Monthly mean currents between the Marquesas and the DRTO between May 2018 and May 2021. Mean ocean circulation with an eastward component is highlighted by red arrows.}
    \label{fig:fig2_drto}
\end{figure}

We analyzed the spatial patterns of hydrodynamic-predicted disease connectivity to the DRTO in November 2020 in more detail as the model suggested this was the month with the highest transmission probability due to high hydrodynamic connectivity. The most probable pathways to the DRTO originated mainly from southern reefs located between the Marquesas and the Western Marquesas, as well as from reefs of the northwestern end of the Western Marquesas (Fig. \ref{fig:fig3_drto}). This suggests that sub-reefs of these areas were more likely to send disease agents to the DRTO in November 2020. Conversely, no connection with large probability originated from the southwestern tip of the Western Marquesas, suggesting that they had very low potential to spread the disease to the DRTO. \\
The reefs with the highest vulnerability indicator value in November 2020 were mostly located on the eastern part of the DRTO, suggesting that these reefs were more likely to be affected during this month (Fig. \ref{fig:fig4_drto}a). More specifically our results suggest that an area of higher vulnerability was present in the vicinity of the locations where SCTLD was observed in May 2021. Vulnerability was the lowest on inner sub-reefs of the DRTO as they are not directly connected to the reefs of the Western Keys. This suggests that the disease must first spread through the outer reefs to reach them. The source indicator values confirm that disease transmission to the DRTO is more likely to originate from southern reefs at the interface between the Marquesas and the Western Marquesas, and reefs on the northwestern end of the Western Marquesas.

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig3.jpg}
    \caption{Edges with largest connection probability to each sub-reef of the DRTO from the Marquesas and the Western Marquesas in November 2020. Connections mostly originated from the southern reefs between the Marquesas and Western Marquesas, and from the northwestern reefs of the Western Marquesas. Reefs are highlighted in light gray while islands are shown in dark gray.}
    \label{fig:fig3_drto}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig4.jpg}
    \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020. Eastrn reefs of the DRTO appear most vulnerable to the disease. Locations where the SCTLD was observed in the DRTO are indicated by red stars (b) Source indicator value in November 2020 for every sub-reef of the Marquesas and the Western Marquesas. Northwestern reefs of the Western Marquesas and southern reefs between the Marquesas and the Western Marquesas show a higher transmission potential to the DRTO. Main islands are shown in dark grey, separation between the Marquesas and the Western Marquesas is indicated by a dashed line.}
    \label{fig:fig4_drto}
\end{figure}

\section{Discussion and conclusion}
Our results show that the eddies generated by the LC/FC system strongly influence the connectivity between the Marquesas and the DRTO. Mesoscale eddies modify the westward coastal currents by breaking them through strong cross-shore flow regime \citep{kourafalou2012florida}, as illustrated for October 2020 in Fig. \ref{fig:fig1_drto}. \cite{lee1994evolution} hypothesized that such cross-shore flow could impact the transport of larvae spawned in the Florida Keys and cause their retention on the West Florida Shelf for periods on the order of 6 months. As Tortugas eddies may remain stationary near the DRTO for up to 120 days \citep{fratantoni1998influence}, they may thus act as barriers for the exchanges between the Marquesas and the DRTO for extended periods of time. Moreover, the passage of mesoscale eddies was identified as one of the physical mechanisms likely to deliver larvae from the DRTO to the Upper Keys, therefore opposing exchanges from the Marquesas to the DRTO. Our results confirm the role of mesoscale eddies as potential barriers to connectivity between these two regions as most months during which Tortugas eddies were present correspond to a vanishing connectivity indicator. The fact that no eddies were observed at the beginning of our simulation might be related to the identification method, based solely on visual assessment of the residual currents. A comprehensive identification of all Tortugas eddies during our simulated period would require the analysis of sea surface temperature and elevation, which is beyond the scope of the present study. While our approach might miss some Tortugas gyres, it filters out submesoscale eddies with lifespan of the order of several days. Consequently, when horizontally elongated mesoscale cyclonic streamlines are present in the computed monthly-averaged currents, they can be identified as Tortugas eddies with sufficient certainty.

The connectivity from the Marquesas to the DRTO does not seem to follow any clear seasonality and exhibits high interannual variability. Although a longer simulation period might be required to rigorously assess the presence of seasonal variations, no clear seasonal patterns were observed over the 3-year simulated period for both the connectivity indicator and the mean currents (Fig. \ref{fig:fig2_drto}). In this regard, the boundary of the West Florida Shelf significantly differs from the inner shelf, which exhibits a robust cycle between summer and winter months \citep{liu2012seasonal}. Furthermore, our results indicate that the connectivity from the Marquesas to the DRTO strongly varies from year to year. This is best illustrated in March-September 2020, during which no connectivity was observed for an extended period of time and the dominant currents between the Marquesas and the DRTO had an eastward component. Such behavior was not observed during the rest of the simulation. These results are consistent with the strong variability of the LC/FC system, which strongly impacts circulation near the DRTO. \cite{lee1994evolution} and \cite{fratantoni1998influence} first described the influence of the LC penetration and ring shedding dynamics in the GoM on the formation of eddies near the DRTO. \cite{kourafalou2018physical} then further identified that LC eddy separation events were related to the formation of southward currents over the Pulley Ridge reefs and the perturbation of the slope stratification on the Southwestern Florida Shelf. The detachment of anticyclonic eddies from the main LC affects in turn the position of the FC, which strongly impacts eddy activity near the DRTO and the Marquesas, either through the formation of local eddies or the retention of remote eddies near the DRTO \citep{kourafalou2012florida}. These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability and generate complex eddy to eddy interactions \citep{kourafalou2012florida}. Furthermore, the DRTO area was identified as a region of significant importance for the ecology of the West Florida Shelf, as contacts with the LC cause the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local,liu2016offshore}. \cite{weisberg2017loop} further hypothesized that these contacts might in turn impact the penetration of the LC in the GoM.

We have shown that the ocean circulation, and in particular the presence or absence of eddies shed from the Florida Current, significantly correlates with the timing of initial SCTLD observations and likely transmission to the DRTO. Intense eddy activity between February 2020 and May 2021 would have largely prevented the exchange of disease agents westward, with the exceptions of November and December 2020. During those two months, the residual circulation on the southwestern Florida shelf turned westward, and thus would have allowed disease agents to reach the DRTO. The transmission of disease agents to the DRTO was most likely to occur in November 2020, when the connectivity indicator was the largest. The developed vulnerability indicator indicates that sub-reefs on the eastern part of the DRTO were more likely to be affected by the disease. These results are consistent with observation as the locations where SCTLD was observed in the DRTO are located near sub-reefs identified as highly vulnerable by our model \citep{kramer2019map}. Nonetheless, the area predicted to be vulnerable exceeds the extent of the region where signs of the disease were first observed. These discrepancies might be explained by the fact that our vulnerability indicator accounts for all potential connections to a given sub-reef. In practice, these connections only yield the transmission of disease agents if the sub-reef they originate from is affected by SCTLD. Our model might therefore overestimate the vulnerability of some sub-reefs. Moreover, due to the lack of data on coral distribution in FCR, an uniform coral density was assumed over all reefs. The impact of some reefs with low coral cover might thus be overestimated by the model. Nonetheless, we argue that our model remains a valuable tool to assess the (absence of) connectivity between the Marquesas and the DRTO.

Model results suggest a time lag of 5 to 6 months between the moment the hydrodynamics predicted that disease agents reached the DRTO and the moment SCTLD symptoms were observed in the DRTO. This seems to contradict the results of previous modeling and experimental studies that found SCTLD transmission times of 10 days or less \citep{dobbelaere2020coupled, meiling2021variable}. However, several elements might explain a slower development of the disease on the DRTO. First, in situ conditions might significantly differ from the ones of a controlled tank environment, potentially leading to differences in the time required for the first disease lesions to appear. Furthermore, disease transmission in the Florida Keys might occur faster as reefs can receive disease agents from several affected neighboring reefs. In the case of disease agents reaching the fully susceptible population of the DRTO, such interactions with neighboring affected colonies do not take place and the outbreak might therefore require more time to develop. Furthermore, our results suggest that corals of the DRTO could only be exposed to agents coming from the Marquesas in November and December 2020, while exposure to incoming disease agents was continuous in the Keys due to the density of the reef system. Additionally, the epidemiological parameters of \cite{dobbelaere2020coupled} were calibrated on species-averaged prevalence data in the Florida Keys, which might not be representative of the colony composition and environmental conditions in the DRTO. Finally, observational data is dependent on the frequency and spatial distribution of field missions and might therefore give an incomplete picture of the situation on the field. For example, fine scale observations of SCTLD in the Lower Florida Keys found low disease incidence (< 10 colonies) starting in October, and then disease incidence peaked in April 2019 \citep{williams2021fine}. Alternatively, one might argue that this lag between the modeled transmission of the disease agents to the DRTO and the first observations of the SCTLD indicates that the propagation of the disease was not caused by local currents. In that case, another vector, such as ballast waters, might have to be considered. However, such investigations are beyond the scope of the present study.

It is important to note that the time and location of the first observations of an outbreak do not necessarily correspond to its initiation. An epidemic might grow unnoticed during its first stages, as it first develops at a linear rate before growing exponentially. Tools such as the ones developed in the present study might therefore have helped to mitigate the SCTLD during the first stages of its development in the DRTO, before May 2021. Furthermore, this study has shown that strong eddy activity near the DRTO was likely to interrupt connectivity from the Western Florida Keys to the DRTO. Therefore, monitoring eddy activity through satellite observations of sea surface temperature and sea surface elevation anomaly \citep{kourafalou2018physical} might serve as a proxy for reef managers to estimate connectivity between these two areas. The implications of our connectivity results are not limited to the propagation of disease agents and apply to the spread of any type of propagules from the Marquesas and the DRTO. This study therefore brings additional insight on the physical and environmental connectivity in the southwestern part of FCR.

### 2022-01-06 chapters/drto/drto.tex
chapter on drto added to thesis

This chapter is based on the following article:
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%

\item \textbf{T. Dobbelaere}, D. H. Holstein, E. M. Muller, L. J. Gramer, L. McEachron, S. D. Williams$\&$ E. Hanert, Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas (submitted to \textit{Frontiers in Marine Science})
\end{list}

\begin{abstract}
    For the last seven years, Florida's Coral Reef (FCR) has suffered from widespread and severe coral loss caused by stony coral tissue loss disease (SCTLD). First observed off the coast of Miami-Dade county in 2014, the outbreak has since spread throughout the entirety of FCR and some areas of  the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached the western end of the Marquesas in August 2020. Despite being present about 30 km ($\sim$20 miles) from the Dry Tortugas (DRTO), SCTLD was not reported in this area before May 2021. As SCTLD transmission is likely to be waterborne, here we suggest that this apparently delayed propagation is related to eddy activity near the DRTO under the influence of the Loop Current/Florida Current system. To quantify the impact of the local ocean circulation on the spread of SCTLD from the Marquesas and the DRTO, we evaluated the hydrodynamic-predicted connectivity between these two regions using a high-resolution hydro-epidemiological model between May 2018 and May 2021. Our results suggest that the Marquesas and the DRTO were not connected during February-October 2020 and January-May 2021. These periods coincided with either the occurrence of Tortugas gyres and mean circulation with an eastward component between the Marquesas and the DRTO or the presence of southward currents. Our results suggest that disease agents probably reached the DRTO in November 2020 and that they most likely originated from southern or northwestern reefs of the Marquesas. This study provides novel insight into the role played by the hydrodynamics in the spread of SCTLD within the western-most edge of FCR, and in propagating the disease to uninfected locations.
\end{abstract}

%\newpage

\input{chapters/drto/article}

### 2022-01-06 chapters/irma/irma.tex
chapter on drto added to thesis

@@
 
-\item \textbf{T.
-Dobbelaere}, M.
-Curcic, M.
-Le Hénaff, $\&$ E.
-Hanert, 2022, Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes, \textit{Ocean Modelling},69: 1121-1132.
-\href{https://www.frontiersin.org/article/10.3389/fmars.2020.591881 }{doi: 10.3389/fmars.2020.591881}
+\item \textbf{T. Dobbelaere}, M. Curcic, M. Le Hénaff, $\&$ E. Hanert, Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes (submitted to  \textit{Ocean Modelling})
 \end{list}

### 2022-01-31 chapters/drto/article.tex
draft onset added to manuscript

@@
     \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020.
-Eastrn reefs of the DRTO appear most vulnerable to the disease.
+Eastern reefs of the DRTO appear most vulnerable to the disease.
 Locations where the SCTLD was observed in the DRTO are indicated by red stars (b) Source indicator value in November 2020 for every sub-reef of the Marquesas and the Western Marquesas.
@@
 
-We have shown that the ocean circulation, and in particular the presence or absence of eddies shed from the Florida Current, significantly correlates with the timing of initial SCTLD observations and likely transmission to the DRTO.
-Intense eddy activity between February 2020 and May 2021 would have largely prevented the exchange of disease agents westward, with the exceptions of November and December 2020.
+We have shown that the ocean circulation, and in particular the presence or absence of eddies shed from the Florida Current, coincides with the timing of initial SCTLD observations and likely transmission to the DRTO.
+However, it is important to note that it is currently not possible to determine whether SCTLD is caused by a novel pathogenic agent or simply a variant of a previously existing disease that causes tissue loss on corals.
+The DRTO has experienced episodic outbreaks of tissue loss in years prior to 2020 (see \citealp{frrp2018}).
+These previously documented tissue loss events occurred during the summer months within isolated reef sites and primarily affected one genus of corals, \textit{Orbicella}.
+Although the ecology of SCTLD documented from 2014 to the present differs from these previous reports, until pathogenic agents are identified, the positive identification of a particular coral disease, such as SCTLD, will remain presumptive.
+However, past studies indeed suggest SCTLD is likely caused by a waterborne agent regardless of the particular pathogen at play.
+As such, intense eddy activity between February 2020 and May 2021 would have largely prevented the exchange of disease agents westward, with the exceptions of November and December 2020.
 During those two months, the residual circulation on the southwestern Florida shelf turned westward, and thus would have allowed disease agents to reach the DRTO.

### 2022-01-31 include/abstract.tex
draft onset added to manuscript

\chapter*{Abstract}
%\phantom{Abstract}

The population of corals throughout the world has declined dramatically under the effect of rising water temperature and many coral species are now listed as threatened. In addition to ocean warming, other threats to corals include (but are not limited to) disease outbreaks, intensifying hurricanes and sediment disturbances. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of dredging operations in the Port of Miami, the disease has now spread through the entirety of FCR and has been observed in numerous territories of the Caribbean. Furthermore, hurricanes tend to strike Florida during the reproduction period of the corals, when thousands of larvae are released in the water column. In addition to causing physical harm to corals, hurricanes significantly modify ocean currents and might therefore disturb larval exchanges between reefs. A high resolution ocean model, SLIM, is used to build a biophysical model to study the aforementioned threats to FCR. First, a coupled epidemio-hydrodynamic model is developed to study both the onset and the propagation mechanisms of the SCTLD. Furthermore, a coupled wave-current model is devised to study the impact of Hurricane Irma (2017) on transport processes in the Florida Keys. This study brings new insight on the impact of hydrodynamics on the propagation of SCTLD as well as the impact of hurricane on coral ecosystem.

%The population of corals throughout the world has declined dramatically and many coral species are now listed as threatened Avoiding global warming being no longer an option, reef managers now focus on damage control and restoration strategies. The Florida Reef Tract, currently facing an unprecedented coral disease outbreak and where healthy corals grown in nurseries are replanted on damaged reefs, is a good application case of such reef management strategies. In order to maximize the efficiency of these efforts, a good knowledge of exchanges of infectious materials and larvae between reefs is required. However, connectivity patterns between reefs remain mostly unknown. To address this issue, we will use the high-resolution ocean model SLIM to simulate connectivity patterns between reefs and then identify reefs with the highest restoration potential as well the ones that are most at risk of being infected by the outbreak.

%The dispersal of coral larvae away from their natal habitats is an important process for coral reef ecosystems, but remains poorly understood and hard to gauge. This thesis presents a spatially explicit numerical modelling approach to simulate larval dispersal in Australia's Great Barrier Reef (GBR), and in doing so to estimate connectivity between the thousands of reefs that comprise it. A high resolution, finite element ocean model, SLIM, is used to build a biophysical model of larval dispersal capable of estimating larval exchange between reefs in large sections of the the GBR. Tools from network science are employed to investigate the community structure of the GBR's connectivity network, and these tools are used to reveal the presence of clusters of highly inter-connected reefs. Subsequently, the model is used to estimate the extent of connectivity between reefs of different morphologies and depths in the GBR. Finally, the possible impacts of climate change on connectivity in the GBR are assessed, and estimates are made of the likely changes to coral larval dispersal patterns over the coming century.

%Titan, Saturn’s largest moon, is a unique object in the Solar system as much as it has a substantial atmosphere, a surface with a complex interplay of geological processes and an outer ice shell overlyinga subsurface ocean. The climate on this icy satellite boasts a multi-phase hydrological cycle where methane plays a role similar to that of water on Earth. Surface lakes and seas filled with liquid methane and ethane are found in the polar regions. The subsurface ocean, on the other hand, is filled with liquid water. This liquid layer lying beneath Titan’s surface allows for the large surface deformation observed.

%The first objective of this thesis was to adapt an Earth-based geophysical and environmental model, SLIM (www.climate.be/slim), to Titan’s specific conditions in order to study the tidal motion in the surface lakes and seas. The modified model was applied to the largest lake in the southern hemisphere, Ontario Lacus, and the two largest seas, Kraken and Ligeia Maria, which are located in the northern hemisphere. The predicted surface elevation and velocity fields are part of the data needed to develop an exploration mission focusing on the surface lakes and seas of Titan. The normal modes of Ontario Lacus were also numerically studied. While resonantly forced modes could generate significant liquid motion, the natural periods are much shorter than the period of the astronomical forcings. Therefore, only atmospheric forcings could resonantly force the normal modes. Strong wind conditions corresponding to a stormy event are required to generate a surface elevation resulting in significant shoreline variations.

%Then, the model was used to study the tides of Titan’s global subsurface ocean. To this end, the shallow water equations and the 3D hydrostatic equations under the Boussinesq approximation were modified in order to take into account the ice shell lying at the top of the ocean. The method was adapted to solve the equations on a sphere. The effects of the shell are represented by adding a surface friction and a surface pressure term. The shell decreases the ocean surface elevation and slows down the flow without modifying significantly the global patterns of these fields. The magnitude of these variations depends on the mechanical behaviour of the ice shell. Surface and bottom heat fluxes play a significant role in the liquid motion of the ocean. The interactions between the tidal motion and the thermally driven flow resulting from the surface heat flux were studied by means of the 3D version of SLIM. The surface heat flux significantly impacts the velocity field, both in terms of magnitude and orientation while the influence on the ocean’s surface elevation is small. These results could be useful for astrobiologists paying attention to the ocean habitability and to validate hypotheses about Titan’s internal structure.

### 2022-01-31 chapters/sctld/article.tex
slim+swan paper added
removed duplicated ref

@@
 The mesh resolution depended only on the distance to the coast, but we distinguished between the coastlines along the FRT where we imposed a maximum resolution of 100 m and the other coastlines along which the finest resolution was 2500 m.
-The mesh was generated with the open-source mesh generator GMSH \citep{Geuzaine2009} and is composed of approximately $7 \times 10^5$ elements.
+The mesh was generated with the open-source mesh generator GMSH \citep{geuzaine2009gmsh} and is composed of approximately $7 \times 10^5$ elements.
 The coarsest elements, far away from the FRT, had a characteristic length size of about 10 km.
@@
     \centering
-    \includegraphics[width=.8\textwidth]{chapters/sctld/figures/sctld_validation_corrected.jpg}
+    \includegraphics[width=.9\textwidth]{chapters/sctld/figures/sctld_validation_corrected.jpg}
     \caption{Summary of epidemiological model simulations with calibrated transmission parameters.

### 2022-02-06 onset/main.tex
abstract added

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[preprint,12pt,authoryear]{elsarticle}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage{color,tensor,multirow,siunitx}
\usepackage[onehalfspacing]{setspace}
\usepackage{makecell}
\renewcommand{\cellalign}{cl}
\usepackage{caption}
\usepackage{lipsum}
\captionsetup[figure]{name=Fig., labelfont=bf}
\captionsetup[table]{name=Table, labelfont=bf}

\renewcommand{\cellalign}{cl}

\newcommand{\ds}{\displaystyle}
\newcommand{\nl}{\ \\ }
\newcommand{\ud}{\textrm{ d}}
\newcommand{\bs}{\bigskip}

\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bR}{\mathbf{R}}

\definecolor{orange}{rgb}{1.0, 0.46, 0.09}
\definecolor{red}{rgb}{1,0,0}
\definecolor{blue}{rgb}{0,0,0.8}
\definecolor{green}{rgb}{0,0.5,0}
\newcommand{\emphc}[1]{\emph{\textcolor{red}{#1}}}
\newcommand{\hycom}{\textsc{hycom} }
\newcommand{\ie}{{\it i.e.}\ }
\newcommand{\eg}{{\it e.g.}\ }
\newcommand{\UV}{\mathbf{U}}
\newcommand{\todo}[1]{\textcolor{red}{TO DO: #1}}
\newcommand{\comment}[1]{\textcolor{orange}{#1}}
%\newcommand{\modif}[1]{\textcolor{blue}{#1}}
% \renewcommand{\familydefault}{\sfdefault}

\journal{Marine Pollution Bulletin}

\begin{document}

\begin{frontmatter}

    %% Title, authors and addresses

    %% use the tnoteref command within \title for footnotes;
    %% use the tnotetext command for theassociated footnote;
    %% use the fnref command within \author or \affiliation for footnotes;
    %% use the fntext command for theassociated footnote;
    %% use the corref command within \author for corresponding author footnotes;
    %% use the cortext command for theassociated footnote;
    %% use the ead command for the email address,
    %% and the form \ead[url] for the home page:
    %% \title{Title\tnoteref{label1}}
    %% \tnotetext[label1]{}
    %% \author{Name\corref{cor1}\fnref{label2}}
    %% \ead{email address}
    %% \ead[url]{home page}
    %% \fntext[label2]{}
    %% \cortext[cor1]{}
    %% \affiliation{organization={},
    %%            addressline={}, 
    %%            city={},
    %%            postcode={}, 
    %%            state={},
    %%            country={}}
    %% \fntext[label3]{}

    \title{Early transmission dynamics of SCTLD and its connection with the dredging works of Port of Miami}%Impacts of Hurricane Irma (2017) on ocean transport processes}

    %% use optional labels to link authors explicitly to addresses:
    %% \author[label1,label2]{}
    %% \affiliation[label1]{organization={},
    %%             addressline={},
    %%             city={},
    %%             postcode={},
    %%             state={},
    %%             country={}}
    %%
    %% \affiliation[label2]{organization={},
    %%             addressline={},
    %%             city={},
    %%             postcode={},
    %%             state={},
    %%             country={}}

    \author[eli]{Thomas Dobbelaere\corref{corr}}
    \ead{thomas.dobbelaere@uclouvain.be}
    %\author[rsmas]{Milan Curcic}
    %\author[cimas,aoml]{Matthieu Le H\'enaff}
    \author[eli,immc]{Emmanuel Hanert}
    \cortext[corr]{Corresponding author}

%    \affiliation[eli]{
%        organization={Eath and Life Institute (ELI), UCLouvain},
%        % addressline={}, 
%        city={Louvain-la-Neuve},
%        % postcode={1348}, 
%        % state={},
%        country={Belgium}
%    }
%    \affiliation[rsmas]{
%        organization={Rosenstiel School of Marine and Atmospheric Sciences (RSMAS), University of Miami},
%        % addressline={}, 
%        city={Coral Gables},
%        % postcode={1348}, 
%        state={Florida},
%        country={USA}
%    }
%    \affiliation[cimas]{
%        organization={Cooperative Institute for Marine and Atmospheric
%        Studies (CIMAS), University of Miami},
%        % addressline={}, 
%        city={Miami},
%        % postcode={1348}, 
%        state={Florida},
%        country={USA}
%    }
%    \affiliation[aoml]{
%        organization={Atlantic Oceanographic and Meteorological Laboratory
%        (AOML), NOAA},
%        % addressline={}, 
%        city={Miami},
%        % postcode={1348}, 
%        state={Florida},
%        country={USA}
%    }
    \affiliation[immc]{
        organization={Institute of Mechanics, Materials and Civil Engineering (IMMC), UCLouvain},
        % addressline={}, 
        city={Louvain-la-Neuve},
        % postcode={1348}, 
        % state={},
        country={Belgium}
    }

    \begin{abstract}
        For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD). The outbreak reportedly initiated near Virginia Key in September 2014, during the the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015. The disease then spread to the entirety of FCR, likely through waterborne transmission. Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease. The goal of this study is therefore to evaluate whether sediments produced by the dredging might have been transported to reefs where disease was reported and to assess whether the timing of these inputs of sediment was consistent with the observed spread of the disease. We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment modeled forced by currents from the high-resolution coastal ocean model SLIM. Our result suggested that the dredging had close to no impact on the coral reefs of Virginia Key. However, a significant fraction of the produced sediments reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites. Furthermore, using a previously developed bio-physical model indicated that waterborne transmission from one of these sites to Virginia Key was possible before September 2014. This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst outbreak on record in the Caribbean.
    \end{abstract}

    %%Graphical abstract
    % \begin{graphicalabstract}
    % %\includegraphics{grabs}
    % \end{graphicalabstract}

    %Research highlights
    % \begin{highlights}
    %     \item The coupled SLIM+SWAN model correctly reproduced the hydrodynamics and waves during Irma.
    %    \item Wave radiation stress increased currents by up to 1 m/s during Irma.
	%    \item Wave radiation stress gradients were the largest on the shelf break and over reefs. 
	%   \item Waves could deflect drifting particles by up to 20 km during the hurricane.
	%  \item The Stokes drift had an impact on transport 4 times larger than the wave radiation stress.
    % \end{highlights}

    \begin{keyword}
        Stony coral tissue loss disease \sep Sediments \sep Port of Miami \sep Coastal modeling \sep Sediment plumes 
    %% keywords here, in the form: keyword \sep keyword

    %% PACS codes here, in the form: \PACS code \sep code

    %% MSC codes here, in the form: \MSC code \sep code
    %% or \MSC[2008] code \sep code (2000 is the default)

    \end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}

%  PARAGRAPH DISEASES IN GENERAL
Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region \citep{richardson1998coral, sutherland2004disease, aronson2001white, harvell2007coral, brandt2009dynamics}. One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is stony coral tissue loss disease (SCTLD) \citep{noaa2018}. First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entirety of FCR \citep{muller2020spatial,dobbelaere2022} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}. Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}. Furthermore, recent studies showed evidence that sediments act as vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

% POM
Some of the first signs of SCTLD were reported in September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}. In this study, a relationship was found between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key. It was therefore hypothesized that the epidemic started near Virginia Key and then radiated to the neighboring reef, both north and south. However, earlier signs of diseases were reported north of Virginia Key in June 2014, at the monitoring site of N. Sunny Isles \citep{precht2016unprecedented}. All these observations occurred during the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015. The dredging was monitored twice-weekly at 26 permanent monitoring stations established within the Miami-Dade County, making it one of the most complete datasets related to a dredging project \citep{gintert2019regional}. Interestingly, signs of disease were also reported at one these monitoring stations in May 2014. While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located 4.7 nautical miles offshore. However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}. The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report}. Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.

% SEDIMENTS
Sediments released by dredging can affect the biological functions of corals in numerous ways  through turbidity and sedimentation \citep{erftemeijer2012environmental, jones2015effects}. Increased turbidity caused by the suspended sediments reduces the light available to symbiotic zooxanthellae, leading to reduced coral cover and growth. Sedimentation, on the other hand, can cause smothering or burial of coral polyps \citep{erftemeijer2012environmental}. Furthermore, both sedimentation and turbidity can significantly reduce larval recruitment by inhibiting settlement and reducing larval survival in the water column \citep{jones2015effects}. These effects are stronger with fine-grained sediments, as they cause a stronger light reduction \citep{fourney2017additive}. Additionally, fined-grained sediments such as silts have high nutrient contents, which can lead to an increased microbial activity, eventually causing anoxic conditions in the immediate vicinity of corals \citep{weber2012mechanisms}. As they release finer sediments over significantly longer periods than natural events such as hurricanes, dredging activities can thus be more harmful to corals and reef habitat compared to other types of sedimentation \citep{cunning2019extensive}.

% ON ENTRE DANS LE VIF DU SUJET
Nonetheless, \cite{gintert2019regional} argued that the reported coral mortality during the dredging project was dominated by the regional outbreak of SCTLD. Further, the study suggested that the onset of the disease might have been linked to a leaking discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant (WWTP) located off Virginia Key. However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease was transported to the monitoring site of Virginia Key on sediments released by the dredging. This possibility can be evaluated using a hydrodynamic model simulating the transport of sediments produced during the dredging project. As coastal reef ecosystems are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures, such a model would require high spatial resolution to accurately represent the transport of sediments at reef-scale. In this context, unstructured-mesh models are best suited, as they can easily adapt to the topography \citep{fringer2019future} and can capture small-scale circulation features around reefs and islands \citep{lambrechts2008multi}.

The goal of this study is therefore to simulate the trajectories of the sediments released during the entirety of the deepening of the PoM shipping channel using a high-resolution hydrodynamic model coupled with a sediment transport model. Specifically, we will attempt to answer the following questions: (1) Which reefs were impacted by the dredging ? (2) Is the impact on these reefs consistent with the observed timing of the onset of SCTLD ?

% === METHODS === %
\section{Methods}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/fig_mesh_onset.png}
    \caption{\textbf{A}: Model mesh near the dredged channel. Elements have a characteristic length of 100 m over reefs (in light grey) and along the coasts (in dark gray). The monitoring sites considered in the present study are shown by blue dots. The date where SCTLD was first observed at these sites is given between brackets. The Ocean Dredge Material Disposal Site (ODMDS) is shown in red. \textbf{B}: Close up view of the dredged channel. The locations of the different types of dredging that took place during the expansion of PoM are shown by colored markers. \textbf{C:} Snapshot of the modeled currents in the vicinity of the dredged channel. Small-scale flow features such as the acceleration of currents between reefs and between islands are well captured by the model.}
    \label{fig:onset_mesh}
\end{figure}

The hydrodynamics of the entirety of FCR was modeled using the high resolution unstructured-mesh model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022}. SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features. The mesh used in this study was built following the same methodology as \cite{dobbelaere2022}, with a local refinement near PoM and in the Bay of Biscayne to achieve a resolution of 100 m in the vicinity of the dredged channel (Fig. \ref{fig:onset_mesh}A). It was made up of approximately $3.5\times 10^5$ triangles and was generated with the seamsh\footnote{\url{https://pypi.org/project/seamsh/}} Python library, which is based on the the open-source mesh generator GMSH \citep{geuzaine2009gmsh}. The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the first observation of SCTLD by \cite{precht2016unprecedented}. Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the accelerations between reefs and between islands.

The transport of sediments released from the channel was then modeled using a Lagrangian particle tracking model, forced by SLIM velocity field. The sediment model is inspired by the Particle Transport Model (PTM), developed by the US Army Corps of Engineers \citep{macdonald2006ptm}. In this model, particles undergo a combination of horizontal and vertical motions. The vertical is mostly driven by gravity, with heavier particles sinking faster. Once they have settled, particles can be resuspended when shear stress exceed the critical Schields parameter, as parameterized by \cite{soulsby1997threshold}. The horizontal motion of the suspended particles is derived from the 2D model velocity by assuming a vertical log profile, following a quasi-3D approach. When sediment particles enter the near-bed zone, their horizontal velocity is greatly reduced and sediments are transported with the bedload. The 2D model of this particle tracker can also be used to model the transport of neutrally buoyant material driven by mean current within the water column \citep{dobbelaere2020coupled}. This model was used to model the impact of wastewater leaking from the discharge pipe mentioned by \cite{gintert2019regional}. Without clear information about the location of the leak, particles were continuously released every 100 m along a straight line between Miami Central District Municipal WWTP, located on Virginia Key and the ocean discharge outfall located at 25$^\circ$44'31''N 80$^\circ$05'10''W \citep{koopman2006ocean}.

As sediment dispersion is dependent on the grain size, we modeled the dispersal of five classes of sediments to represent to impact of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m. We performed a  different simulation for each class, with the grain size randomly drawn from a uniform distribution over the corresponding size range. The density of each sediment particle was derived from their size using the formula of \cite{hamilton1982sound}. Furthermore, all particles were differentiated based on the type of dredge that produced them. Five types of dredge were considered in our modeling study (Fig. \ref{fig:onset_mesh}B): ($a$) Texas cutterhead (TX), ($b$) non-conventional dredging, \ie TX with suction mechanism turned off (NonConv), ($c$) Spider Barge (SB), ($d$) Terrapin Island hopper (TI), and ($e$) Dredge 55 clamshell (D55).

Dredging operations performed during the expansion of PoM were characterized in our dataset by a date, a location and  a type of dredging operation (Fig. \ref{fig:onset_mesh}B). In the absence of information about the exact time of the dredging, sediment particles were released from the dredging location during a whole day at a rate of 80 particles/hour in the model. To account for the motion of spider barges between the dredging site and disposal site, particles were released every 500 m along a straight line joining the dredging location to the ODMDS (see Fig. \ref{fig:onset_mesh}A) for every dredging operation labelled as SB.

The outputs of the sediment model were then used to evaluate the turbidity and sedimentation generated by the expansion of PoM over coral reefs. The occurrence of high turbidity over reefs was assessed based on the concentration of suspended sediment particles in the model. This concentration was computed by counting the number of suspended particles inside the cells of a regular 200 m $\times$ 200 m grid over our computational domain. The modeled occurrence of plume was then compared against daily data of plume detection. This data set was derived from satellite imagery by \cite{cunning2019extensive}\footnote{data sets available at \url{https://github.com/jrcunning/pom-dredge/}} following the methods of \cite{barnes2015sediment} at sites located within 15 km of the dredged channel. As in these two previous studies, we computed  the simulated plume frequency by dividing the number of days during which plumes occurred by the total number of simulated days for all grid cells. The impact of sedimentation was quantified by computing the cumulated concentration of settled particles within the same computational grid. This cumulated concentration was normalized by the total number of simulated time steps to obtained an averaged concentration of deposited particles over the simulated period. Higher values of this indicator would indicate larger sediment deposition over longer cumulated time.

The potential impact of dredging on the onset of the outbreak was more specifically assessed at five monitoring sites were signs of disease were reported. Four sites were reefs where disease was reported in 2014 by \cite{precht2016unprecedented}. The fifth site was monitoring station HBSC1-CP of the monitoring of the expansion of PoM, where signs of disease were reported in May 2014 (Fig. \ref{fig:onset_mesh}A) \comment{[Photos in appendix ? ]}. The impact of dredging was assessed by counting the number of sediment particles originating from each dredging that were transported within 500 m of all five site. This number was then divided by the total number of sediment particles released by each type of dredge. Larger values of this indicator would suggest a greater impact of a given type of dredging at a given monitoring sites. This same indicator was used to evaluate the impact of particles released along the discharge pipe at the site of Virginia Key.

Finally, as previous studies showed evidence of waterborne transmission of SCTLD \citep{aeby2019pathogenesis, dobbelaere2020coupled,eaton2021measuring, meiling2021variable}, there is a possibility that the disease propagated to Virginia Key from other diseased reefs affected prior to September 2014. To evaluate this possibility, we computed monthly disease connectivity matrices following the methodology of \cite{dobbelaere2020coupled} during our simulated period. These connectivity matrices can be interpreted as large graphs whose vertices are sub-areas of reefs and whose edges represent disease connectivity pathways. Evaluating the possibility of disease propagation from any reef $A$ to any reef $B$ is therefore equivalent to evaluating the existence of paths, \ie sequences of connected vertices in the network starting from $A$ and reaching $B$. As computing all possible paths is not computationally tractable, we limited ourselves to the computation of shortest paths from any given reefs to the Virginia Key monitoring site. This was performed using the function \texttt{get\_all\_shortest\_paths} of the Python \texttt{python-igraph} package \citep{csardi2006igraph}. Such function requires the definition of a weight $w_{ij}$ for the edge connecting reef $i$ to reef $j$. We chose $w_{ij} = 1-\tilde{C}_{ij}$, where $\tilde{C}_{ij}$ is the probability of disease propagation from reef $i$ to reef $j$, so that "shorter" edges of the networks (\ie connectivity pathways with smaller weights) correspond to connections with larger disease propagation probability. The probability of a given path was then defined as the mean connection probability of the edges composing this path.

%%%%%%%%%%%%%%%%%%%
% --- RESULTS --- %
%%%%%%%%%%%%%%%%%%%
\section{Results}

5-50 $\mu$m was the only grain size that produced plumes consistent with the observations of \cite{barnes2015sediment} and \cite{cunning2019extensive} (Fig. \ref{fig:onset_depo}A). For larger grain sizes, particles settled in the direct vicinity of the channel. For these heavier particles, suspended sediments were only observed offshore, closer to the Florida Current, where current velocity was sufficiently strong to prevent deposition. This suggests that the observed turbidity was mostly due to fine silts. The modeled occurrence of plumes within 15 km of the dredged matched the presence/absence data of \cite{cunning2019extensive} in 82.7\% of cases and the total area where plumes were observed was about 230 km$^2$, consistent with the $\sim228$ km$^2$ estimated by \cite{barnes2015sediment}. Modeled plumes mostly occurred north of the dredged channel, as particles were driven northward under the action of the Florida Current, with plume frequencies of about 20\% near the monitoring site of N. Sunny Isles. Plume frequencies was more limited over offshore but reached of up to 50\% over inshore reefs. These inner reefs correspond to regions of important sedimentation (Fig. \ref{fig:onset_depo}B), which is consistent with the correlation between plumes and sedimentation highlighted in \cite{cunning2019extensive}. The plume frequency was about 1\% at site of HBSC1-CP while no plume occurred at other southern monitoring sites during the simulation.

% \begin{table}
%     \centering
%     \begin{tabular}{|l|cccc|}
%         \hline
%          & Threshold         & Accuracy & False positive  & False negative \\
%          & [particles/m$^2$] & [\%]     & [\%]            & [\%]  \\
%         \hline
%         5-50 $\mu$m    & 21 & 86.678 & 2.912 & 10.410 \\
%         50-100 $\mu$m  & 20 & 83.058 & 8.525 & 8.417 \\
%         100-200 $\mu$m & 30 & 84.814 & 7.133 & 8.053 \\
%         200-300 $\mu$m & 18 & 84.459 & 7.988 & 7.553 \\
%         300-400 $\mu$m & 13 & 84.615 & 8.140 & 7.244 \\
%         \hline
%     \end{tabular}
%     \caption{Validation of the modeled presence of plumes against plume observations from satellite imagery. Overall, the model agrees well with observations}
%     \label{tab:onset_val} 
% \end{table}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/deposition_plumes.png}
    \caption{\textbf{A}: Plume frequency for grain size in the range 5-50 $\mu$m. The averaged concentration of deposited sediments are shown for grain sizes more likely to carry the disease: 5-50 $\mu$m (\textbf{B}) and 50-100 $\mu$m (\textbf{C}). Most modeled turbidity and sedimentation occurred north of the dredged channel. Non negligible sediment deposition occurred at site HBSC1-CP.}
    \label{fig:onset_depo}    
\end{figure}

Deposition results are shown for grain sizes corresponding to silts, which are more likely to carry organic matter and therefore more likely to carry SCTLD agents (Fig. \ref{fig:onset_depo}B,C). Sedimentation mostly occurred on reefs located north of the dredge channel. For grain sizes smaller than 5-50 $\mu$m, sediments mostly settled on inshore reefs while sedimentation mostly took place on offshore reefs for grain size larger than 50$\mu$m. For increasing grain sizes, the spatial distribution of sedimentation shifted southward, with a decrease of sedimentation near N. Sunny Isles and an increase of the concentration of settled sediments near the sites of HBSC1-CP and Virginia Key. The normalized cumulated deposition was around 0.01 particles/m$^2$ at site HBSC1-CP for all grain sizes while it increased from $2.5\times 10^{-7}$ to $1.6\times 10^{-3}$ with grain size at Virginia Kay. For all grain sizes, no sedimentation occurred near Rainbow Reef and Emerald Reef. However, sediments settled at similar latitudes west to these sites for grain sizes below 100 $\mu$m.

The evaluation of the impact of dredging at the monitoring sites indicated that dredging did not significantly impact the stations located south to the dredge channel, except at site HBSC1-CP (Fig. \ref{fig:onset_bar}). In contrast, up to 40\% of the released sediment particles reached the northern site of N. Sunny Isles. The impact of dredging at this site decreased with sediment size and the fraction of sediment particles reaching the N. Sunny Isles dropped below 2\% for grain sizes larger than 200 $\mu$m. For grain sizes below however, 50 $\mu$m, N. Sunny Isles was heavily impact by TX dredging, with 188,485 particles reaching the site. This is 50\% larger than the cumulated number of particles produced by other dredging operations reaching N. Sunny Isles for the same grain size. Such large impact can be explained by the fact that TX was one of the most frequent type of dredging during the simulations, with 268 simulated operations (Fig. \ref{fig:onset_bar}E). However, SB events occurred at an equivalent frequency but caused 4 times fewer sediment particles to reach N. Sunny Isles. For grain sizes below 100 $\mu$m, between 10\% and 30\% of the sediments released by non conventional dredging reached N. Sunny Isles. This fraction dropped below  1\% for grain sizes larger than 200$\mu$m. The fraction of sediments produced by non conventional dredging reaching site HBSC1-CP remained between 1\% and 5\% for all grain sizes. For grain sizes below 100 $\mu$m, D55 and NonConv were the strongest sources of sediments reaching HBSC1-CP. Above 100 $\mu$m, the strongest source of sediments to the site became TI. The fraction of sediments produced by Spider Barges (SB) reaching HBSC1-CP remained negligible for all grain sizes. No simulated sediment particles reached the sites of Rainbow Reef and Emerald Reef in all simulations. However, some particles reached Virginia Key for grain sizes larger than 100 $\mu$m. These sediments particles almost solely originated from Spider Barges. Nonetheless, the impact of the dredging on Virginia Key remained limited with less than 1$\%$ of the released particles reaching the site.

\begin{figure}
    \centering
    \includegraphics[width=.85\textwidth]{figures/aggregated_with_absolute.png}
    \includegraphics[width=.85\textwidth]{figures/timeline.png}
    \caption{Fraction and absolute number of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N. Sunnny Isles grain sized of 5-50 $\mu$m (\textbf{A},\textbf{C}) and 50-100 $\mu$m (\textbf{B},\textbf{D}). \textbf{E}: Temporal distribution of the simulated dredging operations. With the exception of HBSC1-CP, sites located south to the dredged channel were barely impacted by the dredging. However, a significant fraction of the released sediment particles reached the northern site of N. Sunny Isles, with 188,485 sediment particles produced by TX for grain sizes in the range 5-50 $\mu$m. The number of simulated dredging operations is given between brackets for each dredging type}
    \label{fig:onset_bar}
\end{figure}

Applying the same analysis to the discharge pipe of Miami Central District Municipal WWTP suggests that only wastewater released from a leak in the direct vicinity of the Virginia Key were likely to impact the site. Fig. \ref{fig:onset_pipe} shows the fraction of released wastewater particles that reached the site of Virginia key for every 100 m segments of the discharge pipe. The fraction of particles that drifted within 500 of Virginia Key was: 100\% for particles released within 500 m of the sites; 0-1\% for particles released between 500 and 1.5 km away from the site; and 0\% for all other sources of wastewater particles. As HBSC1-CP was also located near the discharge pipe, a similar analysis was conducted for this monitoring site and gave similar results.

\begin{figure}
    \centering
    \includegraphics[width=.7\textwidth]{figures/pipeline.png}
    \caption{Fraction of wastewater particles reaching the site of Virginia Key for potential leak positions located every 100 m along the discharge pipe. Only particles released in the direct vicinity of Virginia Key reached the monitoring site}
    \label{fig:onset_pipe}
\end{figure}

As signs of disease were observed at site HBSC1-CP in May 2014, before the first observations of SCTLD in Virginia Key in September 2014, we assessed the presence of shortest paths from HBSC1-CP to Virginia Key in the modeled monthly disease connectivity networks between May and September 2014 (Fig. \ref{fig:onset_path}). We found connectivity pathways connecting the two sites during all months of May-September 2014, except July 2014. This suggests that there was a possibility of disease propagation from HBSC1-CP to Virginia Key during most of these 5 months. However, we found no direct pathway connecting the two sites. Southern intermediary reefs were systematically needed as stepping stones for the propagation of the disease. This suggests that several months might have been required for disease agents to reach Virginia Key from HBSC1-CP. Moreover, shortest paths in June, August and September 2014 had many sub-reef stepping stones in common. These similar connectivity patterns indicate favorable conditions for disease propagation over several months from HBSC1-CP to Virginia Key during this period.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/fig_paths.png}
    \caption{Shortest path from HBSC1-CP to Virginia Key in the monthly disease connectivity networks between May and September 2014. July 2014 was the only month without modeled connectivity between the two sites between May and September 2014. Southern intermediary reefs were needed as stepping stones for the propagation of the disease from HBSC1-CP to Virginia Key.}
    \label{fig:onset_path}
\end{figure}

% === DISCUSSION === %
\section{Discussion}

% SUMMARY
The quasi-3D sediment model forced by currents from the high-resolution coastal ocean model SLIM reproduced sediment dynamics consistent with plume observations derived from satellite imagery. It suggested that sediment particles produced by dredging were mostly transported northward under the influence of the Florida Current. No sediment particles reached the southern sites of Rainbow Reef and Emerald Reef while dredging had limited impact on Virginia Key, identified as the site where the outbreak of SCTLD was initiated in September 2014. Furthermore, our results indicate that it is unlikely that the site was affected by a leaking discharge pipe of Miami Central District Municipal WWTP, as suggested by \cite{gintert2019regional}. However, signs of disease were observed prior to September 2014 at the sites of N. Sunny Isles and HBSC1-CP, where the impact of dredging was greater. Up to 30\% of silt particles produced by non-conventional dredging reached coral reefs of N. Sunny Isles while 5\% of them reached site HBSC1-CP, where significant sedimentation occurred. Furthermore, analysis of monthly disease connectivity networks indicate disease transmission from HBSC1-CP to the site of Virginia Key was possible through waterborne transmission.

% LARGE IMPACT OF DREDGING
This study confirms previous reports about the widespread impact of the expansion of PoM. As in \cite{barnes2015sediment}, our modeled predicted plumes over a total area about 230 km$2$. Furthermore, the model results reproduced the presence/absence within 15 km of the dredged channel derived \cite{cunning2019extensive} with an accuracy of 82.7\%. Such good agreement was only obtained for grain sizes in the range 5-50 $\mu$m, suggesting that fine silts were the main drivers of the turbidity generated by dredging, which is consistent with previous studies \citep{fourney2017additive}. Despite this good agreement, the distribution of plumes predicted by our model tended to be shifted northward compared to these previous studies. These discrepancies might be explained by the fact that our approaches is solely based on sediment concentration. However, turbidity depends on many local factors such as the water content phytoplankton organic matters \citep{gray2000comparability,thackston2000improved}. As reported by \cite{cunning2019extensive}, the plumes mostly occurred in areas of high sedimentation. As such areas were mostly located over coral reefs, our results suggests that the expansion of PoM might have harmed coral populations over an extensive through increased turbidity and sedimentation. This sedimentation mostly took place on inshore reefs for grain sizes smaller than 50$\mu$ m and on offshore reefs for larger grain sizes. Sediment settlement might be significantly harmful to offshore coral reefs populations as they are usually less accustomed to sedimentation than their inshore  counterparts \citep{wolanski2005fine}.

% LIMITATION + SHOW THAT N. SUNNY ISLE AND HBSC1-CP WERE SIGNIFICANTLY IMPACTED
A limitation of our study is that the amount released sediments might be overestimated, as conventional and non-conventional dredging are treated in the same way in the model. For all dredging types, sediment particles were released at the same rate in the model. Although conventional dredging was reported to release fine-grained sediments in the water column through dewatering and overflow from barges \citep{jones2016assessing}, the quantity of dredged material lost in the water was limited by the use of pumping mechanisms. By contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rocks to be released in the water column. The numbers of particles reaching the monitoring sites shown in Fig. \ref{fig:onset_bar}C,D were therefore likely overestimated for all sources but non conventional dredging. However, the fractions given in Fig. \ref{fig:onset_bar}A,B remain valid as they are relative to the total number of sediments released in the water columns. Despite these overestimations, the coral reefs of N. Sunny Isles appeared to be highly sensitive to fine-grained sediments produced by dredging (Fig. \ref{fig:onset_bar}). Furthermore, the non-conventional rock-chopping that took place between December 2013 and May 2014, was a non-negligible source of sediments to N. Sunny Isles, as up to 30\% of the released particles reached the site. As sediments have the potential to act as a SCTLD vector \citep{rosales2020rhodobacterales, studivan2022reef}, the dredging might therefore have contributed to the observed onset of the disease on the reefs of Sunny Isles in June 2014. Moreover, up to 5\% of the fined-grained sediments produced by non-conventional dredging reached the site HBSC1-CP. Although this represents a more limited quantity of sediments, HBSC1 was within 2 km from the sites where non-conventional dredging took place, in a region where the model predicted an important sedimentation. Therefore, although fewer sediments reached HBSC1-CP, they had a higher probability to settle and to be in direct contact with corals. In addition to smothering corals and diverting their energy through sediment removal \citep{erftemeijer2012environmental}, these settled sediments are more likely to transmit the disease to the reefs.

% POSSIBILITY OF DISEASE PROPAGATION THROUGH WATERBORNE TRANSMISSION
In all simulations, the expansion of PoM had limited direct impact on Virginia Key, identified as the site where the outbreak initiated in September 2014 \cite{precht2016unprecedented}. For all grain sizes, the fraction of sediment particles reaching Virginia Key remained below 1\%. It is therefore unlikely that sediments produced by the deepening of the channel brought the disease to the site. Furthermore, our simulation of leaking wastewater along the discharge pipe of Miami Central District Municipal WWTP suggests that currents tended to flush wastewater particles away from Virginia Key. It was therefore unlikely that wastewater affected the site, except if the leak was located within hundreds of meters from the site. However, signs of disease were observed prior to September 2014 at the sites of N. Sunny Isles and HBSC1-CP, both of which were impacted by the dredging. Several studies showed evidence of waterborne transmission of SCTLD \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring,meiling2021variable}. Disease agents might therefore have been transported by currents from one of the diseased sites to Virginia Key. As site HBSC-CP was closest to Virginia Key, we built and analyzed monthly disease connectivity networks and found connectivity pathways from HBSC1 to Virginia Key in May, June, August and September 2014. This implied that the propagation of the disease from HBSC1-CP to Virginia Key through hydrodynamics-driven transport of disease agents was possible during these four months. However, these pathways used reefs located further southward as stepping stones. Therefore, the propagation of outbreak to reach Virginia Key starting from HBSC1-CP required these intermediary reefs to get first infected by disease agents released by disease colonies of HBSC1-CP. Once diseased, these colonies would then send diseased agents to affect the next colonies of the connectivity pathway until the outbreak reached Virginia Key. Assuming an averaged transmission time of the order of 5-10 days \citep{dobbelaere2020coupled}, disease propagation from HBSC1-CP to Virginia Key might have required several months. This is consistent with the 5 month period separating reports of diseased at the two sites. Moreover, the fact that June, August and September exhibited similar connectivity pathways suggests that disease propagation could indeed have occurred over several months without being interrupted by changing hydrodynamic conditions.

\section{Conclusion}

In the present study, we evaluated the impact of the dredging operations that took place the impact of the deepening of the Port of Miami shipping channel on the neighboring coral reefs. This evaluation was performed using a quasi-3D sediment modeled forced by currents from the high-resolution coastal ocean model SLIM. As the first signs of the SCTLD outbreak were reported during these expansion works near Virginia Key and knowing that sediments may serve as SCTLD vector, we aimed at answering two main questions. First, we wanted to evaluate the fraction of sediments produced by the deepening that reaching the site of Virginia Key. Second, we evaluated the possibility of waterborne transmission of SCTLD to Virginia Key from other reefs impacted by the dredging. Doing so, a special attention was accorded to the sediments produced by non conventional rock chopping during which pumping mechanisms were turned off, causing all chopped rocks to be directly released in the water column.

Our results suggest that the dredging works had close to no impact on the coral reefs of Virginia Key, indicating that disease transmission trough the produced sediments was extremely unlikely. However, our results show the sediments released by the expansion works had a non-negligible impact on the sites of N. Sunny Isles and HBSC1-CP, where signs of disease were observed before the outbreak was first reported at Virginia Key, indicating that the sediments might have played a role in the onset of the diseased at these two sites. Furthermore, using a previously developed bio-physical model that successfully reproduced the observed spread of SCTLD, we show that there was a possibility of waterborne disease propagation from HBSC1-CP to Virginia Key.

This study suggests that the expansion of the Port of Miami might have played a role in the onset of the outbreak of SCTLD in sites where the disease was reported during the first half of 2014. Furthermore, it shows that the disease could then have spread from these sites to Virginia Key, from where SCTLD was then reported to propagate both north and south to the neighboring reefs.

\bibliographystyle{elsarticle-harv} 
\bibliography{./biblio.bib}

\end{document}
\endinput

### 2022-02-07 onset/main.tex
modification fig 1
commentaires du patron - par I

@@
 
-    \title{Early transmission dynamics of SCTLD and its connection with the dredging works of Port of Miami}%Impacts of Hurricane Irma (2017) on ocean transport processes}
+    \title{Early transmission dynamics of SCTLD and its connection with the PortMiami Deep Dredge Project}%Impacts of Hurricane Irma (2017) on ocean transport processes}
 
@@
     \ead{thomas.dobbelaere@uclouvain.be}
+    \author[]{other co-authors TBD}
     %\author[rsmas]{Milan Curcic}
@@
 
-%    \affiliation[eli]{
-%        organization={Eath and Life Institute (ELI), UCLouvain},
-%        % addressline={}, 
-%        city={Louvain-la-Neuve},
-%        % postcode={1348}, 
-%        % state={},
-%        country={Belgium}
-%    }
+   \affiliation[eli]{
+       organization={Eath and Life Institute (ELI), UCLouvain},
+       % addressline={}, 
+       city={Louvain-la-Neuve},
+       % postcode={1348}, 
+       % state={},
+       country={Belgium}
+   }
 %    \affiliation[rsmas]{
@@
 Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease.
-The goal of this study is therefore to evaluate whether sediments produced by the dredging might have been transported to reefs where disease was reported and to assess whether the timing of these inputs of sediment was consistent with the observed spread of the disease.
-We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment modeled forced by currents from the high-resolution coastal ocean model SLIM.
-Our result suggested that the dredging had close to no impact on the coral reefs of Virginia Key.
-However, a significant fraction of the produced sediments reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites.
-Furthermore, using a previously developed bio-physical model indicated that waterborne transmission from one of these sites to Virginia Key was possible before September 2014.
-This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst outbreak on record in the Caribbean.
+The goal of this study is therefore to evaluate whether sediments produced or resuspended during dredging operations might have been transported to reefs where disease was reported and to assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
+We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM.
+Our result suggested that the dredging had close to no direct impact on the coral reefs of Virginia Key.
+However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites.
+Furthermore, using a biophysical disease agent transport model suggests th   at waterborne transmission from one of these sites to Virginia Key was possible before September 2014.
+This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.
     \end{abstract}
@@
     \begin{keyword}
-        Stony coral tissue loss disease \sep Sediments \sep Port of Miami \sep Coastal modeling \sep Sediment plumes 
+        Stony coral tissue loss disease \sep Sediments \sep PortMiami \sep Coastal modeling \sep Sediment plumes 
     %% keywords here, in the form: keyword \sep keyword
@@
 Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region \citep{richardson1998coral, sutherland2004disease, aronson2001white, harvell2007coral, brandt2009dynamics}.
-One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is stony coral tissue loss disease (SCTLD) \citep{noaa2018}.
-First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entirety of FCR \citep{muller2020spatial,dobbelaere2022} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
+One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}.
+First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
 Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}.
-Furthermore, recent studies showed evidence that sediments act as vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
+Furthermore, recent studies showed evidence that sediments can act as  a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
 
@@
 Some of the first signs of SCTLD were reported in September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}.
-In this study, a relationship was found between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
-It was therefore hypothesized that the epidemic started near Virginia Key and then radiated to the neighboring reef, both north and south.
-However, earlier signs of diseases were reported north of Virginia Key in June 2014, at the monitoring site of N.
+They derived a relationship between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
+It was therefore hypothesized that the epidemic started near Virginia Key and then spread to the neighboring reef, both north and south.
+However, earlier signs of diseases were already reported north of Virginia Key in June 2014, at the monitoring site of N.
 Sunny Isles \citep{precht2016unprecedented}.
@@
 The dredging was monitored twice-weekly at 26 permanent monitoring stations established within the Miami-Dade County, making it one of the most complete datasets related to a dredging project \citep{gintert2019regional}.
-Interestingly, signs of disease were also reported at one these monitoring stations in May 2014.
+Interestingly, disease signs were also reported at one of these monitoring stations in May 2014.
+
 While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located 4.7 nautical miles offshore.
 However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
-The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report}.
+The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report} \comment{[find better ref ?]}.
 Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.
@@
 These effects are stronger with fine-grained sediments, as they cause a stronger light reduction \citep{fourney2017additive}.
-Additionally, fined-grained sediments such as silts have high nutrient contents, which can lead to an increased microbial activity, eventually causing anoxic conditions in the immediate vicinity of corals \citep{weber2012mechanisms}.
+Additionally, fine-grained sediments such as silts have high nutrient contents, which can lead to an increased microbial activity, eventually causing anoxic conditions in the immediate vicinity of corals \citep{weber2012mechanisms}.
 As they release finer sediments over significantly longer periods than natural events such as hurricanes, dredging activities can thus be more harmful to corals and reef habitat compared to other types of sedimentation \citep{cunning2019extensive}.
@@
 Nonetheless, \cite{gintert2019regional} argued that the reported coral mortality during the dredging project was dominated by the regional outbreak of SCTLD.
-Further, the study suggested that the onset of the disease might have been linked to a leaking discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant (WWTP) located off Virginia Key.
+Further, they suggested that the onset of the disease might have been linked to a leaking discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant (WWTP) located off Virginia Key.
 However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease was transported to the monitoring site of Virginia Key on sediments released by the dredging.
 This possibility can be evaluated using a hydrodynamic model simulating the transport of sediments produced during the dredging project.
-As coastal reef ecosystems are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures, such a model would require high spatial resolution to accurately represent the transport of sediments at reef-scale.
-In this context, unstructured-mesh models are best suited, as they can easily adapt to the topography \citep{fringer2019future} and can capture small-scale circulation features around reefs and islands \citep{lambrechts2008multi}.
+As coastal reef ecosystems are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures, such a model would require high spatial resolution to accurately represent the transport of sediments at the reef-scale.
+In this context, unstructured-mesh models are particularly well suited, as they can easily adapt to the topography \citep{fringer2019future} and can capture small-scale circulation features around reefs and islands \citep{lambrechts2008multi}.
 
-The goal of this study is therefore to simulate the trajectories of the sediments released during the entirety of the deepening of the PoM shipping channel using a high-resolution hydrodynamic model coupled with a sediment transport model.
-Specifically, we will attempt to answer the following questions: (1) Which reefs were impacted by the dredging ?
+The goal of this study is therefore to simulate the trajectories of the sediments released during the entirety of the PortMiami Deep Dredge Project (PMDDP) using a high-resolution hydrodynamic model coupled with a sediment transport model.
+Specifically, our goal is to answer the following questions: (1) Which reefs were impacted by PMDDP ?
 (2) Is the impact on these reefs consistent with the observed timing of the onset of SCTLD ?
@@
 The date where SCTLD was first observed at these sites is given between brackets.
-The Ocean Dredge Material Disposal Site (ODMDS) is shown in red.
+The Ocean Dredge Material Disposal Site (ODMDS) is shown in red and the discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant in green.
 \textbf{B}: Close up view of the dredged channel.
@@
 \textbf{C:} Snapshot of the modeled currents in the vicinity of the dredged channel.
-Small-scale flow features such as the acceleration of currents between reefs and between islands are well captured by the model.}
+Small-scale flow features such as the acceleration of currents between reefs and islands are well reproduced by the model.
+\todo{refaire la figure}}
     \label{fig:onset_mesh}
@@
 
-The hydrodynamics of the entirety of FCR was modeled using the high resolution unstructured-mesh model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022}.
+The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022}.
 SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features.
@@
 The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the first observation of SCTLD by \cite{precht2016unprecedented}.
-Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the accelerations between reefs and between islands.
+Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the flow acceleration between reefs and islands.
 
-The transport of sediments released from the channel was then modeled using a Lagrangian particle tracking model, forced by SLIM velocity field.
+The transport of sediments released from the channel was then modeled using a Lagrangian particle tracking model, forced by the simulated currents.
 The sediment model is inspired by the Particle Transport Model (PTM), developed by the US Army Corps of Engineers \citep{macdonald2006ptm}.
@@
 The vertical is mostly driven by gravity, with heavier particles sinking faster.
-Once they have settled, particles can be resuspended when shear stress exceed the critical Schields parameter, as parameterized by \cite{soulsby1997threshold}.
-The horizontal motion of the suspended particles is derived from the 2D model velocity by assuming a vertical log profile, following a quasi-3D approach.
+Once they have settled, particles can be resuspended when shear stress exceeds   the critical Schields parameter, as parameterized by \cite{soulsby1997threshold}.
+The horizontal motion of the suspended particles is derived from the 2D model velocity by assuming a vertical log profile, hence yielding a quasi-3D approach.
 When sediment particles enter the near-bed zone, their horizontal velocity is greatly reduced and sediments are transported with the bedload.
-The 2D model of this particle tracker can also be used to model the transport of neutrally buoyant material driven by mean current within the water column \citep{dobbelaere2020coupled}.
-This model was used to model the impact of wastewater leaking from the discharge pipe mentioned by \cite{gintert2019regional}.
-Without clear information about the location of the leak, particles were continuously released every 100 m along a straight line between Miami Central District Municipal WWTP, located on Virginia Key and the ocean discharge outfall located at 25$^\circ$44'31''N 80$^\circ$05'10''W \citep{koopman2006ocean}.
 
-As sediment dispersion is dependent on the grain size, we modeled the dispersal of five classes of sediments to represent to impact of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m.
+As sediment dispersion is dependent on the grain size, we modeled the dispersal of five sediment classes to represent to impact of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m.
 We performed a  different simulation for each class, with the grain size randomly drawn from a uniform distribution over the corresponding size range.
@@
 The modeled occurrence of plume was then compared against daily data of plume detection.
-This data set was derived from satellite imagery by \cite{cunning2019extensive}\footnote{data sets available at \url{https://github.com/jrcunning/pom-dredge/}} following the methods of \cite{barnes2015sediment} at sites located within 15 km of the dredged channel.
+This dataset was derived from satellite imagery by \cite{cunning2019extensive}\footnote{datasets available at \url{https://github.com/jrcunning/pom-dredge/}} following the methods of \cite{barnes2015sediment} at sites located within 15 km of the dredged channel.
 As in these two previous studies, we computed  the simulated plume frequency by dividing the number of days during which plumes occurred by the total number of simulated days for all grid cells.
@@
 
-The potential impact of dredging on the onset of the outbreak was more specifically assessed at five monitoring sites were signs of disease were reported.
+The potential impact of dredging on the onset of the SLD outbreak was more specifically assessed at five monitoring sites were signs of disease were reported.
 Four sites were reefs where disease was reported in 2014 by \cite{precht2016unprecedented}.
-The fifth site was monitoring station HBSC1-CP of the monitoring of the expansion of PoM, where signs of disease were reported in May 2014 (Fig.
-\ref{fig:onset_mesh}A) \comment{[Photos in appendix ?
-]}.
+The fifth site was monitoring station HBSC1-CP of the monitoring of the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
+\ref{fig:onset_mesh}A, see pictures in Appendix).
 The impact of dredging was assessed by counting the number of sediment particles originating from each dredging that were transported within 500 m of all five site.
-This number was then divided by the total number of sediment particles released by each type of dredge.
-Larger values of this indicator would suggest a greater impact of a given type of dredging at a given monitoring sites.
-This same indicator was used to evaluate the impact of particles released along the discharge pipe at the site of Virginia Key.
+This distance was chosen to match the scale of the 500 m $\times$ 500 m reef polygons used in the model \citep{dobbelaere2020coupled}.
+The obtained number of particles was then divided by the total number of sediment particles released by each type of dredge.
+Larger values of this indicator would suggest a greater impact of a given type of dredging operation at a given monitoring sites.
 
@@
 To evaluate this possibility, we computed monthly disease connectivity matrices following the methodology of \cite{dobbelaere2020coupled} during our simulated period.
-These connectivity matrices can be interpreted as large graphs whose vertices are sub-areas of reefs and whose edges represent disease connectivity pathways.
-Evaluating the possibility of disease propagation from any reef $A$ to any reef $B$ is therefore equivalent to evaluating the existence of paths, \ie sequences of connected vertices in the network starting from $A$ and reaching $B$.
+These connectivity matrices can be interpreted as large graphs whose vertices are 500 m $\times$ 500 m reef areas and whose edges represent disease connectivity pathways.
+Evaluating the possibility of disease propagation from any sub-reef $A$ to any sub-reef $B$ is therefore equivalent to evaluating the existence of paths, \ie sequences of connected vertices in the network starting from $A$ and reaching $B$.
 As computing all possible paths is not computationally tractable, we limited ourselves to the computation of shortest paths from any given reefs to the Virginia Key monitoring site.
@@
 The probability of a given path was then defined as the mean connection probability of the edges composing this path.
+
+Finally, the 2D mode of the particle tracking model was used to assess the impact of wastewater leaking from the discharge pipe mentioned by \cite{gintert2019regional}.
+This model simulates the transport of neutrally buoyant material driven by mean current within the water column \citep{dobbelaere2020coupled}.
+Without clear information about the location of the leak, particles were continuously released every 100 m along a straight line between Miami Central District Municipal WWTP, located on Virginia Key and the ocean discharge outfall located at 25$^\circ$44'31''N 80$^\circ$05'10''W \citep{koopman2006ocean}.
+The impact of wastewater released from the different potential leak locations was then evaluated by computing the fraction of released wastewater particles that were transported within 500 m of the site of Virginia Key.
 
@@
 For larger grain sizes, particles settled in the direct vicinity of the channel.
-For these heavier particles, suspended sediments were only observed offshore, closer to the Florida Current, where current velocity was sufficiently strong to prevent deposition.
+For these heavier particles, suspended sediments were only observed offshore, closer to the Florida Current, where the current velocity was sufficiently strong to prevent deposition.
 This suggests that the observed turbidity was mostly due to fine silts.
@@
 Sunny Isles.
-Plume frequencies was more limited over offshore but reached of up to 50\% over inshore reefs.
-These inner reefs correspond to regions of important sedimentation (Fig.
+Plume frequencies was lower ($\sim 5$\%) over offshore reefs but reached of up to 50\% over alongshore reefs.
+These alongshore reefs correspond to regions of important sedimentation (Fig.
 \ref{fig:onset_depo}B), which is consistent with the correlation between plumes and sedimentation highlighted in \cite{cunning2019extensive}.
@@
 
-Deposition results are shown for grain sizes corresponding to silts, which are more likely to carry organic matter and therefore more likely to carry SCTLD agents (Fig.
+Deposition results are shown for grain sizes corresponding to silts, which are more likely to carry organic matter and therefore more likely to carry SCTLD agents \citep{erftemeijer2012environmental}(Fig.
 \ref{fig:onset_depo}B,C).
@@
 For grain sizes smaller than 5-50 $\mu$m, sediments mostly settled on inshore reefs while sedimentation mostly took place on offshore reefs for grain size larger than 50$\mu$m.
-For increasing grain sizes, the spatial distribution of sedimentation shifted southward, with a decrease of sedimentation near N.
-Sunny Isles and an increase of the concentration of settled sediments near the sites of HBSC1-CP and Virginia Key.
+When increasing grain sizes, the spatial distribution of sedimentation shifted southward, with a decrease of sedimentation near N.
+Sunny Isles and an increase of the concentration of settled sediments near HBSC1-CP and Virginia Key.
 The normalized cumulated deposition was around 0.01 particles/m$^2$ at site HBSC1-CP for all grain sizes while it increased from $2.5\times 10^{-7}$ to $1.6\times 10^{-3}$ with grain size at Virginia Kay.
@@
 
-The evaluation of the impact of dredging at the monitoring sites indicated that dredging did not significantly impact the stations located south to the dredge channel, except at site HBSC1-CP (Fig.
+The evaluation of the impact of dredging at the monitoring sites indicated that dredging did not significantly impact the stations located south of the dredge channel, except at site HBSC1-CP (Fig.
 \ref{fig:onset_bar}).
@@
 Sunny Isles dropped below 2\% for grain sizes larger than 200 $\mu$m.
-For grain sizes below however, 50 $\mu$m, N.
-Sunny Isles was heavily impact by TX dredging, with 188,485 particles reaching the site.
-This is 50\% larger than the cumulated number of particles produced by other dredging operations reaching N.
-Sunny Isles for the same grain size.
+However, for grain sizes finer than 50 $\mu$m, N.
+Sunny Isles was heavily impact by TX dredging.
+Indeed, the number of particles produced by TX reaching N.Sunny Isles was 50\% larger than the cumulated number of particles produced by other dredging operations reaching the monitoring site for the same grain size.
 Such large impact can be explained by the fact that TX was one of the most frequent type of dredging during the simulations, with 268 simulated operations (Fig.
 \ref{fig:onset_bar}E).
-However, SB events occurred at an equivalent frequency but caused 4 times fewer sediment particles to reach N.
+However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles to reach N.
 Sunny Isles.
@@
 The fraction of sediments produced by non conventional dredging reaching site HBSC1-CP remained between 1\% and 5\% for all grain sizes.
-For grain sizes below 100 $\mu$m, D55 and NonConv were the strongest sources of sediments reaching HBSC1-CP.
-Above 100 $\mu$m, the strongest source of sediments to the site became TI.
-The fraction of sediments produced by Spider Barges (SB) reaching HBSC1-CP remained negligible for all grain sizes.
-No simulated sediment particles reached the sites of Rainbow Reef and Emerald Reef in all simulations.
+For grain sizes below 100 $\mu$m, D55 and NonConv were the main sources of sediments reaching HBSC1-CP.
+Above 100 $\mu$m, the main source of sediments to the site became TI.
+The fraction of sediments produced by SB reaching HBSC1-CP remained negligible for all grain sizes.
+No sediment particles reached the sites of Rainbow Reef and Emerald Reef for all grain sizes.
 However, some particles reached Virginia Key for grain sizes larger than 100 $\mu$m.
-These sediments particles almost solely originated from Spider Barges.
+These sediments particles almost entirely originated from SB.
 Nonetheless, the impact of the dredging on Virginia Key remained limited with less than 1$\%$ of the released particles reaching the site.
@@
 However, a significant fraction of the released sediment particles reached the northern site of N.
-Sunny Isles, with 188,485 sediment particles produced by TX for grain sizes in the range 5-50 $\mu$m.
+Sunny Isles.
 The number of simulated dredging operations is given between brackets for each dredging type}
@@
 
-Applying the same analysis to the discharge pipe of Miami Central District Municipal WWTP suggests that only wastewater released from a leak in the direct vicinity of the Virginia Key were likely to impact the site.
+Applying the same analysis to the discharge pipe of Miami Central District Municipal WWTP suggests that only wastewater released from a leak in the direct vicinity of the Virginia Key could have impacted the site.
 Fig.
 \ref{fig:onset_pipe} shows the fraction of released wastewater particles that reached the site of Virginia key for every 100 m segments of the discharge pipe.
-The fraction of particles that drifted within 500 of Virginia Key was: 100\% for particles released within 500 m of the sites; 0-1\% for particles released between 500 and 1.5 km away from the site; and 0\% for all other sources of wastewater particles.
+The fraction of particles that drifted within 500 m of Virginia Key was: 100\% for particles released within 500 m of the site; 0-1\% for particles released between 500 m and 1.5 km away from the site; and 0\% for all other sources of wastewater particles.
 As HBSC1-CP was also located near the discharge pipe, a similar analysis was conducted for this monitoring site and gave similar results.
@@
     \caption{Fraction of wastewater particles reaching the site of Virginia Key for potential leak positions located every 100 m along the discharge pipe.
-Only particles released in the direct vicinity of Virginia Key reached the monitoring site}
+Only particles released in the direct vicinity of Virginia Key could reach the monitoring site.}
     \label{fig:onset_pipe}
@@
 \ref{fig:onset_path}).
-We found connectivity pathways connecting the two sites during all months of May-September 2014, except July 2014.
+We found connectivity pathways connecting the two sites during all months of May-September 2014, with the exception of July.
 This suggests that there was a possibility of disease propagation from HBSC1-CP to Virginia Key during most of these 5 months.
@@
 % SUMMARY
-The quasi-3D sediment model forced by currents from the high-resolution coastal ocean model SLIM reproduced sediment dynamics consistent with plume observations derived from satellite imagery.
-It suggested that sediment particles produced by dredging were mostly transported northward under the influence of the Florida Current.
-No sediment particles reached the southern sites of Rainbow Reef and Emerald Reef while dredging had limited impact on Virginia Key, identified as the site where the outbreak of SCTLD was initiated in September 2014.
+The quasi-3D sediment model forced by currents from the high-resolution coastal ocean model SLIM reproduced a sediment dynamics consistent with plume observations derived from satellite imagery.
+It suggests that sediment particles produced by dredging were mostly transported northward under the influence of the Florida Current.
+No sediment particles reached the southern sites of Rainbow Reef and Emerald Reef while dredging had limited direct impact on Virginia Key, identified as the site where the outbreak of SCTLD was initiated in September 2014.
 Furthermore, our results indicate that it is unlikely that the site was affected by a leaking discharge pipe of Miami Central District Municipal WWTP, as suggested by \cite{gintert2019regional}.
@@
 % LARGE IMPACT OF DREDGING
-This study confirms previous reports about the widespread impact of the expansion of PoM.
-As in \cite{barnes2015sediment}, our modeled predicted plumes over a total area about 230 km$2$.
-Furthermore, the model results reproduced the presence/absence within 15 km of the dredged channel derived \cite{cunning2019extensive} with an accuracy of 82.7\%.
+Our study confirms previous reports about the widespread impact of the expansion of PoM.
+As in \cite{barnes2015sediment}, the total area of plumes in our model was about 230 km$^2$.
+Furthermore, the model results reproduced the presence/absence within 15 km of the dredged channel derived by \cite{cunning2019extensive} with an accuracy of 82.7\%.
 Such good agreement was only obtained for grain sizes in the range 5-50 $\mu$m, suggesting that fine silts were the main drivers of the turbidity generated by dredging, which is consistent with previous studies \citep{fourney2017additive}.
-Despite this good agreement, the distribution of plumes predicted by our model tended to be shifted northward compared to these previous studies.
+Despite this good agreement, the distribution of plumes predicted by our model tends to be shifted northward compared to these previous studies.
 These discrepancies might be explained by the fact that our approaches is solely based on sediment concentration.
-However, turbidity depends on many local factors such as the water content phytoplankton organic matters \citep{gray2000comparability,thackston2000improved}.
+However, turbidity depends on many local factors such as the water content of phytoplankton and organic matters \citep{gray2000comparability,thackston2000improved}.
 As reported by \cite{cunning2019extensive}, the plumes mostly occurred in areas of high sedimentation.
-As such areas were mostly located over coral reefs, our results suggests that the expansion of PoM might have harmed coral populations over an extensive through increased turbidity and sedimentation.
-This sedimentation mostly took place on inshore reefs for grain sizes smaller than 50$\mu$ m and on offshore reefs for larger grain sizes.
+As such areas were mostly located over coral reefs, our results suggests that the expansion of PoM might have harmed coral populations over an extended area through increased turbidity and sedimentation.
+This sedimentation mostly took place on alongshore reefs for grain sizes smaller than 50$\mu$m and on offshore reefs for coarser grain sizes.
 Sediment settlement might be significantly harmful to offshore coral reefs populations as they are usually less accustomed to sedimentation than their inshore  counterparts \citep{wolanski2005fine}.
@@
 % LIMITATION + SHOW THAT N. SUNNY ISLE AND HBSC1-CP WERE SIGNIFICANTLY IMPACTED
-A limitation of our study is that the amount released sediments might be overestimated, as conventional and non-conventional dredging are treated in the same way in the model.
+A limitation of our study is that the amount of released sediments might be overestimated, as conventional and non-conventional dredging are treated in the same way in the model.
 For all dredging types, sediment particles were released at the same rate in the model.
 Although conventional dredging was reported to release fine-grained sediments in the water column through dewatering and overflow from barges \citep{jones2016assessing}, the quantity of dredged material lost in the water was limited by the use of pumping mechanisms.
-By contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rocks to be released in the water column.
-The numbers of particles reaching the monitoring sites shown in Fig.
-\ref{fig:onset_bar}C,D were therefore likely overestimated for all sources but non conventional dredging.
+By contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rock particles to be released in the water column.
+The numbers of particles reaching the monitoring sites were therefore likely overestimated for all sources but non conventional dredging.
 However, the fractions given in Fig.
@@
 Sunny Isles, as up to 30\% of the released particles reached the site.
-As sediments have the potential to act as a SCTLD vector \citep{rosales2020rhodobacterales, studivan2022reef}, the dredging might therefore have contributed to the observed onset of the disease on the reefs of Sunny Isles in June 2014.
+As sediments have the potential to act as a SCTLD vector \citep{rosales2020rhodobacterales, studivan2022reef}, the dredging might therefore have contributed to the observed onset of the disease on the reefs of N.
+Sunny Isles in June 2014.
 Moreover, up to 5\% of the fined-grained sediments produced by non-conventional dredging reached the site HBSC1-CP.
@@
 It is therefore unlikely that sediments produced by the deepening of the channel brought the disease to the site.
-Furthermore, our simulation of leaking wastewater along the discharge pipe of Miami Central District Municipal WWTP suggests that currents tended to flush wastewater particles away from Virginia Key.
+Furthermore, our simulation of leaking wastewater along the discharge pipe of Miami Central District Municipal WWTP suggests that currents were likely to flush wastewater away from Virginia Key.
 It was therefore unlikely that wastewater affected the site, except if the leak was located within hundreds of meters from the site.
@@
 Disease agents might therefore have been transported by currents from one of the diseased sites to Virginia Key.
-As site HBSC-CP was closest to Virginia Key, we built and analyzed monthly disease connectivity networks and found connectivity pathways from HBSC1 to Virginia Key in May, June, August and September 2014.
+As site HBSC-CP was the closest to Virginia Key, we built and analyzed monthly disease connectivity networks and found connectivity pathways from HBSC1 to Virginia Key in May, June, August and September 2014.
 This implied that the propagation of the disease from HBSC1-CP to Virginia Key through hydrodynamics-driven transport of disease agents was possible during these four months.
@@
 In the present study, we evaluated the impact of the dredging operations that took place the impact of the deepening of the Port of Miami shipping channel on the neighboring coral reefs.
-This evaluation was performed using a quasi-3D sediment modeled forced by currents from the high-resolution coastal ocean model SLIM.
+This evaluation was performed using a quasi-3D sediment model forced by high-resolution currents from the coastal ocean model SLIM.
 As the first signs of the SCTLD outbreak were reported during these expansion works near Virginia Key and knowing that sediments may serve as SCTLD vector, we aimed at answering two main questions.
-First, we wanted to evaluate the fraction of sediments produced by the deepening that reaching the site of Virginia Key.
+First, we wanted to evaluate the fraction of sediments produced by the PortMiami Deep Dredge Project that reached the site of Virginia Key.
 Second, we evaluated the possibility of waterborne transmission of SCTLD to Virginia Key from other reefs impacted by the dredging.
-Doing so, a special attention was accorded to the sediments produced by non conventional rock chopping during which pumping mechanisms were turned off, causing all chopped rocks to be directly released in the water column.
+Doing so, we paid a special attention to the sediments produced by non conventional rock chopping during which pumping mechanisms were turned off, causing all chopped rock particles to be directly released in the water column.
 
-Our results suggest that the dredging works had close to no impact on the coral reefs of Virginia Key, indicating that disease transmission trough the produced sediments was extremely unlikely.
+Our results suggest that the dredging works had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission trough the produced sediments was extremely unlikely.
 However, our results show the sediments released by the expansion works had a non-negligible impact on the sites of N.
@@
 Furthermore, it shows that the disease could then have spread from these sites to Virginia Key, from where SCTLD was then reported to propagate both north and south to the neighboring reefs.
+Our work brings new insight on the role played by a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
 

### 2022-02-07 onset/main.tex
remarques du patron - part II

@@
 Such large impact can be explained by the fact that TX was one of the most frequent type of dredging during the simulations, with 268 simulated operations (Fig.
-\ref{fig:onset_bar}E).
+\ref{fig:onset_bar}C).
 However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles to reach N.
@@
     \centering
-    \includegraphics[width=.85\textwidth]{figures/aggregated_with_absolute.png}
-    \includegraphics[width=.85\textwidth]{figures/timeline.png}
-    \caption{Fraction and absolute number of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N.
-Sunnny Isles grain sized of 5-50 $\mu$m (\textbf{A},\textbf{C}) and 50-100 $\mu$m (\textbf{B},\textbf{D}).
-\textbf{E}: Temporal distribution of the simulated dredging operations.
+    \includegraphics[width=\textwidth]{figures/aggregated_new.png}
+    \includegraphics[width=\textwidth]{figures/timeline.png}
+    \caption{Fraction of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N.
+Sunnny Isles with grain sized of 5-50 $\mu$m (\textbf{A}) and 50-100 $\mu$m (\textbf{B}).
+\textbf{C}: Temporal distribution of the simulated dredging operations.
+The dates of the first reported disease signs at each monitoring site are shown in brown.
 With the exception of HBSC1-CP, sites located south to the dredged channel were barely impacted by the dredging.
@@
 Furthermore, it shows that the disease could then have spread from these sites to Virginia Key, from where SCTLD was then reported to propagate both north and south to the neighboring reefs.
-Our work brings new insight on the role played by a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
+Our work brings new insight on the impact of a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
 

### 2022-02-07 chapters/drto/drto.tex
last chapter added to manuscript

@@
 
-\item \textbf{T.
-Dobbelaere}, D.
-H.
-Holstein, E.
-M.
-Muller, L.
-J.
-Gramer, L.
-McEachron, S.
-D.
-Williams$\&$ E.
-Hanert, Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas (submitted to \textit{Frontiers in Marine Science})
+\item \textbf{Dobbelaere, T.}, Holstein, D.
+M., Muller, E.
+M., Gramer, L.
+J., McEachron L., Williams, S.
+D., $\&$ Hanert, E.
+Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas.
+accepted for publication in \textit{Frontiers in Marine Science}.\\ \href{https://doi.org/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.
+
 \end{list}

### 2022-02-07 chapters/irma/irma.tex
last chapter added to manuscript

@@
 }%
-
-\item \textbf{T. Dobbelaere}, M. Curcic, M. Le Hénaff, $\&$ E. Hanert, Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes (submitted to  \textit{Ocean Modelling})
+\item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., & Hanert, E.
+(2022).
+Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes.
+\textit{Ocean Modelling}, 101947.
+\href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.1016/j.ocemod.2022.101947}.
 \end{list}

### 2022-02-07 chapters/onset/article.tex
last chapter added to manuscript

--- 2022-02-07 onset/main.tex
@@
-%% 
-%% Copyright 2007-2020 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
-
-\documentclass[preprint,12pt,authoryear]{elsarticle}
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage[utf8]{inputenc}
-\usepackage{url,hyperref,lineno,microtype,subcaption}
-\usepackage{color,tensor,multirow,siunitx}
-\usepackage[onehalfspacing]{setspace}
-\usepackage{makecell}
-\renewcommand{\cellalign}{cl}
-\usepackage{caption}
-\usepackage{lipsum}
-\captionsetup[figure]{name=Fig., labelfont=bf}
-\captionsetup[table]{name=Table, labelfont=bf}
-
-\renewcommand{\cellalign}{cl}
-
-\newcommand{\ds}{\displaystyle}
-\newcommand{\nl}{\ \\ }
-\newcommand{\ud}{\textrm{ d}}
-\newcommand{\bs}{\bigskip}
-
-\newcommand{\bu}{\mathbf{u}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\be}{\mathbf{e}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bk}{\mathbf{k}}
-\newcommand{\bn}{\mathbf{n}}
-\newcommand{\bR}{\mathbf{R}}
-
-\definecolor{orange}{rgb}{1.0, 0.46, 0.09}
-\definecolor{red}{rgb}{1,0,0}
-\definecolor{blue}{rgb}{0,0,0.8}
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand{\emphc}[1]{\emph{\textcolor{red}{#1}}}
-\newcommand{\hycom}{\textsc{hycom} }
-\newcommand{\ie}{{\it i.e.}\ }
-\newcommand{\eg}{{\it e.g.}\ }
-\newcommand{\UV}{\mathbf{U}}
-\newcommand{\todo}[1]{\textcolor{red}{TO DO: #1}}
-\newcommand{\comment}[1]{\textcolor{orange}{#1}}
-%\newcommand{\modif}[1]{\textcolor{blue}{#1}}
-% \renewcommand{\familydefault}{\sfdefault}
-
-\journal{Marine Pollution Bulletin}
-
-\begin{document}
-
-\begin{frontmatter}
-
-    %% Title, authors and addresses
-
-    %% use the tnoteref command within \title for footnotes;
-    %% use the tnotetext command for theassociated footnote;
-    %% use the fnref command within \author or \affiliation for footnotes;
-    %% use the fntext command for theassociated footnote;
-    %% use the corref command within \author for corresponding author footnotes;
-    %% use the cortext command for theassociated footnote;
-    %% use the ead command for the email address,
-    %% and the form \ead[url] for the home page:
-    %% \title{Title\tnoteref{label1}}
-    %% \tnotetext[label1]{}
-    %% \author{Name\corref{cor1}\fnref{label2}}
-    %% \ead{email address}
-    %% \ead[url]{home page}
-    %% \fntext[label2]{}
-    %% \cortext[cor1]{}
-    %% \affiliation{organization={},
-    %%            addressline={}, 
-    %%            city={},
-    %%            postcode={}, 
-    %%            state={},
-    %%            country={}}
-    %% \fntext[label3]{}
-
-    \title{Early transmission dynamics of SCTLD and its connection with the PortMiami Deep Dredge Project}%Impacts of Hurricane Irma (2017) on ocean transport processes}
-
-    %% use optional labels to link authors explicitly to addresses:
-    %% \author[label1,label2]{}
-    %% \affiliation[label1]{organization={},
-    %%             addressline={},
-    %%             city={},
-    %%             postcode={},
-    %%             state={},
-    %%             country={}}
-    %%
-    %% \affiliation[label2]{organization={},
-    %%             addressline={},
-    %%             city={},
-    %%             postcode={},
-    %%             state={},
-    %%             country={}}
-
-    \author[eli]{Thomas Dobbelaere\corref{corr}}
-    \ead{thomas.dobbelaere@uclouvain.be}
-    \author[]{other co-authors TBD}
-    %\author[rsmas]{Milan Curcic}
-    %\author[cimas,aoml]{Matthieu Le H\'enaff}
-    \author[eli,immc]{Emmanuel Hanert}
-    \cortext[corr]{Corresponding author}
-
-   \affiliation[eli]{
-       organization={Eath and Life Institute (ELI), UCLouvain},
-       % addressline={}, 
-       city={Louvain-la-Neuve},
-       % postcode={1348}, 
-       % state={},
-       country={Belgium}
-   }
-%    \affiliation[rsmas]{
-%        organization={Rosenstiel School of Marine and Atmospheric Sciences (RSMAS), University of Miami},
-%        % addressline={}, 
-%        city={Coral Gables},
-%        % postcode={1348}, 
-%        state={Florida},
-%        country={USA}
-%    }
-%    \affiliation[cimas]{
-%        organization={Cooperative Institute for Marine and Atmospheric
-%        Studies (CIMAS), University of Miami},
-%        % addressline={}, 
-%        city={Miami},
-%        % postcode={1348}, 
-%        state={Florida},
-%        country={USA}
-%    }
-%    \affiliation[aoml]{
-%        organization={Atlantic Oceanographic and Meteorological Laboratory
-%        (AOML), NOAA},
-%        % addressline={}, 
-%        city={Miami},
-%        % postcode={1348}, 
-%        state={Florida},
-%        country={USA}
-%    }
-    \affiliation[immc]{
-        organization={Institute of Mechanics, Materials and Civil Engineering (IMMC), UCLouvain},
-        % addressline={}, 
-        city={Louvain-la-Neuve},
-        % postcode={1348}, 
-        % state={},
-        country={Belgium}
-    }
-
-    \begin{abstract}
-        For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD).
-The outbreak reportedly initiated near Virginia Key in September 2014, during the the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015.
-The disease then spread to the entirety of FCR, likely through waterborne transmission.
-Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease.
-The goal of this study is therefore to evaluate whether sediments produced or resuspended during dredging operations might have been transported to reefs where disease was reported and to assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
-We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM.
-Our result suggested that the dredging had close to no direct impact on the coral reefs of Virginia Key.
-However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites.
-Furthermore, using a biophysical disease agent transport model suggests th   at waterborne transmission from one of these sites to Virginia Key was possible before September 2014.
-This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.
-    \end{abstract}
-
-    %%Graphical abstract
-    % \begin{graphicalabstract}
-    % %\includegraphics{grabs}
-    % \end{graphicalabstract}
-
-    %Research highlights
-    % \begin{highlights}
-    %     \item The coupled SLIM+SWAN model correctly reproduced the hydrodynamics and waves during Irma.
-    %    \item Wave radiation stress increased currents by up to 1 m/s during Irma.
-	%    \item Wave radiation stress gradients were the largest on the shelf break and over reefs. 
-	%   \item Waves could deflect drifting particles by up to 20 km during the hurricane.
-	%  \item The Stokes drift had an impact on transport 4 times larger than the wave radiation stress.
-    % \end{highlights}
-
-    \begin{keyword}
-        Stony coral tissue loss disease \sep Sediments \sep PortMiami \sep Coastal modeling \sep Sediment plumes 
-    %% keywords here, in the form: keyword \sep keyword
-
-    %% PACS codes here, in the form: \PACS code \sep code
-
-    %% MSC codes here, in the form: \MSC code \sep code
-    %% or \MSC[2008] code \sep code (2000 is the default)
-
-    \end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
 \section{Introduction}
@@
 However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
-The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report} \comment{[find better ref ?]}.
+The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report}.
 Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.
@@
 	\centering
-    \includegraphics[width=\textwidth]{figures/fig_mesh_onset.png}
+	\includegraphics[width=\textwidth]{chapters/onset/figures/fig_mesh_onset.png}
 	\caption{\textbf{A}: Model mesh near the dredged channel.
@@
 \textbf{C:} Snapshot of the modeled currents in the vicinity of the dredged channel.
-Small-scale flow features such as the acceleration of currents between reefs and islands are well reproduced by the model.
-\todo{refaire la figure}}
+Small-scale flow features such as the acceleration of currents between reefs and islands are well reproduced by the model.}
 	\label{fig:onset_mesh}
@@
 	\centering
-    \includegraphics[width=\textwidth]{figures/deposition_plumes.png}
+	\includegraphics[width=\textwidth]{chapters/onset/figures/deposition_plumes.png}
 	\caption{\textbf{A}: Plume frequency for grain size in the range 5-50 $\mu$m.
@@
 	\centering
-    \includegraphics[width=\textwidth]{figures/aggregated_new.png}
-    \includegraphics[width=\textwidth]{figures/timeline.png}
+	\includegraphics[width=\textwidth]{chapters/onset/figures/aggregated_new.png}
+	\includegraphics[width=\textwidth]{chapters/onset/figures/timeline.png}
 	\caption{Fraction of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N.
@@
 	\centering
-    \includegraphics[width=.7\textwidth]{figures/pipeline.png}
+	\includegraphics[width=.7\textwidth]{chapters/onset/figures/pipeline.png}
 	\caption{Fraction of wastewater particles reaching the site of Virginia Key for potential leak positions located every 100 m along the discharge pipe.
@@
 	\centering
-    \includegraphics[width=\textwidth]{figures/fig_paths.png}
+	\includegraphics[width=\textwidth]{chapters/onset/figures/fig_paths.png}
 	\caption{Shortest path from HBSC1-CP to Virginia Key in the monthly disease connectivity networks between May and September 2014.
@@
 Our work brings new insight on the impact of a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
-
-\bibliographystyle{elsarticle-harv} 
-\bibliography{./biblio.bib}
-
-\end{document}
-\endinput

### 2022-02-07 chapters/onset/onset.tex
last chapter added to manuscript

--- 2022-02-07 onset/main.tex
@@
-%% 
-%% Copyright 2007-2020 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
+This chapter is based on the following article:
+\begin{list}{}{%
+\setlength{\topsep}{0pt}%
+\setlength{\leftmargin}{0.23in}%
+\setlength{\listparindent}{-0.23in}%
+\setlength{\itemindent}{-0.23in}%
+\setlength{\parsep}{\parskip}%
+}%
 
-\documentclass[preprint,12pt,authoryear]{elsarticle}
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage[utf8]{inputenc}
-\usepackage{url,hyperref,lineno,microtype,subcaption}
-\usepackage{color,tensor,multirow,siunitx}
-\usepackage[onehalfspacing]{setspace}
-\usepackage{makecell}
-\renewcommand{\cellalign}{cl}
-\usepackage{caption}
-\usepackage{lipsum}
-\captionsetup[figure]{name=Fig., labelfont=bf}
-\captionsetup[table]{name=Table, labelfont=bf}
-
-\renewcommand{\cellalign}{cl}
-
-\newcommand{\ds}{\displaystyle}
-\newcommand{\nl}{\ \\ }
-\newcommand{\ud}{\textrm{ d}}
-\newcommand{\bs}{\bigskip}
-
-\newcommand{\bu}{\mathbf{u}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\be}{\mathbf{e}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bk}{\mathbf{k}}
-\newcommand{\bn}{\mathbf{n}}
-\newcommand{\bR}{\mathbf{R}}
-
-\definecolor{orange}{rgb}{1.0, 0.46, 0.09}
-\definecolor{red}{rgb}{1,0,0}
-\definecolor{blue}{rgb}{0,0,0.8}
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand{\emphc}[1]{\emph{\textcolor{red}{#1}}}
-\newcommand{\hycom}{\textsc{hycom} }
-\newcommand{\ie}{{\it i.e.}\ }
-\newcommand{\eg}{{\it e.g.}\ }
-\newcommand{\UV}{\mathbf{U}}
-\newcommand{\todo}[1]{\textcolor{red}{TO DO: #1}}
-\newcommand{\comment}[1]{\textcolor{orange}{#1}}
-%\newcommand{\modif}[1]{\textcolor{blue}{#1}}
-% \renewcommand{\familydefault}{\sfdefault}
-
-\journal{Marine Pollution Bulletin}
-
-\begin{document}
-
-\begin{frontmatter}
-
-    %% Title, authors and addresses
-
-    %% use the tnoteref command within \title for footnotes;
-    %% use the tnotetext command for theassociated footnote;
-    %% use the fnref command within \author or \affiliation for footnotes;
-    %% use the fntext command for theassociated footnote;
-    %% use the corref command within \author for corresponding author footnotes;
-    %% use the cortext command for theassociated footnote;
-    %% use the ead command for the email address,
-    %% and the form \ead[url] for the home page:
-    %% \title{Title\tnoteref{label1}}
-    %% \tnotetext[label1]{}
-    %% \author{Name\corref{cor1}\fnref{label2}}
-    %% \ead{email address}
-    %% \ead[url]{home page}
-    %% \fntext[label2]{}
-    %% \cortext[cor1]{}
-    %% \affiliation{organization={},
-    %%            addressline={}, 
-    %%            city={},
-    %%            postcode={}, 
-    %%            state={},
-    %%            country={}}
-    %% \fntext[label3]{}
-
-    \title{Early transmission dynamics of SCTLD and its connection with the PortMiami Deep Dredge Project}%Impacts of Hurricane Irma (2017) on ocean transport processes}
-
-    %% use optional labels to link authors explicitly to addresses:
-    %% \author[label1,label2]{}
-    %% \affiliation[label1]{organization={},
-    %%             addressline={},
-    %%             city={},
-    %%             postcode={},
-    %%             state={},
-    %%             country={}}
-    %%
-    %% \affiliation[label2]{organization={},
-    %%             addressline={},
-    %%             city={},
-    %%             postcode={},
-    %%             state={},
-    %%             country={}}
-
-    \author[eli]{Thomas Dobbelaere\corref{corr}}
-    \ead{thomas.dobbelaere@uclouvain.be}
-    \author[]{other co-authors TBD}
-    %\author[rsmas]{Milan Curcic}
-    %\author[cimas,aoml]{Matthieu Le H\'enaff}
-    \author[eli,immc]{Emmanuel Hanert}
-    \cortext[corr]{Corresponding author}
-
-   \affiliation[eli]{
-       organization={Eath and Life Institute (ELI), UCLouvain},
-       % addressline={}, 
-       city={Louvain-la-Neuve},
-       % postcode={1348}, 
-       % state={},
-       country={Belgium}
-   }
-%    \affiliation[rsmas]{
-%        organization={Rosenstiel School of Marine and Atmospheric Sciences (RSMAS), University of Miami},
-%        % addressline={}, 
-%        city={Coral Gables},
-%        % postcode={1348}, 
-%        state={Florida},
-%        country={USA}
-%    }
-%    \affiliation[cimas]{
-%        organization={Cooperative Institute for Marine and Atmospheric
-%        Studies (CIMAS), University of Miami},
-%        % addressline={}, 
-%        city={Miami},
-%        % postcode={1348}, 
-%        state={Florida},
-%        country={USA}
-%    }
-%    \affiliation[aoml]{
-%        organization={Atlantic Oceanographic and Meteorological Laboratory
-%        (AOML), NOAA},
-%        % addressline={}, 
-%        city={Miami},
-%        % postcode={1348}, 
-%        state={Florida},
-%        country={USA}
-%    }
-    \affiliation[immc]{
-        organization={Institute of Mechanics, Materials and Civil Engineering (IMMC), UCLouvain},
-        % addressline={}, 
-        city={Louvain-la-Neuve},
-        % postcode={1348}, 
-        % state={},
-        country={Belgium}
-    }
+\item \textbf{Dobbelaere T.}, other co-authors TBD, $\&$ E.
+Hanert, 2022, Early transmission dynamics of SCTLD and its connection with the PortMiami Deep Dredge Project.
+(planned submission to \textit{Marine Pollution Bulletin})
+\end{list}
 
@@
 
-    %%Graphical abstract
-    % \begin{graphicalabstract}
-    % %\includegraphics{grabs}
-    % \end{graphicalabstract}
+%\newpage
 
-    %Research highlights
-    % \begin{highlights}
-    %     \item The coupled SLIM+SWAN model correctly reproduced the hydrodynamics and waves during Irma.
-    %    \item Wave radiation stress increased currents by up to 1 m/s during Irma.
-	%    \item Wave radiation stress gradients were the largest on the shelf break and over reefs. 
-	%   \item Waves could deflect drifting particles by up to 20 km during the hurricane.
-	%  \item The Stokes drift had an impact on transport 4 times larger than the wave radiation stress.
-    % \end{highlights}
-
-    \begin{keyword}
-        Stony coral tissue loss disease \sep Sediments \sep PortMiami \sep Coastal modeling \sep Sediment plumes 
-    %% keywords here, in the form: keyword \sep keyword
-
-    %% PACS codes here, in the form: \PACS code \sep code
-
-    %% MSC codes here, in the form: \MSC code \sep code
-    %% or \MSC[2008] code \sep code (2000 is the default)
-
-    \end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
-
-%  PARAGRAPH DISEASES IN GENERAL
-Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region \citep{richardson1998coral, sutherland2004disease, aronson2001white, harvell2007coral, brandt2009dynamics}.
-One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}.
-First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
-Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}.
-Furthermore, recent studies showed evidence that sediments can act as  a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
-
-% POM
-Some of the first signs of SCTLD were reported in September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}.
-They derived a relationship between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
-It was therefore hypothesized that the epidemic started near Virginia Key and then spread to the neighboring reef, both north and south.
-However, earlier signs of diseases were already reported north of Virginia Key in June 2014, at the monitoring site of N.
-Sunny Isles \citep{precht2016unprecedented}.
-All these observations occurred during the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015.
-The dredging was monitored twice-weekly at 26 permanent monitoring stations established within the Miami-Dade County, making it one of the most complete datasets related to a dredging project \citep{gintert2019regional}.
-Interestingly, disease signs were also reported at one of these monitoring stations in May 2014.
-
-While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located 4.7 nautical miles offshore.
-However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
-The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report} \comment{[find better ref ?]}.
-Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.
-
-% SEDIMENTS
-Sediments released by dredging can affect the biological functions of corals in numerous ways  through turbidity and sedimentation \citep{erftemeijer2012environmental, jones2015effects}.
-Increased turbidity caused by the suspended sediments reduces the light available to symbiotic zooxanthellae, leading to reduced coral cover and growth.
-Sedimentation, on the other hand, can cause smothering or burial of coral polyps \citep{erftemeijer2012environmental}.
-Furthermore, both sedimentation and turbidity can significantly reduce larval recruitment by inhibiting settlement and reducing larval survival in the water column \citep{jones2015effects}.
-These effects are stronger with fine-grained sediments, as they cause a stronger light reduction \citep{fourney2017additive}.
-Additionally, fine-grained sediments such as silts have high nutrient contents, which can lead to an increased microbial activity, eventually causing anoxic conditions in the immediate vicinity of corals \citep{weber2012mechanisms}.
-As they release finer sediments over significantly longer periods than natural events such as hurricanes, dredging activities can thus be more harmful to corals and reef habitat compared to other types of sedimentation \citep{cunning2019extensive}.
-
-% ON ENTRE DANS LE VIF DU SUJET
-Nonetheless, \cite{gintert2019regional} argued that the reported coral mortality during the dredging project was dominated by the regional outbreak of SCTLD.
-Further, they suggested that the onset of the disease might have been linked to a leaking discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant (WWTP) located off Virginia Key.
-However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease was transported to the monitoring site of Virginia Key on sediments released by the dredging.
-This possibility can be evaluated using a hydrodynamic model simulating the transport of sediments produced during the dredging project.
-As coastal reef ecosystems are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures, such a model would require high spatial resolution to accurately represent the transport of sediments at the reef-scale.
-In this context, unstructured-mesh models are particularly well suited, as they can easily adapt to the topography \citep{fringer2019future} and can capture small-scale circulation features around reefs and islands \citep{lambrechts2008multi}.
-
-The goal of this study is therefore to simulate the trajectories of the sediments released during the entirety of the PortMiami Deep Dredge Project (PMDDP) using a high-resolution hydrodynamic model coupled with a sediment transport model.
-Specifically, our goal is to answer the following questions: (1) Which reefs were impacted by PMDDP ?
-(2) Is the impact on these reefs consistent with the observed timing of the onset of SCTLD ?
-
-% === METHODS === %
-\section{Methods}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/fig_mesh_onset.png}
-    \caption{\textbf{A}: Model mesh near the dredged channel.
-Elements have a characteristic length of 100 m over reefs (in light grey) and along the coasts (in dark gray).
-The monitoring sites considered in the present study are shown by blue dots.
-The date where SCTLD was first observed at these sites is given between brackets.
-The Ocean Dredge Material Disposal Site (ODMDS) is shown in red and the discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant in green.
-\textbf{B}: Close up view of the dredged channel.
-The locations of the different types of dredging that took place during the expansion of PoM are shown by colored markers.
-\textbf{C:} Snapshot of the modeled currents in the vicinity of the dredged channel.
-Small-scale flow features such as the acceleration of currents between reefs and islands are well reproduced by the model.
-\todo{refaire la figure}}
-    \label{fig:onset_mesh}
-\end{figure}
-
-The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022}.
-SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features.
-The mesh used in this study was built following the same methodology as \cite{dobbelaere2022}, with a local refinement near PoM and in the Bay of Biscayne to achieve a resolution of 100 m in the vicinity of the dredged channel (Fig.
-\ref{fig:onset_mesh}A).
-It was made up of approximately $3.5\times 10^5$ triangles and was generated with the seamsh\footnote{\url{https://pypi.org/project/seamsh/}} Python library, which is based on the the open-source mesh generator GMSH \citep{geuzaine2009gmsh}.
-The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the first observation of SCTLD by \cite{precht2016unprecedented}.
-Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the flow acceleration between reefs and islands.
-
-The transport of sediments released from the channel was then modeled using a Lagrangian particle tracking model, forced by the simulated currents.
-The sediment model is inspired by the Particle Transport Model (PTM), developed by the US Army Corps of Engineers \citep{macdonald2006ptm}.
-In this model, particles undergo a combination of horizontal and vertical motions.
-The vertical is mostly driven by gravity, with heavier particles sinking faster.
-Once they have settled, particles can be resuspended when shear stress exceeds   the critical Schields parameter, as parameterized by \cite{soulsby1997threshold}.
-The horizontal motion of the suspended particles is derived from the 2D model velocity by assuming a vertical log profile, hence yielding a quasi-3D approach.
-When sediment particles enter the near-bed zone, their horizontal velocity is greatly reduced and sediments are transported with the bedload.
-
-As sediment dispersion is dependent on the grain size, we modeled the dispersal of five sediment classes to represent to impact of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m.
-We performed a  different simulation for each class, with the grain size randomly drawn from a uniform distribution over the corresponding size range.
-The density of each sediment particle was derived from their size using the formula of \cite{hamilton1982sound}.
-Furthermore, all particles were differentiated based on the type of dredge that produced them.
-Five types of dredge were considered in our modeling study (Fig.
-\ref{fig:onset_mesh}B): ($a$) Texas cutterhead (TX), ($b$) non-conventional dredging, \ie TX with suction mechanism turned off (NonConv), ($c$) Spider Barge (SB), ($d$) Terrapin Island hopper (TI), and ($e$) Dredge 55 clamshell (D55).
-
-Dredging operations performed during the expansion of PoM were characterized in our dataset by a date, a location and  a type of dredging operation (Fig.
-\ref{fig:onset_mesh}B).
-In the absence of information about the exact time of the dredging, sediment particles were released from the dredging location during a whole day at a rate of 80 particles/hour in the model.
-To account for the motion of spider barges between the dredging site and disposal site, particles were released every 500 m along a straight line joining the dredging location to the ODMDS (see Fig.
-\ref{fig:onset_mesh}A) for every dredging operation labelled as SB.
-
-The outputs of the sediment model were then used to evaluate the turbidity and sedimentation generated by the expansion of PoM over coral reefs.
-The occurrence of high turbidity over reefs was assessed based on the concentration of suspended sediment particles in the model.
-This concentration was computed by counting the number of suspended particles inside the cells of a regular 200 m $\times$ 200 m grid over our computational domain.
-The modeled occurrence of plume was then compared against daily data of plume detection.
-This dataset was derived from satellite imagery by \cite{cunning2019extensive}\footnote{datasets available at \url{https://github.com/jrcunning/pom-dredge/}} following the methods of \cite{barnes2015sediment} at sites located within 15 km of the dredged channel.
-As in these two previous studies, we computed  the simulated plume frequency by dividing the number of days during which plumes occurred by the total number of simulated days for all grid cells.
-The impact of sedimentation was quantified by computing the cumulated concentration of settled particles within the same computational grid.
-This cumulated concentration was normalized by the total number of simulated time steps to obtained an averaged concentration of deposited particles over the simulated period.
-Higher values of this indicator would indicate larger sediment deposition over longer cumulated time.
-
-The potential impact of dredging on the onset of the SLD outbreak was more specifically assessed at five monitoring sites were signs of disease were reported.
-Four sites were reefs where disease was reported in 2014 by \cite{precht2016unprecedented}.
-The fifth site was monitoring station HBSC1-CP of the monitoring of the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
-\ref{fig:onset_mesh}A, see pictures in Appendix).
-The impact of dredging was assessed by counting the number of sediment particles originating from each dredging that were transported within 500 m of all five site.
-This distance was chosen to match the scale of the 500 m $\times$ 500 m reef polygons used in the model \citep{dobbelaere2020coupled}.
-The obtained number of particles was then divided by the total number of sediment particles released by each type of dredge.
-Larger values of this indicator would suggest a greater impact of a given type of dredging operation at a given monitoring sites.
-
-Finally, as previous studies showed evidence of waterborne transmission of SCTLD \citep{aeby2019pathogenesis, dobbelaere2020coupled,eaton2021measuring, meiling2021variable}, there is a possibility that the disease propagated to Virginia Key from other diseased reefs affected prior to September 2014.
-To evaluate this possibility, we computed monthly disease connectivity matrices following the methodology of \cite{dobbelaere2020coupled} during our simulated period.
-These connectivity matrices can be interpreted as large graphs whose vertices are 500 m $\times$ 500 m reef areas and whose edges represent disease connectivity pathways.
-Evaluating the possibility of disease propagation from any sub-reef $A$ to any sub-reef $B$ is therefore equivalent to evaluating the existence of paths, \ie sequences of connected vertices in the network starting from $A$ and reaching $B$.
-As computing all possible paths is not computationally tractable, we limited ourselves to the computation of shortest paths from any given reefs to the Virginia Key monitoring site.
-This was performed using the function \texttt{get\_all\_shortest\_paths} of the Python \texttt{python-igraph} package \citep{csardi2006igraph}.
-Such function requires the definition of a weight $w_{ij}$ for the edge connecting reef $i$ to reef $j$.
-We chose $w_{ij} = 1-\tilde{C}_{ij}$, where $\tilde{C}_{ij}$ is the probability of disease propagation from reef $i$ to reef $j$, so that "shorter" edges of the networks (\ie connectivity pathways with smaller weights) correspond to connections with larger disease propagation probability.
-The probability of a given path was then defined as the mean connection probability of the edges composing this path.
-
-Finally, the 2D mode of the particle tracking model was used to assess the impact of wastewater leaking from the discharge pipe mentioned by \cite{gintert2019regional}.
-This model simulates the transport of neutrally buoyant material driven by mean current within the water column \citep{dobbelaere2020coupled}.
-Without clear information about the location of the leak, particles were continuously released every 100 m along a straight line between Miami Central District Municipal WWTP, located on Virginia Key and the ocean discharge outfall located at 25$^\circ$44'31''N 80$^\circ$05'10''W \citep{koopman2006ocean}.
-The impact of wastewater released from the different potential leak locations was then evaluated by computing the fraction of released wastewater particles that were transported within 500 m of the site of Virginia Key.
-
-%%%%%%%%%%%%%%%%%%%
-% --- RESULTS --- %
-%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-5-50 $\mu$m was the only grain size that produced plumes consistent with the observations of \cite{barnes2015sediment} and \cite{cunning2019extensive} (Fig.
-\ref{fig:onset_depo}A).
-For larger grain sizes, particles settled in the direct vicinity of the channel.
-For these heavier particles, suspended sediments were only observed offshore, closer to the Florida Current, where the current velocity was sufficiently strong to prevent deposition.
-This suggests that the observed turbidity was mostly due to fine silts.
-The modeled occurrence of plumes within 15 km of the dredged matched the presence/absence data of \cite{cunning2019extensive} in 82.7\% of cases and the total area where plumes were observed was about 230 km$^2$, consistent with the $\sim228$ km$^2$ estimated by \cite{barnes2015sediment}.
-Modeled plumes mostly occurred north of the dredged channel, as particles were driven northward under the action of the Florida Current, with plume frequencies of about 20\% near the monitoring site of N.
-Sunny Isles.
-Plume frequencies was lower ($\sim 5$\%) over offshore reefs but reached of up to 50\% over alongshore reefs.
-These alongshore reefs correspond to regions of important sedimentation (Fig.
-\ref{fig:onset_depo}B), which is consistent with the correlation between plumes and sedimentation highlighted in \cite{cunning2019extensive}.
-The plume frequency was about 1\% at site of HBSC1-CP while no plume occurred at other southern monitoring sites during the simulation.
-
-% \begin{table}
-%     \centering
-%     \begin{tabular}{|l|cccc|}
-%         \hline
-%          & Threshold         & Accuracy & False positive  & False negative \\
-%          & [particles/m$^2$] & [\%]     & [\%]            & [\%]  \\
-%         \hline
-%         5-50 $\mu$m    & 21 & 86.678 & 2.912 & 10.410 \\
-%         50-100 $\mu$m  & 20 & 83.058 & 8.525 & 8.417 \\
-%         100-200 $\mu$m & 30 & 84.814 & 7.133 & 8.053 \\
-%         200-300 $\mu$m & 18 & 84.459 & 7.988 & 7.553 \\
-%         300-400 $\mu$m & 13 & 84.615 & 8.140 & 7.244 \\
-%         \hline
-%     \end{tabular}
-%     \caption{Validation of the modeled presence of plumes against plume observations from satellite imagery. Overall, the model agrees well with observations}
-%     \label{tab:onset_val} 
-% \end{table}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/deposition_plumes.png}
-    \caption{\textbf{A}: Plume frequency for grain size in the range 5-50 $\mu$m.
-The averaged concentration of deposited sediments are shown for grain sizes more likely to carry the disease: 5-50 $\mu$m (\textbf{B}) and 50-100 $\mu$m (\textbf{C}).
-Most modeled turbidity and sedimentation occurred north of the dredged channel.
-Non negligible sediment deposition occurred at site HBSC1-CP.}
-    \label{fig:onset_depo}    
-\end{figure}
-
-Deposition results are shown for grain sizes corresponding to silts, which are more likely to carry organic matter and therefore more likely to carry SCTLD agents \citep{erftemeijer2012environmental}(Fig.
-\ref{fig:onset_depo}B,C).
-Sedimentation mostly occurred on reefs located north of the dredge channel.
-For grain sizes smaller than 5-50 $\mu$m, sediments mostly settled on inshore reefs while sedimentation mostly took place on offshore reefs for grain size larger than 50$\mu$m.
-When increasing grain sizes, the spatial distribution of sedimentation shifted southward, with a decrease of sedimentation near N.
-Sunny Isles and an increase of the concentration of settled sediments near HBSC1-CP and Virginia Key.
-The normalized cumulated deposition was around 0.01 particles/m$^2$ at site HBSC1-CP for all grain sizes while it increased from $2.5\times 10^{-7}$ to $1.6\times 10^{-3}$ with grain size at Virginia Kay.
-For all grain sizes, no sedimentation occurred near Rainbow Reef and Emerald Reef.
-However, sediments settled at similar latitudes west to these sites for grain sizes below 100 $\mu$m.
-
-The evaluation of the impact of dredging at the monitoring sites indicated that dredging did not significantly impact the stations located south of the dredge channel, except at site HBSC1-CP (Fig.
-\ref{fig:onset_bar}).
-In contrast, up to 40\% of the released sediment particles reached the northern site of N.
-Sunny Isles.
-The impact of dredging at this site decreased with sediment size and the fraction of sediment particles reaching the N.
-Sunny Isles dropped below 2\% for grain sizes larger than 200 $\mu$m.
-However, for grain sizes finer than 50 $\mu$m, N.
-Sunny Isles was heavily impact by TX dredging.
-Indeed, the number of particles produced by TX reaching N.Sunny Isles was 50\% larger than the cumulated number of particles produced by other dredging operations reaching the monitoring site for the same grain size.
-Such large impact can be explained by the fact that TX was one of the most frequent type of dredging during the simulations, with 268 simulated operations (Fig.
-\ref{fig:onset_bar}E).
-However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles to reach N.
-Sunny Isles.
-For grain sizes below 100 $\mu$m, between 10\% and 30\% of the sediments released by non conventional dredging reached N.
-Sunny Isles.
-This fraction dropped below  1\% for grain sizes larger than 200$\mu$m.
-The fraction of sediments produced by non conventional dredging reaching site HBSC1-CP remained between 1\% and 5\% for all grain sizes.
-For grain sizes below 100 $\mu$m, D55 and NonConv were the main sources of sediments reaching HBSC1-CP.
-Above 100 $\mu$m, the main source of sediments to the site became TI.
-The fraction of sediments produced by SB reaching HBSC1-CP remained negligible for all grain sizes.
-No sediment particles reached the sites of Rainbow Reef and Emerald Reef for all grain sizes.
-However, some particles reached Virginia Key for grain sizes larger than 100 $\mu$m.
-These sediments particles almost entirely originated from SB.
-Nonetheless, the impact of the dredging on Virginia Key remained limited with less than 1$\%$ of the released particles reaching the site.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.85\textwidth]{figures/aggregated_with_absolute.png}
-    \includegraphics[width=.85\textwidth]{figures/timeline.png}
-    \caption{Fraction and absolute number of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N.
-Sunnny Isles grain sized of 5-50 $\mu$m (\textbf{A},\textbf{C}) and 50-100 $\mu$m (\textbf{B},\textbf{D}).
-\textbf{E}: Temporal distribution of the simulated dredging operations.
-With the exception of HBSC1-CP, sites located south to the dredged channel were barely impacted by the dredging.
-However, a significant fraction of the released sediment particles reached the northern site of N.
-Sunny Isles.
-The number of simulated dredging operations is given between brackets for each dredging type}
-    \label{fig:onset_bar}
-\end{figure}
-
-Applying the same analysis to the discharge pipe of Miami Central District Municipal WWTP suggests that only wastewater released from a leak in the direct vicinity of the Virginia Key could have impacted the site.
-Fig.
-\ref{fig:onset_pipe} shows the fraction of released wastewater particles that reached the site of Virginia key for every 100 m segments of the discharge pipe.
-The fraction of particles that drifted within 500 m of Virginia Key was: 100\% for particles released within 500 m of the site; 0-1\% for particles released between 500 m and 1.5 km away from the site; and 0\% for all other sources of wastewater particles.
-As HBSC1-CP was also located near the discharge pipe, a similar analysis was conducted for this monitoring site and gave similar results.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.7\textwidth]{figures/pipeline.png}
-    \caption{Fraction of wastewater particles reaching the site of Virginia Key for potential leak positions located every 100 m along the discharge pipe.
-Only particles released in the direct vicinity of Virginia Key could reach the monitoring site.}
-    \label{fig:onset_pipe}
-\end{figure}
-
-As signs of disease were observed at site HBSC1-CP in May 2014, before the first observations of SCTLD in Virginia Key in September 2014, we assessed the presence of shortest paths from HBSC1-CP to Virginia Key in the modeled monthly disease connectivity networks between May and September 2014 (Fig.
-\ref{fig:onset_path}).
-We found connectivity pathways connecting the two sites during all months of May-September 2014, with the exception of July.
-This suggests that there was a possibility of disease propagation from HBSC1-CP to Virginia Key during most of these 5 months.
-However, we found no direct pathway connecting the two sites.
-Southern intermediary reefs were systematically needed as stepping stones for the propagation of the disease.
-This suggests that several months might have been required for disease agents to reach Virginia Key from HBSC1-CP.
-Moreover, shortest paths in June, August and September 2014 had many sub-reef stepping stones in common.
-These similar connectivity patterns indicate favorable conditions for disease propagation over several months from HBSC1-CP to Virginia Key during this period.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/fig_paths.png}
-    \caption{Shortest path from HBSC1-CP to Virginia Key in the monthly disease connectivity networks between May and September 2014.
-July 2014 was the only month without modeled connectivity between the two sites between May and September 2014.
-Southern intermediary reefs were needed as stepping stones for the propagation of the disease from HBSC1-CP to Virginia Key.}
-    \label{fig:onset_path}
-\end{figure}
-
-% === DISCUSSION === %
-\section{Discussion}
-
-% SUMMARY
-The quasi-3D sediment model forced by currents from the high-resolution coastal ocean model SLIM reproduced a sediment dynamics consistent with plume observations derived from satellite imagery.
-It suggests that sediment particles produced by dredging were mostly transported northward under the influence of the Florida Current.
-No sediment particles reached the southern sites of Rainbow Reef and Emerald Reef while dredging had limited direct impact on Virginia Key, identified as the site where the outbreak of SCTLD was initiated in September 2014.
-Furthermore, our results indicate that it is unlikely that the site was affected by a leaking discharge pipe of Miami Central District Municipal WWTP, as suggested by \cite{gintert2019regional}.
-However, signs of disease were observed prior to September 2014 at the sites of N.
-Sunny Isles and HBSC1-CP, where the impact of dredging was greater.
-Up to 30\% of silt particles produced by non-conventional dredging reached coral reefs of N.
-Sunny Isles while 5\% of them reached site HBSC1-CP, where significant sedimentation occurred.
-Furthermore, analysis of monthly disease connectivity networks indicate disease transmission from HBSC1-CP to the site of Virginia Key was possible through waterborne transmission.
-
-% LARGE IMPACT OF DREDGING
-Our study confirms previous reports about the widespread impact of the expansion of PoM.
-As in \cite{barnes2015sediment}, the total area of plumes in our model was about 230 km$^2$.
-Furthermore, the model results reproduced the presence/absence within 15 km of the dredged channel derived by \cite{cunning2019extensive} with an accuracy of 82.7\%.
-Such good agreement was only obtained for grain sizes in the range 5-50 $\mu$m, suggesting that fine silts were the main drivers of the turbidity generated by dredging, which is consistent with previous studies \citep{fourney2017additive}.
-Despite this good agreement, the distribution of plumes predicted by our model tends to be shifted northward compared to these previous studies.
-These discrepancies might be explained by the fact that our approaches is solely based on sediment concentration.
-However, turbidity depends on many local factors such as the water content of phytoplankton and organic matters \citep{gray2000comparability,thackston2000improved}.
-As reported by \cite{cunning2019extensive}, the plumes mostly occurred in areas of high sedimentation.
-As such areas were mostly located over coral reefs, our results suggests that the expansion of PoM might have harmed coral populations over an extended area through increased turbidity and sedimentation.
-This sedimentation mostly took place on alongshore reefs for grain sizes smaller than 50$\mu$m and on offshore reefs for coarser grain sizes.
-Sediment settlement might be significantly harmful to offshore coral reefs populations as they are usually less accustomed to sedimentation than their inshore  counterparts \citep{wolanski2005fine}.
-
-% LIMITATION + SHOW THAT N. SUNNY ISLE AND HBSC1-CP WERE SIGNIFICANTLY IMPACTED
-A limitation of our study is that the amount of released sediments might be overestimated, as conventional and non-conventional dredging are treated in the same way in the model.
-For all dredging types, sediment particles were released at the same rate in the model.
-Although conventional dredging was reported to release fine-grained sediments in the water column through dewatering and overflow from barges \citep{jones2016assessing}, the quantity of dredged material lost in the water was limited by the use of pumping mechanisms.
-By contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rock particles to be released in the water column.
-The numbers of particles reaching the monitoring sites were therefore likely overestimated for all sources but non conventional dredging.
-However, the fractions given in Fig.
-\ref{fig:onset_bar}A,B remain valid as they are relative to the total number of sediments released in the water columns.
-Despite these overestimations, the coral reefs of N.
-Sunny Isles appeared to be highly sensitive to fine-grained sediments produced by dredging (Fig.
-\ref{fig:onset_bar}).
-Furthermore, the non-conventional rock-chopping that took place between December 2013 and May 2014, was a non-negligible source of sediments to N.
-Sunny Isles, as up to 30\% of the released particles reached the site.
-As sediments have the potential to act as a SCTLD vector \citep{rosales2020rhodobacterales, studivan2022reef}, the dredging might therefore have contributed to the observed onset of the disease on the reefs of N.
-Sunny Isles in June 2014.
-Moreover, up to 5\% of the fined-grained sediments produced by non-conventional dredging reached the site HBSC1-CP.
-Although this represents a more limited quantity of sediments, HBSC1 was within 2 km from the sites where non-conventional dredging took place, in a region where the model predicted an important sedimentation.
-Therefore, although fewer sediments reached HBSC1-CP, they had a higher probability to settle and to be in direct contact with corals.
-In addition to smothering corals and diverting their energy through sediment removal \citep{erftemeijer2012environmental}, these settled sediments are more likely to transmit the disease to the reefs.
-
-% POSSIBILITY OF DISEASE PROPAGATION THROUGH WATERBORNE TRANSMISSION
-In all simulations, the expansion of PoM had limited direct impact on Virginia Key, identified as the site where the outbreak initiated in September 2014 \cite{precht2016unprecedented}.
-For all grain sizes, the fraction of sediment particles reaching Virginia Key remained below 1\%.
-It is therefore unlikely that sediments produced by the deepening of the channel brought the disease to the site.
-Furthermore, our simulation of leaking wastewater along the discharge pipe of Miami Central District Municipal WWTP suggests that currents were likely to flush wastewater away from Virginia Key.
-It was therefore unlikely that wastewater affected the site, except if the leak was located within hundreds of meters from the site.
-However, signs of disease were observed prior to September 2014 at the sites of N.
-Sunny Isles and HBSC1-CP, both of which were impacted by the dredging.
-Several studies showed evidence of waterborne transmission of SCTLD \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring,meiling2021variable}.
-Disease agents might therefore have been transported by currents from one of the diseased sites to Virginia Key.
-As site HBSC-CP was the closest to Virginia Key, we built and analyzed monthly disease connectivity networks and found connectivity pathways from HBSC1 to Virginia Key in May, June, August and September 2014.
-This implied that the propagation of the disease from HBSC1-CP to Virginia Key through hydrodynamics-driven transport of disease agents was possible during these four months.
-However, these pathways used reefs located further southward as stepping stones.
-Therefore, the propagation of outbreak to reach Virginia Key starting from HBSC1-CP required these intermediary reefs to get first infected by disease agents released by disease colonies of HBSC1-CP.
-Once diseased, these colonies would then send diseased agents to affect the next colonies of the connectivity pathway until the outbreak reached Virginia Key.
-Assuming an averaged transmission time of the order of 5-10 days \citep{dobbelaere2020coupled}, disease propagation from HBSC1-CP to Virginia Key might have required several months.
-This is consistent with the 5 month period separating reports of diseased at the two sites.
-Moreover, the fact that June, August and September exhibited similar connectivity pathways suggests that disease propagation could indeed have occurred over several months without being interrupted by changing hydrodynamic conditions.
-
-\section{Conclusion}
-
-In the present study, we evaluated the impact of the dredging operations that took place the impact of the deepening of the Port of Miami shipping channel on the neighboring coral reefs.
-This evaluation was performed using a quasi-3D sediment model forced by high-resolution currents from the coastal ocean model SLIM.
-As the first signs of the SCTLD outbreak were reported during these expansion works near Virginia Key and knowing that sediments may serve as SCTLD vector, we aimed at answering two main questions.
-First, we wanted to evaluate the fraction of sediments produced by the PortMiami Deep Dredge Project that reached the site of Virginia Key.
-Second, we evaluated the possibility of waterborne transmission of SCTLD to Virginia Key from other reefs impacted by the dredging.
-Doing so, we paid a special attention to the sediments produced by non conventional rock chopping during which pumping mechanisms were turned off, causing all chopped rock particles to be directly released in the water column.
-
-Our results suggest that the dredging works had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission trough the produced sediments was extremely unlikely.
-However, our results show the sediments released by the expansion works had a non-negligible impact on the sites of N.
-Sunny Isles and HBSC1-CP, where signs of disease were observed before the outbreak was first reported at Virginia Key, indicating that the sediments might have played a role in the onset of the diseased at these two sites.
-Furthermore, using a previously developed bio-physical model that successfully reproduced the observed spread of SCTLD, we show that there was a possibility of waterborne disease propagation from HBSC1-CP to Virginia Key.
-
-This study suggests that the expansion of the Port of Miami might have played a role in the onset of the outbreak of SCTLD in sites where the disease was reported during the first half of 2014.
-Furthermore, it shows that the disease could then have spread from these sites to Virginia Key, from where SCTLD was then reported to propagate both north and south to the neighboring reefs.
-Our work brings new insight on the role played by a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
-
-\bibliographystyle{elsarticle-harv} 
-\bibliography{./biblio.bib}
-
-\end{document}
-\endinput
+\input{chapters/onset/article}

### 2022-02-07 chapters/sctld/article.tex
last chapter added to manuscript

@@
 As with any modeling study, it is important to understand the assumptions on which the model is based.
-Here, we have used a 2D barotropic ocean model forced by the 3D model HYCOM \citep{Chassignet2007} in order to indirectly represent baroclinic phenomena.
+Here, we have used a 2D barotropic ocean model forced by the 3D model HYCOM \citep{chassignet2007hycom} in order to indirectly represent baroclinic phenomena.
 Such a model is well-suited to simulate the fate of neutrally-buoyant material in shallow regions.

### 2022-02-07 chapters/sctld/sctld.tex
last chapter added to manuscript

@@
 
-\item \textbf{T.
-Dobbelaere}, E.
-M.
-Muller, L.
-J.
-Gramer, D.
-M.
-Holstein, $\&$ E.
-Hanert, 2020, Coupled Epidemio-Hydrodynamic Modeling to Understand the Spread of a Deadly Coral Disease in Florida, \textit{Frontiers in Marine Science},69: 1121-1132.
-\href{https://www.frontiersin.org/article/10.3389/fmars.2020.591881 }{doi: 10.3389/fmars.2020.591881}
+\item \textbf{Dobbelaere, T.}, Muller, E.
+M., Gramer, L.
+J., Holstein, D.
+M., $\&$ Hanert, E.
+(2020).
+Coupled epidemio-hydrodynamic modeling to understand the spread of a deadly coral disease in Florida.
+\textit{Frontiers in Marine Science},1016.
+\href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.3389/fmars.2020.591881}.
 \end{list}

### 2022-02-07 thesis.tex
slim+swan paper added
chapter on drto added to thesis
last chapter added to manuscript

@@
 
-\newcommand{\hycom}{\textsc{hycom} }
-\newcommand{\ie}{\textit{i.e}} 
-\newcommand{\eg}{\textit{e.g.}}
-
 \begin{document}
@@
 \input{chapters/sctld/sctld}
-%%
-% \mychapter{Paper Dry Tortugas} \label{chap:ontario}
-% \chaptermark{Paper DRTO}  
-%   \input{chapters/ontario/ontario}
-%
-% \mychapter{Paper onset} \label{chap:eigenmode}
-% \chaptermark{Paper onset}  
-%   \input{chapters/eigenmode/eigenmode}
-%%%   
+\mychapter{Transmission of SCTLD from the Marquesas to the Dry Tortugas}\label{chap:drto}
+\chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
+\input{chapters/drto/drto}
+\mychapter{The onset of the SCTLD outbreak and the PortMiami Deep Dredge Project} \label{chap:onset}
+\chaptermark{Onset of SCTLD and PortMiami Deep Dredge Project}  
+\input{chapters/onset/onset}
+\mychapter{Impacts of hurricanes on wave-induced ocean transport processes} \label{chap:irma}
+\chaptermark{Wave-induced transport under hurricanes}  
+\input{chapters/irma/irma}
 %%%%%%%%%%%

### 2022-02-08 chapters/onset/article.tex
updated ref for onset

@@
 One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}.
-First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
+First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022connecting} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
 Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}.
@@
 However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
-The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report}.
+The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{usace2017}.
 Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.
@@
 
-The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022}.
+The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022connecting}.
 SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features.
-The mesh used in this study was built following the same methodology as \cite{dobbelaere2022}, with a local refinement near PoM and in the Bay of Biscayne to achieve a resolution of 100 m in the vicinity of the dredged channel (Fig.
+The mesh used in this study was built following the same methodology as \cite{dobbelaere2020coupled}, with a local refinement near PoM and in the Bay of Biscayne to achieve a resolution of 100 m in the vicinity of the dredged channel (Fig.
 \ref{fig:onset_mesh}A).

### 2022-02-08 chapters/intro/main.tex
debut d'intro

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references
\documentclass[12pt ,a4paper]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage{color,tensor,multirow,siunitx}
\usepackage[onehalfspacing]{setspace}
\usepackage{makecell}
\usepackage{graphicx}
\renewcommand{\cellalign}{cl}
% \usepackage{times}
\usepackage{caption}
\usepackage{lipsum}
\captionsetup[figure]{name=Fig., labelfont=bf}
\captionsetup[table]{name=Table, labelfont=bf}

\renewcommand{\cellalign}{cl}

\newcommand{\ds}{\displaystyle}
\newcommand{\nl}{\ \\ }
\newcommand{\ud}{\textrm{ d}}
\newcommand{\bs}{\bigskip}

\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bR}{\mathbf{R}}

\definecolor{orange}{rgb}{1.0, 0.46, 0.09}
\definecolor{red}{rgb}{1,0,0}
\definecolor{blue}{rgb}{0,0,0.8}
\definecolor{green}{rgb}{0,0.5,0}
\newcommand{\emphc}[1]{\emph{\textcolor{red}{#1}}}
\newcommand{\hycom}{\textsc{hycom} }
\newcommand{\ie}{{\it i.e.}\ }
\newcommand{\eg}{{\it e.g.}\ }
\newcommand{\UV}{\mathbf{U}}
\newcommand{\todo}[1]{\textcolor{red}{TO DO: #1}}
\newcommand{\comment}[1]{\textcolor{orange}{#1}}
\usepackage[authoryear,round]{natbib}
%\newcommand{\modif}[1]{\textcolor{blue}{#1}}
% \renewcommand{\familydefault}{\sfdefault}
\usepackage{times}

\begin{document}
\section{Coral reefs}

\section{Coral reefs of Florida}

The Florida Reef Tract (FRT) spans over 580 km from the Dry Tortugas west of the Florida Keys to the St. Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf}. These reefs have a fauna and species richness typical of Caribbean reefs, including more than 40 species of stony corals \citep{banks2008reef,jackson2014status}. The northern half-section of the FRT consists of relic, Holocene framework reefs and indurated sand ridges harboring a rich but non-framebuilding fauna. Reefs in this area are separated into three shore-parallel reefs (inner, middle, outer) separated by sandy plains. Benthic cover is generally denser on the middle and outer reefs while the inner reef harbors some large patches of dense \textit{Acropora cervicornis}. The bouldering \textit{Montastraea cavernosa} is the dominant reef builder, but living hard coral cover is pretty low (below 6\%) \citep{banks2008reef}. The southern half-section is composed of chain of limestone islands (the Keys), sandbars and ancient coral reefs, where active carbonate deposition is taking place \citep{hoffmeister1968geology}.

Florida's Coral Reef (FCR) is of considerable importance to the economy of the state. \cite{johns2003socio} estimated the contribution of recreational users of natural and artificial reefs over the period June 2000 to May 2001 to have been US\$2.3 billion in sales and US\$1.1 billion in income. Moreover, recreational use of the reefs was evaluated to bring 36,500 full and part-time jobs. Focusing on natural reefs only, \cite{brander2013total} estimated the total coral reef values Florida to be US\$174 millions. However, this value might be underestimated as non-use value and indirect use values, such as support for coastal fisheries and coastal protection were not taken into account.

However, FCR has been heavily impacted by human activities including densely populated coastlines, high visitor numbers, polluted terrestrial run-off and overfishing \citep{jackson2014status}. Florida’s coral reefs have therefore declined significantly over the past decades due to anthropogenic impacts and numerous disease outbreaks \citep{jackson2014status}.

The Caribbean is considered as a disease "hotspot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral reef diseases in this area \citep{harvell2007coral}. Although they only represent 8\% of the total global coral reef area \citep{spalding1997new}, over 70\% of all disease and syndrome reports originate from the Caribbean.

Florida is a prime landfall target for tropical cyclones from June through November. Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}. This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}. basin Hurricanes can influence the biota of coral reefs in many ways. The most obvious is wholesale destruction but hurricanes can also positively impact corals by facilitating the asexual recruitment of some species. However, some possible. Furthermore, hurricane-generated wave and currents have been hypothesized to transport silt and clay shoreward from deeper areas, causing burial of nearshore reefs on the continental shelf \citep{miller2008effects}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/hurricane_strikes.jpg}
    \caption{}
    \label{inro:landfall}
\end{figure}

\bibliographystyle{apalike} 
\bibliography{./biblio.bib}

\end{document}

### 2022-02-08 chapters/intro/main.tex
intro - added section on coral reefs

@@
 %\newcommand{\modif}[1]{\textcolor{blue}{#1}}
-% \renewcommand{\familydefault}{\sfdefault}
-\usepackage{times}
+\renewcommand{\familydefault}{\sfdefault}
+% \usepackage{times}
 
 \begin{document}
-\section{Coral reefs}
+
+\section{Concerning coral reefs}
+
+Coral reefs are among the most productive and biologically diverse ecosystems on Earth \citep{connell1978diversity, moberg1999ecological} and supply food, income and coastal protection to vast numbers of people.
+Although they account for less than 0.5\% of the ocean floor \citep{spalding1997new}, almost a third of the known marine biodiversity is found on coral reefs \citep{moberg1999ecological}.
+Reef-related fisheries constitute around 10\% of the fish consumed by humans \citep{smith1978coral} and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
+Through calcification, corals create the structure of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
+This complex reef framework  dissipates waves, protecting shorelines from currents and storms and preventing land loss due to erosion \citep{ferrario2014effectiveness,elliff2017coral}.
+Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass ecosystems \citep{moberg1999ecological}.
+
+The calcifying process of reef building corals heavily depends on their symbiosis with the microalgae zooxanthellae.
+These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the coral host with most of the energy needed to meet its metabolic demands \citep{muscatine1977reef}.
+However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the cellular mechanisms that maintain this association, resulting in th expulsion of the endosymbionts \citep{hoegh2007coral}.
+This process, called bleaching, causes the coral to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}.
+Bleached corals are physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high levels of coral mortality \citep{hughes2018spatial}
 
@@
 
-The Florida Reef Tract (FRT) spans over 580 km from the Dry Tortugas west of the Florida Keys to the St.
+The Florida Reef Tract (FRT) spans over 580 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St.
 Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf}.
@@
 The bouldering \textit{Montastraea cavernosa} is the dominant reef builder, but living hard coral cover is pretty low (below 6\%) \citep{banks2008reef}.
-The southern half-section is composed of chain of limestone islands (the Keys), sandbars and ancient coral reefs, where active carbonate deposition is taking place \citep{hoffmeister1968geology}.
+The southern half-section isa chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO.
+The Upper Keys are remnants of ancient coral reefs and the Lower Keys are sand bars \citep{hoffmeister1968geology}.
+The DRTO are made of small circular reef banks whose coral populations are among the most preserved of the continental United States \citep{hine2008coral, kourafalou2018physical}.
+Moreover, they are believed to be important sources of recruitment for coral-reef fishes in the Florida Keys \citep{domeier2004potential}.
+Since 2001, the Florida Keys and the DRTO harbor the largest network of no-take marine reserve in North America \citep{ault2006building}.
 
 Florida's Coral Reef (FCR) is of considerable importance to the economy of the state.
-\cite{johns2003socio} estimated the contribution of recreational users of natural and artificial reefs over the period June 2000 to May 2001 to have been US\$2.3 billion in sales and US\$1.1 billion in income.
+\cite{johns2003socio} estimated the contribution of recreational users of natural and artificial reefs over the period June 2000 to May 2001 to have been US\$2.3 billions in sales and US\$1.1 billion in income.
 Moreover, recreational use of the reefs was evaluated to bring 36,500 full and part-time jobs.
 Focusing on natural reefs only, \cite{brander2013total} estimated the total coral reef values Florida to be US\$174 millions.
-However, this value might be underestimated as non-use value and indirect use values, such as support for coastal fisheries and coastal protection were not taken into account.
+However, this value might be underestimated as non-use value and indirect use values, such as support for coastal fisheries \citep{ault2006building} and coastal protection \citep{ferrario2014effectiveness} were not taken into account.
 
-However, FCR has been heavily impacted by human activities including densely populated coastlines, high visitor numbers, polluted terrestrial run-off and overfishing \citep{jackson2014status}.
-Florida’s coral reefs have therefore declined significantly over the past decades due to anthropogenic impacts and numerous disease outbreaks \citep{jackson2014status}.
+Nevertheless, FCR has been heavily impacted by human activities including densely populated coastlines, high visitor numbers, polluted terrestrial run-off and overfishing \citep{jackson2014status}.
+Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
+For instance, the dredging conducted during the deepening of the Port of Miami shipping channel were reported to cause the death of $>$ 560, 000 corals within 0.5 km of the channel \citep{cunning2019extensive}.
 
-The Caribbean is considered as a disease "hotspot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral reef diseases in this area \citep{harvell2007coral}.
-Although they only represent 8\% of the total global coral reef area \citep{spalding1997new}, over 70\% of all disease and syndrome reports originate from the Caribbean.
+The Caribbean is considered a disease "hot spot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral reef diseases in this area \citep{green2000significance, harvell2007coral}.
 

### 2022-02-09 chapters/intro/main.tex
On continue toujours l'intro

@@
 \usepackage[authoryear,round]{natbib}
+\setlength{\bibsep}{0.0pt}
 %\newcommand{\modif}[1]{\textcolor{blue}{#1}}
@@
 
-\section{Concerning coral reefs}
+\section{Coral reefs: a rich but threatened ecosystem}
 
@@
 This complex reef framework  dissipates waves, protecting shorelines from currents and storms and preventing land loss due to erosion \citep{ferrario2014effectiveness,elliff2017coral}.
-Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass ecosystems \citep{moberg1999ecological}.
+Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass and mangrove ecosystems \citep{moberg1999ecological}.
 
@@
 These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the coral host with most of the energy needed to meet its metabolic demands \citep{muscatine1977reef}.
-However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the cellular mechanisms that maintain this association, resulting in th expulsion of the endosymbionts \citep{hoegh2007coral}.
+However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral}.
 This process, called bleaching, causes the coral to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}.
-Bleached corals are physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high levels of coral mortality \citep{hughes2018spatial}
+Bleached corals are physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high coral mortality \citep{hughes2018spatial}.
+Decline in coral health and cover pushes the reef ecosystem to a macroalgal-dominated state \citep{hughes2003climate,mumby2007thresholds}.
+Once a "tipping point" is exceeded, returning to a coral-dominated state becomes difficult \citep{mumby2007thresholds,graham2015predicting}.
 
-\section{Coral reefs of Florida}
+Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2017coral}: it is estimated  that 30\% have already been severely damaged, and close to 60\% may be lost by 2030 \citep{hughes2003climate}.
+This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} is caused by both global and local anthropogenic stressors.
+The main global stressor is climate change induced ocean warming, that cause recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
+Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
+Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
 
-The Florida Reef Tract (FRT) spans over 580 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St.
+Due to the combined damages from of all these stressors, it is now impossible to return reefs to their past configurations \citep{hughes2017coral}.
+Instead, the challenge is now to maintain the biological functions of coral reefs through quickly changing environmental conditions.
+This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting}.
+The aim of the present dissertation was therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support ecosystem management in Florida, the third largest barrier reef of the world.
+
+\section{Identifying the threats to coral survival in \\Florida}
+
+The Florida Reef Tract (FRT) spans over approximately 577 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St.
 Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf}.
@@
 Benthic cover is generally denser on the middle and outer reefs while the inner reef harbors some large patches of dense \textit{Acropora cervicornis}.
-The bouldering \textit{Montastraea cavernosa} is the dominant reef builder, but living hard coral cover is pretty low (below 6\%) \citep{banks2008reef}.
+The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is pretty low (below 6\%) \citep{banks2008reef}.
 The southern half-section isa chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO.
@@
 Moreover, they are believed to be important sources of recruitment for coral-reef fishes in the Florida Keys \citep{domeier2004potential}.
-Since 2001, the Florida Keys and the DRTO harbor the largest network of no-take marine reserve in North America \citep{ault2006building}.
+
+\todo{Add figure with map of the FRT}
 
@@
 
-Nevertheless, FCR has been heavily impacted by human activities including densely populated coastlines, high visitor numbers, polluted terrestrial run-off and overfishing \citep{jackson2014status}.
+Nevertheless, FCR has been heavily impacted by the human activities of the major metropolitan area of greater Miami.
+These impacts include densely populated coastlines, high visitor numbers, polluted terrestrial run-off and overfishing \citep{jackson2014status}.
+For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to cause the death of $>$560,000 corals \citep{cunning2019extensive}.
 Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
-For instance, the dredging conducted during the deepening of the Port of Miami shipping channel were reported to cause the death of $>$ 560, 000 corals within 0.5 km of the channel \citep{cunning2019extensive}.
+Populations of previously dominant reef-building acroporids almost totally disappeared throughout the Caribbean and Florida \citep{aronson2001white}, and coral cover dropped from 40 to 60\% in 1975 to $<$7\% in the Florida Keys \citep{jackson2014status} and $<$3\% in the northern section of the FRT \citep{walton2018impacts}.
+Furthermore, since 2014, many coral species have experienced widespread and severe decline caused by the ongoing outbreak of stony coral tissue loss disease (SCTLD).
+For instance, the abundance \textit{M.
+cavernosa}, once one of the most abundant species of the FRT, decreased by 45\% between 2015 and 2016 \citep{walton2018impacts}.
 
-The Caribbean is considered a disease "hot spot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral reef diseases in this area \citep{green2000significance, harvell2007coral}.
+Coral disease outbreaks are frequent in the Caribbean, which is considered a disease "hot spot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral diseases in the area \citep{green2000significance, harvell2007coral}.
+For instance, populations of  \textit{Acropora} spp.
+decreased by $>$95\% throughout the Caribbean during numerous outbreaks of white-band disease in the 1970-1990s \citep{aronson2001white}.
+Other ajor outbreaks include the white plague and black band diseases, that respectively affect 22 and 42 corals species and caused significant declines in reef-building coral populations in the Caribbean \citep{bruckner2003field,miller2009coral, muller2011black}.
 
-Florida is a prime landfall target for tropical cyclones from June through November.
+The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}.
+First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects}.
+The outbreak affects at least 24 species of scleractinian coral and often results in whole colony mortality \citep{precht2016unprecedented, walton2018impacts}.
+Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both.
+In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton.
+The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution of reports consistent with the case definition suggests that SCTLD is the largest coral disease outbreak on record in Florida.
+Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring,meiling2021variable}.
+Furthermore, recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
+
+\todo{Add figure with pictures of SCTLD.}
+
+Additionally, Florida is a prime landfall target for tropical cyclones from June through November (Fig.
+\ref{inro:landfall}).
 Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}.
 This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}.
-basin Hurricanes can influence the biota of coral reefs in many ways.
-The most obvious is wholesale destruction but hurricanes can also positively impact corals by facilitating the asexual recruitment of some species.
-However, some possible.
-Furthermore, hurricane-generated wave and currents have been hypothesized to transport silt and clay shoreward from deeper areas, causing burial of nearshore reefs on the continental shelf \citep{miller2008effects}
+Hurricanes are a major agent of coral mortality in the Caribbean that contributed to teh decline of acroporids \citep{gardner2003long,aronson2001white}.
+They cause a wholesale destruction of the reef biota on their passage and can induced coral burial through sediment resuspension \citep{banks2008reef, miller2008effects}.
+Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}.
+Furthermore, these
 
@@
 
+\section{Thesis objectives}
+
+\section{Modeling the hydrodynamics of the FRT}
+
 \bibliographystyle{apalike} 

### 2022-02-09 chapters/intro/main.tex
modifs avant le sik

@@
 The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution of reports consistent with the case definition suggests that SCTLD is the largest coral disease outbreak on record in Florida.
-Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring,meiling2021variable}.
-Furthermore, recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
+Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
 
@@
 \ref{inro:landfall}).
+Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds > 209 km/h) \citep{neumann1999tropical}.
 Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}.
@@
 Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}.
-Furthermore, these
+Furthermore, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
+This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine}.
+Hurricane might therefore interrupt larval exchanges between otherwise connected reef but also create new connectivity pathways.
+As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricane on wave-induced processes could inform the management of reefs under future climatic conditions.
+Furthermore, as incoming storm-generated waves dissipate over reefs, better understanding the protective role of coral reefs might promote their restoration as tool to reduce coastal flooding hazard \citep{roelvink2021coral}.
 
@@
 
-\section{Modeling the hydrodynamics of the FRT}
+This thesis focuses on two main threats influencing the future and survival of FCR: the current outbreak of SCTLD and the hurricanes.
+More specifically, our goal is to develop modeling tools to answer four main questions: (1) how does the SCTLD spread from reef to reef ?
+(2) how can we explain the timing of this spread ?
+(3)
+
+Answering these questions require the development
+
+\section*{Modeling the hydrodynamics of the FRT}
 

### 2022-02-10 chapters/intro/intro.tex
intro added to main manuscript

--- 2022-02-09 chapters/intro/main.tex
@@
-%% 
-%% Copyright 2007-2020 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
-\documentclass[12pt ,a4paper]{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage[utf8]{inputenc}
-\usepackage{url,hyperref,lineno,microtype,subcaption}
-\usepackage{color,tensor,multirow,siunitx}
-\usepackage[onehalfspacing]{setspace}
-\usepackage{makecell}
-\usepackage{graphicx}
-\renewcommand{\cellalign}{cl}
-% \usepackage{times}
-\usepackage{caption}
-\usepackage{lipsum}
-\captionsetup[figure]{name=Fig., labelfont=bf}
-\captionsetup[table]{name=Table, labelfont=bf}
-
-\renewcommand{\cellalign}{cl}
-
-\newcommand{\ds}{\displaystyle}
-\newcommand{\nl}{\ \\ }
-\newcommand{\ud}{\textrm{ d}}
-\newcommand{\bs}{\bigskip}
-
-\newcommand{\bu}{\mathbf{u}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\be}{\mathbf{e}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bk}{\mathbf{k}}
-\newcommand{\bn}{\mathbf{n}}
-\newcommand{\bR}{\mathbf{R}}
-
-\definecolor{orange}{rgb}{1.0, 0.46, 0.09}
-\definecolor{red}{rgb}{1,0,0}
-\definecolor{blue}{rgb}{0,0,0.8}
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand{\emphc}[1]{\emph{\textcolor{red}{#1}}}
-\newcommand{\hycom}{\textsc{hycom} }
-\newcommand{\ie}{{\it i.e.}\ }
-\newcommand{\eg}{{\it e.g.}\ }
-\newcommand{\UV}{\mathbf{U}}
-\newcommand{\todo}[1]{\textcolor{red}{TO DO: #1}}
-\newcommand{\comment}[1]{\textcolor{orange}{#1}}
-\usepackage[authoryear,round]{natbib}
-\setlength{\bibsep}{0.0pt}
-%\newcommand{\modif}[1]{\textcolor{blue}{#1}}
-\renewcommand{\familydefault}{\sfdefault}
-% \usepackage{times}
-
-\begin{document}
-
 \section{Coral reefs: a rich but threatened ecosystem}
 
-Coral reefs are among the most productive and biologically diverse ecosystems on Earth \citep{connell1978diversity, moberg1999ecological} and supply food, income and coastal protection to vast numbers of people.
+\todo{Talk about acidifcation}
+
+Coral reefs are among the most productive and biologically diverse ecosystems on Earth \citep{connell1978diversity, moberg1999ecological}.
+They supply food, income and coastal protection to vast numbers of people (Fig.
+\ref{intro:corals}).
 Although they account for less than 0.5\% of the ocean floor \citep{spalding1997new}, almost a third of the known marine biodiversity is found on coral reefs \citep{moberg1999ecological}.
 Reef-related fisheries constitute around 10\% of the fish consumed by humans \citep{smith1978coral} and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
-Through calcification, corals create the structure of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
+Through calcification, corals are the backbone of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
 This complex reef framework  dissipates waves, protecting shorelines from currents and storms and preventing land loss due to erosion \citep{ferrario2014effectiveness,elliff2017coral}.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=.8\textwidth]{chapters/intro/figures/anthropocene.png}
+    \caption{Goods and services of coral reefs and their linkage with human activities. Source: \cite{hughes2017coral}}
+    \label{intro:corals}
+\end{figure}
+
 The calcifying process of reef building corals heavily depends on their symbiosis with the microalgae zooxanthellae.
-These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the coral host with most of the energy needed to meet its metabolic demands \citep{muscatine1977reef}.
+These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the corals with most of the energy needed to meet their metabolic demands \citep{muscatine1977reef}.
 However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral}.
@@
 Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2017coral}: it is estimated  that 30\% have already been severely damaged, and close to 60\% may be lost by 2030 \citep{hughes2003climate}.
-This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} is caused by both global and local anthropogenic stressors.
+This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} (Fig.
+\ref{intro:cover}) is caused by both global and local anthropogenic stressors.
 The main global stressor is climate change induced ocean warming, that cause recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
-Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
-Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
+Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral disease outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
+Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
 
-Due to the combined damages from of all these stressors, it is now impossible to return reefs to their past configurations \citep{hughes2017coral}.
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{chapters/intro/figures/cover.jpeg}
+    \caption{Evolution of the absolute coral cover in the Caribbean. Source: \cite{gardner2003long}}
+    \label{intro:cover}
+\end{figure}
+
+Due to the combined damages from of all these stressors, it is almost impossible to return reefs to their past configurations \citep{hughes2017coral}.
 Instead, the challenge is now to maintain the biological functions of coral reefs through quickly changing environmental conditions.
@@
 Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf}.
-These reefs have a fauna and species richness typical of Caribbean reefs, including more than 40 species of stony corals \citep{banks2008reef,jackson2014status}.
-The northern half-section of the FRT consists of relic, Holocene framework reefs and indurated sand ridges harboring a rich but non-framebuilding fauna.
-Reefs in this area are separated into three shore-parallel reefs (inner, middle, outer) separated by sandy plains.
+These reefs harbor a rich Caribbean fauna, including more than 40 species of stony corals \citep{banks2008reef,jackson2014status}.
+The northern half-section of the FRT consists of relic, Holocene framework reefs and indurated sand ridges, divided into three shore-parallel reefs (inner, middle, outer) separated by sandy plains.
 Benthic cover is generally denser on the middle and outer reefs while the inner reef harbors some large patches of dense \textit{Acropora cervicornis}.
-The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is pretty low (below 6\%) \citep{banks2008reef}.
-The southern half-section isa chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO.
+The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is fairly low (below 3\%)\citep{banks2008reef,walton2018impacts}.
+The southern half-section is a chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO.
 The Upper Keys are remnants of ancient coral reefs and the Lower Keys are sand bars \citep{hoffmeister1968geology}.
@@
 Focusing on natural reefs only, \cite{brander2013total} estimated the total coral reef values Florida to be US\$174 millions.
-However, this value might be underestimated as non-use value and indirect use values, such as support for coastal fisheries \citep{ault2006building} and coastal protection \citep{ferrario2014effectiveness} were not taken into account.
+However, this value might be underestimated as non-use and indirect use values, such as support for coastal fisheries \citep{ault2006building} and coastal protection \citep{ferrario2014effectiveness} were not taken into account.
 
@@
 Furthermore, since 2014, many coral species have experienced widespread and severe decline caused by the ongoing outbreak of stony coral tissue loss disease (SCTLD).
-For instance, the abundance \textit{M.
-cavernosa}, once one of the most abundant species of the FRT, decreased by 45\% between 2015 and 2016 \citep{walton2018impacts}.
+For instance, populations of \textit{M.
+cavernosa}, once one of the most abundant species of the FRT, decreased by 45\% in abundance between 2015 and 2016 \citep{walton2018impacts}.
+
+\todo{Add map of the propagation of SCTLD}
 
@@
 decreased by $>$95\% throughout the Caribbean during numerous outbreaks of white-band disease in the 1970-1990s \citep{aronson2001white}.
-Other ajor outbreaks include the white plague and black band diseases, that respectively affect 22 and 42 corals species and caused significant declines in reef-building coral populations in the Caribbean \citep{bruckner2003field,miller2009coral, muller2011black}.
+Other major outbreaks include the white plague and black band diseases, that respectively affect 22 and 42 corals species and caused significant declines in reef-building coral populations in the Caribbean \citep{bruckner2003field,miller2009coral, muller2011black}.
 
@@
 Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both.
-In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton.
-The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution of reports consistent with the case definition suggests that SCTLD is the largest coral disease outbreak on record in Florida.
+In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig.
+\ref{intro:sctld}).
+The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution suggest that SCTLD might be the largest coral disease outbreak on record in Florida.
 Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
 
-\todo{Add figure with pictures of SCTLD.}
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{chapters/intro/figures/sctld.jpg}
+    \caption{Brain coral affected by SCTLD in Looe Key, Florida.
+The disease is leaving a white band of recently dead skeleton in contrast to the healthy, yellow/brown tissue.
+Source: Florida Fish and Wildlife Research Institute}
+    \label{intro:sctld}
+\end{figure}
 
@@
 \ref{inro:landfall}).
-Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds > 209 km/h) \citep{neumann1999tropical}.
+Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds $>$ 209 km/h) \citep{neumann1999tropical}.
 Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}.
 This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}.
-Hurricanes are a major agent of coral mortality in the Caribbean that contributed to teh decline of acroporids \citep{gardner2003long,aronson2001white}.
-They cause a wholesale destruction of the reef biota on their passage and can induced coral burial through sediment resuspension \citep{banks2008reef, miller2008effects}.
+Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,aronson2001white}.
+They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects}.
 Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}.
-Furthermore, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
+Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
 This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine}.
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/hurricane_strikes.jpg}
-    \caption{}
+    \includegraphics[width=\textwidth]{chapters/intro/figures/hurricane_strikes.jpg}
+    \caption{Total number of major hurricanes strikes in Florida between 1900-2010.
+Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
     \label{inro:landfall}
@@
 This thesis focuses on two main threats influencing the future and survival of FCR: the current outbreak of SCTLD and the hurricanes.
-More specifically, our goal is to develop modeling tools to answer four main questions: (1) how does the SCTLD spread from reef to reef ?
-(2) how can we explain the timing of this spread ?
-(3)
+More specifically, our goal is to develop modeling tools to answer four main questions.
+First, as SCTLD was expected to propagate through waterborne transmission, we developed a coupled epidemio-hydrodynamic model to reproduce the observed spread of the outbreak through the FRT.
+Such a model provides novel insight into the management of the epidemic, the identification of its causative agent, and its mode of transmission.
+Second, we used this model to explain the timing of the propagation of SCTLD within the FRT.
+More specifically, we investigated whether the modeled hydrodynamics were explanatory of the apparent stalling of the outbreak between the Marquesas and the DRTO in 2020.
+Third, as sediments can act as vector for the disease, we evaluated the impact of the PMDDP on the reported onset of SCTLD in 2014 using a sediment transport model.
+We assessed whether sediments produced or resuspended during dredging operations might have been transported to reefs where disease was reported and evaluated whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
+Finally, we developed a coupled wave-current model to evaluate the impact of hurricanes on wave-induced transport processes.
+This model was applied to Irma, a major hurricane that made landfall in the Florida Keys in 2017, during the reproduction period of several coral species.
 
-Answering these questions require the development
+Answering these four questions required the use of an ocean model able to capture both the large-scale ocean features influencing circulation along the FRT as well as small-scale flow features down to the reef scale.
+A description of the main hydrodynamic features of the FRT as well as an overview of the modeling tools used in this thesis are given in the next section.
 
 \section*{Modeling the hydrodynamics of the FRT}
+\todo{Description of circulation in the FRT and SLIM ?}
 
-\bibliographystyle{apalike} 
-\bibliography{./biblio.bib}
+\section*{Outline}
+This thesis is mainly a collection of papers describing our answers to the four main research questions.
+Chapter 2 describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT.
+Chapter 3 explains how this model was used to study the apparent stalling of the spread of the SCTLD between the Marquesas and the DRTO.
+Chapter 4 investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014.
+Finally, chapter describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
 
-\end{document}
+\section*{Supporting publications}
+
+\begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{0.23in}%
+    \setlength{\listparindent}{-0.23in}%
+    \setlength{\itemindent}{-0.23in}%
+    \setlength{\parsep}{\parskip}%
+    }%
+
+    \item \textbf{Dobbelaere, T.}, Muller, E. M., Gramer, L. J., Holstein, D. M., \& Hanert, E.
+    (2020). Coupled epidemio-hydrodynamic modeling to understand the spread
+    of a deadly coral disease in Florida. \textit{Frontiers in Marine Science}, 7, 1016. \href{https://www.frontiersin.org/articles/10.3389/fmars.2020.591881/full}{doi: 10.3389/fmars.2020.591881}
+
+    \item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., \& Hanert, E.
+(2022).
+Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes.
+\textit{Ocean Modelling}, 101947.
+\href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.1016/j.ocemod.2022.101947}
+
+    \item \textbf{Dobbelaere, T.}, Holstein, D.
+M., Muller, E.
+M., Gramer, L.
+J., McEachron L., Williams, S.
+D., \& Hanert, E.
+(2022).
+Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas.
+\textit{Frontiers in Marine Science}, 9.
+\href{https://www.frontiersin.org/article/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.
+
+    \item Purkis, S.
+J., Oehlert, A.
+M., \textbf{Dobbelaere, T.}, Hanert, E., \& Harris, P.
+(2020), Always a White Christmas in the Bahamas - Ocean Chemistry and Hydrodynamics Focus Winter Mud Production on Great Bahama Bank, \textit{Sedimentology}, in revision.
+
+    \item Alaerts, L., \textbf{Dobbelaere, T.}, Gravinese, P. M., \& Hanert, E. (2022), Climate change will fragment Florida stone crab communities. \textit{Frontiers in Marine Science}, submitted.
+
+    \item Hanert, E., Aboobacker, V.M., Veerasingam, S, \textbf{Dobbelaere, T.}, Vallaeys, V. \&
+    Vethamony, P.
+(2021) A multiscale ocean modelling system for the central Arabian Gulf: From basin-scale circulation patterns to flow-structure interactions.
+\textit{Estuarine, Coastal and Shelf Science}, submitted.
+
+    \item Anselain, T., \textbf{Dobbelaere, T.}, Heggy, E.
+\& Hanert, E.
+(2022) Qatar oil spill vulnerability threatens both its national water security and the global gas market.
+\textit{Nature Communications}, in preparation.
+
+    \item Holstein, D.
+M., \textbf{Dobbelaere, T.}, Gramer, L.
+J., McEachron, L., Muller E.
+M., Williams, S.
+D.
+\&S Hanert, E.
+(2022) Coral restoration in the age of emergent disease: Balancing the benefits and risks of population connectivity.
+\textit{Conservation Letters}.
+in preparation.
+
+    \item Lopez-Gamundi, C., \textbf{Dobbelaere, T.}, Hanert, E., Harris, P.,Eberli G., and Purkis, S.
+(2022) Simulating Sedimentation on the Great Bahama Bank – Sources, Sinks, and Storm.
+\textit{Sedimentology}.
+submitted.
+
+\end{list}

### 2022-02-10 chapters/intro/main.tex
intro added to main manuscript

@@
 
-Coral reefs are among the most productive and biologically diverse ecosystems on Earth \citep{connell1978diversity, moberg1999ecological} and supply food, income and coastal protection to vast numbers of people.
+\todo{Talk about acidifcation}
+
+Coral reefs are among the most productive and biologically diverse ecosystems on Earth \citep{connell1978diversity, moberg1999ecological}.
+They supply food, income and coastal protection to vast numbers of people (Fig.
+\ref{intro:corals}).
 Although they account for less than 0.5\% of the ocean floor \citep{spalding1997new}, almost a third of the known marine biodiversity is found on coral reefs \citep{moberg1999ecological}.
 Reef-related fisheries constitute around 10\% of the fish consumed by humans \citep{smith1978coral} and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
-Through calcification, corals create the structure of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
+Through calcification, corals are the backbone of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
 This complex reef framework  dissipates waves, protecting shorelines from currents and storms and preventing land loss due to erosion \citep{ferrario2014effectiveness,elliff2017coral}.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=.8\textwidth]{figures/anthropocene.png}
+    \caption{Goods and services of coral reefs and their linkage with human activities. Source: \cite{hughes2017coral}}
+    \label{intro:corals}
+\end{figure}
+
 The calcifying process of reef building corals heavily depends on their symbiosis with the microalgae zooxanthellae.
-These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the coral host with most of the energy needed to meet its metabolic demands \citep{muscatine1977reef}.
+These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the corals with most of the energy needed to meet their metabolic demands \citep{muscatine1977reef}.
 However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral}.
@@
 Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2017coral}: it is estimated  that 30\% have already been severely damaged, and close to 60\% may be lost by 2030 \citep{hughes2003climate}.
-This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} is caused by both global and local anthropogenic stressors.
+This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} (Fig.
+\ref{intro:cover}) is caused by both global and local anthropogenic stressors.
 The main global stressor is climate change induced ocean warming, that cause recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
-Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
-Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
+Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral disease outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
+Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
 
-Due to the combined damages from of all these stressors, it is now impossible to return reefs to their past configurations \citep{hughes2017coral}.
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{figures/cover.jpeg}
+    \caption{Evolution of the absolute coral cover in the Caribbean. Source: \cite{gardner2003long}}
+    \label{intro:cover}
+\end{figure}
+
+Due to the combined damages from of all these stressors, it is almost impossible to return reefs to their past configurations \citep{hughes2017coral}.
 Instead, the challenge is now to maintain the biological functions of coral reefs through quickly changing environmental conditions.
@@
 Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both.
-In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton.
-The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution of reports consistent with the case definition suggests that SCTLD is the largest coral disease outbreak on record in Florida.
+In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig.
+\ref{intro:sctld}).
+The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution suggest that SCTLD might be the largest coral disease outbreak on record in Florida.
 Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
 
-\todo{Add figure with pictures of SCTLD.}
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/sctld.jpg}
+    \caption{Brain coral affected by SCTLD in Looe Key, Florida.
+The disease is leaving a white band of recently dead skeleton in contrast to the healthy, yellow/brown tissue.
+Source: Florida Fish and Wildlife Research Institute}
+    \label{intro:sctld}
+\end{figure}
 
@@
     \includegraphics[width=\textwidth]{figures/hurricane_strikes.jpg}
-    \caption{}
+    \caption{Total number of major hurricanes strikes in Florida between 1900-2010.
+Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
     \label{inro:landfall}
@@
 This thesis focuses on two main threats influencing the future and survival of FCR: the current outbreak of SCTLD and the hurricanes.
-More specifically, our goal is to develop modeling tools to answer four main questions: (1) how does the SCTLD spread from reef to reef ?
-(2) how can we explain the timing of this spread ?
-(3)
+More specifically, our goal is to develop modeling tools to answer four main questions.
+First, as SCTLD was expected to propagate through waterborne transmission, we developed a coupled epidemio-hydrodynamic model to reproduce the observed spread of the outbreak through the FRT.
+Such a model provides novel insight into the management of the epidemic, the identification of its causative agent, and its mode of transmission.
+Second, we used this model to explain the timing of the propagation of SCTLD within the FRT.
+More specifically, we investigated whether the modeled hydrodynamics were explanatory of the apparent stalling of the outbreak between the Marquesas and the DRTO in 2020.
+Third, as sediments can act as vector for the disease, we evaluated the impact of the PMDDP on the reported onset of SCTLD in 2014 using a sediment transport model.
+We assessed whether sediments produced or resuspended during dredging operations might have been transported to reefs where disease was reported and evaluated whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
+Finally, we developed a coupled wave-current model to evaluate the impact of hurricanes on wave-induced transport processes.
+This model was applied to Irma, a major hurricane that made landfall in the Florida Keys in 2017, during the reproduction period of several coral species.
 
-Answering these questions require the development
+Answering these four questions required the use of an ocean model able to capture both the large-scale ocean features influencing circulation along the FRT as well as small-scale flow features down to the reef scale.
+A description of the main hydrodynamic features of the FRT as well as an overview of the modeling tools used in this thesis are given in the next section.
 
 \section*{Modeling the hydrodynamics of the FRT}
+\todo{Description of circulation in the FRT and SLIM ?}
 
+\section*{Outline}
+This thesis is mainly a collection of papers describing our answers to the four main research questions.
+Chapter 2 describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT.
+Chapter 3 explains how this model was used to study the apparent stalling of the spread of the SCTLD between the Marquesas and the DRTO.
+Chapter 4 investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014.
+Finally, chapter describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
+
+\section*{Supporting publications}
+
+\begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{0.23in}%
+    \setlength{\listparindent}{-0.23in}%
+    \setlength{\itemindent}{-0.23in}%
+    \setlength{\parsep}{\parskip}%
+    }%
+
+    \item \textbf{Dobbelaere, T.}, Muller, E. M., Gramer, L. J., Holstein, D. M., \& Hanert, E.
+    (2020). Coupled epidemio-hydrodynamic modeling to understand the spread
+    of a deadly coral disease in Florida. \textit{Frontiers in Marine Science}, 7, 1016. \href{https://www.frontiersin.org/articles/10.3389/fmars.2020.591881/full}{doi: 10.3389/fmars.2020.591881}
+
+    \item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., \& Hanert, E.
+(2022).
+Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes.
+\textit{Ocean Modelling}, 101947.
+\href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.1016/j.ocemod.2022.101947}
+
+    \item \textbf{Dobbelaere, T.}, Holstein, D.
+M., Muller, E.
+M., Gramer, L.
+J., McEachron L., Williams, S.
+D., \& Hanert, E.
+(2022).
+Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas.
+\textit{Frontiers in Marine Science}, 9.
+\href{https://www.frontiersin.org/article/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.
+
+    \item Purkis, S.
+J., Oehlert, A.
+M., \textbf{Dobbelaere, T.}, Hanert, E., \& Harris, P.
+(2020), Always a White Christmas in the Bahamas - Ocean Chemistry and Hydrodynamics Focus Winter Mud Production on Great Bahama Bank, \textit{Sedimentology}, in revision.
+
+    \item Alaerts, L., \textbf{Dobbelaere, T.}, Gravinese, P. M., \& Hanert, E. (2022), Climate change will fragment Florida stone crab communities. \textit{Frontiers in Marine Science}, submitted.
+
+    \item Hanert, E., Aboobacker, V.M., Veerasingam, S, \textbf{Dobbelaere, T.}, Vallaeys, V. \&
+    Vethamony, P.
+(2021) A multiscale ocean modelling system for the central Arabian Gulf: From basin-scale circulation patterns to flow-structure interactions.
+\textit{Estuarine, Coastal and Shelf Science}, submitted.
+
+    \item Anselain, T., \textbf{Dobbelaere, T.}, Heggy, E.
+\& Hanert, E.
+(2022) Qatar oil spill vulnerability threatens both its national water security and the global gas market.
+\textit{Nature Communications}, in preparation.
+
+    \item Holstein, D.
+M., \textbf{Dobbelaere, T.}, Gramer, L.
+J., McEachron, L., Muller E.
+M., Williams, S.
+D.
+\&S Hanert, E.
+(2022) Coral restoration in the age of emergent disease: Balancing the benefits and risks of population connectivity.
+\textit{Conservation Letters}.
+in preparation.
+
+    \item Lopez-Gamundi, C., \textbf{Dobbelaere, T.}, Hanert, E., Harris, P.,Eberli G., and Purkis, S.
+(2022) Simulating Sedimentation on the Great Bahama Bank – Sources, Sinks, and Storm.
+\textit{Sedimentology}.
+submitted.
+
+\end{list}
+
+\newpage 
 \bibliographystyle{apalike} 

### 2022-02-10 chapters/drto/article.tex
last chapter added to manuscript
appendices added to manuscript

@@
     \centering
-    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig4.jpg}
+    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig4.png}
     \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020.
@@
 This study therefore brings additional insight on the physical and environmental connectivity in the southwestern part of FCR.
+
+\begin{subappendices}
+	\section{Validation of the model}
+
+	SLIM outputs were validated against field measurements from tide gauges and ADCP in the Florida Keys and on the shelf (Fig.
+\ref{fig:a1}).
+Here, we show the comparison between observed and simulated sea surface elevation and current speed for the month of November 2020.
+The model accurately reproduces the observed sea surface elevation at all tide gauges (Fig.
+\ref{fig:a2}) with an RMSE of less than 8 cm.
+The current velocity on the shelf is also  well reproduced by the model, despite a slight underestimation of the southward component of the observed currents at mooring C10 and C22, located on the 25 m and 70 m isobaths respectively (Fig.
+\ref{fig:a3}).
+Using the vector correlation analysis of \citep{kundu1976ekman}, we obtained a correlation factor of 0.9 and a veering angle of 10$^\circ$ between the modeled currents and the observations at station C10.
+A correlation factor of 0.75 and a veering angle of 8° were obtained at mooring C22.
+Such values are similar to the performances of previous modeling studies on the West Florida Shelf \cite{liu2020impacts}.
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=.8\textwidth]{chapters/drto/figures/a1.JPEG}
+		\caption{Mesh of the computational domain with the location of the stations used for the validation of the model outputs. Land is shown in dark gray.}
+		\label{fig:a1}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=\textwidth]{chapters/drto/figures/a2.png}
+		\caption{Comparison of SLIM sea surface elevation against tide gauge elevation in November 2020. The tidal signal is well reproduced at all stations with an RMSE of less than 8 cm.}
+		\label{fig:a2}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=\textwidth]{chapters/drto/figures/a3.png}
+		\caption{Comparison of SLIM currents with depth-averaged ADCP measurements at mooring C10 and C22 in November 2020.
+Despite a slight underestimation of their southward component, currents and their oscillations are well reproduced by the model.}
+		\label{fig:a3}
+	\end{figure}
+
+	SLIM relies on its coupling with HYCOM GoM to accurately reproduce the Loop Current (LC)-Florida Current (FC) system.
+HYCOM’s surface fields were therefore compared with sea surface temperature data from the Group for High Resolution Sea Surface Temperature (GHRSST; \url{https://www.ghrsst.org/}).
+HYCOM reproduces well the penetration of the LC in the Gulf of Mexico (Fig.
+\ref{fig:a4}), which is key to simulating the eddy activity linked to the meandering of the FC near the DRTO.
+
+		\begin{figure}
+		\centering
+		\includegraphics[width=\textwidth]{chapters/drto/figures/a4.JPEG}
+		\caption{Comparison of (\textbf{A}) satellite-derived sea surface temperature from the Group for High Resolution Sea Surface Temperature (GHRSST) against (\textbf{B}) sea surface temperature, (\textbf{C}) sea surface height (SSH) and (\textbf{D}) surface current magnitude modeled by HYCOM.
+All panels are snapshots of November 15, 2019 at 0000 UTC.
+The model reproduces well the penetration of the LC in the Gulf of Mexico.}
+		\label{fig:a4}
+	\end{figure}
+
+\end{subappendices}

### 2022-02-10 chapters/conclusions/conclusions.tex
started conclusion

\section{Conclusions}
The previous chapters of this thesis have presented an effort to build a modeling framework to better understand the propagation and onset of the SCTLD, as well as the impacts of major hurricanes on the transport processes over coral reefs. This model had to be able to accurately capture ocean circulation at different scales to study transport processes in the FRT. First, the large-scale Loop Current-Florida Current system had to be accurately reproduced to correctly capture its impact on eddy formation near the DRTO and the Florida Keys. Second, the model had do resolve ocean circulation at the reef-scale to capture recirculation and acceleration of currents between reefs and islands. This was performed by coupling the multi-scale ocean model SLIM to a connectivity-based epidemiological model (chapter \ref{chap:sctld}), a sediment transport model (chapter \ref{chap:onset}) and a spectral wave model (chapter \ref{chap:irma}).

The developed coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by modeling the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents. It confirmed the result of previous ex situ studies that showed evidence of waterborne transmission of the SCTLD with an averaged transmission time of the order of 10 days. Furthermore, by using coral resistance to the disease as a parameter, our model results suggested that, on average, corals had pretty low resistance against the disease. This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived based field observations on large scale systems. As such, models informed and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as a barrier reef.

This epidemiological model was then used to build networks of potential exchanges of disease agents between reefs during a period of three years. The analysis of these networks showed that exchanges from the western end of the DRTO were interrupted during most of 2020. This interruption was likely linked to eddy activity near the DRTO and was consistent with the apparent stalling of the outbreak in the region in 2020. This confirms the role played the Loop Current/Florida Current system in the control of the connectivity in the Florida Keys and the DRTO. Furthermore, location predicted to be more vulnerable to SCTLD in the disease networks were consistent with the observed geographic distribution of disease coral in the DRTO. This further highlights that modeled hydrodynamics are highly explanatory of the propagation of the disease.

The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that a high-resolution biophysical is a valuable tool to understand the spread of SCTLD in Florida. The model successfully reproduced the observed propagation of the disease in 2018-2019 and allowed to link the observed stalling of SCTLD to hydodynamic features.

\section{Perspectives for future works}

\subsection*{Improvement of the epidemiological model}

\subsection*{Better understanding the protective role of corals against storm waves}

### 2022-02-11 chapters/conclusions/conclusions.tex
conclusion v1

@@
 Second, the model had do resolve ocean circulation at the reef-scale to capture recirculation and acceleration of currents between reefs and islands.
-This was performed by coupling the multi-scale ocean model SLIM to a connectivity-based epidemiological model (chapter \ref{chap:sctld}), a sediment transport model (chapter \ref{chap:onset}) and a spectral wave model (chapter \ref{chap:irma}).
+This was performed by coupling the multi-scale ocean model SLIM to a connectivity-based epidemiological model (chapter \ref{chap:sctld}), a sediment transport model (chapter \ref{chap:onset}) and a spectral wave model (chapter~\ref{chap:irma}).
 
@@
 
-The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that a high-resolution biophysical is a valuable tool to understand the spread of SCTLD in Florida.
-The model successfully reproduced the observed propagation of the disease in 2018-2019 and allowed to link the observed stalling of SCTLD to hydodynamic features.
+The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that the developed high-resolution biophysical is a valuable tool to understand the spread of SCTLD in Florida.
+The model successfully reproduced the observed propagation of the disease in 2018-2019 and allowed to link the observed stalling of SCTLD to observable hydrodynamic features.
+Furthermore, the model allowed to deduce potential characteristics of the disease agent and its vector.
+This highlights the potential of  model to understand the driving mechanics of ongoing outbreaks.
+Furthermore, indicators such as the vulnerablity and source indicators developed in chaper \ref{chap:drto} could be useful to predict and/or mitigate the spread of SCTLD in other territories of the Caribbean.
+
+Using a sediment transport model, we investigated the impact of the PMDDP on the onset of the SCTLD outbreak in 2014.
+Our results suggest that the dredging operations had close to no direct impact at Virginia Key monitoring site, identified as the first affected reef.
+However, reefs affected earlier could have been reached by chopped rock particles produced at the dredged channel.
+Moreover, our epidemiological model suggested that disease propagation from these reefs to Virginia Key was possible.
+Sediments released by the PDMMP might therefore have impacted the initiation of one of the worst coral disease outbreaks to date in the Caribbean.
+These results could have repercussions on the planning of future major dredging projects in the area.
+FCR is routinely suffering from important anthropogenic stresses caused by the intense activity of the area of Miami.
+Hence, coastal developments harm coral reefs in two ways: first, by increasing water turbidity and sedimentation during the dredging, and then by causing an increase in pollution and shipping activities.
+Such adverse impacts should be taken into account when planning dredging operations.
+The results of chapter \ref{chap:onset} suggest that the potential to initiate a coral disease outbreak should now also be taken into account when considering future dredging project.
+
+Finally, we developed a couple wave-current model to study the impact of Hurricane Irma on transport processes in the Florida Keys.
+This tropical cyclone was selected because it hit coral reefs during their reproduction period.
+This coupled modeled correctly reproduced the observed wave and hydrodynamics during the passage of the hurricane.
+Furthermore, model results suggested that an import dissipation occurred over coral reefs through depth-induced wave-breaking and bottom friction.
+This important dissipation produced a large wave radiation stress gradients, which caused differences in modeled velocity of up to 1 m/s between the coupled and uncoupled model runs.
+However, wave-induced transport was dominated by the Stokes drift, which yielded an impact four times larger than the radiation stress gradient.
+Results of chapter \ref{chap:irma} strongly advocates for the use of coupled wave-currents to accurately model the dispersal of coral larvae under storm conditions.
+Although we built the tools to study the impact of hurricane on the exchanges of larvae between reefs by developping the coupled SLIM+SWAN, we did not evaluate the impact of Irma on reef connectivity by lack of time.
+
+In conclusion, we answered the four research questions defined in chapter \ref{chap:intro}:
+
+\begin{list}{}{%
+    % \setlength{\listparindent}{-0.23in}%
+    \setlength{\leftmargin}{0in}%
+    % \setlength{\itemindent}{-0.23in} 
+    }
+    \item \textbf{1. Can we explain the observed spread of SCTLD throughout Florida using an epidemio-hydrodynamic model ?}
+    \begin{list}{}{\setlength{\topsep}{0pt}}
+        \item Yes, by modeling the transport of disease agents as neutrally buoyant material within the water column, and by calibrating a connectivity-based epidemiological model with field data, we managed to reproduce the observed spread of SCTLD through the FRT.
+    \end{list}
+    \item \textbf{2. What caused the apparent stalling of the propagation of SCTLD ?}
+    \begin{list}{}{\setlength{\topsep}{0pt}}
+        \item Our results suggest that there was an interruption of hydrodynamic-driven exchanges from the Florida Keys to the DRTO during most of 2020, which prevented the propagation of disease agents westward.
+This interruption of the disease connectivity pathways to the DRTO was likely related to the local eddy activity generated by the meandering of the Florida Current.
+    \end{list}
+    \item \textbf{3. What was the impact of the PMDDP on the onset of the oubreak of SCTLD~?}
+    \begin{list}{}{\setlength{\topsep}{0pt}}
+        \item Modeling of sediment transport suggests that the chopped rock particles produced by the dredging activities reached reefs where signs of disease were reported during the first half of 2014.
+Furthermore, the modeled disease connectivity suggested that disease propagation from these sites to Virginia Key was possible prior to September 2014.
+The sediments released by the PMDDP might thus have influenced the observed onset of the SCTLD outreak at this site.
+    \end{list}
+    \item  \textbf{4. What is the effect of hurricane-induced wave-current interactions on transport processes and should they be taken into account when modeling the dispersal of coral larvae ?}
+    \begin{list}{}{\setlength{\topsep}{0pt}}
+        \item Results of a coupled wave-current model suggest that wave-induced transport under storm conditions dominated by Stokes drift can yield differences in particle trajectories of up to 20 km.
+Wave-current interactions should thus be taken into to accurately model the dispersal of coral larvae during hurricanes.
+    \end{list}
+\end{list}
 
@@
 
+\subsection*{Combining disease and larval connectivity}
+All chapters dedicated to SCTLD rely on the concept of connectivity.
+In this context, connectivity negatively impacts coral reefs as it yields the propagation of disease agents.
+Larval connectivity, on the other hand, promotes the resilience of the reef network as it allows the recolonization of depopulated reefs after extreme events such as mass bleaching, hurricanes or disease outbreaks.
+Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration.
+Identifying reefs with large out-degree our centrality in the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.
+Furthermore, knowledge about larval allows to build robust and sustainable networks of connected marine protected areas.
+Connectivity appears thus to be a double-edged sword.
+As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.
+This combined information would allow to identify reefs that have high potential for restoration and low vulnerability to the disease.
+A possible measure of the restoration potential is the (weighted) out-degree of reefs in the larval connectivity network, as reefs with strong and numerous outgoing connections in the network are more likely to send many larvae to many connected reefs.
+Vulnerability to disease can be evaluated using the (weighted) in-degree of reefs in the disease connectivity network, as reefs with strong and numerous incoming connections are more likely to receive more disease agents from many connected reefs.
+Therefore, reefs with high restoration potential and low vulnerability would correspond to reefs with large larval(weighted) out-degree and low disease (weighted) in-degree.
+This approach have been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.
+
 \subsection*{Improvement of the epidemiological model}
+A limitation of the epidemiological model developed in chapter \ref{chap:sctld} is that it uses species-averaged parameters and assumes an uniform distribution of coral reefs throughout the FRT.
+The fact that our model correctly reproduced the observed spread of SCTLD for a well-defined range of values of coral resistance to the disease highlights the important impact of this parameter on the propagation of the disease.
+However, coral susceptibility to SCTLD can significantly vary between species and ranges from highly susceptible (\eg~\textit{Dichocoenia stokesii}, \textit{Meandrina meandrites}) to intermediately susceptible (\eg~\textit{Orbicella faveolata}, \textit{Montastrea cavernosa}) and weakly susceptible (\eg~\textit{Acropora Palmata}, \textit{Acropora cervicornis}).
+A first improvement could therefore be to consider classes corals with different susceptibilities and therefore different infection and removal parameter values in the model.
+Such improvement would be technically challenging as it would increase the number of variables of model.
+The development of the model could therefore require the use of more advance graph theory tools such as the multilayer networks \citep{kivela2014multilayer,pilosof2017multilayer}.
+Furthermore, increasing the number of parameters will thus make their calibration more complex.
+Moreover, this new modeling framework would require estimates of the distribution of the different coral species across the FRT.
+Again, compiling this data and deriving realistic estimates would represent a significant challenge.
+Adding species-specific or location-specific parameters in the model would also allow the evaluate the impact of disease mitigating actions.
+Lesion treatment using antibiotics could for instance be taken into account in the model by locally reducing the transmission and/or removal parameters.
+The model parameters could also be made time-dependent as SCTLD progression was reported to slow down during warmer months in the US Virgin Islands \citep{meiling20203d}.
+Another improvement would be the inclusion of additional processes in the model.
+For instance, reinfection by SCTLD is currently not possible as corals are "removed" once they are affected by the disease.
+Given the high mortality of the disease, the "removed" state of the model was assumed equivalent to "dead".
+However, one might want to discriminate between coral recovery and mortality.
+However, again, adding processes implies an increase in the number of parameters to be calibrated.
 
 \subsection*{Better understanding the protective role of corals against storm waves}
+Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipating over reefs.
+The modeled significant wave height experiences a twofold reduction over the barrier reef during the passage of the hurricane (Fig.
+\ref{ccl:swh}).
+This highlights the important protective role of coral reefs against coastal flood hazard.
+Better understanding this role can motivate coral reef restoration efforts.
+For instance, \citep{ferrario2014effectiveness} estimated that coral reefs could provide wave attenuation comparable to artificial defenses such as breakwaters for a lower cost.
+However, modelling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes.
+The smallest scales that model such as SLIM and SWAN can simulate is limited by the intrinsic assumptions in the model formulations, i.e., hydrostatic for SLIM and phase-averaged for SWAN.
+These assumptions are no longer valid as soon as the horizontal scales of motion become comparable to the vertical scales \citep{marshall1997hydrostatic}.
+For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modelling framework.
+Nonhydrostatic models are more expensive to run as they require a much finer spatial resolution (~1 m) and have a more complex physics.
+They can therefore often be used only in rather small areas and over short periods of time \citep{fringer2019future}.
+One promising approach to tackle this issue would be to use machine learning algorithms to locally emulate the dynamics of a complex high-resolution model and hence speed up the computational process \citep{kasim2021building}.
+Machine learning is based on the principle of extracting information from data, which can either be observational data or high-resolution model outputs, and deriving relationships between datasets without the need for physics-based models.
+This procedure has recently been applied to ocean modelling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{chapters/conclusions/figures/swh_figure.png}
+    \caption{Snapshot of the significant wave height modeled by the coupled SLIM+SWAN model during Hurricane Irma, on Sept 10, 2017 at 1200 UTC.
+The path of the hurricane is shown by a black line with squares.
+Land is shown in dark gray and reefs in light gray.
+Significant wave attenuation is visible between the fringing coral reefs and the islands.}
+    \label{ccl:swh}
+\end{figure}

### 2022-02-11 chapters/intro/intro.tex
started conclusion
conclusion v1

@@
 
-\section{Identifying the threats to coral survival in \\Florida}
+\section{Identifying the threats to coral survival in Florida}
 
@@
 This thesis focuses on two main threats influencing the future and survival of FCR: the current outbreak of SCTLD and the hurricanes.
-More specifically, our goal is to develop modeling tools to answer four main questions.
+These threats were studied by fulfilling four main objectives.
 First, as SCTLD was expected to propagate through waterborne transmission, we developed a coupled epidemio-hydrodynamic model to reproduce the observed spread of the outbreak through the FRT.
@@
 
+These objectives can be summarized into four main questions:
+\begin{enumerate}
+    \item Can we explain the observed spread of SCTLD throughout Florida using an epidemio-hydrodynamic model ?
+    \item What caused the apparent stalling of the propagation of SCTLD ? 
+    \item What was the impact of the PMDDP on the onset of the oubreak of SCTLD~?
+    \item  What is the effect of hurricane-induced wave-current interactions on transport processes and should they be taken into account when modeling the dispersal of coral larvae ?
+\end{enumerate}     
 Answering these four questions required the use of an ocean model able to capture both the large-scale ocean features influencing circulation along the FRT as well as small-scale flow features down to the reef scale.
@@
 \end{list}
+\newpage 
+\section{List of acronyms and abbreviations}
+
+\begingroup
+    \renewcommand{\arraystretch}{1.3}
+    \begin{tabular}{ll}
+        \textbf{DRTO}  & Dry Tortugas \\
+        \textbf{ECMWF} &  \\
+        \textbf{FCR}   & Florida's Coral Reef \\
+        \textbf{FRT}   & Florida Reef Tract  \\
+        \textbf{HYCOM} & \\
+        \textbf{NOAA}  & National Oceanographic and Atmospheric Administration \\
+        \textbf{PoM}   & Port of Miami \\
+        \textbf{PMDDP} & PortMiami Deep Dredge Project \\
+        \textbf{RS}    & Radiation stress \\
+        \textbf{SCTLD} & Stony coral tissue loss disease \\
+        \textbf{SLIM}  & \\
+        \textbf{SWAN}  &
+    \end{tabular}
+\endgroup

### 2022-02-11 chapters/onset/article.tex
appendices added to manuscript
conclusion v1

@@
 
-The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022connecting}.
+The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys2020fine,dobbelaere2020coupled,dobbelaere2022connecting}.
 SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features.
@@
 The fifth site was monitoring station HBSC1-CP of the monitoring of the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
-\ref{fig:onset_mesh}A, see pictures in Appendix).
+\ref{fig:onset_mesh}A, see Appendic\ref{onset:appendice} ).
 The impact of dredging was assessed by counting the number of sediment particles originating from each dredging that were transported within 500 m of all five site.
@@
 Our work brings new insight on the impact of a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
+
+\begin{subappendices}
+	\section{Reported disease signs at site HBSC1-CP in May 2014}\label{onset:appendice}
+	\includegraphics[width=\textwidth]{chapters/onset/figures/hbsc1_cp.png}
+\end{subappendices}

### 2022-02-11 chapters/sctld/article.tex
conclusion v1

@@
 Unstructured-mesh ocean models offer a potential solution to this resolution issue by locally increasing the model resolution close to reefs and islands \citep{lambrechts2008multi, thomas2014numerical, thomas2015connectivity}, in order to focus the computational resources where they are most needed.
-High resolution bio-physical dispersal models can be used to build the potential connectivity between reefs and therefore approximate exchanges between colonies in the complex topography of the coral reef systems \citep{frys20}.
+High resolution bio-physical dispersal models can be used to build the potential connectivity between reefs and therefore approximate exchanges between colonies in the complex topography of the coral reef systems \citep{frys2020fine}.
 
@@
 \slim has already been successfully applied in complex coastal systems such as the Great Barrier Reef \citep{lambrechts2008multi, thomas2014numerical} and is well suited to shallow-water flows.
-Details of the model formulation and validation are provided in \cite{frys20}.
+Details of the model formulation and validation are provided in \cite{frys2020fine}.
 
@@
 The results shown in Fig.
-\ref{fig:results} were obtained by removing the large reef located North to Vaca key, denoted Vaca reef in \cite{frys20}, from our reef polygons.
+\ref{fig:results} were obtained by removing the large reef located North to Vaca key, denoted Vaca reef in \cite{frys2020fine}, from our reef polygons.
 Preliminary simulations showed that this reef had close to no impact on the modeled spread of the disease to the rest of the FRT, as it sent very few disease agents to southerly and easterly neighboring reefs.
@@
 Although such parametric estimation of surface and bottom currents adds additional uncertainty to those simulations, using a 2D model as the basis for the study allowed for reef-scale resolution throughout the whole FRT with available computational resources.
-Such high-resolution allowed the simulation to explicitly represent recirculation eddies around islands and reefs, that significantly impact the weighted connectivity length as well as the local retention of pathogenic material on the reefs \citep{frys20}.
+Such high-resolution allowed the simulation to explicitly represent recirculation eddies around islands and reefs, that significantly impact the weighted connectivity length as well as the local retention of pathogenic material on the reefs \citep{frys2020fine}.
 

### 2022-02-11 chapters/conclusions/conclusions.tex
added figures to intro paragraph about hydro

@@
 For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modelling framework.
-Nonhydrostatic models are more expensive to run as they require a much finer spatial resolution (~1 m) and have a more complex physics.
+Nonhydrostatic models are more expensive to run as they require a much finer spatial resolution ($\sim$ 1 m) and have a more complex physics.
 They can therefore often be used only in rather small areas and over short periods of time \citep{fringer2019future}.
@@
 This procedure has recently been applied to ocean modelling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}.
+Additionally, this multi-physics modeling frameworks could be uses to better capture the motion of coral larvae over reef.
+This would improve the modeled coral connectivity, as the the initial motion of larvae over reefs then drives larval transport following mass spawning events and initiate coral connectivity.
 

### 2022-02-11 chapters/intro/intro.tex
added figures to intro paragraph about hydro

@@
 The Florida Reef Tract (FRT) spans over approximately 577 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St.
-Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf}.
+Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf} (Fig.
+\ref{intro:frt}).
 These reefs harbor a rich Caribbean fauna, including more than 40 species of stony corals \citep{banks2008reef,jackson2014status}.
@@
 The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is fairly low (below 3\%)\citep{banks2008reef,walton2018impacts}.
-The southern half-section is a chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO.
+The southern half-section is a chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO on the the boundary of the West Florida Shelf (WFS).
 The Upper Keys are remnants of ancient coral reefs and the Lower Keys are sand bars \citep{hoffmeister1968geology}.
@@
 
-\todo{Add figure with map of the FRT}
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{chapters/intro/figures/frt.png}
+    \caption{Map of Florida Reef Tract with major features and divisions. Source: \cite{kupfner2019untapped}}
+    \label{intro:frt}
+\end{figure}
 
@@
 
-\todo{Add map of the propagation of SCTLD}
-
 Coral disease outbreaks are frequent in the Caribbean, which is considered a disease "hot spot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral diseases in the area \citep{green2000significance, harvell2007coral}.
@@
 The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}.
-First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects}.
+First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects} (Fig.
+\ref{intro:propagation}).
 The outbreak affects at least 24 species of scleractinian coral and often results in whole colony mortality \citep{precht2016unprecedented, walton2018impacts}.
@@
 Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{chapters/intro/figures/fig_sctld.png}
+    \caption{Propagation of the SCTLD troughout the FRT between 2014 and 2021}
+    \label{intro:propagation}
+\end{figure}
 
@@
 Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,aronson2001white}.
-They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects}.
+They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects, tweel2014contribution}.
 Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}.
@@
 
-\section*{Modeling the hydrodynamics of the FRT}
-\todo{Description of circulation in the FRT and SLIM ?}
+\section{Modeling the hydrodynamics of the FRT}
+The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean.
+The large-scale circulation in this region is dominated by the interplay of the Florida Current (FC) and the Loop Current (LC).
+The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC).
+The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream.
+The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}.
+The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT.
+These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability.
+The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}.
+These eddies significantly impact the ecology of the southwest Florida Shelf as they have been identified as a mechanism causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}.
+Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}.
+These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}.
+On the northernmost section of the FRT however, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}.
+The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}
+
+\todo{Figure LC, GoM Florida Straits, etc.}
+
+The aforementioned large-scale ocean circulation significantly impact transport processes in the FRT.
+The modeling tools developed in this thesis should therefore be able to capture them accurately.
+Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation.
+For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}.
+Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}.
+Unstructured-mesh models offer a well-suited solution to tackle the multi-scale nature of flow features in the FRT.
 
@@
 This thesis is mainly a collection of papers describing our answers to the four main research questions.
-Chapter 2 describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT.
-Chapter 3 explains how this model was used to study the apparent stalling of the spread of the SCTLD between the Marquesas and the DRTO.
-Chapter 4 investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014.
-Finally, chapter describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
+Chapter \ref{chap:sctld} describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT.
+Chapter \ref{chap:drto} explains how this model was used to study the apparent stalling of the spread of the SCTLD between the Marquesas and the DRTO.
+Chapter \ref{chap:onset} investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014.
+Finally, chapter \ref{chap:irma} describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
 
@@
 D.
-\&S Hanert, E.
+\& Hanert, E.
 (2022) Coral restoration in the age of emergent disease: Balancing the benefits and risks of population connectivity.
@@
 \begingroup
+    \setlength{\tabcolsep}{20pt}
     \renewcommand{\arraystretch}{1.3}
@@
         \textbf{DRTO}  & Dry Tortugas \\
-        \textbf{ECMWF} &  \\
+        \textbf{ECMWF} & European Centre for Medium-Range Weather Forecast \\
+        \textbf{FC}    & Florida Current \\
         \textbf{FCR}   & Florida's Coral Reef \\
         \textbf{FRT}   & Florida Reef Tract  \\
-        \textbf{HYCOM} & \\
+        \textbf{GoM}   & Gulf of Mexico \\
+        \textbf{HYCOM} & HYbrid Coordinate Ocean Model \\
+        \textbf{LC}    & Loop Current \\
         \textbf{NOAA}  & National Oceanographic and Atmospheric Administration \\
@@
         \textbf{SCTLD} & Stony coral tissue loss disease \\
-        \textbf{SLIM}  & \\
-        \textbf{SWAN}  &
+        \textbf{SLIM}  & Second-generation Louvain-la-Neuve Ice-ocean Model\\
+        \textbf{SWAN}  & Simulating WAves Nearshore \\
+        \textbf{WFS}   & West Florida Shelf
     \end{tabular}

### 2022-02-11 chapters/intro/intro.tex
finished hydro part in intro

@@
 
-\section{Modeling the hydrodynamics of the FRT}
+\section{Modeling hydrodynamics and transport processes in the FRT}
 The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean.
@@
 Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}.
-Unstructured-mesh models offer a well-suited solution to tackle the multi-scale nature of flow features in the FRT.
+Unstructured-mesh models offer a well-suited solution to cature the multi-scale nature of flow features in the FRT in a computationally tractable way.
+They allow to locally increase the model resolution close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
+Nested grid models are another possible approach \citep{warner2010development} but they lack the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
+
+In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}).
+This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}.
+Moreover, by coupling SLIM with the HYbrid Coordinate Ocean Model (HYCOM-GOMl0.04, \citealp{chassignet2007hycom}), SLIM was able to indirectly represent the baroclinic dynamics of the FC.
+Using this coupling, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs.
+In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents.
+To answer our different research questions, we adapted this model to evaluate the transport of the causative agent of the SCTLD and sediments particles.
+
+However, at the beginning this thesis, SLIM was not able to represent wave-current interactions under storm conditions.
+Heavy winds generate large wind-waves that disturb ocean conditions and interact non-linearly with currents \citep{liu2020impacts,wu2011fvcom}.
+The wave-generated currents cause water level variations near shorelines and wave breaking points which, in turn, affect the motion and evolution of the waves \citep{longuet1970longshore, sikiric2013coupling}.
+Capturing wave-current interactions hence requires the calculation of the full directional wave spectrum in order to correctly reproduce the dynamics of wind-driven surface waves.
+This is usually achieved by spectral wave models, which describe the evolution of the wave energy spectrum.
+An objective of this thesis was therefore the coupling of SLIM with the spectral wave model Simulating WAve Nearshore (SWAN, \citealp{booij1999third}) SWAN was specifically developed for coastal applications and relies on numerical schemes adapted to small-scale, shallow water regions.
 
@@
 
+\newpage
 \section*{Supporting publications}

### 2022-02-11 include/title.tex
slim+swan paper added
chapter on drto added to thesis
draft onset added to manuscript
finished hydro part in intro

@@
 \centering
-\LARGE\bfseries How modeling tools can inform coral reef management in the Florida Reef Tract
+\LARGE\bfseries Better understanding transport processes to support Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
 \end{minipage}
@@
 in partial fulfillment of the requirements for the degree of\\ [2.ex]
-\Large Docteur en Sciences de l'ingénieur
+\Large Doctor in Engineering Science
 
@@
 \footnotesize{
-\hspace{-1.2cm}
+\hspace{-1cm}
 \begin{tabular}{ll}
@@
 Dr. Jonathan Lambrechts            & Université catholique de Louvain\\
-???                                & ??? \\
-???                                & ??? \\ 
+Prof. Claire B. Paris-Limouzy      & University of Miami \\
+Dr. Christopher Thomas             & Université catholique de Louvain \footnote{Present address: Risk Management Solutions, London, UK}
 \end{tabular}

### 2022-02-12 chapters/conclusions/conclusions.tex
commentaires du patron - part I

@@
 \section{Conclusions}
-The previous chapters of this thesis have presented an effort to build a modeling framework to better understand the propagation and onset of the SCTLD, as well as the impacts of major hurricanes on the transport processes over coral reefs.
-This model had to be able to accurately capture ocean circulation at different scales to study transport processes in the FRT.
+In this thesis, we have developed a framework to better understand the propagation and onset of the SCTLD, as well as the impact of major hurricanes onctransport processes over coral reefs.
+This model had to be able to accurately capture ocean circulation at different scales to study transport processes in the FCR.
 First, the large-scale Loop Current-Florida Current system had to be accurately reproduced to correctly capture its impact on eddy formation near the DRTO and the Florida Keys.
-Second, the model had do resolve ocean circulation at the reef-scale to capture recirculation and acceleration of currents between reefs and islands.
+Second, the model had do reproduce ocean circulation at the reef-scale to capture recirculation and acceleration of currents between reefs and islands.
 This was performed by coupling the multi-scale ocean model SLIM to a connectivity-based epidemiological model (chapter \ref{chap:sctld}), a sediment transport model (chapter \ref{chap:onset}) and a spectral wave model (chapter~\ref{chap:irma}).
 
-The developed coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by modeling the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents.
-It confirmed the result of previous ex situ studies that showed evidence of waterborne transmission of the SCTLD with an averaged transmission time of the order of 10 days.
-Furthermore, by using coral resistance to the disease as a parameter, our model results suggested that, on average, corals had pretty low resistance against the disease.
-This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived based field observations on large scale systems.
-As such, models informed and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as a barrier reef.
+Our coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by modeling the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents.
+It confirmed the result of previous ex-situ studies that showed evidence of waterborne transmission of the SCTLD with an average transmission time of the order of 10 days.
+Furthermore, by using coral resistance to the disease as a parameter, our model results suggested that, on average, corals had a rather low resistance against the disease.
+This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived from field observations on large scale systems.
+As such, models informed and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as the FCR.
 
-This epidemiological model was then used to build networks of potential exchanges of disease agents between reefs during a period of three years.
-The analysis of these networks showed that exchanges from the western end of the DRTO were interrupted during most of 2020.
+This epidemiological model was then used to build networks of potential exchanges of disease agents between $\sim 1.6\times10^4$ sub-reef areas during a period of three years.
+The analysis of these networks showed that exchanges from the Marquesas to the DRTO were interrupted during most of 2020.
 This interruption was likely linked to eddy activity near the DRTO and was consistent with the apparent stalling of the outbreak in the region in 2020.
 This confirms the role played the Loop Current/Florida Current system in the control of the connectivity in the Florida Keys and the DRTO.
-Furthermore, location predicted to be more vulnerable to SCTLD in the disease networks were consistent with the observed geographic distribution of disease coral in the DRTO.
+Furthermore, sub-reefs predicted to be more vulnerable to SCTLD in the disease networks were consistent with the observed geographic distribution of disease coral in the DRTO.
 This further highlights that modeled hydrodynamics are highly explanatory of the propagation of the disease.
 
-The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that the developed high-resolution biophysical is a valuable tool to understand the spread of SCTLD in Florida.
-The model successfully reproduced the observed propagation of the disease in 2018-2019 and allowed to link the observed stalling of SCTLD to observable hydrodynamic features.
-Furthermore, the model allowed to deduce potential characteristics of the disease agent and its vector.
-This highlights the potential of  model to understand the driving mechanics of ongoing outbreaks.
+The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that our high-resolution biophysical is a valuable tool to understand the spread of SCTLD in Florida.
+The model successfully reproduced the observed propagation of the disease in 2018-2019 and allowed us to link the observed stalling of SCTLD to observable hydrodynamic features.
+Furthermore, the model allowed us to deduce potential characteristics of the disease agent and its vector.
+This highlights the potential of  model to understand the driving mechanics of ongoing marine disease outbreaks.
 Furthermore, indicators such as the vulnerablity and source indicators developed in chaper \ref{chap:drto} could be useful to predict and/or mitigate the spread of SCTLD in other territories of the Caribbean.
@@
 Our results suggest that the dredging operations had close to no direct impact at Virginia Key monitoring site, identified as the first affected reef.
-However, reefs affected earlier could have been reached by chopped rock particles produced at the dredged channel.
-Moreover, our epidemiological model suggested that disease propagation from these reefs to Virginia Key was possible.
-Sediments released by the PDMMP might therefore have impacted the initiation of one of the worst coral disease outbreaks to date in the Caribbean.
-These results could have repercussions on the planning of future major dredging projects in the area.
+However, reefs affected earlier could have been infected by chopped rock particles produced by non conventional dredging operations in the channel.
+Moreover, our epidemiological model suggests that disease propagation from these reefs to Virginia Key was possible.
+Sediments released by the PDMMP might therefore have initiated one of the worst coral disease outbreaks to date in the Caribbean.
+These results could have consequences on the planning of future major dredging projects in the area such as the Port Everglade Dredging Project\footnote{https://www.porteverglades.net/construction/harbor-improvements/, last consulted on February 12, 2022}.
 FCR is routinely suffering from important anthropogenic stresses caused by the intense activity of the area of Miami.
@@
 Finally, we developed a couple wave-current model to study the impact of Hurricane Irma on transport processes in the Florida Keys.
-This tropical cyclone was selected because it hit coral reefs during their reproduction period.
-This coupled modeled correctly reproduced the observed wave and hydrodynamics during the passage of the hurricane.
-Furthermore, model results suggested that an import dissipation occurred over coral reefs through depth-induced wave-breaking and bottom friction.
+This coupled model correctly reproduced the observed wave and hydrodynamics during the passage of the hurricane.
+Furthermore, model results suggest that an import dissipation occurred over coral reefs through depth-induced wave-breaking and bottom friction.
 This important dissipation produced a large wave radiation stress gradients, which caused differences in modeled velocity of up to 1 m/s between the coupled and uncoupled model runs.
@@
 Results of chapter \ref{chap:irma} strongly advocates for the use of coupled wave-currents to accurately model the dispersal of coral larvae under storm conditions.
-Although we built the tools to study the impact of hurricane on the exchanges of larvae between reefs by developping the coupled SLIM+SWAN, we did not evaluate the impact of Irma on reef connectivity by lack of time.
+Although we built the tools to study the impact of hurricane on the exchanges of larvae between reefs by developing the coupled SLIM+SWAN, we did not evaluate the impact of Irma on coral connectivity by lack of time.
 
@@
     \begin{list}{}{\setlength{\topsep}{0pt}}
-        \item Yes, by modeling the transport of disease agents as neutrally buoyant material within the water column, and by calibrating a connectivity-based epidemiological model with field data, we managed to reproduce the observed spread of SCTLD through the FRT.
+        \item The observed spread of the SCTLD through the FRT was successfully reproduced by modeling the transport of disease agents as neutrally buoyant material within the water column, and by calibrating a connectivity-based epidemiological model with field data.
+This therefore suggests that the disease spread through waterborne transmission.
     \end{list}
@@
     \begin{list}{}{\setlength{\topsep}{0pt}}
-        \item Our results suggest that there was an interruption of hydrodynamic-driven exchanges from the Florida Keys to the DRTO during most of 2020, which prevented the propagation of disease agents westward.
+        \item Our results suggest that there was an interruption of hydrodynamic-driven exchanges from the Florida Keys and the Marquesas to the DRTO during most of 2020, which prevented the propagation of disease agents westward.
 This interruption of the disease connectivity pathways to the DRTO was likely related to the local eddy activity generated by the meandering of the Florida Current.
@@
     \begin{list}{}{\setlength{\topsep}{0pt}}
-        \item Modeling of sediment transport suggests that the chopped rock particles produced by the dredging activities reached reefs where signs of disease were reported during the first half of 2014.
+        \item Modeling of sediment transport suggests that the chopped rock particles produced by the non conventional dredging activities reached reefs where signs of disease were first reported during the first half of 2014.
 Furthermore, the modeled disease connectivity suggested that disease propagation from these sites to Virginia Key was possible prior to September 2014.
-The sediments released by the PMDDP might thus have influenced the observed onset of the SCTLD outreak at this site.
+The sediments released by the PMDDP might thus have triggered the onset of the SCTLD outreak at this site.
     \end{list}

### 2022-02-12 include/title.tex
commentaires du patron - part I

@@
 \centering
-\LARGE\bfseries Better understanding transport processes to support Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
+\LARGE\bfseries Better understanding transport processes to inform the management of Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
 \end{minipage}

### 2022-02-13 chapters/conclusions/conclusions.tex
figure gom + acidification

@@
 All chapters dedicated to SCTLD rely on the concept of connectivity.
-In this context, connectivity negatively impacts coral reefs as it yields the propagation of disease agents.
-Larval connectivity, on the other hand, promotes the resilience of the reef network as it allows the recolonization of depopulated reefs after extreme events such as mass bleaching, hurricanes or disease outbreaks.
+In this context, connectivity negatively impacts coral reefs as it drives the propagation of disease agents.
+However, connectivity acts as a double-edged sword.
+Larval connectivity promotes the resilience of the reef network as it allows the recolonization of depopulated reefs after extreme events such as mass bleaching, hurricanes or disease outbreaks.
 Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration.
 Identifying reefs with large out-degree our centrality in the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.
-Furthermore, knowledge about larval allows to build robust and sustainable networks of connected marine protected areas.
-Connectivity appears thus to be a double-edged sword.
+Furthermore, knowledge about larval connectivity would allow us to build robust and sustainable networks of connected marine protected areas.
 As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.
-This combined information would allow to identify reefs that have high potential for restoration and low vulnerability to the disease.
+This combined information would allow us to identify reefs that have high potential for restoration and low vulnerability to the disease.
 A possible measure of the restoration potential is the (weighted) out-degree of reefs in the larval connectivity network, as reefs with strong and numerous outgoing connections in the network are more likely to send many larvae to many connected reefs.
@@
 
-\subsection*{Improvement of the epidemiological model}
+\subsection*{Improving the epidemiological model}
 A limitation of the epidemiological model developed in chapter \ref{chap:sctld} is that it uses species-averaged parameters and assumes an uniform distribution of coral reefs throughout the FRT.
@@
 However, coral susceptibility to SCTLD can significantly vary between species and ranges from highly susceptible (\eg~\textit{Dichocoenia stokesii}, \textit{Meandrina meandrites}) to intermediately susceptible (\eg~\textit{Orbicella faveolata}, \textit{Montastrea cavernosa}) and weakly susceptible (\eg~\textit{Acropora Palmata}, \textit{Acropora cervicornis}).
-A first improvement could therefore be to consider classes corals with different susceptibilities and therefore different infection and removal parameter values in the model.
+A first improvement could therefore be to consider classes od corals with different susceptibilities and hence use different infection and removal parameter values in the model.
 Such improvement would be technically challenging as it would increase the number of variables of model.
@@
 Again, compiling this data and deriving realistic estimates would represent a significant challenge.
-Adding species-specific or location-specific parameters in the model would also allow the evaluate the impact of disease mitigating actions.
+Adding species-specific or location-specific parameters in the model would also allow the evaluation of the impact of disease mitigating actions.
 Lesion treatment using antibiotics could for instance be taken into account in the model by locally reducing the transmission and/or removal parameters.
@@
 However, one might want to discriminate between coral recovery and mortality.
-However, again, adding processes implies an increase in the number of parameters to be calibrated.
+Yet, again, adding processes implies an increase in the number of parameters to be calibrated.
+This modifications could lead to an endemic disease dynamics.
 
 \subsection*{Better understanding the protective role of corals against storm waves}
-Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipating over reefs.
-The modeled significant wave height experiences a twofold reduction over the barrier reef during the passage of the hurricane (Fig.
+Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipated over reefs.
+The modeled significant wave height experiences a twofold (\todo{check that}) reduction over the barrier reef during the passage of the hurricane (Fig.
 \ref{ccl:swh}).
-This highlights the important protective role of coral reefs against coastal flood hazard.
+This highlights the important protective role of coral reefs against coastal flooding hazard.
 Better understanding this role can motivate coral reef restoration efforts.
-For instance, \citep{ferrario2014effectiveness} estimated that coral reefs could provide wave attenuation comparable to artificial defenses such as breakwaters for a lower cost.
+For instance, \cite{ferrario2014effectiveness} estimated that coral reefs could provide wave attenuation comparable to artificial defenses such as breakwaters for a much lower cost.
 However, modelling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes.
@@
 For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modelling framework.
-Nonhydrostatic models are more expensive to run as they require a much finer spatial resolution ($\sim$ 1 m) and have a more complex physics.
+Nonhydrostatic models are more expensive to	 run as they require a much finer spatial resolution of the order of a meter and have a more complex physics.
 They can therefore often be used only in rather small areas and over short periods of time \citep{fringer2019future}.

### 2022-02-13 chapters/intro/intro.tex
commentaires du patron - part I
figure gom + acidification

@@
 \section{Coral reefs: a rich but threatened ecosystem}
-
-\todo{Talk about acidifcation}
 
@@
 
-Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2017coral}: it is estimated  that 30\% have already been severely damaged, and close to 60\% may be lost by 2030 \citep{hughes2003climate}.
+Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2007coral}: it is estimated  that 30\% have already been severely damaged, and close to 60\% may be lost by 2030 \citep{hughes2003climate}.
 This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} (Fig.
 \ref{intro:cover}) is caused by both global and local anthropogenic stressors.
-The main global stressor is climate change induced ocean warming, that cause recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
+The main global stressors are climate change induced ocean warming and acidification.
+Ocean warming cause recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
 Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral disease outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
+Furthermore, the increase of the concentration of atmospheric carbon dioxyde causes a decrease of the pH and carbonate concentration of seawater.
+As a result, coral calcification and growth decrease, which inhibits their capacity to build structurally complex habitats \citep{hoegh2007coral}.
 Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
@@
 \begin{enumerate}
-    \item Can we explain the observed spread of SCTLD throughout Florida using an epidemio-hydrodynamic model ?
+    \item How did the SCTLD spread through the FRT ?
     \item What caused the apparent stalling of the propagation of SCTLD ? 
@@
 
-\todo{Figure LC, GoM Florida Straits, etc.}
+\begin{figure}
+	\centering
+	\includegraphics[width=.9\textwidth]{chapters/intro/figures/fig_gom.png}
+	\caption{}
+	\label{intro:gom}
+\end{figure}
 
@@
         \textbf{ECMWF}  & European Centre for Medium-Range Weather Forecast \\
+        \textbf{D55}    & Dredge 55 clamshell \\
         \textbf{FC}     & Florida Current \\
@@
         \textbf{GoM}    & Gulf of Mexico \\
+        \textbf{HURDAT} & HURricane DATabases \\
         \textbf{HYCOM}  & HYbrid Coordinate Ocean Model \\
@@
         \textbf{RS}     & Radiation stress \\
+        \textbf{SB}     & Spider Barge \\
         \textbf{SCTLD}  & Stony coral tissue loss disease \\
@@
         \textbf{SWAN}   & Simulating WAves Nearshore \\
+        \textbf{TI}     & Terrapin Island hopper \\
+        \textbf{TX}     & Texas cutterhead \\
         \textbf{WFS}    & West Florida Shelf

### 2022-02-13 chapters/onset/article.tex
figure gom + acidification

@@
 Five types of dredge were considered in our modeling study (Fig.
-\ref{fig:onset_mesh}B): ($a$) Texas cutterhead (TX), ($b$) non-conventional dredging, \ie TX with suction mechanism turned off (NonConv), ($c$) Spider Barge (SB), ($d$) Terrapin Island hopper (TI), and ($e$) Dredge 55 clamshell (D55).
+\ref{fig:onset_mesh}B): (\textit{a}) Texas cutterhead (TX), (\textit{b}) non-conventional dredging, \ie~TX with suction mechanism turned off (NonConv), (\textit{c}) Spider Barge (SB), (\textit{d}) Terrapin Island hopper (TI), and (\textit{e}) Dredge 55 clamshell (D55).
 

### 2022-02-13 include/title.tex
figure gom + acidification

--- 2022-02-11 include/title.tex
@@
 \centering
-\LARGE\bfseries Better understanding transport processes to support Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
+\LARGE\bfseries Modeling transport processes to inform the management of Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
 \end{minipage}

### 2022-02-13 thesis.tex
updated ref for onset
intro added to main manuscript
conclusion v1
figure gom + acidification

@@
 \usepackage[pdftex,linktocpage]{hyperref}		%For hyperlinks within doc, and setting doc info
-\hypersetup{pdftitle={How modeling tools can inform coral reef management in Florida}}
+\hypersetup{pdftitle={Modeling transport processes to inform the managment of Florida's Coral Reef}}
 \hypersetup{pdfauthor={Thomas DOBBELAERE}}
-\hypersetup{pdfsubject={How modeling tools can inform coral reef management in Florida}}
+\hypersetup{pdfsubject={Modeling transport processes to inform the managment of Florida's Coral Reef}}
 \hypersetup{pdfkeywords={Biophysical modeling, Florida Coral Reef, SCTLD, SLIM}}
@@
 % \usepackage{cmbright}				%TODO: this doesn't compile when I add it *** Font: Computer modern bright
-% \usepackage{sfmath}				%Display all maths with sans serif font
+%\usepackage{sfmath}				%Display all maths with sans serif font
+\renewcommand\familydefault{\sfdefault}
 

### 2022-02-13 chapters/conclusions/conclusions.tex
few last things

@@
     }
-    \item \textbf{1. Can we explain the observed spread of SCTLD throughout Florida using an epidemio-hydrodynamic model ?}
+    \item \textbf{1. How did the SCTLD spread through the FRT ?}
     \begin{list}{}{\setlength{\topsep}{0pt}}
@@
 Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipated over reefs.
-The modeled significant wave height experiences a twofold (\todo{check that}) reduction over the barrier reef during the passage of the hurricane (Fig.
+The modeled significant wave height under the center of the hurricane reduced from 8 m to $<1$ m over the barrier reef during the passage of the hurricane (Fig.
 \ref{ccl:swh}).
@@
     \centering
-    \includegraphics[width=\textwidth]{chapters/conclusions/figures/swh_figure.png}
-    \caption{Snapshot of the significant wave height modeled by the coupled SLIM+SWAN model during Hurricane Irma, on Sept 10, 2017 at 1200 UTC.
+    \includegraphics[width=\textwidth]{chapters/conclusions/figures/swh_curve.png}
+    \caption{\textbf{A}: Snapshot of the significant wave height modeled by the coupled SLIM+SWAN model during Hurricane Irma, on Sept 10, 2017 at 1200 UTC.
 The path of the hurricane is shown by a black line with squares.
 Land is shown in dark gray and reefs in light gray.
-Significant wave attenuation is visible between the fringing coral reefs and the islands.}
+\textbf{B}: Evolution of the significant wave height under the center of Irma.
+Important wave attenuation is visible between the fringing coral reefs and the islands.}
     \label{ccl:swh}
 \end{figure}
+
+% Last nice chapter
+The work presented in this thesis has demonstrated the capabilities of modeling tools to inform the management of Florida's Coral Reef.
+Models allow the evaluation of hypotheses derived from field observations on large-scale systems and can help to deduce the characteristics and underlying mechanisms of complex processes.
+Furthermore, they can be used in a predictive manner to evaluate the impacts of changing environmental conditions on a given ecosystem.
+Additionally, the framework developed in this thesis relies on the same tools as already widely used larval transport models and could be combined with them to optimize the design of coral reef protection and restoration campaigns.
+Given this high potential, further research to improve the model and synergistic work with field an lab studies may well yield results which could help maintaining the biological functions of coral reefs as long as possible through the Anthropocene.

### 2022-02-13 chapters/intro/intro.tex
few last things

@@
 The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC).
-The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream.
+The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream (Fig.
+\ref{intro:gom}).
 The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}.
@@
 The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}.
-These eddies significantly impact the ecology of the southwest Florida Shelf as they have been identified as a mechanism causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}.
+These eddies significantly impact the ecology of the southwest Florida Shelf by causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}.
 Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}.
@@
 	\includegraphics[width=.9\textwidth]{chapters/intro/figures/fig_gom.png}
-	\caption{}
+	\caption{An example of the Loop Current and the Florida Current on January 12, 2019 as shown by HYCOM.}
 	\label{intro:gom}
@@
 In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents.
-To answer our different research questions, we adapted this model to evaluate the transport of the causative agent of the SCTLD and sediments particles.
+To answer our different research questions, we adapted this larval transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.
 
-However, at the beginning this thesis, SLIM was not able to represent wave-current interactions under storm conditions.
+At the beginning this thesis, SLIM was not able to represent wave-current interactions under storm conditions.
 Heavy winds generate large wind-waves that disturb ocean conditions and interact non-linearly with currents \citep{liu2020impacts,wu2011fvcom}.
@@
 This is usually achieved by spectral wave models, which describe the evolution of the wave energy spectrum.
-An objective of this thesis was therefore the coupling of SLIM with the spectral wave model Simulating WAve Nearshore (SWAN, \citealp{booij1999third}) SWAN was specifically developed for coastal applications and relies on numerical schemes adapted to small-scale, shallow water regions.
+An objective of this thesis was therefore the coupling of SLIM with the spectral wave model Simulating WAve Nearshore (SWAN, \citealp{booij1999third}), specifically developed for coastal applications and relies on numerical schemes adapted to small-scale, shallow water regions.
+Further details on the models and their coupling is given in chapter \ref{chap:irma}.
 
@@
         \textbf{NOAA}   & National Oceanographic and Atmospheric Administration \\
+        \textbf{ODMDS}	& Ocean Dredge Material Disposal Site \\
         \textbf{PoM}    & Port of Miami \\
@@
         \textbf{TX}     & Texas cutterhead \\
+        \textbf{WWTP}	& Wastewater Treatment Plant \\
         \textbf{WFS}    & West Florida Shelf

### 2022-02-13 chapters/onset/article.tex
few last things

@@
 First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022connecting} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
-Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}.
-Furthermore, recent studies showed evidence that sediments can act as  a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
+Although the causative agent of the disease remains unknown, the hydrodynamics is likely to play an important role in its propagation as both modeling studies and ex situ experiments showed evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}.
+Furthermore, recent studies showed that sediments can act as  a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.
 
@@
 The dredging was monitored twice-weekly at 26 permanent monitoring stations established within the Miami-Dade County, making it one of the most complete datasets related to a dredging project \citep{gintert2019regional}.
-Interestingly, disease signs were also reported at one of these monitoring stations in May 2014.
+Interestingly, disease signs were also reported at one of these monitoring stations in May 2014 (see supplementary material \ref{onset:appendice}).
 
@@
 However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
-The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{usace2017}.
+\citep{usace2017} provideda back-of-the-envelope estimation that this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel .
 Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.
@@
 Further, they suggested that the onset of the disease might have been linked to a leaking discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant (WWTP) located off Virginia Key.
-However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease was transported to the monitoring site of Virginia Key on sediments released by the dredging.
-This possibility can be evaluated using a hydrodynamic model simulating the transport of sediments produced during the dredging project.
-As coastal reef ecosystems are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures, such a model would require high spatial resolution to accurately represent the transport of sediments at the reef-scale.
+However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease was transported to the monitoring site of Virginia Key by sediments released during the dredging.
+This scenario can be evaluated using a sediment transport model.
+As coastal reef ecosystems are characterized by the complex topography of the coastline and the presence of islands, reefs and artificial structures, such a model would require a fine spatial resolution to accurately represent the transport of sediments at the reef-scale.
 In this context, unstructured-mesh models are particularly well suited, as they can easily adapt to the topography \citep{fringer2019future} and can capture small-scale circulation features around reefs and islands \citep{lambrechts2008multi}.
 
-The goal of this study is therefore to simulate the trajectories of the sediments released during the entirety of the PortMiami Deep Dredge Project (PMDDP) using a high-resolution hydrodynamic model coupled with a sediment transport model.
-Specifically, our goal is to answer the following questions: (1) Which reefs were impacted by PMDDP ?
-(2) Is the impact on these reefs consistent with the observed timing of the onset of SCTLD ?
+The goal of this study is therefore to simulate the dispersed sediments during the entirety of the PortMiami Deep Dredge Project (PMDDP) using a high-resolution hydrodynamic model coupled with a sediment transport model.
+Specifically, our goal is to answer the following questions: (1) Which reefs were impacted by the PMDDP and was this impact consistent with the observed timing of the onset of SCTLD ?
+(2) Was disease transmission to Virginia Key from other diseased reefs possible before September 2014 ?
 
@@
 It was made up of approximately $3.5\times 10^5$ triangles and was generated with the seamsh\footnote{\url{https://pypi.org/project/seamsh/}} Python library, which is based on the the open-source mesh generator GMSH \citep{geuzaine2009gmsh}.
-The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the first observation of SCTLD by \cite{precht2016unprecedented}.
-Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the flow acceleration between reefs and islands.
+The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the observation of SCTLD on Virginia Key by \cite{precht2016unprecedented}.
+Using such a fine mesh resolution, we simulated fine-scale details of the ocean currents, such as the flow acceleration between reefs and islands (Fig.
+\ref{fig:onset_mesh}C).
 
-The transport of sediments released from the channel was then modeled using a Lagrangian particle tracking model, forced by the simulated currents.
+The transport of sediments released from dredging operations along the channel was then modeled using a Lagrangian particle tracking model, forced by the simulated currents.
 The sediment model is inspired by the Particle Transport Model (PTM), developed by the US Army Corps of Engineers \citep{macdonald2006ptm}.
@@
 
-The outputs of the sediment model were then used to evaluate the turbidity and sedimentation generated by the expansion of PoM over coral reefs.
+The outputs of the sediment model were then used to evaluate the turbidity and sedimentation generated by the different dredging operations over coral reefs.
 The occurrence of high turbidity over reefs was assessed based on the concentration of suspended sediment particles in the model.
@@
 This cumulated concentration was normalized by the total number of simulated time steps to obtained an averaged concentration of deposited particles over the simulated period.
-Higher values of this indicator would indicate larger sediment deposition over longer cumulated time.
+Higher values of this indicator would indicate larger sediment deposition over a longer cumulated time.
 
@@
 Four sites were reefs where disease was reported in 2014 by \cite{precht2016unprecedented}.
-The fifth site was monitoring station HBSC1-CP of the monitoring of the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
-\ref{fig:onset_mesh}A, see Appendic\ref{onset:appendice} ).
+The fifth site was the monitoring station HBSC1-CP, used throughout the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
+\ref{fig:onset_mesh}A, see pictures of coral colony in appendix \ref{onset:appendice}).
 The impact of dredging was assessed by counting the number of sediment particles originating from each dredging that were transported within 500 m of all five site.
 This distance was chosen to match the scale of the 500 m $\times$ 500 m reef polygons used in the model \citep{dobbelaere2020coupled}.
-The obtained number of particles was then divided by the total number of sediment particles released by each type of dredge.
+The resulting numbers of particles was then divided by the total number of sediment particles released by each type of dredging operation.
 Larger values of this indicator would suggest a greater impact of a given type of dredging operation at a given monitoring sites.
@@
 To evaluate this possibility, we computed monthly disease connectivity matrices following the methodology of \cite{dobbelaere2020coupled} during our simulated period.
-These connectivity matrices can be interpreted as large graphs whose vertices are 500 m $\times$ 500 m reef areas and whose edges represent disease connectivity pathways.
+These connectivity matrices can be interpreted as large graphs whose vertices are 500 m $\times$ 500 m sub-reef polygons and whose edges represent disease connectivity pathways.
 Evaluating the possibility of disease propagation from any sub-reef $A$ to any sub-reef $B$ is therefore equivalent to evaluating the existence of paths, \ie sequences of connected vertices in the network starting from $A$ and reaching $B$.
@@
 This was performed using the function \texttt{get\_all\_shortest\_paths} of the Python \texttt{python-igraph} package \citep{csardi2006igraph}.
-Such function requires the definition of a weight $w_{ij}$ for the edge connecting reef $i$ to reef $j$.
+Such a function requires the definition of a weight $w_{ij}$ for the edge connecting reef $i$ to reef $j$.
 We chose $w_{ij} = 1-\tilde{C}_{ij}$, where $\tilde{C}_{ij}$ is the probability of disease propagation from reef $i$ to reef $j$, so that "shorter" edges of the networks (\ie connectivity pathways with smaller weights) correspond to connections with larger disease propagation probability.
@@
 
-5-50 $\mu$m was the only grain size that produced plumes consistent with the observations of \cite{barnes2015sediment} and \cite{cunning2019extensive} (Fig.
+Only grain sizes in the 5-50 $\mu$m range produced plumes consistent with the observations of \cite{barnes2015sediment} and \cite{cunning2019extensive} (Fig.
 \ref{fig:onset_depo}A).
@@
 The modeled occurrence of plumes within 15 km of the dredged matched the presence/absence data of \cite{cunning2019extensive} in 82.7\% of cases and the total area where plumes were observed was about 230 km$^2$, consistent with the $\sim228$ km$^2$ estimated by \cite{barnes2015sediment}.
-Modeled plumes mostly occurred north of the dredged channel, as particles were driven northward under the action of the Florida Current, with plume frequencies of about 20\% near the monitoring site of N.
+Modeled plumes mostly occurred north of the dredged channel, as particles were driven northward under the action of the Florida Current, with plume occurring during 20\% of the simulated days near the monitoring site of N.
 Sunny Isles.
-Plume frequencies was lower ($\sim 5$\%) over offshore reefs but reached of up to 50\% over alongshore reefs.
+Plume frequencies were lower ($\sim 5$\%) over offshore reefs but reached of up to 50\% over alongshore reefs.
 These alongshore reefs correspond to regions of important sedimentation (Fig.
 \ref{fig:onset_depo}B), which is consistent with the correlation between plumes and sedimentation highlighted in \cite{cunning2019extensive}.
-The plume frequency was about 1\% at site of HBSC1-CP while no plume occurred at other southern monitoring sites during the simulation.
+The plume frequency was about 1\% at site of HBSC1-CP while no plume reached other southern monitoring sites during the simulation.
 
@@
 For all grain sizes, no sedimentation occurred near Rainbow Reef and Emerald Reef.
-However, sediments settled at similar latitudes west to these sites for grain sizes below 100 $\mu$m.
+However, sediments settled at similar latitudes west of these sites for grain sizes below 100 $\mu$m.
 
@@
 Sunny Isles.
-The impact of dredging at this site decreased with sediment size and the fraction of sediment particles reaching the N.
+The impact of dredging at this site decreased with increasing sediment size and the fraction of sediment particles reaching the N.
 Sunny Isles dropped below 2\% for grain sizes larger than 200 $\mu$m.
 However, for grain sizes finer than 50 $\mu$m, N.
-Sunny Isles was heavily impact by TX dredging.
+Sunny Isles was heavily impacted by TX dredging.
 Indeed, the number of particles produced by TX reaching N.Sunny Isles was 50\% larger than the cumulated number of particles produced by other dredging operations reaching the monitoring site for the same grain size.
-Such large impact can be explained by the fact that TX was one of the most frequent type of dredging during the simulations, with 268 simulated operations (Fig.
+Such large impact can be explained by the fact that TX was one of the most frequent type of dredging during the simulations, with 268 simulated operations out of 734 (Fig.
 \ref{fig:onset_bar}C).
-However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles to reach N.
+However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles that reached N.
 Sunny Isles.
@@
 Fig.
-\ref{fig:onset_pipe} shows the fraction of released wastewater particles that reached the site of Virginia key for every 100 m segments of the discharge pipe.
+\ref{fig:onset_pipe} shows the fraction of released wastewater particles that reached the site of Virginia Key for every 100 m segments of the discharge pipe.
 The fraction of particles that drifted within 500 m of Virginia Key was: 100\% for particles released within 500 m of the site; 0-1\% for particles released between 500 m and 1.5 km away from the site; and 0\% for all other sources of wastewater particles.
@@
 It suggests that sediment particles produced by dredging were mostly transported northward under the influence of the Florida Current.
-No sediment particles reached the southern sites of Rainbow Reef and Emerald Reef while dredging had limited direct impact on Virginia Key, identified as the site where the outbreak of SCTLD was initiated in September 2014.
+No sediment particles reached the southern sites of Rainbow Reef and Emerald Reef while dredging had limited direct impact on Virginia Key, identified as the site where the outbreak of SCTLD was initiated in September 2014 \citep{precht2016unprecedented}.
 Furthermore, our results indicate that it is unlikely that the site was affected by a leaking discharge pipe of Miami Central District Municipal WWTP, as suggested by \cite{gintert2019regional}.
@@
 Sunny Isles and HBSC1-CP, where the impact of dredging was greater.
-Up to 30\% of silt particles produced by non-conventional dredging reached coral reefs of N.
+Up to 30\% of silt particles produced by non-conventional dredging operations reached coral reefs of N.
 Sunny Isles while 5\% of them reached site HBSC1-CP, where significant sedimentation occurred.
-Furthermore, analysis of monthly disease connectivity networks indicate disease transmission from HBSC1-CP to the site of Virginia Key was possible through waterborne transmission.
+Furthermore, monthly disease connectivity networks indicate that disease agents could be transmitted from HBSC1-CP to the site of Virginia Key.
 
@@
 By contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rock particles to be released in the water column.
-The numbers of particles reaching the monitoring sites were therefore likely overestimated for all sources but non conventional dredging.
+The numbers of particles reaching the monitoring sites were therefore likely underestimated for non conventional dredging operations.
 However, the fractions given in Fig.
-\ref{fig:onset_bar}A,B remain valid as they are relative to the total number of sediments released in the water columns.
-Despite these overestimations, the coral reefs of N.
+\ref{fig:onset_bar}A,B remain valid as they are relative to the total number of sediments released in the water column.
+The coral reefs of N.
 Sunny Isles appeared to be highly sensitive to fine-grained sediments produced by dredging (Fig.
@@
 Sunny Isles in June 2014.
-Moreover, up to 5\% of the fined-grained sediments produced by non-conventional dredging reached the site HBSC1-CP.
+Moreover, up to 5\% of the fined-grained sediments produced by non-conventional dredging reached the site HBSC1-CP before May 2014, when the disease was first reported there.
 Although this represents a more limited quantity of sediments, HBSC1 was within 2 km from the sites where non-conventional dredging took place, in a region where the model predicted an important sedimentation.
@@
 % POSSIBILITY OF DISEASE PROPAGATION THROUGH WATERBORNE TRANSMISSION
-In all simulations, the expansion of PoM had limited direct impact on Virginia Key, identified as the site where the outbreak initiated in September 2014 \cite{precht2016unprecedented}.
+In all simulations, the expansion of PoM had limited direct impact on Virginia Key, identified as the site where the outbreak initiated in September 2014 \citep{precht2016unprecedented}.
 For all grain sizes, the fraction of sediment particles reaching Virginia Key remained below 1\%.
@@
 This implied that the propagation of the disease from HBSC1-CP to Virginia Key through hydrodynamics-driven transport of disease agents was possible during these four months.
-However, these pathways used reefs located further southward as stepping stones.
+All these pathways used reefs located further southward as stepping stones.
 Therefore, the propagation of outbreak to reach Virginia Key starting from HBSC1-CP required these intermediary reefs to get first infected by disease agents released by disease colonies of HBSC1-CP.
@@
 Assuming an averaged transmission time of the order of 5-10 days \citep{dobbelaere2020coupled}, disease propagation from HBSC1-CP to Virginia Key might have required several months.
-This is consistent with the 5 month period separating reports of diseased at the two sites.
+This is consistent with the 5-month period separating disease reports at these two sites.
 Moreover, the fact that June, August and September exhibited similar connectivity pathways suggests that disease propagation could indeed have occurred over several months without being interrupted by changing hydrodynamic conditions.
@@
 This evaluation was performed using a quasi-3D sediment model forced by high-resolution currents from the coastal ocean model SLIM.
-As the first signs of the SCTLD outbreak were reported during these expansion works near Virginia Key and knowing that sediments may serve as SCTLD vector, we aimed at answering two main questions.
-First, we wanted to evaluate the fraction of sediments produced by the PortMiami Deep Dredge Project that reached the site of Virginia Key.
+As the first signs of SCTLD were reported during this dredge project and since sediments may be a SCTLD vector, we aimed at answering two main questions.
+First, we wanted to evaluate the fraction of sediments produced by the PortMiami Deep Dredge Project that reached Virginia Key and other monitoring sites where the disease was reported.
 Second, we evaluated the possibility of waterborne transmission of SCTLD to Virginia Key from other reefs impacted by the dredging.
@@
 
-Our results suggest that the dredging works had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission trough the produced sediments was extremely unlikely.
-However, our results show the sediments released by the expansion works had a non-negligible impact on the sites of N.
-Sunny Isles and HBSC1-CP, where signs of disease were observed before the outbreak was first reported at Virginia Key, indicating that the sediments might have played a role in the onset of the diseased at these two sites.
+Our results suggest dredging operations had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission by sediments was extremely unlikely.
+However, our results show the sediments released by the dredging operations had a non-negligible impact on the sites of N.
+Sunny Isles and HBSC1-CP, where signs of disease were observed before the outbreak was first reported at Virginia Key.
+This suggests that the sediments might have triggered the onset of the diseased at these two sites.
 Furthermore, using a previously developed bio-physical model that successfully reproduced the observed spread of SCTLD, we show that there was a possibility of waterborne disease propagation from HBSC1-CP to Virginia Key.

### 2022-02-13 chapters/onset/onset.tex
updated ref for onset
few last things

@@
 
-\item \textbf{Dobbelaere T.}, other co-authors TBD, $\&$ E.
-Hanert, 2022, Early transmission dynamics of SCTLD and its connection with the PortMiami Deep Dredge Project.
+\item \textbf{Dobbelaere, T.}, other co-authors TBD, \& Hanert, E.
+Early transmission dynamics of a deadly coral disease and its connection with the PortMiami Deep Dredge Project.
 (planned submission to \textit{Marine Pollution Bulletin})
@@
     For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD).
-The outbreak reportedly initiated near Virginia Key in September 2014, during the the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015.
+The outbreak reportedly initiated near Virginia Key in September 2014, during the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015.
 The disease then spread to the entirety of FCR, likely through waterborne transmission.
 Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease.
-The goal of this study is therefore to evaluate whether sediments produced or resuspended during dredging operations might have been transported to reefs where disease was reported and to assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
+Here we evaluate whether sediments produced or resuspended during dredging operations could have reached reefs where the disease was first reported and assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
 We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM.
-Our result suggested that the dredging had close to no direct impact on the coral reefs of Virginia Key.
+A particular attention was paid to the non conventional dredging operations performed without pumping the produced chopped rock particles.
+Our result suggest that the dredging had close to no direct impact on the coral reefs of Virginia Key.
 However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites.
-Furthermore, using a biophysical disease agent transport model suggests th   at waterborne transmission from one of these sites to Virginia Key was possible before September 2014.
+Furthermore, using a biophysical disease agent transport model, we show that waterborne transmission from one of these sites to Virginia Key was possible before September 2014.
 This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.

### 2022-02-13 include/abstract.tex
few last things

@@
 
-The population of corals throughout the world has declined dramatically under the effect of rising water temperature and many coral species are now listed as threatened.
-In addition to ocean warming, other threats to corals include (but are not limited to) disease outbreaks, intensifying hurricanes and sediment disturbances.
+Coral populations have declined dramatically worldwide under the combined effects of climate change and local anthropogenic stressors.
+In addition to ocean warming and acidification, threats to coral reefs in the Caribbean include frequent and virulent disease outbreaks, and intensifying hurricanes.
 In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD).
-First observed in 2014 during the monitoring of dredging operations in the Port of Miami, the disease has now spread through the entirety of FCR and has been observed in numerous territories of the Caribbean.
-Furthermore, hurricanes tend to strike Florida during the reproduction period of the corals, when thousands of larvae are released in the water column.
-In addition to causing physical harm to corals, hurricanes significantly modify ocean currents and might therefore disturb larval exchanges between reefs.
-A high resolution ocean model, SLIM, is used to build a biophysical model to study the aforementioned threats to FCR.
-First, a coupled epidemio-hydrodynamic model is developed to study both the onset and the propagation mechanisms of the SCTLD.
-Furthermore, a coupled wave-current model is devised to study the impact of Hurricane Irma (2017) on transport processes in the Florida Keys.
-This study brings new insight on the impact of hydrodynamics on the propagation of SCTLD as well as the impact of hurricane on coral ecosystem.
+First observed in 2014 during the monitoring of the PortMiami Deep Dredge Project (PMDDP), the disease has now spread through the entire FCR and has been reported in numerous territories of the Caribbean.
+However, the propagation of the disease through FCR seemed to slow down when it reached its southwestern end, between the Marquesas and the Dry Tortugas (DRTO).
+Although the causative agent of the disease remains unknown, studies showed that its transmission was likely waterborne and that sediments could act as SCTLD vectors.
+The hydrodynamics should therefore be highly explanatory of its spread.
+Furthermore, Florida is a prime landfall target for hurricanes during the reproduction period of corals.
+In addition to causing wholesale destruction of the reefs, hurricanes might thus also impact larval dispersal through wind-wave-induced currents.
+Here, we use the high resolution coastal ocean model SLIM to capture transport processes in the FCR, and study the impact of SCTLD and hurricanes to coral reefs.
+First, a coupled epidemio-hydrodynamic model is developed to reproduce and understand the observed spread of SCTLD in Florida.
+A sediment transport model is then used to evaluate the impact of the PMDDP on the observed onset of the disease in 2014.
+Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys.
+Assuming the dispersal of disease agents within neutrally material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling to eddy activity near the DRTO.
+Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease.
+Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes.
+This thesis highlights the potential of models informed and confronted against field knowledge as powerful tools to inform the management of complex ecosystems such as the FCR.
 

### 2022-02-13 include/thanks.tex
few last things

 \phantom{thanks}
%\textit{Remerciements à }
\chapter*{Remerciements}

Mes premiers remerciements vont à mon promoteur, le professeur Emmanuel Hanert, le grand chef qui m'a guidé et accompagné tout au long de cette thèse. D'une énergie et surtout d'un optimisme à toute épreuve (si toutes ses prédictions s'étaient avérées exactes, cette thèse aurait été finie en 1 ans), il m’a poussé et soutenu durant les quatre dernières années. En outre, il a toujours réussi à accomplir le miracle de dégager du temps dans son agenda bien chargé pour me conseiller ou relire et commenter mon travail avec minutie.

Je remercie également le groupe de vie ENGE, qui m'a accueilli durant les bientôt quatre années de cette thèse. Ce groupe hétéroclite m'a fourni un cadre de travail convivial, agrémenté de pauses café, de verres à la quinzaine et de soupers de Noël endiablés. Ce fut un plaisir de les avoir comme collègues et je ne pourrai désormais plus manger de burrito sans penser à eux. Parmi eux, je tiens particulièrement à remercier mon fidèle voisin de bureau et excellent camarade de conférence, Antoine Saint-Amand. Il m'a guidé et accompagné au cours de mon épopée dans la jungle de la thèse et a été une source intarissable de bonnes histoires et anecdotes. De plus, son engagement à la ville comme à l’université ainsi que son dévouement pour ses étudiants ont été une source d’inspiration. Il me reste à remercier l’autre mouscronnois du bureau, Valentin Vallaeys, pour son humour et sa propension à inonder ses collègues de sucreries. En outre, son don inégalé pour poser les bonnes questions m'a plus d'une fois permis d’entrevoir une issue à un problème apparemment insoluble l’instant d’avant.

Je tiens également à remercier tous les autres membres de la team SLIM. Ce fut un plaisir de travailler avec eux au développement de ce modèle. Parmi eux, je tiens à remercier spécialement Jonathan Lambrechts, qui rend l’impossible possible en 3 coups de clavier. Malgré les nombreuses sollicitations qu’il reçoit, il s'est toujours rendu disponible pour me débugger et cette thèse aurait été techniquement impossible sans lui. À ce propos, un échange téléphonique particulièrement éclairant me revient en mémoire: “Salut Jon, est-ce que tu as un peu de temps pour m’aider ? Non, mais dis toujours.”... 5 minutes plus tard, le problème était réglé.

I would also like to thank the team of collaborators and co-authors that started in 2019 with Lew Gramer, later joined by Erinn Muller and Dan Holstein, and has been growing ever since. It has been a real pleasure to work with them on all these exciting projects. I would like to extend my thanks to Pr Sam Purkis and Ceci Lopez-Gamundi. Although unraveling the mysteries of the Great Bahama Bank was not directly linked to the subject of this thesis, it has been a pleasure to join them in this geological quest. Et tant qu’on est à Miami, je remercie également Matthieu le Hénaff, qui garde un œil sur nos projets floridiens en coulisse. Ses conseils avisés ainsi que les personnes avec lesquelles il nous a mis en contact ont été d’un grand secours.

Il convient également de remercier ceux qui n'ont probablement pas contribué à me rendre plus productif mais ont assurément rendu ces années de thèse beaucoup plus amusantes. Tout d'abord, il me faut remercier les 36 "voix" du kot Carrefour. Cette grande famille m'a permis de rencontrer des gens d'origines et de cultures variées tout en me fournissant un lieu de vie calme (à l’exception de quelques folies) pour mener à bien cette thèse. Les prochains remerciements sont pour l’équipe de l'Oenokot, qui réussit à convaincre tout le monde (même mes anciens étudiants) que je suis un étudiant en BAC2 de droit. Des remerciements tout particuliers s'imposent pour le dénominateur commun à ces deux bandes de joyeux lurons, Redge, qui me supporte depuis presque 5 ans malgré des premiers contacts difficiles au kot Erasmus. Enfin, je remercie la Revue des Ingénieurs, qui offre un auditoire à mes blagues depuis 9 ans et qui, jusque dans les derniers jours de la rédaction de cette thèse, m’a fourni quelques bons prétextes à la procrastination.

\vspace{3em}

Ce travail a été financé par le Fonds de la Recherche Scientifique au moyen d’une bourse FRIA.

### 2022-02-13 thesis.tex
few last things

@@
 %%%%%% 
-\mychapter{Modeling the spread of SCTLD in Florida} \label{chap:sctld}
+\mychapter{The mechanisms of the spread of SCTLD in Florida} \label{chap:sctld}
 \chaptermark{Modeling the spread of SCTLD in Florida}  
 \input{chapters/sctld/sctld}
-\mychapter{Transmission of SCTLD from the Marquesas to the Dry Tortugas}\label{chap:drto}
+\mychapter{The apparent stalling of the spread of SCTLD in the Marquesas}\label{chap:drto}
 \chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
@@
 \input{chapters/onset/onset}
-\mychapter{Impacts of hurricanes on wave-induced ocean transport processes} \label{chap:irma}
+\mychapter{The impacts of hurricanes on transport processes in the Florida Keys} \label{chap:irma}
 \chaptermark{Wave-induced transport under hurricanes}  

### 2022-02-13 chapters/conclusions/conclusions.tex
commentaires du patron

@@
 \section{Conclusions}
-In this thesis, we have developed a framework to better understand the propagation and onset of the SCTLD, as well as the impact of major hurricanes onctransport processes over coral reefs.
-This model had to be able to accurately capture ocean circulation at different scales to study transport processes in the FCR.
+In this thesis, we have developed a modeling framework to better understand the propagation and onset of the SCTLD, as well as the impact of major hurricanes on transport processes over coral reefs.
+Our model had to be able to accurately capture ocean circulation patterns at different scales to study transport processes in the FCR.
 First, the large-scale Loop Current-Florida Current system had to be accurately reproduced to correctly capture its impact on eddy formation near the DRTO and the Florida Keys.
-Second, the model had do reproduce ocean circulation at the reef-scale to capture recirculation and acceleration of currents between reefs and islands.
+Second, the model had to reproduce ocean circulation at the reef-scale to capture the recirculation and acceleration of currents between reefs and islands.
 This was performed by coupling the multi-scale ocean model SLIM to a connectivity-based epidemiological model (chapter \ref{chap:sctld}), a sediment transport model (chapter \ref{chap:onset}) and a spectral wave model (chapter~\ref{chap:irma}).
 
-Our coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by modeling the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents.
+Our coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by representing the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents.
 It confirmed the result of previous ex-situ studies that showed evidence of waterborne transmission of the SCTLD with an average transmission time of the order of 10 days.
@@
 This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived from field observations on large scale systems.
-As such, models informed and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as the FCR.
+As such, models informed by and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as the FCR.
 
-This epidemiological model was then used to build networks of potential exchanges of disease agents between $\sim 1.6\times10^4$ sub-reef areas during a period of three years.
+This epidemiological model was then used to build networks of potential exchanges of disease agents between $\sim 1.6\times10^4$ sub-reef areas composing the FRT during a period of three years.
 The analysis of these networks showed that exchanges from the Marquesas to the DRTO were interrupted during most of 2020.
 This interruption was likely linked to eddy activity near the DRTO and was consistent with the apparent stalling of the outbreak in the region in 2020.
-This confirms the role played the Loop Current/Florida Current system in the control of the connectivity in the Florida Keys and the DRTO.
-Furthermore, sub-reefs predicted to be more vulnerable to SCTLD in the disease networks were consistent with the observed geographic distribution of disease coral in the DRTO.
-This further highlights that modeled hydrodynamics are highly explanatory of the propagation of the disease.
+This confirms the role played the Loop Current/Florida Current system in the control of the connectivity pathways in the Florida Keys and the DRTO.
+Furthermore, sub-reefs predicted to be more vulnerable to SCTLD in the disease networks were consistent with the observed geographic distribution of diseased corals in the DRTO.
+This further highlights that the modeled hydrodynamics is highly explanatory of the propagation of the disease.
 
-The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that our high-resolution biophysical is a valuable tool to understand the spread of SCTLD in Florida.
+The results of chapters \ref{chap:sctld} and \ref{chap:drto} show that our high-resolution biophysical model is a valuable tool to understand the spread of SCTLD in Florida.
 The model successfully reproduced the observed propagation of the disease in 2018-2019 and allowed us to link the observed stalling of SCTLD to observable hydrodynamic features.
 Furthermore, the model allowed us to deduce potential characteristics of the disease agent and its vector.
-This highlights the potential of  model to understand the driving mechanics of ongoing marine disease outbreaks.
+This highlights the potential of biophysical models to understand the driving mechanics of ongoing marine disease outbreaks.
 Furthermore, indicators such as the vulnerablity and source indicators developed in chaper \ref{chap:drto} could be useful to predict and/or mitigate the spread of SCTLD in other territories of the Caribbean.
@@
 Finally, we developed a couple wave-current model to study the impact of Hurricane Irma on transport processes in the Florida Keys.
-This coupled model correctly reproduced the observed wave and hydrodynamics during the passage of the hurricane.
-Furthermore, model results suggest that an import dissipation occurred over coral reefs through depth-induced wave-breaking and bottom friction.
+This coupled model correctly reproduced the observed waves and current dynamics during the passage of the hurricane.
+Furthermore, model results suggest that an important dissipation occurred over coral reefs through depth-induced wave-breaking and bottom friction.
 This important dissipation produced a large wave radiation stress gradients, which caused differences in modeled velocity of up to 1 m/s between the coupled and uncoupled model runs.
@@
 In this context, connectivity negatively impacts coral reefs as it drives the propagation of disease agents.
-However, connectivity acts as a double-edged sword.
-Larval connectivity promotes the resilience of the reef network as it allows the recolonization of depopulated reefs after extreme events such as mass bleaching, hurricanes or disease outbreaks.
+However, connectivity is actually a double-edged sword.
+Larval connectivity promotes the resilience of the reef network as it allows the recolonization of disturbed reefs following extreme events such as mass bleaching, hurricanes or disease outbreaks.
 Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration.
-Identifying reefs with large out-degree our centrality in the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.
+Identifying reefs that are both good sources and sinks of larvae the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.
 Furthermore, knowledge about larval connectivity would allow us to build robust and sustainable networks of connected marine protected areas.
 As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.
-This combined information would allow us to identify reefs that have high potential for restoration and low vulnerability to the disease.
-A possible measure of the restoration potential is the (weighted) out-degree of reefs in the larval connectivity network, as reefs with strong and numerous outgoing connections in the network are more likely to send many larvae to many connected reefs.
+This combined information would allow us to identify reefs that are good larval sources and have low vulnerability to the disease.
+A possible measure of larval supply to the network is the (weighted) out-degree of reefs in the larval connectivity network, as reefs with strong and numerous outgoing connections in the network are more likely to send many larvae to many connected reefs.
 Vulnerability to disease can be evaluated using the (weighted) in-degree of reefs in the disease connectivity network, as reefs with strong and numerous incoming connections are more likely to receive more disease agents from many connected reefs.
-Therefore, reefs with high restoration potential and low vulnerability would correspond to reefs with large larval(weighted) out-degree and low disease (weighted) in-degree.
+Therefore, reefs with high restoration potential would correspond to reefs with large larval(weighted) out-degree and low disease (weighted) in-degree.
 This approach have been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.
@@
 \subsection*{Improving the epidemiological model}
-A limitation of the epidemiological model developed in chapter \ref{chap:sctld} is that it uses species-averaged parameters and assumes an uniform distribution of coral reefs throughout the FRT.
+A limitation of the epidemiological model developed in chapter \ref{chap:sctld} is that it uses species-averaged parameters and assumes an uniform distribution of coral species throughout the FRT.
 The fact that our model correctly reproduced the observed spread of SCTLD for a well-defined range of values of coral resistance to the disease highlights the important impact of this parameter on the propagation of the disease.
 However, coral susceptibility to SCTLD can significantly vary between species and ranges from highly susceptible (\eg~\textit{Dichocoenia stokesii}, \textit{Meandrina meandrites}) to intermediately susceptible (\eg~\textit{Orbicella faveolata}, \textit{Montastrea cavernosa}) and weakly susceptible (\eg~\textit{Acropora Palmata}, \textit{Acropora cervicornis}).
-A first improvement could therefore be to consider classes od corals with different susceptibilities and hence use different infection and removal parameter values in the model.
+A first improvement could therefore be to consider classes of corals with different susceptibilities and hence use different infection and removal parameter values in the model.
 Such improvement would be technically challenging as it would increase the number of variables of model.
@@
 The work presented in this thesis has demonstrated the capabilities of modeling tools to inform the management of Florida's Coral Reef.
-Models allow the evaluation of hypotheses derived from field observations on large-scale systems and can help to deduce the characteristics and underlying mechanisms of complex processes.
+Models allow the evaluation of hypotheses derived from field observations on large-scale systems and can help to deducing the characteristics and underlying mechanisms of complex processes.
 Furthermore, they can be used in a predictive manner to evaluate the impacts of changing environmental conditions on a given ecosystem.

### 2022-02-13 chapters/intro/intro.tex
commentaires du patron

@@
 \ref{intro:corals}).
-Although they account for less than 0.5\% of the ocean floor \citep{spalding1997new}, almost a third of the known marine biodiversity is found on coral reefs \citep{moberg1999ecological}.
-Reef-related fisheries constitute around 10\% of the fish consumed by humans \citep{smith1978coral} and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
+While they account for less than 0.5\% of the ocean floor \citep{spalding1997new}, they are home for almost a third of the marine biodiversity \citep{moberg1999ecological}.
+Reef-related fisheries constitute around 10\% of the fish consumed by humans \citep{smith1978coral}(\todo{more recent stuff}) and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
 Through calcification, corals are the backbone of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
-This complex reef framework  dissipates waves, protecting shorelines from currents and storms and preventing land loss due to erosion \citep{ferrario2014effectiveness,elliff2017coral}.
+This complex reef framework  dissipates most of incoming waves energy, protects shorelines from storms and floods, and prevents coastal erosion \citep{ferrario2014effectiveness,elliff2017coral}.
 Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass and mangrove ecosystems \citep{moberg1999ecological}.
@@
 
-Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2007coral}: it is estimated  that 30\% have already been severely damaged, and close to 60\% may be lost by 2030 \citep{hughes2003climate}.
-This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} (Fig.
+Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2007coral} (\todo{more recent ref}): it is estimated  that 30\% have already been severely damaged, and almost all could be lost under a 1.5$^\circ$C warming \citep{hughes2003climate,dixon2022future}.
+This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} (\todo{more recent ref}) (Fig.
 \ref{intro:cover}) is caused by both global and local anthropogenic stressors.
 The main global stressors are climate change induced ocean warming and acidification.
-Ocean warming cause recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
+Ocean warming causes recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
 Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral disease outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
-Furthermore, the increase of the concentration of atmospheric carbon dioxyde causes a decrease of the pH and carbonate concentration of seawater.
+Moreover, the increase of the concentration of atmospheric carbon dioxyde causes a decrease of the pH and carbonate concentration of seawater.
 As a result, coral calcification and growth decrease, which inhibits their capacity to build structurally complex habitats \citep{hoegh2007coral}.
@@
 Focusing on natural reefs only, \cite{brander2013total} estimated the total coral reef values Florida to be US\$174 millions.
-However, this value might be underestimated as non-use and indirect use values, such as support for coastal fisheries \citep{ault2006building} and coastal protection \citep{ferrario2014effectiveness} were not taken into account.
+However, this value might be strongly underestimated as non-use and indirect use values, such as support for coastal fisheries \citep{ault2006building} and coastal protection \citep{ferrario2014effectiveness} were not taken into account.
 
 Nevertheless, FCR has been heavily impacted by the human activities of the major metropolitan area of greater Miami.
-These impacts include densely populated coastlines, high visitor numbers, polluted terrestrial run-off and overfishing \citep{jackson2014status}.
-For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to cause the death of $>$560,000 corals \citep{cunning2019extensive}.
+These impacts include densely populated coastlines, large numbers of visitors, polluted terrestrial run-off and overfishing \citep{jackson2014status}.
+For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to have caused the death of $>$560,000 corals \citep{cunning2019extensive}.
 Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
@@
 
-Coral disease outbreaks are frequent in the Caribbean, which is considered a disease "hot spot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral diseases in the area \citep{green2000significance, harvell2007coral}.
+Coral disease outbreaks are frequent in the Caribbean, which is considered as a disease "hot spot" because of the fast emergence, high prevalence, wide geographic distribution, and virulence of coral diseases in the area \citep{green2000significance, harvell2007coral}.
 For instance, populations of  \textit{Acropora} spp.
@@
 Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both.
-In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig.
+In some cases, tissues bordering areas of chronic tissue loss show indistinct bands of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig.
 \ref{intro:sctld}).
@@
 Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
-This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine}.
-Hurricane might therefore interrupt larval exchanges between otherwise connected reef but also create new connectivity pathways.
-As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricane on wave-induced processes could inform the management of reefs under future climatic conditions.
+This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine} (\todo{other ref}).
+Hurricane might therefore interrupt larval exchanges between otherwise connected reefs but also create new connectivity pathways.
+As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricanes on wave-induced processes could inform the management of reefs under future climatic conditions.
 Furthermore, as incoming storm-generated waves dissipate over reefs, better understanding the protective role of coral reefs might promote their restoration as tool to reduce coastal flooding hazard \citep{roelvink2021coral}.
@@
 
-This thesis focuses on two main threats influencing the future and survival of FCR: the current outbreak of SCTLD and the hurricanes.
+This thesis focuses on two main threats that could jeopardize the future and survival of FCR: the current outbreak of SCTLD and hurricanes.
 These threats were studied by fulfilling four main objectives.
@@
 These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}.
-On the northernmost section of the FRT however, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}.
+However, on the northernmost section of the FRT, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}.
 The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}
@@
 The modeling tools developed in this thesis should therefore be able to capture them accurately.
-Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation.
+Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation features.
 For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}.
 Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}.
-Unstructured-mesh models offer a well-suited solution to cature the multi-scale nature of flow features in the FRT in a computationally tractable way.
+Unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way.
 They allow to locally increase the model resolution close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
 Nested grid models are another possible approach \citep{warner2010development} but they lack the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
+\todo{say something about HYCOM-GoM et HYCOM-FKeys...}
 
@@
 
-At the beginning this thesis, SLIM was not able to represent wave-current interactions under storm conditions.
-Heavy winds generate large wind-waves that disturb ocean conditions and interact non-linearly with currents \citep{liu2020impacts,wu2011fvcom}.
+At the beginning of this thesis, SLIM was not able to represent wave-current interactions under storm conditions.
+Heavy winds generate large wind-waves that disturb ocean conditions and interact non-linearly with the currents \citep{liu2020impacts,wu2011fvcom}.
 The wave-generated currents cause water level variations near shorelines and wave breaking points which, in turn, affect the motion and evolution of the waves \citep{longuet1970longshore, sikiric2013coupling}.
@@
 \section*{Outline}
-This thesis is mainly a collection of papers describing our answers to the four main research questions.
+This thesis is mainly a collection of publications describing our answers to the four main research questions.
 Chapter \ref{chap:sctld} describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT.
@@
 Chapter \ref{chap:onset} investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014.
-Finally, chapter \ref{chap:irma} describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
+Chapter \ref{chap:irma} describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
+Finally, chapter \ref{chap:conclusion} presents the conclusions of this thesis and perspectives for further works.
 
@@
 
+    \item Lopez-Gamundi, C., \textbf{Dobbelaere, T.}, Hanert, E., Harris, P.,Eberli G., and Purkis, S.
+(2022) Simulating Sedimentation on the Great Bahama Bank – Sources, Sinks, and Storm.
+\textit{Sedimentology}.
+submitted.
+
     \item Anselain, T., \textbf{Dobbelaere, T.}, Heggy, E.
@@
 in preparation.
-
-    \item Lopez-Gamundi, C., \textbf{Dobbelaere, T.}, Hanert, E., Harris, P.,Eberli G., and Purkis, S.
-(2022) Simulating Sedimentation on the Great Bahama Bank – Sources, Sinks, and Storm.
-\textit{Sedimentology}.
-submitted.
 

### 2022-02-13 include/abstract.tex
commentaires du patron

@@
 Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys.
-Assuming the dispersal of disease agents within neutrally material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling to eddy activity near the DRTO.
+Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current-Florida Current system.
 Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease.
 Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes.
-This thesis highlights the potential of models informed and confronted against field knowledge as powerful tools to inform the management of complex ecosystems such as the FCR.
+This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as the FCR.
 

### 2022-02-13 include/title.tex
few last things
commentaires du patron

@@
 \begin{minipage}[b]{0.3\textwidth}
-\hspace*{-0.8cm}\includegraphics[width=5.5cm]{./figures/logo_ucl.jpg}\vspace{0.85cm}
+\hspace*{-0.8cm}\includegraphics[width=5.5cm]{./figures/logo_ucl.jpg}\vspace{0.35cm}
 %\hspace*{3.4cm}\includegraphics[width=4.5cm]{./figures/orb.png}\vspace*{.1cm}
@@
 \begin{minipage}[b]{0.69\textwidth}
-
 	\begin{center}
@@
 
-\vspace*{1.2cm}
+\vspace*{2cm}
 \hspace{-0.8cm}
@@
 \begin{tabular}{ll}
-Prof. Emmanuel Hanert (Supervisor) & Université catholique de Louvain \\
-Prof. Eric Deleersnijder           & Université catholique de Louvain \\
-Dr. Jonathan Lambrechts            & Université catholique de Louvain\\
-Prof. Claire B. Paris-Limouzy      & University of Miami \\
-Dr. Christopher Thomas             & Université catholique de Louvain \footnote{Present address: Risk Management Solutions, London, UK}
+Prof. Emmanuel Hanert (Supervisor) & UCLouvain \\
+Prof. Eric Deleersnijder           & UCLouvain\\
+Dr. Jonathan Lambrechts            & UCLouvain \\
+Prof. Claire B. Paris-Limouzy      & University of Miami, USA \\
+Dr. Christopher Thomas             & Risk Management Solutions, UK \\
+Prof. Marnik Vanclooster           & UCLouvain
 \end{tabular}
@@
 \centering
-\large Louvain-la-Neuve, March 2022
+\large Louvain-la-Neuve, February 16, 2022
 \end{minipage}

### 2022-02-13 thesis.tex
commentaires du patron

--- 2022-02-13 thesis.tex
@@
   \input{include/abstract}
+  \clearemptydoublepage
+  %\input{include/thanks}
   %\clearemptydoublepage
-  \input{include/thanks}
-  \clearemptydoublepage
   \tableofcontents
@@
 %%%%%% 
-\mychapter{Modeling the spread of SCTLD in Florida} \label{chap:sctld}
-\chaptermark{Modeling the spread of SCTLD in Florida}  
+\mychapter{Processes driving the spread of SCTLD through the FRT} \label{chap:sctld}
+\chaptermark{Processes driving the spread of SCTLD in the FRT}  
 \input{chapters/sctld/sctld}
-\mychapter{Transmission of SCTLD from the Marquesas to the Dry Tortugas}\label{chap:drto}
+\mychapter{Transmission dynamics of SCTLD from the Marquesas and to the DRTO}\label{chap:drto}
 \chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
@@
 \input{chapters/onset/onset}
-\mychapter{Impacts of hurricanes on wave-induced ocean transport processes} \label{chap:irma}
+\mychapter{Impacts of hurricanes on transport processes in the Florida Keys} \label{chap:irma}
 \chaptermark{Wave-induced transport under hurricanes}  

### 2022-02-14 chapters/intro/intro.tex
ref added

@@
 While they account for less than 0.5\% of the ocean floor \citep{spalding1997new}, they are home for almost a third of the marine biodiversity \citep{moberg1999ecological}.
-Reef-related fisheries constitute around 10\% of the fish consumed by humans \citep{smith1978coral}(\todo{more recent stuff}) and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
+Reef-related fisheries account to up to 40\% of the total marine
+fish landed value of some regions \citep{teh2013global} and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
 Through calcification, corals are the backbone of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
@@
 
-Despite their importance, coral reefs have experienced a massive system-wide decline \citep{hoegh2007coral} (\todo{more recent ref}): it is estimated  that 30\% have already been severely damaged, and almost all could be lost under a 1.5$^\circ$C warming \citep{hughes2003climate,dixon2022future}.
-This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,bruno2007regional} (\todo{more recent ref}) (Fig.
+Despite their importance, coral reefs have experienced a massive system-wide decline: it is estimated  that 50\% have already been severely damaged, and almost all could be lost under a 1.5$^\circ$C warming \citep{hoegh2019people,dixon2022future}.
+This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,pandolfi2011projecting,bruno2007regional,perry2013caribbean,dove2020ocean} (Fig.
 \ref{intro:cover}) is caused by both global and local anthropogenic stressors.
@@
 This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}.
-Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,aronson2001white}.
+Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,gardner2005hurricanes,aronson2001white}.
 They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects, tweel2014contribution}.
@@
 Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
-This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine} (\todo{other ref}).
+This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{hendee1998champ}.
 Hurricane might therefore interrupt larval exchanges between otherwise connected reefs but also create new connectivity pathways.

### 2022-02-14 chapters/intro/intro.tex
added sentence about FKeyS-HYCOM

@@
 Unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way.
-They allow to locally increase the model resolution close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
+They allow to locally refine the model resolution to $\sim~100$ m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
+In contrast, the best resolution currently available in the Florida Keys using structured-grid models is $\sim900$ m (FKeyS-HYCOM, \citealp{kourafalou2012florida}).
 Nested grid models are another possible approach \citep{warner2010development} but they lack the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
-\todo{say something about HYCOM-GoM et HYCOM-FKeys...}
 

### 2022-02-14 chapters/intro/intro.tex
added a few refs

@@
 fish landed value of some regions \citep{teh2013global} and hundreds of millions of people depend on coral reefs for their livelihood or part of their protein intake \citep{salvat1992coral,hoegh2019people}.
-Through calcification, corals are the backbone of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological}.
+Through calcification, corals are the backbone of a complex three-dimensional habitat that serves as spawning, breeding and feeding area for many species \citep{moberg1999ecological,rogers2014vulnerability}.
 This complex reef framework  dissipates most of incoming waves energy, protects shorelines from storms and floods, and prevents coastal erosion \citep{ferrario2014effectiveness,elliff2017coral}.
-Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass and mangrove ecosystems \citep{moberg1999ecological}.
+Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass and mangrove ecosystems \citep{moberg1999ecological,guannel2016power}.
 
@@
 These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the corals with most of the energy needed to meet their metabolic demands \citep{muscatine1977reef}.
-However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral}.
+However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral,van2022global}.
 This process, called bleaching, causes the coral to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}.
@@
 
-Despite their importance, coral reefs have experienced a massive system-wide decline: it is estimated  that 50\% have already been severely damaged, and almost all could be lost under a 1.5$^\circ$C warming \citep{hoegh2019people,dixon2022future}.
-This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,pandolfi2011projecting,bruno2007regional,perry2013caribbean,dove2020ocean} (Fig.
+Despite their importance, coral reefs have experienced a massive system-wide decline: it is estimated  that 50\% have already been severely damaged \citep{hoegh2019people}, and almost all could be lost under a 1.5$^\circ$C warming \citep{dixon2022future}.
+This global decline in live coral cover \citep{gardner2003long,pandolfi2003global,pandolfi2011projecting,perry2013caribbean,dietzel2021population} (Fig.
 \ref{intro:cover}) is caused by both global and local anthropogenic stressors.
-The main global stressors are climate change induced ocean warming and acidification.
-Ocean warming causes recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial}.
-Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral disease outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean}.
+The main global stressors are climate change induced ocean warming and acidification \citep{dove2020ocean,figueiredo2021global,dixon2022future}.
+Ocean warming causes recurrent mass bleaching events at an increasing frequency \citep{connell199730, hughes2018spatial,van2022global}.
+Furthermore, thermal stress can increase the susceptibility of corals to disease, leading to an increase in the incidence of coral disease outbreaks \citep{harvell2002climate,bruno2007thermal,muller2012caribbean,howells2020annual}.
 Moreover, the increase of the concentration of atmospheric carbon dioxyde causes a decrease of the pH and carbonate concentration of seawater.
-As a result, coral calcification and growth decrease, which inhibits their capacity to build structurally complex habitats \citep{hoegh2007coral}.
-Local stressors are numerous and include pollution \citep{loya1980effects}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,sheppard2017biology}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical}.
+As a result, coral calcification and growth decrease, which inhibits their capacity to build structurally complex habitats \citep{hoegh2007coral,albright2016reversal}.
+Local stressors are numerous and include pollution \citep{loya1980effects,van2011chemical}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,fabricius2005effects}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical,shantz2020overfishing}.
 
@@
 Instead, the challenge is now to maintain the biological functions of coral reefs through quickly changing environmental conditions.
-This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting}.
+This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting,kleypas2021designing}.
 The aim of the present dissertation was therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support ecosystem management in Florida, the third largest barrier reef of the world.
@@
 For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to have caused the death of $>$560,000 corals \citep{cunning2019extensive}.
-Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
+Florida's coral reefs have therefore
+d significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
 Populations of previously dominant reef-building acroporids almost totally disappeared throughout the Caribbean and Florida \citep{aronson2001white}, and coral cover dropped from 40 to 60\% in 1975 to $<$7\% in the Florida Keys \citep{jackson2014status} and $<$3\% in the northern section of the FRT \citep{walton2018impacts}.

### 2022-02-14 chapters/conclusions/conclusions.tex
encore plus de refs

@@
 This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived from field observations on large scale systems.
-As such, models informed by and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as the FCR.
+As such, models informed by and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as the FCR \citep{foster2012connectivity}.
 
@@
 Larval connectivity promotes the resilience of the reef network as it allows the recolonization of disturbed reefs following extreme events such as mass bleaching, hurricanes or disease outbreaks.
-Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration.
+Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration \citep{botsford2009connectivity,mumby2011reserve}.
 Identifying reefs that are both good sources and sinks of larvae the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.

### 2022-02-14 chapters/conclusions/conclusions.tex
relecture I

@@
 Our model had to be able to accurately capture ocean circulation patterns at different scales to study transport processes in the FCR.
-First, the large-scale Loop Current-Florida Current system had to be accurately reproduced to correctly capture its impact on eddy formation near the DRTO and the Florida Keys.
+First, the large-scale Loop Current/Florida Current system had to be accurately reproduced to correctly capture its impact on eddy formation near the DRTO and the Florida Keys.
 Second, the model had to reproduce ocean circulation at the reef-scale to capture the recirculation and acceleration of currents between reefs and islands.
@@
 This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived from field observations on large scale systems.
-As such, models informed by and confronted against field knowledge are a powerful tool to inform the management of complex ecosystems such as the FCR \citep{foster2012connectivity}.
+As such, models informed by and confronted against field knowledge \citep{foster2012connectivity} are a powerful tool to inform the management of complex ecosystems such as the FCR.
 
@@
 
-Finally, we developed a couple wave-current model to study the impact of Hurricane Irma on transport processes in the Florida Keys.
+Finally, we developed a coupled wave-current model to study the impact of Hurricane Irma on transport processes in the Florida Keys.
 This coupled model correctly reproduced the observed waves and current dynamics during the passage of the hurricane.
@@
 However, wave-induced transport was dominated by the Stokes drift, which yielded an impact four times larger than the radiation stress gradient.
-Results of chapter \ref{chap:irma} strongly advocates for the use of coupled wave-currents to accurately model the dispersal of coral larvae under storm conditions.
+Results of chapter \ref{chap:irma} strongly advocate for the use of coupled wave-currents to accurately model the dispersal of coral larvae under storm conditions.
 Although we built the tools to study the impact of hurricane on the exchanges of larvae between reefs by developing the coupled SLIM+SWAN, we did not evaluate the impact of Irma on coral connectivity by lack of time.
@@
     \begin{list}{}{\setlength{\topsep}{0pt}}
-        \item Results of a coupled wave-current model suggest that wave-induced transport under storm conditions dominated by Stokes drift can yield differences in particle trajectories of up to 20 km.
-Wave-current interactions should thus be taken into to accurately model the dispersal of coral larvae during hurricanes.
+        \item Results of a coupled wave-current model suggest that wave-induced transport under storm conditions, dominated by Stokes drift, can yield differences in particle trajectories of up to 20 km.
+Wave-current interactions should thus be taken into account to accurately model the dispersal of coral larvae during hurricanes.
     \end{list}
@@
 Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration \citep{botsford2009connectivity,mumby2011reserve}.
-Identifying reefs that are both good sources and sinks of larvae the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.
+Identifying reefs that are both good sources and sinks of larvae in the connectivity network indicates the optimal sites for coral outplants in the context of active restoration.
 Furthermore, knowledge about larval connectivity would allow us to build robust and sustainable networks of connected marine protected areas.
-As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.
+As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine,figueiredo2021global}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.
 This combined information would allow us to identify reefs that are good larval sources and have low vulnerability to the disease.
@@
 A first improvement could therefore be to consider classes of corals with different susceptibilities and hence use different infection and removal parameter values in the model.
-Such improvement would be technically challenging as it would increase the number of variables of model.
-The development of the model could therefore require the use of more advance graph theory tools such as the multilayer networks \citep{kivela2014multilayer,pilosof2017multilayer}.
-Furthermore, increasing the number of parameters will thus make their calibration more complex.
+Such a development could require the use of more advanced graph theory tools such as multilayer networks \citep{kivela2014multilayer,pilosof2017multilayer}.
+Furthermore, this improvement would increase the number of variables and parameters of the model, which  would make its calibration more complex.
 Moreover, this new modeling framework would require estimates of the distribution of the different coral species across the FRT.
-Again, compiling this data and deriving realistic estimates would represent a significant challenge.
-Adding species-specific or location-specific parameters in the model would also allow the evaluation of the impact of disease mitigating actions.
+Compiling this data and deriving realistic estimates would represent a significant challenge.
+Interestingly, adding species-specific or location-specific parameters in the model would also allow the evaluation of the impact of disease mitigating actions.
 Lesion treatment using antibiotics could for instance be taken into account in the model by locally reducing the transmission and/or removal parameters.
@@
 Yet, again, adding processes implies an increase in the number of parameters to be calibrated.
-This modifications could lead to an endemic disease dynamics.
+These modifications could lead to an endemic disease dynamics.
 
 \subsection*{Better understanding the protective role of corals against storm waves}
-Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipated over reefs.
+Simulation results from the coupled wave-current model suggest that storm waves were significantly dissipated over reefs.
 The modeled significant wave height under the center of the hurricane reduced from 8 m to $<1$ m over the barrier reef during the passage of the hurricane (Fig.
@@
 However, modelling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes.
-The smallest scales that model such as SLIM and SWAN can simulate is limited by the intrinsic assumptions in the model formulations, i.e., hydrostatic for SLIM and phase-averaged for SWAN.
+The smallest scales that models such as SLIM and SWAN can simulate is limited by the intrinsic assumptions in the model formulations, i.e., hydrostatic for SLIM and phase-averaged for SWAN.
 These assumptions are no longer valid as soon as the horizontal scales of motion become comparable to the vertical scales \citep{marshall1997hydrostatic}.
@@
 This procedure has recently been applied to ocean modelling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}.
-Additionally, this multi-physics modeling frameworks could be uses to better capture the motion of coral larvae over reef.
-This would improve the modeled coral connectivity, as the the initial motion of larvae over reefs then drives larval transport following mass spawning events and initiate coral connectivity.
+Additionally, this multi-physics modeling frameworks could be used to better capture the motion of coral larvae over reef.
+This would improve the modeled coral connectivity, as the initial motion of larvae over reefs then drives larval transport following mass spawning events, and initiate coral connectivity.
 

### 2022-02-14 chapters/drto/article.tex
relecture I

@@
 This suggests that the disease must first spread through the outer reefs to reach them.
-The source indicator values confirm that disease transmission to the DRTO is more likely to originate from southern reefs at the interface between the Marquesas and the Western Marquesas, and reefs on the northwestern end of the Western Marquesas.
+The source indicator values confirm that disease transmission to the DRTO is more likely to originate from southern reefs at the interface between the Marquesas and the Western Marquesas, and reefs on the northwestern end of the Western Marquesas (Fig.
+\ref{fig:fig4_drto}b).
 

### 2022-02-14 chapters/intro/intro.tex
encore plus de refs
relecture I

@@
 For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to have caused the death of $>$560,000 corals \citep{cunning2019extensive}.
-Florida's coral reefs have therefore
-d significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
+Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
 Populations of previously dominant reef-building acroporids almost totally disappeared throughout the Caribbean and Florida \citep{aronson2001white}, and coral cover dropped from 40 to 60\% in 1975 to $<$7\% in the Florida Keys \citep{jackson2014status} and $<$3\% in the northern section of the FRT \citep{walton2018impacts}.
@@
 Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}.
-Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
+However, these upwelling also reduce sea surface temperature and can thus alleviate the thermal stress experienced by corals \citep{wilkinson2008status}.
+Finally, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}.
 This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{hendee1998champ}.
@@
 The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT.
-These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability.
+These variations of the FC position have been shown to have no seasonal pattern, and a strong interannual variability.
 The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}.
-These eddies significantly impact the ecology of the southwest Florida Shelf by causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}.
+These eddies significantly impact the ecology of the southwest Florida Shelf by causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution,limouzy1997translocation,sponaugle2005florida, kourafalou2012florida}.
 Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}.
@@
 
-The aforementioned large-scale ocean circulation significantly impact transport processes in the FRT.
+The aforementioned large-scale ocean circulation significantly impacts transport processes in the FRT.
 The modeling tools developed in this thesis should therefore be able to capture them accurately.
@@
 Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}.
-Unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way.
+A reference ocean model of the FRT is the 3D HYbrid Coordinate Ocean Model (HYCOM, \citealp{chassignet2007hycom}), or more specifically the nested high resolution model FKEYS-HYCOM, that covers the Florida Straits, South Florida and Keys with a $\sim$900 m resolution \citep{kourafalou2012florida}.
+This model is embedded within the coarser GoM-HYCOM, that models the whole GoM and Florida Straits using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}.
+However as a structured-grid model, the finer resolution of HYCOM is limited by computational resources and the nested grid approach lacks the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
+In contrast, unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way.
 They allow to locally refine the model resolution to $\sim~100$ m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
-In contrast, the best resolution currently available in the Florida Keys using structured-grid models is $\sim900$ m (FKeyS-HYCOM, \citealp{kourafalou2012florida}).
-Nested grid models are another possible approach \citep{warner2010development} but they lack the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
 
@@
 This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}.
-Moreover, by coupling SLIM with the HYbrid Coordinate Ocean Model (HYCOM-GOMl0.04, \citealp{chassignet2007hycom}), SLIM was able to indirectly represent the baroclinic dynamics of the FC.
+Moreover, by coupling SLIM with GOM-HYCOM, the model was able to indirectly represent the baroclinic dynamics of the FC.
 Using this coupling, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs.
-In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents.
+In this study, coral larvae were modeled using a Lagrangian particle tracking model \citep{van2018lagrangian} forced by SLIM currents.
 To answer our different research questions, we adapted this larval transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.

### 2022-02-14 include/abstract.tex
relecture I

@@
 Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys.
-Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current-Florida Current system.
+Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current/Florida Current system.
 Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease.

### 2022-02-15 chapters/conclusions/conclusions.tex
commentaires de Xaymara + qqs figures refaites

@@
 Additionally, this multi-physics modeling frameworks could be used to better capture the motion of coral larvae over reef.
-This would improve the modeled coral connectivity, as the initial motion of larvae over reefs then drives larval transport following mass spawning events, and initiate coral connectivity.
+This would improve the modeled coral connectivity, as the initial motion of larvae over reefs then drives larval transport following mass spawning events, and initiates coral connectivity.
 

### 2022-02-15 chapters/intro/intro.tex
commentaires de Xaymara + qqs figures refaites

@@
 As a result, coral calcification and growth decrease, which inhibits their capacity to build structurally complex habitats \citep{hoegh2007coral,albright2016reversal}.
-Local stressors are numerous and include pollution \citep{loya1980effects,van2011chemical}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,fabricius2005effects}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical,shantz2020overfishing}.
+Local stressors are numerous and include pollution \citep{loya1980effects,van2011chemical}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, jones2015effects,jones2019sedimentcunning2019extensive}, nutrient run-off \citep{hughes2003climate,fabricius2005effects}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical,shantz2020overfishing}.
 

### 2022-02-15 chapters/onset/article.tex
commentaires du patron
commentaires de Xaymara + qqs figures refaites

@@
 Some of the first signs of SCTLD were reported in September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}.
-They derived a relationship between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
+In this study, a relationship was derived between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
 It was therefore hypothesized that the epidemic started near Virginia Key and then spread to the neighboring reef, both north and south.
@@
 % SEDIMENTS
-Sediments released by dredging can affect the biological functions of corals in numerous ways  through turbidity and sedimentation \citep{erftemeijer2012environmental, jones2015effects}.
+Sediments released by dredging can affect the biological functions of corals in numerous ways  through turbidity and sedimentation \citep{erftemeijer2012environmental,jones2015effects}.
 Increased turbidity caused by the suspended sediments reduces the light available to symbiotic zooxanthellae, leading to reduced coral cover and growth.
-Sedimentation, on the other hand, can cause smothering or burial of coral polyps \citep{erftemeijer2012environmental}.
+Sedimentation, on the other hand, can cause smothering or burial of coral polyps \citep{erftemeijer2012environmental,jones2015effects,jones2019sediment}.
 Furthermore, both sedimentation and turbidity can significantly reduce larval recruitment by inhibiting settlement and reducing larval survival in the water column \citep{jones2015effects}.
-These effects are stronger with fine-grained sediments, as they cause a stronger light reduction \citep{fourney2017additive}.
+These effects are stronger with fine-grained sediments, as they cause a stronger light reduction \citep{storlazzi2015influence,fourney2017additive}.
 Additionally, fine-grained sediments such as silts have high nutrient contents, which can lead to an increased microbial activity, eventually causing anoxic conditions in the immediate vicinity of corals \citep{weber2012mechanisms}.
@@
 Further, they suggested that the onset of the disease might have been linked to a leaking discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant (WWTP) located off Virginia Key.
-However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease was transported to the monitoring site of Virginia Key by sediments released during the dredging.
+However, as sediments can act as vector for SCTLD \citep{studivan2022reef}, there is also a possibility that the causative agents of the disease were transported to the monitoring site of Virginia Key by sediments released during the dredging.
 This scenario can be evaluated using a sediment transport model.
@@
 
-As sediment dispersion is dependent on the grain size, we modeled the dispersal of five sediment classes to represent to impact of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m.
+As sediment dispersion is dependent on the grain size, we modeled the dispersal of five sediment classes to represent impacts of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m.
 We performed a  different simulation for each class, with the grain size randomly drawn from a uniform distribution over the corresponding size range.
@@
 \ref{fig:onset_mesh}A, see pictures of coral colony in appendix \ref{onset:appendice}).
-The impact of dredging was assessed by counting the number of sediment particles originating from each dredging that were transported within 500 m of all five site.
+The impact of dredging was assessed by counting the total number of sediment particles originating from each dredging that were transported within 500 m of all five site.
 This distance was chosen to match the scale of the 500 m $\times$ 500 m reef polygons used in the model \citep{dobbelaere2020coupled}.
@@
 	\includegraphics[width=\textwidth]{chapters/onset/figures/timeline.png}
-	\caption{Fraction of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N.
+	\caption{Fraction of the total number of sediment particles released by each type of dredge that drifted within 500 of HBSC1-CP and N.
 Sunnny Isles with grain sized of 5-50 $\mu$m (\textbf{A}) and 50-100 $\mu$m (\textbf{B}).
 \textbf{C}: Temporal distribution of the simulated dredging operations.
-The dates of the first reported disease signs at each monitoring site are shown in brown.
+The dates of the first reported disease signs at each monitoring site are shown with dotted vertical lines in dark red.
 With the exception of HBSC1-CP, sites located south to the dredged channel were barely impacted by the dredging.
@@
 
-Our results suggest dredging operations had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission by sediments was extremely unlikely.
+Our results suggest that the dredging operations had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission by sediments was extremely unlikely.
 However, our results show the sediments released by the dredging operations had a non-negligible impact on the sites of N.
@@
 	\section{Reported disease signs at site HBSC1-CP in May 2014}\label{onset:appendice}
+	\begin{figure}
+		\centering
 		\includegraphics[width=\textwidth]{chapters/onset/figures/hbsc1_cp.png}
+		\caption{Pictures of a monitored colony at site HBSC1-CP. The first signs of the disease probably appeared between May 25 and May 30. By June 4, the whole colony was completely dead}
+	\end{figure}
 \end{subappendices}

### 2022-02-15 include/title.tex
ref added
added a few refs
commentaires de Xaymara + qqs figures refaites

@@
 
-\vspace*{2cm}
+\vspace*{3cm}
 \hspace{-0.8cm}
@@
 \centering
-\LARGE\bfseries Modeling transport processes to inform the management of Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
+\LARGE\bfseries Florida's Coral Reef - Impact of ocean transport on management approaches % How modeling tools can inform coral reef management in the Florida Reef Tract
 \end{minipage}
@@
 
-\vspace*{1.5cm}
+\vspace*{0.5cm}
 \begin{minipage}{\textwidth}
@@
 Dr. Christopher Thomas               & Risk Management Solutions, UK \\
-Prof. Marnik Vanclooster           & UCLouvain
+Prof. Marnik Vanclooster (President) & UCLouvain
 \end{tabular}
 }
+
+\textcolor{white}{Et pour la tradition, Régis Malpas \& Marc-Henry Comans}
+
 \end{center}

### 2022-02-15 thesis.tex
commentaires de Xaymara + qqs figures refaites

@@
 \usepackage[pdftex,linktocpage]{hyperref}		%For hyperlinks within doc, and setting doc info
-\hypersetup{pdftitle={Modeling transport processes to inform the managment of Florida's Coral Reef}}
+\hypersetup{pdftitle={Florida's Coral Reef - Impact of ocean transport on management approaches}}
 \hypersetup{pdfauthor={Thomas DOBBELAERE}}
-\hypersetup{pdfsubject={Modeling transport processes to inform the managment of Florida's Coral Reef}}
+\hypersetup{pdfsubject={Florida's Coral Reef - Impact of ocean transport on management approaches}}
 \hypersetup{pdfkeywords={Biophysical modeling, Florida Coral Reef, SCTLD, SLIM}}

### 2022-02-15 chapters/conclusions/conclusions.tex
relecture II

@@
 This interruption was likely linked to eddy activity near the DRTO and was consistent with the apparent stalling of the outbreak in the region in 2020.
-This confirms the role played the Loop Current/Florida Current system in the control of the connectivity pathways in the Florida Keys and the DRTO.
+This confirms the role played by the Loop Current/Florida Current system in the control of the connectivity pathways in the Florida Keys and the DRTO.
 Furthermore, sub-reefs predicted to be more vulnerable to SCTLD in the disease networks were consistent with the observed geographic distribution of diseased corals in the DRTO.
@@
 This highlights the potential of biophysical models to understand the driving mechanics of ongoing marine disease outbreaks.
-Furthermore, indicators such as the vulnerablity and source indicators developed in chaper \ref{chap:drto} could be useful to predict and/or mitigate the spread of SCTLD in other territories of the Caribbean.
+Furthermore, metrics such as the vulnerablity and source indicators developed in chaper \ref{chap:drto} could be useful to predict and/or mitigate the spread of SCTLD in other territories of the Caribbean.
 
 Using a sediment transport model, we investigated the impact of the PMDDP on the onset of the SCTLD outbreak in 2014.
-Our results suggest that the dredging operations had close to no direct impact at Virginia Key monitoring site, identified as the first affected reef.
+Our results suggest that the dredging operations had close to no direct impact at Virginia Key monitoring site, identified as the reef from which the disease started its propagation.
 However, reefs affected earlier could have been infected by chopped rock particles produced by non conventional dredging operations in the channel.
@@
 Such adverse impacts should be taken into account when planning dredging operations.
-The results of chapter \ref{chap:onset} suggest that the potential to initiate a coral disease outbreak should now also be taken into account when considering future dredging project.
+The results of chapter \ref{chap:onset} suggest that the potential to initiate a coral disease outbreak should now also be taken into account when considering future dredging projects.
 
@@
 Results of chapter \ref{chap:irma} strongly advocate for the use of coupled wave-currents to accurately model the dispersal of coral larvae under storm conditions.
-Although we built the tools to study the impact of hurricane on the exchanges of larvae between reefs by developing the coupled SLIM+SWAN, we did not evaluate the impact of Irma on coral connectivity by lack of time.
+Although we built the tools to study the impact of hurricane on the exchanges of larvae between reefs by developing the coupled SLIM+SWAN model, we did not evaluate the impact of Irma on coral connectivity by lack of time.
 
@@
 Vulnerability to disease can be evaluated using the (weighted) in-degree of reefs in the disease connectivity network, as reefs with strong and numerous incoming connections are more likely to receive more disease agents from many connected reefs.
-Therefore, reefs with high restoration potential would correspond to reefs with large larval(weighted) out-degree and low disease (weighted) in-degree.
+Therefore, reefs with high restoration potential would correspond to reefs with large larval (weighted) out-degree and low disease (weighted) in-degree.
 This approach have been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.
@@
 For instance, \cite{ferrario2014effectiveness} estimated that coral reefs could provide wave attenuation comparable to artificial defenses such as breakwaters for a much lower cost.
-However, modelling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes.
+However, modeling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes.
 The smallest scales that models such as SLIM and SWAN can simulate is limited by the intrinsic assumptions in the model formulations, i.e., hydrostatic for SLIM and phase-averaged for SWAN.
 These assumptions are no longer valid as soon as the horizontal scales of motion become comparable to the vertical scales \citep{marshall1997hydrostatic}.
-For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modelling framework.
+For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modeling framework.
 Nonhydrostatic models are more expensive to	 run as they require a much finer spatial resolution of the order of a meter and have a more complex physics.
@@
 Machine learning is based on the principle of extracting information from data, which can either be observational data or high-resolution model outputs, and deriving relationships between datasets without the need for physics-based models.
-This procedure has recently been applied to ocean modelling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}.
+This procedure has recently been applied to ocean modeling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}.
 Additionally, this multi-physics modeling frameworks could be used to better capture the motion of coral larvae over reef.
@@
 The work presented in this thesis has demonstrated the capabilities of modeling tools to inform the management of Florida's Coral Reef.
-Models allow the evaluation of hypotheses derived from field observations on large-scale systems and can help to deducing the characteristics and underlying mechanisms of complex processes.
+Models allow the evaluation of hypotheses derived from field observations on large-scale systems and can help infer the characteristics and underlying mechanisms of complex processes.
 Furthermore, they can be used in a predictive manner to evaluate the impacts of changing environmental conditions on a given ecosystem.
-Additionally, the framework developed in this thesis relies on the same tools as already widely used larval transport models and could be combined with them to optimize the design of coral reef protection and restoration campaigns.
-Given this high potential, further research to improve the model and synergistic work with field an lab studies may well yield results which could help maintaining the biological functions of coral reefs as long as possible through the Anthropocene.
+Additionally, the framework developed in this thesis relies on the same widely used tools as larval transport models.
+Both approaches could thus be easily combined to optimize the design of coral reef protection and restoration campaigns.
+The development of such modeling tools coupled with field an lab studies may thus inform reef management to maintain the biological functions of coral reefs through the Anthropocene.

### 2022-02-15 chapters/intro/intro.tex
relecture II

--- 2022-02-14 chapters/intro/intro.tex
@@
 This complex reef framework  dissipates most of incoming waves energy, protects shorelines from storms and floods, and prevents coastal erosion \citep{ferrario2014effectiveness,elliff2017coral}.
-Additionally, wave dissipation creates lagoon and sedimentary environment, providing favorable conditions for the growth of seagrass and mangrove ecosystems \citep{moberg1999ecological,guannel2016power}.
+Additionally, wave dissipation creates lagoons and sedimentary environments, providing favorable conditions for the growth of seagrass and mangrove ecosystems \citep{moberg1999ecological,guannel2016power}.
 
@@
 This process, called bleaching, causes the coral to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}.
-Bleached corals are physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high coral mortality \citep{hughes2018spatial}.
+Bleached corals are then physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high coral mortality \citep{hughes2018spatial}.
 Decline in coral health and cover pushes the reef ecosystem to a macroalgal-dominated state \citep{hughes2003climate,mumby2007thresholds}.
@@
 As a result, coral calcification and growth decrease, which inhibits their capacity to build structurally complex habitats \citep{hoegh2007coral,albright2016reversal}.
-Local stressors are numerous and include pollution \citep{loya1980effects,van2011chemical}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, cunning2019extensive}, nutrient run-off \citep{hughes2003climate,fabricius2005effects}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical,shantz2020overfishing}.
+Local stressors are numerous and include pollution \citep{loya1980effects,van2011chemical}, coastal development and the associated sediment and wastewater fluxes \citep{erftemeijer2012environmental, jones2015effects,jones2019sediment,cunning2019extensive}, nutrient run-off \citep{hughes2003climate,fabricius2005effects}, and overfishing of herbivorous species controlling algae growth \citep{jackson2001historical,shantz2020overfishing}.
 
@@
 This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting,kleypas2021designing}.
-The aim of the present dissertation was therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support ecosystem management in Florida, the third largest barrier reef of the world.
+The aim of the present dissertation was therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support the management of Florida's Coral Reef, the third largest barrier reef of the world.
 
@@
 
-The Florida Reef Tract (FRT) spans over approximately 577 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St.
+The Florida Reef Tract (FRT) spans over approximately 577 km from the Dry Tortugas (DRTO), west of the Florida Keys, to the St.
 Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf} (Fig.
@@
 The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is fairly low (below 3\%)\citep{banks2008reef,walton2018impacts}.
-The southern half-section is a chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO on the the boundary of the West Florida Shelf (WFS).
+The southern half-section is a chain of limestone islands (the Keys), that extends from the southern tip of the Florida mainland southwest to the DRTO on the boundary of the West Florida Shelf (WFS).
 The Upper Keys are remnants of ancient coral reefs and the Lower Keys are sand bars \citep{hoffmeister1968geology}.
@@
 Moreover, recreational use of the reefs was evaluated to bring 36,500 full and part-time jobs.
-Focusing on natural reefs only, \cite{brander2013total} estimated the total coral reef values Florida to be US\$174 millions.
+Focusing on natural reefs only, \cite{brander2013total} estimated the total coral reef values of Florida to be US\$174 millions.
 However, this value might be strongly underestimated as non-use and indirect use values, such as support for coastal fisheries \citep{ault2006building} and coastal protection \citep{ferrario2014effectiveness} were not taken into account.
@@
 For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to have caused the death of $>$560,000 corals \citep{cunning2019extensive}.
-Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}.
+Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status,walton2018impacts}.
 Populations of previously dominant reef-building acroporids almost totally disappeared throughout the Caribbean and Florida \citep{aronson2001white}, and coral cover dropped from 40 to 60\% in 1975 to $<$7\% in the Florida Keys \citep{jackson2014status} and $<$3\% in the northern section of the FRT \citep{walton2018impacts}.
@@
 A reference ocean model of the FRT is the 3D HYbrid Coordinate Ocean Model (HYCOM, \citealp{chassignet2007hycom}), or more specifically the nested high resolution model FKEYS-HYCOM, that covers the Florida Straits, South Florida and Keys with a $\sim$900 m resolution \citep{kourafalou2012florida}.
-This model is embedded within the coarser GoM-HYCOM, that models the whole GoM and Florida Straits using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}.
+This model is embedded within the coarser GOM-HYCOM, that models the whole GoM and Florida Straits using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}.
 However as a structured-grid model, the finer resolution of HYCOM is limited by computational resources and the nested grid approach lacks the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
@@
 This is usually achieved by spectral wave models, which describe the evolution of the wave energy spectrum.
-An objective of this thesis was therefore the coupling of SLIM with the spectral wave model Simulating WAve Nearshore (SWAN, \citealp{booij1999third}), specifically developed for coastal applications and relies on numerical schemes adapted to small-scale, shallow water regions.
+An objective of this thesis was therefore the coupling of SLIM with the spectral wave model Simulating WAve Nearshore (SWAN, \citealp{booij1999third}), specifically developed for coastal applications and relying on numerical schemes adapted to small-scale, shallow water regions.
 Further details on the models and their coupling is given in chapter \ref{chap:irma}.
@@
 \section*{Outline}
-This thesis is mainly a collection of publications describing our answers to the four main research questions.
+This thesis is mainly a collection of publications describing our answers to our four main research questions.
 Chapter \ref{chap:sctld} describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT.

### 2022-02-15 chapters/irma/article.tex
chapter on drto added to thesis
relecture II

@@
 
-The main questions we want to answer are the following: (1) How important are wave-current interactions during a tropical cyclone?
+The main questions we want to answer in this study are the following: (1) How important are wave-current interactions during a tropical cyclone?
 (2) What effect do they have on the transport of drifting material?
@@
 \end{equation}
-where $N=E/\sigma$ is the wave action density and $E$ is the wave energy spectrum; $\theta$ is the wave propagation direction; $\sigma$ is the intrinsic wave frequency; $\mathbf{c}_g$ is the wave group velocity, $\mathbf{u}=\mathbf{U}/H$ is SLIM depth-averaged current velocity; $c_\theta$ and $c_\sigma$ are the propagation velocities in spectral space due to refraction and shifting in frequency due to variations in depth and currents; and $S_{in}$, $S_{ds}$, and $S_{nl}$ respectively represent wave growth by wind, wave decay and nonlinear transfers of wave energy through four and three-wave interactions, \ie quadruplets and triplets.
+where $N=E/\sigma$ is the wave action density and $E$ is the wave energy spectrum; $\theta$ is the wave propagation direction; $\sigma$ is the intrinsic wave frequency; $\mathbf{c}_g$ is the wave group velocity, $\mathbf{u}=\mathbf{U}/H$ is SLIM depth-averaged current velocity; $c_\theta$ and $c_\sigma$ are the propagation velocities in spectral space due to refraction and shifting in frequency due to variations in depth and currents; and $S_{in}$, $S_{ds}$, and $S_{nl}$ respectively represent wave growth by wind, wave decay and nonlinear transfers of wave energy through four and three-wave interactions, \ie~quadruplets and triplets.
 The wave spectra were discretized with 48 direction bins and 50 frequency bins logarithmically distributed from 0.03 to 2 Hz.
@@
 SLIM's governing equations are integrated using an implicit time integration scheme while SWAN is unconditionally stable \citep{dietrich2012performance}, allowing both models to be run with relatively large time steps.
-In this study, the stationary version of SWAN was used, \ie the first term of Eq.
+In this study, the stationary version of SWAN was used, \ie~the first term of Eq.
 (\ref{eq:swan}) was set to zero.

### 2022-02-15 chapters/onset/article.tex
relecture II

@@
 Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region \citep{richardson1998coral, sutherland2004disease, aronson2001white, harvell2007coral, brandt2009dynamics}.
-One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}.
-First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022connecting} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
+One of the latest and the most damaging outbreaks to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}.
+First observed off the coasts of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022connecting} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}.
 Although the causative agent of the disease remains unknown, the hydrodynamics is likely to play an important role in its propagation as both modeling studies and ex situ experiments showed evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}.
@@
 % POM
-Some of the first signs of SCTLD were reported in September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}.
+Some of the first signs of SCTLD were reported on September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}.
 In this study, a relationship was derived between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
@@
 
-While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located 4.7 nautical miles offshore.
+While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located $\sim$6.7 km offshore.
 However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
-\citep{usace2017} provideda back-of-the-envelope estimation that this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel .
-Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.
+\cite{usace2017} provided a back-of-the-envelope estimation that this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel .
+Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging operations \citep{barnes2015sediment}.
 
@@
 The monitoring sites considered in the present study are shown by blue dots.
-The date where SCTLD was first observed at these sites is given between brackets.
+The date of the first reported signs of SCTLD at these sites is given between brackets.
 The Ocean Dredge Material Disposal Site (ODMDS) is shown in red and the discharge pipe of the Miami Central District Municipal Wastewater Treatment Plant in green.
@@
 In this model, particles undergo a combination of horizontal and vertical motions.
-The vertical is mostly driven by gravity, with heavier particles sinking faster.
+The vertical dynamics is mostly driven by gravity, with heavier particles sinking faster.
 Once they have settled, particles can be resuspended when shear stress exceeds   the critical Schields parameter, as parameterized by \cite{soulsby1997threshold}.
@@
 
-As sediment dispersion is dependent on the grain size, we modeled the dispersal of five sediment classes to represent impacts of fine- to coarse-grained particles: ($i$) 5-50 $\mu$m, ($ii$) 50-100 $\mu$m, ($iii$) 100-200 $\mu$m, ($iv$) 200-300 $\mu$m, and ($v$) 300-400 $\mu$m.
+As sediment dispersion is dependent on the grain size, we modeled the dispersal of five sediment classes to represent impacts of fine- to coarse-grained particles: (\textit{i}) 5-50 $\mu$m, (\textit{ii}) 50-100 $\mu$m, (\textit{iii}) 100-200 $\mu$m, (\textit{iv}) 200-300 $\mu$m, and (\textit{v}) 300-400 $\mu$m.
 We performed a  different simulation for each class, with the grain size randomly drawn from a uniform distribution over the corresponding size range.
@@
 In the absence of information about the exact time of the dredging, sediment particles were released from the dredging location during a whole day at a rate of 80 particles/hour in the model.
-To account for the motion of spider barges between the dredging site and disposal site, particles were released every 500 m along a straight line joining the dredging location to the ODMDS (see Fig.
+To account for the motion of spider barges between the dredging and disposal sites, particles were released every 500 m along a straight line joining the dredging location to the ODMDS (see Fig.
 \ref{fig:onset_mesh}A) for every dredging operation labelled as SB.
@@
 The impact of sedimentation was quantified by computing the cumulated concentration of settled particles within the same computational grid.
-This cumulated concentration was normalized by the total number of simulated time steps to obtained an averaged concentration of deposited particles over the simulated period.
+This cumulated concentration was normalized by the total number of simulated time steps to obtain an averaged concentration of deposited particles over the simulated period.
 Higher values of this indicator would indicate larger sediment deposition over a longer cumulated time.
@@
 The fifth site was the monitoring station HBSC1-CP, used throughout the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
-\ref{fig:onset_mesh}A, see pictures of coral colony in appendix \ref{onset:appendice}).
+\ref{fig:onset_mesh}A, see supplementary material \ref{onset:appendice}).
 The impact of dredging was assessed by counting the total number of sediment particles originating from each dredging that were transported within 500 m of all five site.
 This distance was chosen to match the scale of the 500 m $\times$ 500 m reef polygons used in the model \citep{dobbelaere2020coupled}.
-The resulting numbers of particles was then divided by the total number of sediment particles released by each type of dredging operation.
+The resulting numbers of particles were then divided by the total number of sediment particles released by each type of dredging operation.
 Larger values of this indicator would suggest a greater impact of a given type of dredging operation at a given monitoring sites.
@@
 These connectivity matrices can be interpreted as large graphs whose vertices are 500 m $\times$ 500 m sub-reef polygons and whose edges represent disease connectivity pathways.
-Evaluating the possibility of disease propagation from any sub-reef $A$ to any sub-reef $B$ is therefore equivalent to evaluating the existence of paths, \ie sequences of connected vertices in the network starting from $A$ and reaching $B$.
+Evaluating the possibility of disease propagation from any sub-reef $A$ to any sub-reef $B$ is therefore equivalent to evaluating the existence of paths, \ie~sequences of connected vertices in the network starting from $A$ and reaching $B$.
 As computing all possible paths is not computationally tractable, we limited ourselves to the computation of shortest paths from any given reefs to the Virginia Key monitoring site.
@@
 Such a function requires the definition of a weight $w_{ij}$ for the edge connecting reef $i$ to reef $j$.
-We chose $w_{ij} = 1-\tilde{C}_{ij}$, where $\tilde{C}_{ij}$ is the probability of disease propagation from reef $i$ to reef $j$, so that "shorter" edges of the networks (\ie connectivity pathways with smaller weights) correspond to connections with larger disease propagation probability.
+We chose $w_{ij} = 1-\tilde{C}_{ij}$, where $\tilde{C}_{ij}$ is the probability of disease propagation from reef $i$ to reef $j$, so that "shorter" edges of the networks (\ie~connectivity pathways with smaller weights) correspond to connections with larger disease propagation probability.
 The probability of a given path was then defined as the mean connection probability of the edges composing this path.
@@
 This suggests that the observed turbidity was mostly due to fine silts.
-The modeled occurrence of plumes within 15 km of the dredged matched the presence/absence data of \cite{cunning2019extensive} in 82.7\% of cases and the total area where plumes were observed was about 230 km$^2$, consistent with the $\sim228$ km$^2$ estimated by \cite{barnes2015sediment}.
+The modeled occurrence of plumes within 15 km of the dredged matched the presence/absence data of \cite{cunning2019extensive} in 82.7\% of cases and the total area where plumes were observed was about 230 km$^2$, consistent with the $\sim$228 km$^2$ estimated by \cite{barnes2015sediment}.
 Modeled plumes mostly occurred north of the dredged channel, as particles were driven northward under the action of the Florida Current, with plume occurring during 20\% of the simulated days near the monitoring site of N.
 Sunny Isles.
-Plume frequencies were lower ($\sim 5$\%) over offshore reefs but reached of up to 50\% over alongshore reefs.
+Plume frequencies were lower ($\sim$5\%) over offshore reefs but reached of up to 50\% over alongshore reefs.
 These alongshore reefs correspond to regions of important sedimentation (Fig.
@@
 \ref{fig:onset_bar}C).
-However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles that reached N.
+However, SB events occurred at an equivalent frequency but resulted in 4 times fewer particles reaching N.
 Sunny Isles.
@@
 
-Applying the same analysis to the discharge pipe of Miami Central District Municipal WWTP suggests that only wastewater released from a leak in the direct vicinity of the Virginia Key could have impacted the site.
+Applying the same analysis to the discharge pipe of Miami Central District Municipal WWTP suggested that only wastewater released from a leak in the direct vicinity of the Virginia Key could have impacted the site.
 Fig.
@@
 As in \cite{barnes2015sediment}, the total area of plumes in our model was about 230 km$^2$.
-Furthermore, the model results reproduced the presence/absence within 15 km of the dredged channel derived by \cite{cunning2019extensive} with an accuracy of 82.7\%.
-Such good agreement was only obtained for grain sizes in the range 5-50 $\mu$m, suggesting that fine silts were the main drivers of the turbidity generated by dredging, which is consistent with previous studies \citep{fourney2017additive}.
+Furthermore, the model results reproduced the plume presence/absence within 15 km of the dredged channel derived by \cite{cunning2019extensive} with an accuracy of 82.7\%.
+Such a good agreement was only obtained for grain sizes in the range 5-50 $\mu$m, suggesting that fine silts were the main drivers of the turbidity generated by dredging, which is consistent with previous studies \citep{storlazzi2015influence,fourney2017additive}.
 Despite this good agreement, the distribution of plumes predicted by our model tends to be shifted northward compared to these previous studies.
@@
 % LIMITATION + SHOW THAT N. SUNNY ISLE AND HBSC1-CP WERE SIGNIFICANTLY IMPACTED
-A limitation of our study is that the amount of released sediments might be overestimated, as conventional and non-conventional dredging are treated in the same way in the model.
+A limitation of our study is that conventional and non-conventional dredging are treated in the same way in the model.
 For all dredging types, sediment particles were released at the same rate in the model.
 Although conventional dredging was reported to release fine-grained sediments in the water column through dewatering and overflow from barges \citep{jones2016assessing}, the quantity of dredged material lost in the water was limited by the use of pumping mechanisms.
-By contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rock particles to be released in the water column.
+In contrast, for non-conventional dredging, the suction mechanism was turned off, causing all chopped rock particles to be released in the water column.
 The numbers of particles reaching the monitoring sites were therefore likely underestimated for non conventional dredging operations.
@@
 This implied that the propagation of the disease from HBSC1-CP to Virginia Key through hydrodynamics-driven transport of disease agents was possible during these four months.
-All these pathways used reefs located further southward as stepping stones.
-Therefore, the propagation of outbreak to reach Virginia Key starting from HBSC1-CP required these intermediary reefs to get first infected by disease agents released by disease colonies of HBSC1-CP.
-Once diseased, these colonies would then send diseased agents to affect the next colonies of the connectivity pathway until the outbreak reached Virginia Key.
+All these pathways used reefs located further south as stepping stones.
+Therefore, the propagation of SCTLD to Virginia Key starting from HBSC1-CP required these intermediary reefs to be first infected by disease agents released from HBSC1-CP.
+Once diseased, these colonies would then send disease agents to affect the next colonies of the connectivity pathway until the outbreak reached Virginia Key.
 Assuming an averaged transmission time of the order of 5-10 days \citep{dobbelaere2020coupled}, disease propagation from HBSC1-CP to Virginia Key might have required several months.
@@
 
-In the present study, we evaluated the impact of the dredging operations that took place the impact of the deepening of the Port of Miami shipping channel on the neighboring coral reefs.
+In the present study, we evaluated the impact of the dredging operations that took place during the deepening of the Port of Miami shipping channel on the neighboring coral reefs.
 This evaluation was performed using a quasi-3D sediment model forced by high-resolution currents from the coastal ocean model SLIM.
-As the first signs of SCTLD were reported during this dredge project and since sediments may be a SCTLD vector, we aimed at answering two main questions.
-First, we wanted to evaluate the fraction of sediments produced by the PortMiami Deep Dredge Project that reached Virginia Key and other monitoring sites where the disease was reported.
+As the first signs of SCTLD were reported during the PMDDP and since sediments may be a SCTLD vector, we aimed at answering two main questions.
+First, we wanted to evaluate the fraction of sediments produced by the PMDDP that reached Virginia Key and other monitoring sites where the disease was reported.
 Second, we evaluated the possibility of waterborne transmission of SCTLD to Virginia Key from other reefs impacted by the dredging.
@@
 Our results suggest that the dredging operations had close to no direct impact on the coral reefs of Virginia Key, indicating that disease transmission by sediments was extremely unlikely.
-However, our results show the sediments released by the dredging operations had a non-negligible impact on the sites of N.
+However, our results show that the sediments released by the non conventional dredging operations had a non-negligible impact on the sites of N.
 Sunny Isles and HBSC1-CP, where signs of disease were observed before the outbreak was first reported at Virginia Key.
@@
 
-This study suggests that the expansion of the Port of Miami might have played a role in the onset of the outbreak of SCTLD in sites where the disease was reported during the first half of 2014.
+This study suggests that the expansion of the Port of Miami might have played a role in the onset of the outbreak of SCTLD at sites where the disease was reported during the first half of 2014.
 Furthermore, it shows that the disease could then have spread from these sites to Virginia Key, from where SCTLD was then reported to propagate both north and south to the neighboring reefs.
 Our work brings new insight on the impact of a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
+They suggest that the potential to initiate a coral disease outbreak should be taken into account when considering future dredging projects.
 
@@
 		\includegraphics[width=\textwidth]{chapters/onset/figures/hbsc1_cp.png}
-		\caption{Pictures of a monitored colony at site HBSC1-CP. The first signs of the disease probably appeared between May 25 and May 30. By June 4, the whole colony was completely dead}
+		\caption{Pictures of a monitored colony at site HBSC1-CP. The first signs of the disease appeared between May 25 and May 30. By June 4, the whole colony was completely dead}
 	\end{figure}

### 2022-02-15 chapters/onset/onset.tex
commentaires de Xaymara + qqs figures refaites
relecture II

@@
 The disease then spread to the entirety of FCR, likely through waterborne transmission.
-Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease.
+Although the causative agent of SCTLD remains unknown, there is evidence that sediments can act as vector for the disease.
 Here we evaluate whether sediments produced or resuspended during dredging operations could have reached reefs where the disease was first reported and assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease.
@@
 A particular attention was paid to the non conventional dredging operations performed without pumping the produced chopped rock particles.
-Our result suggest that the dredging had close to no direct impact on the coral reefs of Virginia Key.
+Our results suggest that the dredging had close to no direct impact on the coral reefs of Virginia Key through sediment flux and sedimentation.
 However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites.

### 2022-02-15 chapters/sctld/article.tex
commentaires de Xaymara + qqs figures refaites
relecture II

@@
 where $L_{ij}$ is the distance between origin reef $i$ and destination reef $j$.
-Another measure of the spreading potential of reef $j$ is its out-degree, \ie the product of the number of connections originating from reef $j$ by the quantity of disease agents it sent to the network.
+Another measure of the spreading potential of reef $j$ is its out-degree, \ie~the product of the number of connections originating from reef $j$ by the quantity of disease agents it sent to the network.
 This indicator was obtained by computing the number of non-zero entries of row $i$ in the potential connectivity matrix $C$ and multiplying it with $\sum_j C_{ij}$.
@@
 This indicator is given by $\sum_{j} \tilde{C}_{ij}$.
-Finally, the isolation of reef $i$ in the network was given by its self-recruitment, \ie the fraction of disease agents settling on reef $i$ that originated from reef $i$, computed by $C_{ii}/\sum_jC_{ji}$.
+Finally, the isolation of reef $i$ in the network was given by its self-recruitment, \ie~the fraction of disease agents settling on reef $i$ that originated from reef $i$, computed by $C_{ii}/\sum_jC_{ji}$.
 A large self-recruitment value indicates that little infectious material produced elsewhere settled on the reef and thus that it was isolated from the rest of the network.
@@
 A larger value of threshold $I_0$ corresponds to a greater resistance of corals to the disease, and therefore a slower spread of the disease within reef $j$.
-Coral reproduction and natural (\ie non SCTLD-related) death rates are not taken into account in this model, which amounts to assume that they balance each other.
+Coral reproduction and natural (\ie~non SCTLD-related) death rates are not taken into account in this model, which amounts to assume that they balance each other.
 For this study the same values were used for $\beta$ and $\beta'_0$.
@@
 In order to solve Eqs.
-\ref{eq:epidemio}, initial conditions were needed, \ie fractions of susceptible, infectious and recovered corals at the beginning of the simulated period.
+\ref{eq:epidemio}, initial conditions were needed, \ie~fractions of susceptible, infectious and recovered corals at the beginning of the simulated period.
 This information was constructed from 9 different field-collected datasets: (i) Coral Reef Evaluation and Monitoring Project (CREMP; 2014–2017), (ii) CREMP Presence/Absence Data (CREMP P\_A; 2016–2017), (iii) Southeast Florida Coral Reef Evaluation and Monitoring Project (SECREMP; 2014–2017), (iv) Florida Reef Resilience Program Disturbance Response Monitoring (FRRP; 2014–2017), (v) Hurricane Irma Rapid Reef Assessment (IRMA; 2017, \cite{viehman2018}), (vi) the Southeast Florida Action Network citizen science program (SEAFAN; 2014–2017), and (vii) the Southefrn Coral Disease Margin field effort (2017 and 2018; \cite{neely2018surveying}), (viii) Mote Marine Laboratory’s Field operations data (2018-2019) and (ix) data compiled through the citizen science BleachWatch program (2018).
@@
     \centering
-    \includegraphics[width=.99\linewidth]{chapters/sctld/figures/hull_example_2.jpg}
+    \includegraphics[width=.99\linewidth]{chapters/sctld/figures/hull_example.png}
     \caption{Method used to compute the disease front displacement during a simulated time interval.\textbf{a.} Concave hull $H_0$ of the diseased sub-reefs at the beginning of the simulated period.
@@
 This value, being close to 1, suggests modeled diseased individuals were removed from the system almost as fast as susceptible individuals became diseased.
-This situation caused the fraction of infectious corals on the reefs to remain fairly low (\ie $\leq 0.4\%$) through time.
+This situation caused the fraction of infectious corals on the reefs to remain fairly low (\ie~$\leq 0.4\%$) through time.
 These results suggest that only a small fraction of the colonies caused the disease to spread on the reef during the outbreak.
@@
 The difference between the two values can be explained by field measurement uncertainties as well as the inability to perfectly mimic field conditions in laboratory.
-In the present study, the same values were used for inter- and intra-reef 'infection' rates $\beta$ and $\beta_0'$, \ie disease agents 'infectiousness' was assumed to remain unchanged during their journey from reef to reef.
+In the present study, the same values were used for inter- and intra-reef 'infection' rates $\beta$ and $\beta_0'$, \ie~disease agents 'infectiousness' was assumed to remain unchanged during their journey from reef to reef.
 This assumption is consistent with sensitivity analysis results, which showed that reducing the value of $\beta$ strongly impeded disease propagation through the network.
-This outcome suggests that, to reproduce the observed spread, inter- and intra-reef transmission rates must have similar magnitude, \ie that the causative agent is not degraded while traveling from reef to reef.
+This outcome suggests that, to reproduce the observed spread, inter- and intra-reef transmission rates must have similar magnitude, \ie~that the causative agent is not degraded while traveling from reef to reef.
 Given that the etiology of SCTLD is still unknown, the lack of degradation of the SCTLD causative agent over distances of tens of kilometers between reefs may help to narrow the search.
@@
 Moreover, even when winds are pushing particles along the reef line, these particles spend less time over the same region than particles driven by mean and bottom currents.
-Smaller amounts of particle mass will therefore settle on reef polygons, leading to lower connection strengths in the potential connectivity matrix, \ie lower exchange of infectious material between reefs.
+Smaller amounts of particle mass will therefore settle on reef polygons, leading to lower connection strengths in the potential connectivity matrix, \ie~lower exchange of infectious material between reefs.
 Hence, despite being able to transport the disease over greater distances, surface currents are less efficient to drive the propagation of the disease.

### 2022-02-15 chapters/sctld/sctld.tex
updated ref for onset
relecture II

@@
 J., Holstein, D.
-M., $\&$ Hanert, E.
+M., \& Hanert, E.
 (2020).
 Coupled epidemio-hydrodynamic modeling to understand the spread of a deadly coral disease in Florida.
-\textit{Frontiers in Marine Science},1016.
+\textit{Frontiers in Marine Science}, 7, 1016.
 \href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.3389/fmars.2020.591881}.
@@
 Here we show how a high-resolution bio-physical model coupled with a modified patch Susceptible-Infectious-Removed epidemic model can characterize the potential causative agent(s) of the disease and its vector.
-In the present study, the agent is assumed to be transported within composite material (e.g., coral mucus, dying tissues, and/or resuspended sediments) driven by currents and potentially persisting in the water column for extended periods of time.
+In the present study, the agent is assumed to be transported within composite material (\textit{e.g.}~coral mucus, dying tissues, and/or resuspended sediments) driven by currents and potentially persisting in the water column for extended periods of time.
 In this framework, our simulations suggest that the SCTLD is likely to be propagated within neutrally buoyant material driven by mean barotropic currents.

### 2022-02-15 include/abstract.tex
commentaires de Xaymara + qqs figures refaites
relecture II

@@
 Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes.
-This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as the FCR.
-
-%The population of corals throughout the world has declined dramatically and many coral species are now listed as threatened Avoiding global warming being no longer an option, reef managers now focus on damage control and restoration strategies.
-The Florida Reef Tract, currently facing an unprecedented coral disease outbreak and where healthy corals grown in nurseries are replanted on damaged reefs, is a good application case of such reef management strategies.
-In order to maximize the efficiency of these efforts, a good knowledge of exchanges of infectious materials and larvae between reefs is required.
-However, connectivity patterns between reefs remain mostly unknown.
-To address this issue, we will use the high-resolution ocean model SLIM to simulate connectivity patterns between reefs and then identify reefs with the highest restoration potential as well the ones that are most at risk of being infected by the outbreak.
-
-%The dispersal of coral larvae away from their natal habitats is an important process for coral reef ecosystems, but remains poorly understood and hard to gauge.
-This thesis presents a spatially explicit numerical modelling approach to simulate larval dispersal in Australia's Great Barrier Reef (GBR), and in doing so to estimate connectivity between the thousands of reefs that comprise it.
-A high resolution, finite element ocean model, SLIM, is used to build a biophysical model of larval dispersal capable of estimating larval exchange between reefs in large sections of the the GBR.
-Tools from network science are employed to investigate the community structure of the GBR's connectivity network, and these tools are used to reveal the presence of clusters of highly inter-connected reefs.
-Subsequently, the model is used to estimate the extent of connectivity between reefs of different morphologies and depths in the GBR.
-Finally, the possible impacts of climate change on connectivity in the GBR are assessed, and estimates are made of the likely changes to coral larval dispersal patterns over the coming century.
-
-%Titan, Saturn’s largest moon, is a unique object in the Solar system as much as it has a substantial atmosphere, a surface with a complex interplay of geological processes and an outer ice shell overlyinga subsurface ocean.
-The climate on this icy satellite boasts a multi-phase hydrological cycle where methane plays a role similar to that of water on Earth.
-Surface lakes and seas filled with liquid methane and ethane are found in the polar regions.
-The subsurface ocean, on the other hand, is filled with liquid water.
-This liquid layer lying beneath Titan’s surface allows for the large surface deformation observed.
-
-%The first objective of this thesis was to adapt an Earth-based geophysical and environmental model, SLIM (www.climate.be/slim), to Titan’s specific conditions in order to study the tidal motion in the surface lakes and seas.
-The modified model was applied to the largest lake in the southern hemisphere, Ontario Lacus, and the two largest seas, Kraken and Ligeia Maria, which are located in the northern hemisphere.
-The predicted surface elevation and velocity fields are part of the data needed to develop an exploration mission focusing on the surface lakes and seas of Titan.
-The normal modes of Ontario Lacus were also numerically studied.
-While resonantly forced modes could generate significant liquid motion, the natural periods are much shorter than the period of the astronomical forcings.
-Therefore, only atmospheric forcings could resonantly force the normal modes.
-Strong wind conditions corresponding to a stormy event are required to generate a surface elevation resulting in significant shoreline variations.
-
-%Then, the model was used to study the tides of Titan’s global subsurface ocean.
-To this end, the shallow water equations and the 3D hydrostatic equations under the Boussinesq approximation were modified in order to take into account the ice shell lying at the top of the ocean.
-The method was adapted to solve the equations on a sphere.
-The effects of the shell are represented by adding a surface friction and a surface pressure term.
-The shell decreases the ocean surface elevation and slows down the flow without modifying significantly the global patterns of these fields.
-The magnitude of these variations depends on the mechanical behaviour of the ice shell.
-Surface and bottom heat fluxes play a significant role in the liquid motion of the ocean.
-The interactions between the tidal motion and the thermally driven flow resulting from the surface heat flux were studied by means of the 3D version of SLIM.
-The surface heat flux significantly impacts the velocity field, both in terms of magnitude and orientation while the influence on the ocean’s surface elevation is small.
-These results could be useful for astrobiologists paying attention to the ocean habitability and to validate hypotheses about Titan’s internal structure.
+This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as Florida's Coral Reef.

### 2022-03-17 chapters/conclusions/conclusions.tex
new title

@@
 Our coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by representing the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents.
-It confirmed the result of previous ex-situ studies that showed evidence of waterborne transmission of the SCTLD with an average transmission time of the order of 10 days.
+It confirmed the results of previous ex-situ studies that showed evidence of waterborne transmission of the SCTLD with an average transmission time of the order of 10 days.
 Furthermore, by using coral resistance to the disease as a parameter, our model results suggested that, on average, corals had a rather low resistance against the disease.
-This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived from field observations on large scale systems.
+This confirmation of experimental results by model results illustrates the important potential of models to evaluate hypotheses derived from field observations on large scale systems.
 As such, models informed by and confronted against field knowledge \citep{foster2012connectivity} are a powerful tool to inform the management of complex ecosystems such as the FCR.
@@
 Therefore, reefs with high restoration potential would correspond to reefs with large larval (weighted) out-degree and low disease (weighted) in-degree.
-This approach have been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.
+This approach has been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.
 

### 2022-03-17 chapters/drto/article.tex
another fig modification
new title

@@
 Previous study of the spatio-temporal dynamics of SCTLD between West Palm Beach and Marathon showed that the epizootic propagated at a rate of $\sim$100 m/day to the north and $\sim$92 m/day to the south, hitting the Middle Florida Keys in 2017 \citep{muller2020spatial}.
-The disease was then reported in the Lower Keys in summer 2018 \citep{frrp2018}; near Key West in winter 2018 \citep{frrp2018}; on the easternmost boundary of the Marquesas subregion in October 2019 \citep{frrp2019}; and at its western end, approximately 11 km.
-west of halfmoon shoal, in August 2020 \citep{frrp2020}.
+The disease was then reported in the Lower Keys in summer 2018 \citep{frrp2018}; near Key West in winter 2018 \citep{frrp2018}; on the easternmost boundary of the Marquesas subregion in October 2019 \citep{frrp2019}; and at its western end, approximately 11 km west of halfmoon shoal, in August 2020 \citep{frrp2020}.
 SCTLD was observed propagating faster than 100 m/day within the Marquesas between October 2019 and August 2020 and would have reached the reefs of the Dry Tortugas (DRTO) by February 2021 if it had spread at a similarly high propagation rate \citep{kramer2019map}.
@@
 Stony Coral Tissue Loss Disease data was obtained by using the Florida Reef Resilience Program's Disturbance Response Monitoring Report for 2019 and 2020 \citep{frrp2019,frrp2020} as well as the Atlantic and Gulf Rapid Reef Assessment (AGGRA) resources for SCTLD reported throughout the Caribbean, an open access spatial and temporal database curated by experts in Caribbean coral reef ecology \citep{kramer2019map}.
-Assumptions associated with SCTLD included the following: $(i)$ SCTLD is different from other tissue loss diseases previously reported on corals (i.e.
-white plague, white syndrome) based on its ecology and case definition \citep{noaa2018}, $(ii)$ SCTLD is transmitted, in part, by ocean currents \citep{aeby2019pathogenesis, muller2020spatial,eaton2021measuring} and $(iii)$ SCTLD was along the western most reef area of the Lower Florida Keys in October 2019, reached the western most part of the Marquesas in August 2020, and did not occur in DRTO until May 2021 based on curated data provided by the FRRP reports and the AGGRA open access database.
+Assumptions associated with SCTLD included the following: (\textit{i}) SCTLD is different from other tissue loss diseases previously reported on corals (i.e.
+white plague, white syndrome) based on its ecology and case definition \citep{noaa2018}, (\textit{ii}) SCTLD is transmitted, in part, by ocean currents \citep{aeby2019pathogenesis, muller2020spatial,eaton2021measuring} and (\textit{iii}) SCTLD was along the western most reef area of the Lower Florida Keys in October 2019, reached the western most part of the Marquesas in August 2020, and did not occur in DRTO until May 2021 based on curated data provided by the FRRP reports and the AGGRA open access database.
 
@@
 An entry $C_{ij}$ of the connectivity matrix represents the probability that disease agents produced on sub-reef $i$ settle on sub-reef $j$.
-These sub-reefs were defined by extracting polygons from the “coral reefs and hard bottom” layer of the Unified Florida Reef Tract Map \citep{fwc2017unified} and further dividing them into 500 m $\times$ 500 m squares, generating a total of $16,823$ sub-reef polygons.
+These sub-reefs were defined by extracting polygons from the “coral reefs and hard bottom” layer of the Unified Florida Reef Tract Map \citep{fwc2017unified} and further dividing them into 500 m $\times$ 500 m squares, generating a total of 16,823 sub-reef polygons.
 Since the connectivity matrices had more than 200 million entries, they were not directly interpreted.
@@
 
-To quantify the connectivity between the Marquesas and the DRTO, we defined 3 sub-regions within the sub-reefs of the western part of FCR: $(i)$ the Marquesas, $(ii)$ the Western Marquesas and $(iii)$ the DRTO (Fig.
+To quantify the connectivity between the Marquesas and the DRTO, we defined 3 sub-regions within the sub-reefs of the western part of FCR: (\textit{i}) the Marquesas, (\textit{ii}) the Western Marquesas and (\textit{iii}) the DRTO (Fig.
 \ref{fig:fig1_drto}d).
@@
 We then defined a connectivity indicator by multiplying the number of these direct edges by their mean connection probability.
-This indicator was computed monthly between May 2018 and May 2021 to evaluate the connectivity $(i)$ from the Marquesas to the DRTO, $(ii)$ from the Western Marquesas to the DRTO and $(iii)$ from the Marquesas to the Western Marquesas.
+This indicator was computed monthly between May 2018 and May 2021 to evaluate the connectivity (\textit{i}) from the Marquesas to the DRTO, (\textit{ii}) from the Western Marquesas to the DRTO and (\textit{iii}) from the Marquesas to the Western Marquesas.
 
@@
     \centering
-    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig4.png}
+    \includegraphics[width=\textwidth]{chapters/drto/figures/fig4.png}
     \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020.

### 2022-03-17 chapters/drto/drto.tex
updated ref for onset
new title

@@
 J., McEachron L., Williams, S.
-D., $\&$ Hanert, E.
+D., \& Hanert, E.
+(2022).
 Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas.
-accepted for publication in \textit{Frontiers in Marine Science}.\\ \href{https://doi.org/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.
+\textit{Frontiers in Marine Science}, 9.
+\href{https://www.frontiersin.org/article/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.
 
@@
 \begin{abstract}
-    For the last seven years, Florida's Coral Reef (FCR) has suffered from widespread and severe coral loss caused by stony coral tissue loss disease (SCTLD).
+    For the last seven years, Florida's Coral Reef (FCR) has suffered from widespread and severe coral loss caused by the stony coral tissue loss disease (SCTLD).
 First observed off the coast of Miami-Dade county in 2014, the outbreak has since spread throughout the entirety of FCR and some areas of  the Caribbean.

### 2022-03-17 chapters/intro/intro.tex
new title

@@
 However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral,van2022global}.
-This process, called bleaching, causes the coral to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}.
+This process, called bleaching, causes the corals to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}.
 Bleached corals are then physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high coral mortality \citep{hughes2018spatial}.
@@
 This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting,kleypas2021designing}.
-The aim of the present dissertation was therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support the management of Florida's Coral Reef, the third largest barrier reef of the world.
+The aim of the present dissertation is therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support the management of Florida's Coral Reef, the third largest barrier reef of the world.
 
@@
 The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}.
-First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects} (Fig.
+First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire FRT and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects} (Fig.
 \ref{intro:propagation}).
@@
     \includegraphics[width=\textwidth]{chapters/intro/figures/hurricane_strikes.jpg}
-    \caption{Total number of major hurricanes strikes in Florida between 1900-2010.
+    \caption{Total number of major hurricanes strikes in Florida between 1900 and 2010.
 Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
@@
 A reference ocean model of the FRT is the 3D HYbrid Coordinate Ocean Model (HYCOM, \citealp{chassignet2007hycom}), or more specifically the nested high resolution model FKEYS-HYCOM, that covers the Florida Straits, South Florida and Keys with a $\sim$900 m resolution \citep{kourafalou2012florida}.
-This model is embedded within the coarser GOM-HYCOM, that models the whole GoM and Florida Straits using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}.
+This model is embedded within the coarser GOM-HYCOM, that models the whole GoM and the Straits of Florida using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}.
 However as a structured-grid model, the finer resolution of HYCOM is limited by computational resources and the nested grid approach lacks the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.
 In contrast, unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way.
-They allow to locally refine the model resolution to $\sim~100$ m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
+They allow to locally refine the model resolution to $\sim$100 m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.
 

### 2022-03-17 chapters/irma/irma.tex
updated ref for onset
new title

@@
 }%
-\item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., & Hanert, E.
+\item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., \& Hanert, E.
 (2022).
@@
 Their effect is particularly acute in coastal areas where they cause extensive damage leading to an influx of debris, sediments and waste to the sea.
-However, most operational coastal ocean models do not represent heavy-wind transport processes correctly \modif{if} the hydrodynamics is not coupled with the wind-generated waves.
+However, most operational coastal ocean models do not represent heavy-wind transport processes correctly if the hydrodynamics is not coupled with the wind-generated waves.
 This may lead to significant errors in ocean simulations under tropical cyclone conditions.

### 2022-03-17 chapters/onset/article.tex
new title

@@
 Some of the first signs of SCTLD were reported on September 26th, 2014, near Virginia Key by \cite{precht2016unprecedented}.
-In this study, a relationship was derived between the time of the first outbreak at monitored reefs and their geographic distance to Virginia Key.
-It was therefore hypothesized that the epidemic started near Virginia Key and then spread to the neighboring reef, both north and south.
+In this study, a relationship was derived between the time of the first outbreak at the monitored reefs and their geographic distance to Virginia Key.
+It was therefore hypothesized that the epidemic started near Virginia Key and then spread to the neighboring reefs, both north and south.
 However, earlier signs of diseases were already reported north of Virginia Key in June 2014, at the monitoring site of N.
@@
 
-While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located $\sim$6.7 km offshore.
+While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located $\sim$8.7 km offshore.
 However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}.
@@
 
-The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys2020fine,dobbelaere2020coupled,dobbelaere2022connecting}.
+The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys2020fine,dobbelaere2020coupled,dobbelaere2022impacts,dobbelaere2022connecting}.
 SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features.
@@
 We performed a  different simulation for each class, with the grain size randomly drawn from a uniform distribution over the corresponding size range.
-The density of each sediment particle was derived from their size using the formula of \cite{hamilton1982sound}.
+The density of each sediment particle was derived from its size using the formula of \cite{hamilton1982sound}.
 Furthermore, all particles were differentiated based on the type of dredge that produced them.
@@
 
-Dredging operations performed during the expansion of PoM were characterized in our dataset by a date, a location and  a type of dredging operation (Fig.
+Dredging operations performed during the expansion of PoM were characterized in our dataset by a date, a location and  a type of dredge (Fig.
 \ref{fig:onset_mesh}B).
@@
 Sunny Isles and an increase of the concentration of settled sediments near HBSC1-CP and Virginia Key.
-The normalized cumulated deposition was around 0.01 particles/m$^2$ at site HBSC1-CP for all grain sizes while it increased from $2.5\times 10^{-7}$ to $1.6\times 10^{-3}$ with grain size at Virginia Kay.
+The normalized cumulated deposition was around 0.01 particles/m$^2$ at site HBSC1-CP for all grain sizes while it increased from $2.5\times 10^{-7}$ to $1.6\times 10^{-3}$ with increasing grain size at Virginia Kay.
 For all grain sizes, no sedimentation occurred near Rainbow Reef and Emerald Reef.
@@
 
-The evaluation of the impact of dredging at the monitoring sites indicated that dredging did not significantly impact the stations located south of the dredge channel, except at site HBSC1-CP (Fig.
+The evaluation of the impact of dredging at the monitoring sites indicated that dredging did not significantly impact the stations located south of the dredged channel, except at site HBSC1-CP (Fig.
 \ref{fig:onset_bar}).
@@
 % POSSIBILITY OF DISEASE PROPAGATION THROUGH WATERBORNE TRANSMISSION
-In all simulations, the expansion of PoM had limited direct impact on Virginia Key, identified as the site where the outbreak initiated in September 2014 \citep{precht2016unprecedented}.
+In all simulations, the expansion of PoM had limited direct impact on Virginia Key, identified as the site where the outbreak was initiated in September 2014 \citep{precht2016unprecedented}.
 For all grain sizes, the fraction of sediment particles reaching Virginia Key remained below 1\%.
@@
 Sunny Isles and HBSC1-CP, where signs of disease were observed before the outbreak was first reported at Virginia Key.
-This suggests that the sediments might have triggered the onset of the diseased at these two sites.
+This suggests that the sediments might have triggered the onset of the disease at these two sites.
 Furthermore, using a previously developed bio-physical model that successfully reproduced the observed spread of SCTLD, we show that there was a possibility of waterborne disease propagation from HBSC1-CP to Virginia Key.
 
-This study suggests that the expansion of the Port of Miami might have played a role in the onset of the outbreak of SCTLD at sites where the disease was reported during the first half of 2014.
-Furthermore, it shows that the disease could then have spread from these sites to Virginia Key, from where SCTLD was then reported to propagate both north and south to the neighboring reefs.
+This study suggests that the PMDDP might have played a role in the onset of the outbreak of SCTLD at Virginia Key, from where the disease was then reported to propagate through the entire FCR.
 Our work brings new insight on the impact of a major dredging project in Florida on the onset of one of the worst coral outbreaks on record in the Caribbean.
-They suggest that the potential to initiate a coral disease outbreak should be taken into account when considering future dredging projects.
+It suggests that the potential to initiate a coral disease outbreak should be taken into account when considering future dredging projects.
 

### 2022-03-26 chapters/conclusions/conclusions.tex
modif post-defense - part I

@@
 
+\modif{\section{Developments of the model}}
+
+\modif{Answering the aforementioned research questions required the development of new modeling tools.
+As this thesis heavily relies on Lagrangian transport modeling, several features were added to the particle tracker of SLIM.
+Although the transport of the particles in itself was already implemented in the model, we developed features to characterize the state of the particles and compute connectivity indicators.
+The particle state can be used to represent biological traits such as the competency of coral larvae or abiotic characteristics such as the decaying mass of diseased material or the diameter of sediment particles.
+We developed tools to derive connectivity indicators such as the shortest paths between a pair sub-reef polygons in the obtained  connectivity network.
+Additionally, we developed an epidemiological model using the computed connectivity matrices to simulated the spread of the stony coral tissue loss disease throughout Florida.
+Finally, the particle tracker was made more flexible by adding the capability to transport particles using outputs from larger-scale ocean models such as HYCOM.}
+
+\modif{Additionally, the coupling of SLIM with SWAN required some modifications of SLIM.
+First, we implemented a new parameterization of the wind drag coefficient to account for the formation of spume at the ocean-atmosphere interface in heavy wind conditions.
+Furthermore, a term was added to the model's equations to account for the gradient of atmospheric pressure generated at the center of the hurricane.
+Using SWAN also required the implementation of tools to process wave spectrum and parameters in SLIM.
+The implementation of the coupling between the two models was based on the ADCIRC+SWAN model.
+It required the development of tools to convert SLIM outputs and forcing into SWAN format, and to allow SLIM to read SWAN outputs.
+Compared to the ADCIRC+SWAN, an additional challenge that we had to overcome was the fact that SLIM is based on discontinous Galerkin finite element models while SWAN requires continuous fields on the mesh.
+This was especially critical when defining the communication between the nodes of the partitions of SWAN based on the partitioning of the mesh used in SLIM.}
+
+\modif{However, the implementation of the coupling could be improved, as the two models are currently alternatively run and stopped with different initial and boundary conditions.
+The coupling therefore induces unwanted initialization computational costs each time one of the two model is run.
+Suppressing these initialization costs would strongly reduce the overhead of the coupling.
+Furthermore, a strong limitation of the coupled model is the temporal complexity of SWAN, which runs four times slower than SLIM.
+A way to tackle this issue would be to run SWAN on a smaller, coarser mesh than SLIM to reduce its computation time at the expense of a potentially reduced resolution in the wave outputs.
+Alternatively, as robust estimates of the Stokes drift can be derived from the surface wind speed \citep{ardhuin2009observation}, a one-way coupled wave-current model could be developed using the vortex force formalism.
+In such a model, wave impact on currents would be modeled in SLIM through interactions between a wind-parameterized Stokes drift and the vorticity of the modeled currents, as in \cite{curcic2015explicit}.} \newline
+
 \section{Perspectives for future works}
@@
 These modifications could lead to an endemic disease dynamics.
+
+\modif{\subsection*{Better understanding the impact of hurricanes on corals}}
+\modif{In this thesis, the impact of hurricanes on coral reef ecosystems was mostly evaluated by modeling wave-induced advective transport processes.
+However, when studying the propagation of drifting particles, assessing the impact of hurricanes on the dispersion of the particle cloud is also of interest.
+This could be performed by computing the distance between the positions of the particles of the cloud and its center of mass or by evaluating the evolution of the concentration of particles in the cloud.
+Furthermore, wave-induced transport is not the only way tropical cyclones affect coral reef biota.
+First, they cause mechanic damages to the reefs as they pass over them.
+However, this destruction can have positive consequences by facilitating asexual recruitment and cleaning algal overgrowth on the reefs \citep{banks2008reef}.
+Furthermore, hurricanes increase water turbidity and induce coral burial by resuspending sediments.
+These impacts could be assessed by combining the developed coupled wave-current model with our sediment transport model.
+Moreover, hurricane interact with the Florida Current by disrupting its vertical structure and reducing its transport.
+These structural perturbations might have consequences on eddy activity and baroclinic instabilities near the Florida Keys.
+Our study could therefore be extended by evaluating the impact of Hurricane Irma on the ocean state and the broader circulation patterns in Florida.
+Finally, by modifying the ocean circulation on their passage, hurricanes might influence the propagation of SCTLD through Florida's Coral Reef.
+Hurricane Irma could therefore have acted as a "super spreader" of disease agents or as a barrier to the propagation of the outbreak.
+This effect could be investigated by combining SLIM+SWAN with the developed epidemiological model.} \newline
 

### 2022-03-26 chapters/intro/intro.tex
modif post-defense - part I

@@
     \includegraphics[width=.9\textwidth]{chapters/intro/figures/cover.jpeg}
-    \caption{Evolution of the absolute coral cover in the Caribbean. Source: \cite{gardner2003long}}
+    \caption{Decline of the absolute coral cover in the Caribbean from $\sim$50\% to $\sim$10\% between 1977 and 2001. Source: \cite{gardner2003long}}
     \label{intro:cover}
@@
     \includegraphics[width=\textwidth]{chapters/intro/figures/fig_sctld.png}
-    \caption{Propagation of the SCTLD troughout the FRT between 2014 and 2021}
+    \caption{Observed propagation of the \modif{stony coral tissue loss disease} throughout the \modif{Florida Reef Tract} between 2014 and 2021}
     \label{intro:propagation}
@@
 \begin{enumerate}
-    \item How did the SCTLD spread through the FRT ?
-    \item What caused the apparent stalling of the propagation of SCTLD ? 
-    \item What was the impact of the PMDDP on the onset of the oubreak of SCTLD~?
+    \item How did the \modif{stony coral tissue loss disease} spread through Florida's Coral Reef ?
+    \item What caused the apparent stalling of the propagation of the \modif{disease} ? 
+    \item What was the impact of the \modif{PortMiami Deep Dredge Project} on the onset of the disease oubreak ?
     \item  What is the effect of hurricane-induced wave-current interactions on transport processes and should they be taken into account when modeling the dispersal of coral larvae ?
@@
 
-\section{Modeling hydrodynamics and transport processes in the FRT}
+\section{Modeling hydrodynamics and transport processes in Florida's Coral Reef}
 The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean.
@@
 This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}.
-Moreover, by coupling SLIM with GOM-HYCOM, the model was able to indirectly represent the baroclinic dynamics of the FC.
-Using this coupling, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs.
-In this study, coral larvae were modeled using a Lagrangian particle tracking model \citep{van2018lagrangian} forced by SLIM currents.
+\modif{The 2D version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities.
+This model is therefore well suited for shallow areas where the water-column is well mixed.
+Furthermore, assuming homogeneous vertical profiles reduces the computational cost and allows a finer spatial resolution of the model.
+In deeper areas however, 3D baroclinic models are required to capture the variations of density in the water column due to pressure, temperature and salinity.
+These differences in density cause baroclinic instabilities that generate mesoscale eddies.
+However, 3D baroclinic models are computationally more expensive, which prohibits the use of a reef-scale resolution on the whole Florida Reef Tract.
+As most reefs of the Florida are located in areas shallower than 30 meters [\textbf{INSERT FIGURE}], we coupled the 2D version of SLIM with GOM-HYCOM to indirectly capture the impact of baroclinic instabilities on the connectivity of the Florida Keys.
+Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs.
+The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements, as illustrated in appendix \ref{onset:appendice} and chapter \ref{chap:irma}}.
+
+\modif{new paragraph: Eulerian vs Lagrangian and why Lagrangian} In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents.
 To answer our different research questions, we adapted this larval transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.
+\modif{[add stuff about the bathymetry and the model validation]}
 
@@
 
+\modif{\section{Modeling framework and assumptions}}
+
+\modif{Wave-Current coupling. Radiation stress vs vector force}
+
+\modif{Presentation of the epidemiological model ?}
+
+\modif{Hypotheses of the model/study}\newline
+
 \section*{Outline}
@@
 Chapter \ref{chap:irma} describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
+\modif{By lack of time, the propagation of coral larvae during Irma was not studied in this chapter}.
 Finally, chapter \ref{chap:conclusion} presents the conclusions of this thesis and perspectives for further works.
@@
 \end{list}
-\newpage 
-\section{List of acronyms and abbreviations}
-
-\begingroup
-    \setlength{\tabcolsep}{20pt}
-    \renewcommand{\arraystretch}{1.3}
-    \begin{tabular}{ll}
-        \textbf{DRTO}   & Dry Tortugas \\
-        \textbf{ECMWF}  & European Centre for Medium-Range Weather Forecast \\
-        \textbf{D55}    & Dredge 55 clamshell \\
-        \textbf{FC}     & Florida Current \\
-        \textbf{FCR}    & Florida's Coral Reef \\
-        \textbf{FRT}    & Florida Reef Tract  \\
-        \textbf{GoM}    & Gulf of Mexico \\
-        \textbf{HURDAT} & HURricane DATabases \\
-        \textbf{HYCOM}  & HYbrid Coordinate Ocean Model \\
-        \textbf{LC}     & Loop Current \\
-        \textbf{NOAA}   & National Oceanographic and Atmospheric Administration \\
-        \textbf{ODMDS}	& Ocean Dredge Material Disposal Site \\
-        \textbf{PoM}    & Port of Miami \\
-        \textbf{PMDDP}  & PortMiami Deep Dredge Project \\
-        \textbf{RS}     & Radiation stress \\
-        \textbf{SB}     & Spider Barge \\
-        \textbf{SCTLD}  & Stony coral tissue loss disease \\
-        \textbf{SLIM}   & Second-generation Louvain-la-Neuve Ice-ocean Model\\
-        \textbf{SWAN}   & Simulating WAves Nearshore \\
-        \textbf{TI}     & Terrapin Island hopper \\
-        \textbf{TX}     & Texas cutterhead \\
-        \textbf{WWTP}	& Wastewater Treatment Plant \\
-        \textbf{WFS}    & West Florida Shelf
-    \end{tabular}
-\endgroup

### 2022-03-26 chapters/irma/article.tex
modif post-defense - part I

@@
 We followed the approach of \cite{dietrich2012performance} by characterizing the wave-induced forces on currents using the radiation-stress (RS) gradient formalism, which has been successfully applied in both 2D and 3D coupled wave-current models under storm conditions \citep{hope2013hindcast, sebastian2014characterizing, brown2013depth}.
+\modif{In this formalism, wave-averaged effects are characterized by the divergence of a stress tensor representing the flux of momentum to currents due to surface gravity waves \citep{longuet1964radiation,lane2007wave,kumar2012implementation}}.
 An alternative formalism is the vortex-force (VF) representation \citep{mcwilliams2004asymptotic}, that provides a clearer decomposition of the wave effect \citep{lane2007wave}.
+\modif{In this formalism, wave impact on current is represented by the combination of the gradient of a Bernoulli head representing adjustments to the pressure in incompressible fluids, and the interaction of the Stokes drift with the vorticity of the flow, \textit{i.e.} the vortex force \citep{lane2007wave}}.
 Although both approaches were adopted by coastal modeling communities, there is an ongoing scientific debate over the correctness and applicability of the two concepts \citep{ardhuin2008comments, mellor2013waves, mellor2015combined, ardhuin2017comments}.
@@
 15.
-These initial particle positions were found using backtracking methods \citep{spivakovskaya2005simulation} to ensure that the release particles would intersect the path of Irma during its passage through the Florida Keys.
+These initial particle positions were found using backtracking methods \citep{spivakovskaya2005simulation} \modif{to obtain a probabilistic picture of the release positions from which particles would intersect the path of Irma during its passage through the Florida Keys}.
 We first defined two 25 km$^\text{2}$ circular regions on the trajectory of the hurricane (see red and blue circles in Fig.
@@
     \includegraphics[width=.95\textwidth]{chapters/irma/figures/validation_met_2.png}
-    \caption{Comparison of reconstructed atmospheric pressure (top) and wind speed (bottom)} with field measurements and coarser ECMWF ERA-5 profiles at Vaca Key station.
-The generated hybrid atmospheric forcings better reproduce the observed storm depression while H*wind winds better reproduce the measured peak in wind speed.
+    \caption{Comparison of reconstructed atmospheric pressure (top) and wind speed (bottom) with field measurements and coarser ECMWF ERA-5 profiles at Vaca Key station.
+The generated hybrid atmospheric forcings better reproduce the observed storm depression while H*wind winds better reproduce the measured peak in wind speed.}
     \label{fig:forcings}
@@
     \caption{Difference between the centers of mass of the particle clouds released from the regions highlighted in Fig.
-\ref{fig:init} and advected by the different combinations of Eulerian currents and Stokes drifts described in Table \ref{tab:summary}}.
+\ref{fig:init} and advected by the different combinations of Eulerian currents and Stokes drifts described in Table \ref{tab:summary} \modif{In the top panels, the trajectories of particles advected by Eulerian currents from an uncoupled SLIM run are compared with trajectories obtained with Eulerian currents and Stokes drift from a coupled SLIM+SWAN run (green); Eulerian currents from a coupled SLIM+SWAN run (orange); and Eulerian currents from and uncoupled SLIM run and Stokes drift from a coupled SLIM+SWAN run on the inner (\textbf{A}) and outer shelves (\textbf{B}).
+The bottom panels show the comparison of the trajectories of particles advected by Eulerian currents from an uncoupled SLIM run and Stokes drift from coupled and uncoupled model runs on the inner (\textbf{C}) and outer (\textbf{D}) shelves.}}
     \label{fig:traj}

### 2022-03-26 chapters/sctld/article.tex
modif post-defense - part I

@@
 The polygons of this reef map were then further divided into 500 m $\times$ 500 m squares in order to track the propagation of the disease with a finer geographical resolution, generating a total of 16,823 polygons.
-At the beginning of each simulated month and for each type of current, a total of about $1.5 \times 10^6$ particles were released over all the reef polygons.
+At the beginning of each simulated month and for each type of current, a total of about $1.5 \times 10^6$ particles were released over all the \modif{500$\times$500 sub-reef polygons assuming an uniform coral cover}.
 These particles had a state composed of their polygon of origin as well as their mass, that they lose at a constant rate $\gamma$ as they were moved by surface, mean or bottom currents.
@@
 where $\tilde{C}_{ij}$ is the entry corresponding to reef pair $(i,j)$ of the normalized potential connectivity matrix [-], $A_i$ is the area of reef polygon $i$ [km$^2$], $\sigma$ is the mortality rate [day$^{-1}$], and $\beta$ and $\beta'(I_j)$ are the inter- and intra-reef disease transmission rates [day$^{-1}$], respectively.
-In this model, diseased corals of sub-reef $i$ can 'infect' corals of sub-reef $j$ if there is non-zero probability of disease agents exchange from sub-reef $i$ to sub-reef $j$, given by $\tilde{C}_{ij}$.
+In this model, diseased corals of sub-reef $i$ can 'infect' corals of sub-reef $j$ if there is non-zero probability of disease agents exchange from sub-reef $i$ to sub-reef $j$, given by $\tilde{C}_{ij}$ \modif{, as illustrated in Fig.
+\ref{sctld:network_dir}}.
 Moreover, to account for coral resistance to the disease, the intra-reef transmission function $\beta'(I_j)$ has the shape of a smooth step function of the fraction of infectious corals $I_j$ and is expressed as follows:
@@
 %This assumption seems reasonable since coral populations of the FRT were declining prior to the emergence of the SCTLD outbreak. 
+\begin{figure}
+	\centering
+	\includegraphics[width=.6\textwidth]{chapters/sctld/figures/network_sir.png}
+	\caption{\modif{Illustration of the developed connected SIR model.
+The fraction of infectious corals $I_j$ on sub-reef polygon $j$ infects the fraction of susceptible individuals $S_j$ on the same sub-reef polygon if $I_j \geq I_0$.
+Furthermore, infectious corals on sub-reef $i$ can infect susceptible individuals of sub-reef $j$ if the two sub-reefs are connected in the connectivity matrices ($\tilde{C}_{ij} > 0$)}}.
+	\label{sctld:network_dir}
+\end{figure}
 
@@
     \includegraphics[width=.99\linewidth]{chapters/sctld/figures/hull_example.png}
-    \caption{Method used to compute the disease front displacement during a simulated time interval.\textbf{a.} Concave hull $H_0$ of the diseased sub-reefs at the beginning of the simulated period.
+    \caption{Method used to compute the disease front displacement during a simulated time interval.
+\textbf{a.} Concave hull $H_0$ of the diseased sub-reefs at the beginning of the simulated period.
 \textbf{b.} Concave hull $H_1$ of the sub-reefs diseased during the simulated time interval.

### 2022-03-26 thesis.tex
modif post-defense - part I

@@
   \clearemptydoublepage
-  % \input{include/glossary}
-  % \clearemptydoublepage
+  \input{include/glossary}
+  \clearemptydoublepage
   \mainmatter 
@@
  \input{chapters/intro/intro}
-%%
 %%%%%% 
-\mychapter{Processes driving the spread of SCTLD through the FRT} \label{chap:sctld}
+\mychapter{Processes driving the spread of stony coral tissue loss disease through the Florida Reef Tract} \label{chap:sctld}
 \chaptermark{Processes driving the spread of SCTLD in the FRT}  
 \input{chapters/sctld/sctld}
-\mychapter{Transmission dynamics of SCTLD from the Marquesas and to the DRTO}\label{chap:drto}
+\mychapter{Transmission dynamics of the disease from the Marquesas and to the Dry Tortugas}\label{chap:drto}
 \chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
 \input{chapters/drto/drto}
-\mychapter{The onset of the SCTLD outbreak and the PortMiami Deep Dredge Project} \label{chap:onset}
+\mychapter{The onset of the outbreak and the PortMiami Deep Dredge Project} \label{chap:onset}
 \chaptermark{Onset of SCTLD and PortMiami Deep Dredge Project}  

### 2022-03-27 chapters/intro/intro.tex
modif post-defense part II²

@@
 
-\modif{new paragraph: Eulerian vs Lagrangian and why Lagrangian} In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents.
-To answer our different research questions, we adapted this larval transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.
-\modif{[add stuff about the bathymetry and the model validation]}
+\modif{Transport processes are at the core of this thesis.
+They can be estimated using two general methods: Eulerian tracers and Lagrangian particles.
+Eulerian methods were the earliest methods applied for transport modeling and are still widely used today.
+They offer the advantages of a fixed grid by solving the transport equations directly on the native grid of the model and are easy to implement However, mass conservation issues may arise in problems with high initial gradient of concentrations \citep{spivakovskaya2007lagrangian}.
+The Lagrangian methods, on the other hand, follow virtual particles of zero spatial extent whose movements are modeled with a stochastic differential equation consistent with the advection-diffusion equation \citep{spivakovskaya2007lagrangian,van2018lagrangian}.
+One key difference between the two  methods is their computational cost.
+For each time step, the movements of $N$ virtual particles require $N$ sets of computation, while the advection-diffusion of a tracer takes as many sets of computation as there are discrete grid cells in the model.
+Additionally, Lagrangian methods allow the storage of the entire trajectory history of the virtual particles.
+This enables the \textit{a posteriori} computation of connectivity between different regions of the ocean \citep{thomas2014numerical,frys2020fine}.
+Furthermore, particles can be advected in offline mode when velocity fields are stored in order to reduce computation time, while Eulerian tracers' equations must be solved online with the current velocities.
+Finally, Lagrangian particles can be advected backward in time.
+This reverse-time analysis gives a spatial probability distribution of places from where it is likely that particles originated to reach a given location \citep{spivakovskaya2005simulation}.
+Transport processes were therefore studied in this thesis using Lagrangian particles advected by SLIM currents.
+To answer our different research questions, we adapted SLIM transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.}
 

### 2022-03-27 chapters/intro/intro.tex
modifs post-defense part III

@@
 However, 3D baroclinic models are computationally more expensive, which prohibits the use of a reef-scale resolution on the whole Florida Reef Tract.
-As most reefs of the Florida are located in areas shallower than 30 meters [\textbf{INSERT FIGURE}], we coupled the 2D version of SLIM with GOM-HYCOM to indirectly capture the impact of baroclinic instabilities on the connectivity of the Florida Keys.
+As most reefs of the Florida are located in areas shallower than 30 meters (Fig.
+\ref{intro:bathy}), we coupled the 2D version of SLIM with GOM-HYCOM to indirectly capture the impact of baroclinic instabilities on the connectivity of the Florida Keys.
 Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs.
 The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements, as illustrated in appendix \ref{onset:appendice} and chapter \ref{chap:irma}}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{chapters/intro/figures/bathymetry_figure.png}
+	\caption{\modif{Bathymetry of the Florida Keys and the Dry Tortigas.
+The bathymetry of the model was built by combining three data sources: the General Bathymetric Chart of the Ocean (\href{https://www.gebco.net/}{GEBCO};15 arc-seconds resolution), The Coastal Relief Model (\citealp{noaa2001}; 3 arc-seconds resolution), and NOAA's bathymetric Digital Elevation Model (\citealp{ncei2018}; 1/3 arc-second resolution).}}
+	\label{intro:bathy}
+\end{figure}
 
@@
 
-\modif{\section{Modeling framework and assumptions}}
+%\modif{\section{Modeling framework and assumptions}}
 
-\modif{Wave-Current coupling. Radiation stress vs vector force}
+%\modif{Wave-Current coupling. Radiation stress vs vector force}
 
-\modif{Presentation of the epidemiological model ?}
+%\modif{Presentation of the epidemiological model ?}
 
-\modif{Hypotheses of the model/study}\newline
+%\modif{Hypotheses of the model/study}\newline
 

### 2022-03-28 chapters/conclusions/conclusions.tex
commentaires patron

@@
     }
-    \item \textbf{1. How did the SCTLD spread through the FRT ?}
+    \item \textbf{1. How did the \modif{stony coral tissue loss disease} spread through the \modif{Florida Reef Tract} ?}
     \begin{list}{}{\setlength{\topsep}{0pt}}
@@
     \end{list}
-    \item \textbf{2. What caused the apparent stalling of the propagation of SCTLD ?}
+    \item \textbf{2. What caused the apparent stalling of the propagation of \modif{disease} ?}
     \begin{list}{}{\setlength{\topsep}{0pt}}
@@
     \end{list}
-    \item \textbf{3. What was the impact of the PMDDP on the onset of the oubreak of SCTLD~?}
+    \item \textbf{3. What was the impact of the \modif{PortMiami Deep Dredge Project} on the onset of the \modif{disease outbreak}?}
     \begin{list}{}{\setlength{\topsep}{0pt}}
@@
 
-\modif{\section{Developments of the model}}
+\modif{\section{Our contribution to the model development}}
 
 \modif{Answering the aforementioned research questions required the development of new modeling tools.
-As this thesis heavily relies on Lagrangian transport modeling, several features were added to the particle tracker of SLIM.
+As this thesis relied heavily on Lagrangian transport modeling, several features were added to SLIM's particle tracker.
 Although the transport of the particles in itself was already implemented in the model, we developed features to characterize the state of the particles and compute connectivity indicators.
-The particle state can be used to represent biological traits such as the competency of coral larvae or abiotic characteristics such as the decaying mass of diseased material or the diameter of sediment particles.
-We developed tools to derive connectivity indicators such as the shortest paths between a pair sub-reef polygons in the obtained  connectivity network.
-Additionally, we developed an epidemiological model using the computed connectivity matrices to simulated the spread of the stony coral tissue loss disease throughout Florida.
-Finally, the particle tracker was made more flexible by adding the capability to transport particles using outputs from larger-scale ocean models such as HYCOM.}
+The particle state can be used to represent biological traits such as the competency of coral larvae or abiotic characteristics such as the decay rate of diseased agent mass or the diameter of sediment particles.
+Post-processing features were developed to compute the concentration or the deposition of particles.
+We also developed tools to derive connectivity indicators such as the shortest paths between a pair sub-reef polygons in the connectivity networks.
+Additionally, we developed an epidemiological model using the computed connectivity matrices to simulated the spread of the stony coral tissue loss disease throughout Florida's Coral Reef.
+Finally, the particle tracker was made more flexible by developing new readers for different ocean model outputs such as HYCOM or the Nucleus for European Modelling of the Ocean (NEMO; \citealp{madec2017nemo}).}
 
-\modif{Additionally, the coupling of SLIM with SWAN required some modifications of SLIM.
+\modif{Additionally, coupling SLIM with SWAN required some modifications of SLIM.
 First, we implemented a new parameterization of the wind drag coefficient to account for the formation of spume at the ocean-atmosphere interface in heavy wind conditions.
-Furthermore, a term was added to the model's equations to account for the gradient of atmospheric pressure generated at the center of the hurricane.
+An additional term was also added to the model's equations to account for the gradient of atmospheric pressure generated at the center of the hurricane.
 Using SWAN also required the implementation of tools to process wave spectrum and parameters in SLIM.
-The implementation of the coupling between the two models was based on the ADCIRC+SWAN model.
+The implementation of the coupling between the two models followed the same approach as the ADCIRC+SWAN model.
 It required the development of tools to convert SLIM outputs and forcing into SWAN format, and to allow SLIM to read SWAN outputs.
-Compared to the ADCIRC+SWAN, an additional challenge that we had to overcome was the fact that SLIM is based on discontinous Galerkin finite element models while SWAN requires continuous fields on the mesh.
+Compared to ADCIRC+SWAN, an additional challenge that we had to overcome was the fact that SLIM is based on discontinous Galerkin finite element models while SWAN requires continuous fields on the mesh.
 This was especially critical when defining the communication between the nodes of the partitions of SWAN based on the partitioning of the mesh used in SLIM.}
@@
 Furthermore, wave-induced transport is not the only way tropical cyclones affect coral reef biota.
-First, they cause mechanic damages to the reefs as they pass over them.
+First, they can cause a wholesale destruction of the reefs as they pass over them.
 However, this destruction can have positive consequences by facilitating asexual recruitment and cleaning algal overgrowth on the reefs \citep{banks2008reef}.
@@
 These structural perturbations might have consequences on eddy activity and baroclinic instabilities near the Florida Keys.
-Our study could therefore be extended by evaluating the impact of Hurricane Irma on the ocean state and the broader circulation patterns in Florida.
-Finally, by modifying the ocean circulation on their passage, hurricanes might influence the propagation of SCTLD through Florida's Coral Reef.
+Our study could therefore be extended by evaluating the impact of Hurricane Irma on the ocean state and the broader circulation patterns in Florida although baroclinic models might be required.
+Finally, by modifying the ocean circulation on their passage, hurricanes might influence the connectivity of the Florida Reef Tract.
+The coupled SLIM+SWAN model could thus be used to evaluate the impact of Hurricane Irma on coral larval connectivity as a continuation of the work initiated in chapter \ref{chap:irma}.
+Furthermore, the hurricane-induced transport might have impacted the propagation of SCTLD through Florida's Coral Reef.
 Hurricane Irma could therefore have acted as a "super spreader" of disease agents or as a barrier to the propagation of the outbreak.

### 2022-03-28 chapters/intro/intro.tex
commentaires patron

@@
 In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}).
-This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}.
-\modif{The 2D version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities.
-This model is therefore well suited for shallow areas where the water-column is well mixed.
+This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef [ADD OTHER STUFF] \citep{lambrechts2008multi}.
+\modif{The 2D barotropic version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities.
+This model is therefore well suited for shallow areas where the water column is well mixed.
 Furthermore, assuming homogeneous vertical profiles reduces the computational cost and allows a finer spatial resolution of the model.
-In deeper areas however, 3D baroclinic models are required to capture the variations of density in the water column due to pressure, temperature and salinity.
-These differences in density cause baroclinic instabilities that generate mesoscale eddies.
-However, 3D baroclinic models are computationally more expensive, which prohibits the use of a reef-scale resolution on the whole Florida Reef Tract.
-As most reefs of the Florida are located in areas shallower than 30 meters (Fig.
+However, in deeper areas or in regions where there is an inflow of freshwater, 3D baroclinic models are required to capture the variations of density in the water column due to temperature and salinity gradients.
+These differences in density can cause baroclinic instabilities that generate mesoscale eddies.
+Nonetheless, 3D baroclinic models are computationally more expensive, which prohibits the use of a reef-scale resolution on the whole Florida Reef Tract.
+As most Florida reefs are located in areas shallower than 30 meters (Fig.
 \ref{intro:bathy}), we coupled the 2D version of SLIM with GOM-HYCOM to indirectly capture the impact of baroclinic instabilities on the connectivity of the Florida Keys.
-Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs.
-The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements, as illustrated in appendix \ref{onset:appendice} and chapter \ref{chap:irma}}.
+Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT with SLIM, \ie~the exchanges of larvae between reefs.
+[TODO: move somewhere else]The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements, as illustrated in appendix \ref{onset:appendice} and chapter \ref{chap:irma}}.
 
@@
 Eulerian methods were the earliest methods applied for transport modeling and are still widely used today.
-They offer the advantages of a fixed grid by solving the transport equations directly on the native grid of the model and are easy to implement However, mass conservation issues may arise in problems with high initial gradient of concentrations \citep{spivakovskaya2007lagrangian}.
-The Lagrangian methods, on the other hand, follow virtual particles of zero spatial extent whose movements are modeled with a stochastic differential equation consistent with the advection-diffusion equation \citep{spivakovskaya2007lagrangian,van2018lagrangian}.
+They offer the advantages of a fixed grid by solving the transport equations directly on the native grid of the model and are easy to implement \citep{spivakovskaya2007lagrangian}.
+The Lagrangian methods, on the other hand, follow virtual particles whose movements are modeled with a stochastic differential equation consistent with the advection-diffusion equation \citep{spivakovskaya2007lagrangian,van2018lagrangian}.
 One key difference between the two  methods is their computational cost.
@@
 This enables the \textit{a posteriori} computation of connectivity between different regions of the ocean \citep{thomas2014numerical,frys2020fine}.
-Furthermore, particles can be advected in offline mode when velocity fields are stored in order to reduce computation time, while Eulerian tracers' equations must be solved online with the current velocities.
-Finally, Lagrangian particles can be advected backward in time.
+Finally, Lagrangian particles can be advected backward in time in offline mode, when velocity fields are stored.
 This reverse-time analysis gives a spatial probability distribution of places from where it is likely that particles originated to reach a given location \citep{spivakovskaya2005simulation}.
-Transport processes were therefore studied in this thesis using Lagrangian particles advected by SLIM currents.
+In this thesis, transport processes were therefore studied using Lagrangian particles advected by SLIM currents.
 To answer our different research questions, we adapted SLIM transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.}
@@
 Chapter \ref{chap:irma} describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.
-\modif{By lack of time, the propagation of coral larvae during Irma was not studied in this chapter}.
 Finally, chapter \ref{chap:conclusion} presents the conclusions of this thesis and perspectives for further works.

### 2022-03-28 chapters/irma/article.tex
commentaires patron

@@
 An alternative formalism is the vortex-force (VF) representation \citep{mcwilliams2004asymptotic}, that provides a clearer decomposition of the wave effect \citep{lane2007wave}.
-\modif{In this formalism, wave impact on current is represented by the combination of the gradient of a Bernoulli head representing adjustments to the pressure in incompressible fluids, and the interaction of the Stokes drift with the vorticity of the flow, \textit{i.e.} the vortex force \citep{lane2007wave}}.
+\modif{Here, wave impact on current is represented by the combination of the gradient of a Bernoulli head representing adjustments to the pressure in incompressible fluids, and the interaction of the Stokes drift with the vorticity of the flow, \textit{i.e.} the vortex force \citep{lane2007wave}}.
 Although both approaches were adopted by coastal modeling communities, there is an ongoing scientific debate over the correctness and applicability of the two concepts \citep{ardhuin2008comments, mellor2013waves, mellor2015combined, ardhuin2017comments}.

### 2022-03-28 chapters/onset/article.tex
commentaires patron

@@
 \textbf{B}: Close up view of the dredged channel.
-The locations of the different types of dredging that took place during the expansion of PoM are shown by colored markers.
+The locations of the different types of dredging that took place during the expansion of PortMiami are shown by colored markers.
 \textbf{C:} Snapshot of the modeled currents in the vicinity of the dredged channel.
@@
 Four sites were reefs where disease was reported in 2014 by \cite{precht2016unprecedented}.
-The fifth site was the monitoring station HBSC1-CP, used throughout the expansion of PoM, where signs of disease were already reported in May 2014 (Fig.
+The fifth site was the monitoring station HBSC1-CP, used throughout the expansion of PoM, where signs of disease were already reported in May 2014 (\citealp{dial2015}; Fig.
 \ref{fig:onset_mesh}A, see supplementary material \ref{onset:appendice}).
@@
 		\includegraphics[width=\textwidth]{chapters/onset/figures/hbsc1_cp.png}
-		\caption{Pictures of a monitored colony at site HBSC1-CP. The first signs of the disease appeared between May 25 and May 30. By June 4, the whole colony was completely dead}
+		\caption{Pictures of a monitored colony at site HBSC1-CP \citep{dial2015}. The first signs of the disease appeared between May 25 and May 30. By June 4, the whole colony was completely dead}
 	\end{figure}

### 2022-03-28 chapters/sctld/article.tex
commentaires patron

@@
 When affected by the disease, susceptible individuals become infectious and infect other susceptible individuals until they are removed from the system, either through recovery or death.
+\modif{Other more advanced compartmental models take reinfection into account (SIS) or consider a latent non-infectious exposed ($E$) state between the susceptible and infectious states (SEIR) \citep{Murray07}.
+However, due to the high mortality of the disease and in the absence of clear evidence suggesting a latency period following infection, the SIR model seemed a reasonably simple and generic modeling choice for this study}.
 Such models usually rely on the hypothesis of an homogeneous, well-mixed population.

### 2022-03-29 chapters/drto/article.tex
modifs post-defense part III
modifs part IV

@@
 \ref{fig:fig1_drto}a, inset), while also representing the large-scale flow features within the FC (Fig 1b and c).
-Further details on the model formulation and validation are provided by \cite{frys2020fine}.
+Further details on the model are provided by \cite{frys2020fine}.
+\modif{The model results were validated against tide gauge and ADCP measurements and showed good agreement with observations (see appendix \ref{drto:validation})}.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{chapters/drto/figures/fig4.png}
+    \includegraphics[width=.6\textwidth]{chapters/drto/figures/vulnerability_new.png}
+    \includegraphics[width=.9\textwidth]{chapters/drto/figures/source_new.png}
     \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020.
@@
 The fact that no eddies were observed at the beginning of our simulation might be related to the identification method, based solely on visual assessment of the residual currents.
+\modif{Alternative methods to identify the presence of eddies could have been the computation of the Okubo-Weiss parameter \citep{okubo1970horizontal,weiss1991dynamics} or the extraction of the FC position as the 20$^\circ$C isotherm at 150 m from HYCOM temperature fields, following the approach of \cite{kourafalou2012florida}}.
 A comprehensive identification of all Tortugas eddies during our simulated period would require the analysis of sea surface temperature and elevation, which is beyond the scope of the present study.
+.
 While our approach might miss some Tortugas gyres, it filters out submesoscale eddies with lifespan of the order of several days.
@@
 \begin{subappendices}
-	\section{Validation of the model}
+	\section{Validation of the model}\label{drto:validation}
 

### 2022-03-29 chapters/intro/intro.tex
modifs part IV

@@
 These variations of the FC position have been shown to have no seasonal pattern, and a strong interannual variability.
-The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}.
+The generated meandering and \modif{its interaction with the topology of the Florida Keys} is associated with the presence of \modif{cold} cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}.
 These eddies significantly impact the ecology of the southwest Florida Shelf by causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution,limouzy1997translocation,sponaugle2005florida, kourafalou2012florida}.
-Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}.
+\modif{Furthermore, eddy-induced upwelling of deeper, nutrient-rich waters increase the chlorophyll and zooplankton concentrations in the Florida Keys \citep{lee1992influence}.
+Similarly, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes an inflow of nutrients supporting an ecosystem response \citep{weisberg2003local, liu2016offshore}}.
 These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}.

### 2022-04-05 chapters/intro/intro.tex
paragraph about physics

@@
 
+\modif{The motion of geophysical fluids such as seawater is characterized by the effects of ambient rotation and stratification.
+The ambient rotation is induced by the Earth's spin and generates the Coriolis forces.
+Stratification arises from the presence of waters with different temperature and salinity, and therefore different density.
+Under the action of gravitational forces, heavier waters lower while lighter waters rise, which caused the water column to tend towards a stable stratified equilibrium.
+However, this equilibrium is disrupted in the presence of discontinuities in horizontal velocities induced by wind and water motion.
+These baroclinic instabilities generate internal waves, water mixing, and eddies.
+Eddy activity is especially strong in boundary currents such as the Gulf Stream, where strong horizontal shear occur and where shelf water and deep water masses meet on the shelf break.
+Capturing such processes require baroclinic models that represent the variations of water density along the water column.
+However, in regions where the water column is sufficiently homogeneous and shallow that vertical contrasts are negligible, ocean circulation is accurately approximated by barotropic models.
+Such models use a depth-averaged approach by assuming an homogeneous water column with constant density and are used to model estuaries, coastal regions and shelf seas.
+Both the baroclinic and the barotropic models usually rely on the hydrostatic equilibrium, \textit{i.e.} vertical accelerations are neglected by assuming that gravity balances the vertical component of the pressure gradient force.
+Such assumption is only valid when the horizontal scale of the flow is larger than its vertical scale.}
+
 In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}).

### 2022-04-05 chapters/conclusions/conclusions.tex
additional modifs post physics

@@
 As this thesis relied heavily on Lagrangian transport modeling, several features were added to SLIM's particle tracker.
-Although the transport of the particles in itself was already implemented in the model, we developed features to characterize the state of the particles and compute connectivity indicators.
+Although the transport of the particles in itself was already implemented in the model, we developed new model features to characterize the state of the particles and compute new connectivity indicators.
 The particle state can be used to represent biological traits such as the competency of coral larvae or abiotic characteristics such as the decay rate of diseased agent mass or the diameter of sediment particles.
-Post-processing features were developed to compute the concentration or the deposition of particles.
+Post-processing features were developed to compute the concentration of suspended and deposited particles.
 We also developed tools to derive connectivity indicators such as the shortest paths between a pair sub-reef polygons in the connectivity networks.
@@
 An additional term was also added to the model's equations to account for the gradient of atmospheric pressure generated at the center of the hurricane.
-Using SWAN also required the implementation of tools to process wave spectrum and parameters in SLIM.
+Using SWAN also required the implementation of tools to process the wave spectrum and the associated parameters in SLIM.
 The implementation of the coupling between the two models followed the same approach as the ADCIRC+SWAN model.
@@
 \modif{In this thesis, the impact of hurricanes on coral reef ecosystems was mostly evaluated by modeling wave-induced advective transport processes.
-However, when studying the propagation of drifting particles, assessing the impact of hurricanes on the dispersion of the particle cloud is also of interest.
+However, when studying the propagation of drifting particles, assessing the impact of hurricanes on the diffusion (\ie~spread) of the particle cloud is also of interest.
 This could be performed by computing the distance between the positions of the particles of the cloud and its center of mass or by evaluating the evolution of the concentration of particles in the cloud.
@@
 First, they can cause a wholesale destruction of the reefs as they pass over them.
-However, this destruction can have positive consequences by facilitating asexual recruitment and cleaning algal overgrowth on the reefs \citep{banks2008reef}.
+However, this destruction can sometimes have positive consequences by facilitating asexual recruitment and cleaning algal overgrowth on the reefs \citep{banks2008reef}.
+Moreover, water mixing and upwelling by hurricane can reduce water temperature and alleviate the thermal stress experienced by corals \citep{wilkinson2008status}.
 Furthermore, hurricanes increase water turbidity and induce coral burial by resuspending sediments.
 These impacts could be assessed by combining the developed coupled wave-current model with our sediment transport model.
-Moreover, hurricane interact with the Florida Current by disrupting its vertical structure and reducing its transport.
-These structural perturbations might have consequences on eddy activity and baroclinic instabilities near the Florida Keys.
+Moreover, hurricane interact with the Florida Current by disrupting its vertical structure and reducing its transport \citep{oey2007hurricane}.
+These structural perturbations might have consequences on eddy activity near the Florida Keys and therefore impact nutrient concentration and larval recruitment in the region \citep{lee1992influence}.
 Our study could therefore be extended by evaluating the impact of Hurricane Irma on the ocean state and the broader circulation patterns in Florida although baroclinic models might be required.

### 2022-04-05 chapters/intro/intro.tex
additional modifs post physics

--- 2022-03-29 chapters/intro/intro.tex
@@
     \includegraphics[width=.8\textwidth]{chapters/intro/figures/anthropocene.png}
-    \caption{Goods and services of coral reefs and their linkage with human activities. Source: \cite{hughes2017coral}}
+    \caption{Goods and services of coral reefs and their linkage with human activities.
+\modif{Despite covering less than 0.5\% of the ocean floor, coral reefs are home for almost of third of the marine biodiversity and are a source of food and livelihood to hundreds of millions of people.} Source: \cite{hughes2017coral}}
     \label{intro:corals}
@@
     \includegraphics[width=.9\textwidth]{chapters/intro/figures/cover.jpeg}
-    \caption{Decline of the absolute coral cover in the Caribbean from $\sim$50\% to $\sim$10\% between 1977 and 2001. Source: \cite{gardner2003long}}
+    \caption{Decline of the absolute coral cover in the Caribbean from $\sim$50\% to $\sim$10\% between 1977 and 2001.
+\modif{This illustrates the worldwide decline in live coral cover caused by both global and local stressors} .Source: \cite{gardner2003long}}
     \label{intro:cover}
@@
     \includegraphics[width=\textwidth]{chapters/intro/figures/frt.png}
-    \caption{Map of Florida Reef Tract with major features and divisions. Source: \cite{kupfner2019untapped}}
+    \caption{Map of Florida Reef Tract with major features and divisions.
+\modif{The northern half-section of the Florida Reef Tract consists of relic, Holocene framework reefs with fairly low living hard coral cover.
+The southern hal-section is a chain of limestone islands (the Keys), that extend from the southern tip of Florida to the Dry Tortugas}.
+Source: \cite{kupfner2019untapped}}
     \label{intro:frt}
@@
     \includegraphics[width=\textwidth]{chapters/intro/figures/fig_sctld.png}
-    \caption{Observed propagation of the \modif{stony coral tissue loss disease} throughout the \modif{Florida Reef Tract} between 2014 and 2021}
+    \caption{Observed propagation of the \modif{stony coral tissue loss disease} throughout the \modif{Florida Reef Tract} between 2014 and 2021.
+\modif{The disease was first observed off the coast of Miami in 2014, and then propagated through Florida's Coral Reef both northward and southward.
+The propagation of the disease appeared to slow down in 2020 between the Lower Keys and the Dry Tortugas (DRTO), before it finally reached this area in May 2021.}}
     \label{intro:propagation}
@@
     \caption{Total number of major hurricanes strikes in Florida between 1900 and 2010.
-Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
+\modif{Southern Florida is a prime landfall for tropical cyclones and has been directly hit by 27 of them during the past century.} Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
     \label{inro:landfall}
@@
 
+\modif{The motion of geophysical fluids such as seawater is characterized by the effects of ambient rotation and stratification \citep{cushman2011introduction}.
+The ambient rotation is induced by the Earth's spin and generates the Coriolis forces.
+Stratification arises from the presence of waters with different temperature and salinity, and therefore different density.
+Under the action of gravitational forces, heavier waters lower while lighter waters rise, which caused the water column to tend towards a stable stratified equilibrium \citep{cushman2011introduction}.
+However, this equilibrium is disrupted in the presence of discontinuities in horizontal velocities induced by wind and water motion \citep{miles1959generation}.
+These baroclinic instabilities generate internal waves, water mixing, and eddies \citep{cushman2011introduction}.
+Eddy activity is especially strong in boundary currents such as the Gulf Stream, where strong horizontal shear occur and where shelf water and deep water masses meet on the shelf break \citep{kourafalou2012florida}.
+These baroclinic instabilities were shown to have a significant impact on water composition and larval recruitment in the Florida Keys \citep{lee1992influence}.
+Capturing such processes require baroclinic models that represent the variations of water density along the water column.
+However, in regions where the water column is sufficiently homogeneous and shallow that vertical contrasts are negligible, ocean circulation is accurately approximated by barotropic models.
+Such models use a depth-averaged approach by assuming an homogeneous water column with constant density and are used to model estuaries, coastal regions and shelf seas \citep{tan1992shallow}.
+Both the baroclinic and the barotropic models usually rely on the hydrostatic equilibrium, \textit{i.e.} vertical accelerations are neglected by assuming that gravity balances the vertical component of the pressure gradient force.
+Such assumption is only valid when the horizontal scale of the flow is larger than its vertical scale \citep{cushman2011introduction}.}
+
 In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}).
-This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef [ADD OTHER STUFF] \citep{lambrechts2008multi}.
-\modif{The 2D barotropic version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities.
+This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics \modif{of coastal regions and estuaries such as the whole Great Barrier Reef \citep{lambrechts2008multi}, the Florida Reef Tract \citep{frys2020fine}, the Scheldt estuary \citep{de2012water}, or the Mahakam delta \citep{sassi2011tidal}}.
+The 2D barotropic version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities.
 This model is therefore well suited for shallow areas where the water column is well mixed.
 Furthermore, assuming homogeneous vertical profiles reduces the computational cost and allows a finer spatial resolution of the model.
-However, in deeper areas or in regions where there is an inflow of freshwater, 3D baroclinic models are required to capture the variations of density in the water column due to temperature and salinity gradients.
-These differences in density can cause baroclinic instabilities that generate mesoscale eddies.
-Nonetheless, 3D baroclinic models are computationally more expensive, which prohibits the use of a reef-scale resolution on the whole Florida Reef Tract.
+However, in deeper areas or in regions where there is an inflow of freshwater, 3D baroclinic models are required to capture the effects of density discontinuities in the water column due to temperature and salinity gradients.
+Nonetheless, 3D baroclinic models are computationally more expensive, which currently prohibits the use of a reef-scale resolution on the whole Florida Reef Tract.
 As most Florida reefs are located in areas shallower than 30 meters (Fig.
@@
 Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT with SLIM, \ie~the exchanges of larvae between reefs.
-[TODO: move somewhere else]The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements, as illustrated in appendix \ref{onset:appendice} and chapter \ref{chap:irma}}.
+\modif{The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements \citep{dobbelaere2022connecting,dobbelaere2022impacts}}.
 
@@
 One key difference between the two  methods is their computational cost.
-For each time step, the movements of $N$ virtual particles require $N$ sets of computation, while the advection-diffusion of a tracer takes as many sets of computation as there are discrete grid cells in the model.
+For each time step, the number of operations required to compute the movements of virtual particles is proportional to their number, while the computational cost of the advection-diffusion of a tracer is proportional to the number of discrete grid cells in the model.
 Additionally, Lagrangian methods allow the storage of the entire trajectory history of the virtual particles.
 This enables the \textit{a posteriori} computation of connectivity between different regions of the ocean \citep{thomas2014numerical,frys2020fine}.
-Finally, Lagrangian particles can be advected backward in time in offline mode, when velocity fields are stored.
+Finally, Lagrangian particles can be advected backward in time in offline mode, using an offline velocity field.
 This reverse-time analysis gives a spatial probability distribution of places from where it is likely that particles originated to reach a given location \citep{spivakovskaya2005simulation}.
@@
 
-%\modif{\section{Modeling framework and assumptions}}
-
-%\modif{Wave-Current coupling. Radiation stress vs vector force}
-
-%\modif{Presentation of the epidemiological model ?}
-
-%\modif{Hypotheses of the model/study}\newline
+\modif{As this thesis is mostly based on modeling tools, it relies on the main assumption that a mechanistic model can accurately capture transport processes in Florida's Coral Reef.
+Models are simplifications of the reality and hypotheses must therefore be made on the processes driving the evolution of the system under study.
+Such hypotheses were for instance made by favoring a fine resolution barotropic model over a coarser baroclinic model of the Florida Reef Tract.
+Furthermore, we assumed that if the modeled waves and hydrodynamics accurately reproduced local observations from buoys, tide gauges and ADCP, then the modeled transport that they drive would be reasonably accurate as well.
+According to the popular saying, "all models are wrong but some are useful".
+In the present study, model results are confronted against field observations and experimental results to identify the potential drivers of the dispersion of stressors in Florida's Coral Reef.}
 

### 2022-04-05 chapters/irma/article.tex
additional modifs post physics

@@
     \includegraphics[width=.99\textwidth]{chapters/irma/figures/coupling_v2.png}
-    \caption{Schematic illustration of the coupled SLIM+SWAN model.}
+    \caption{Schematic illustration of the coupled SLIM+SWAN model.
+\modif{Both models run on the same computational cores and mesh.
+SLIM is run first and passes the wind velocity $U_{10}$, water level $\eta$, depth-averaged current $u$ and bottom roughness length $z_0$ to SWAN.
+SWAN then uses these quantities to compute the wave radiation stress gradient $\text{\boldmath$\tau$}_\text{wave}$, that is then passed
+    to SLIM as the force exerted by waves on currents.
+SLIM then uses this quantity to update the value of the surface stress $\text{\boldmath$\tau$}_s$, computed as the combination of wind and wave stresses $\text{\boldmath$\tau$}_\text{wind} + \text{\boldmath$\tau$}_\text{wave}$}}
     \label{fig:coupling}
@@
     \includegraphics[width=.95\textwidth]{chapters/irma/figures/figure_initial_positions.png}
-    \caption{Release regions of the passive particles on the inner and outer shelves on Sept 7 at 0000 UTC (red and blue clouds) obtained by backtracking particles released in the red and blue circular areas during the passage of Irma.}
+    \caption{Release regions of the passive particles on the inner and outer shelves on Sept 7 at 0000 UTC (red and blue clouds) obtained by backtracking particles released in the red and blue circular areas during the passage of Irma.
+\modif{These clouds give a probabilistic picture of the positions from which particles should have been released to intersect the path of the center of the hurricane}}
     \label{fig:init}

### 2022-04-05 compress.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+compressed=${1%.*}_compressed.pdf
+
+du -sh $1
+gs -sDEVICE=pdfwrite -dCompatibilityLevel=1.5 -dNOPAUSE -dQUIET -dBATCH -dPrinted=false -sOutputFile=$compressed $1
+mv $compressed $1
+du -sh $1

### 2022-04-05 chapters/drto/article.tex
modifs mise en page impression

@@
     \centering
-    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig2.jpg}
+    \includegraphics[width=.99\textwidth]{chapters/drto/figures/fig2.jpg}
     \caption{(a) Monthly connectivity indicator between the Marquesas and the DRTO (blue), the Western Marquesas and the DRTO (orange) and the Marquesas and the Western Marquesas (green) between May 2018 and May 2021.
@@
     \centering
-    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig3.jpg}
+    \includegraphics[width=.99\textwidth]{chapters/drto/figures/fig3.jpg}
     \caption{Edges with largest connection probability to each sub-reef of the DRTO from the Marquesas and the Western Marquesas in November 2020.
@@
     \centering
-    \includegraphics[width=.6\textwidth]{chapters/drto/figures/vulnerability_new.png}
-    \includegraphics[width=.9\textwidth]{chapters/drto/figures/source_new.png}
+    \includegraphics[width=.75\textwidth]{chapters/drto/figures/vulnerability_new.png}
+    \includegraphics[width=.99\textwidth]{chapters/drto/figures/source_new.png}
     \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020.

### 2022-04-05 chapters/intro/intro.tex
modifs mise en page impression

@@
 The ambient rotation is induced by the Earth's spin and generates the Coriolis forces.
-Stratification arises from the presence of waters with different temperature and salinity, and therefore different density.
-Under the action of gravitational forces, heavier waters lower while lighter waters rise, which caused the water column to tend towards a stable stratified equilibrium \citep{cushman2011introduction}.
-However, this equilibrium is disrupted in the presence of discontinuities in horizontal velocities induced by wind and water motion \citep{miles1959generation}.
+Stratification arises from the presence of water masses with different temperatures and salinities, and therefore different densities.
+Under the action of the gravitational force, heavier water masses lower while lighter water masses rise, which causes the water column to tend towards a stable stratified equilibrium \citep{cushman2011introduction}.
+However, this equilibrium is disrupted in the presence of velocity gradients induced by wind and water motion \citep{miles1959generation}.
 These baroclinic instabilities generate internal waves, water mixing, and eddies \citep{cushman2011introduction}.
-Eddy activity is especially strong in boundary currents such as the Gulf Stream, where strong horizontal shear occur and where shelf water and deep water masses meet on the shelf break \citep{kourafalou2012florida}.
+Eddy activity is especially strong in boundary currents such as the Gulf Stream, where a strong horizontal shear occurs and where shelf water and deep water masses meet on the shelf break \citep{kourafalou2012florida}.
 These baroclinic instabilities were shown to have a significant impact on water composition and larval recruitment in the Florida Keys \citep{lee1992influence}.
 Capturing such processes require baroclinic models that represent the variations of water density along the water column.
-However, in regions where the water column is sufficiently homogeneous and shallow that vertical contrasts are negligible, ocean circulation is accurately approximated by barotropic models.
-Such models use a depth-averaged approach by assuming an homogeneous water column with constant density and are used to model estuaries, coastal regions and shelf seas \citep{tan1992shallow}.
+However, in regions where the water column is sufficiently homogeneous and shallow so that vertical gradients are negligible, the ocean circulation can be accurately approximated by a barotropic model.
+Such a model uses a depth-averaged approach by assuming an homogeneous water column with a constant density and is used to model coastal regions and shelf seas \citep{tan1992shallow}.
 Both the baroclinic and the barotropic models usually rely on the hydrostatic equilibrium, \textit{i.e.} vertical accelerations are neglected by assuming that gravity balances the vertical component of the pressure gradient force.
@@
 In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}).
-This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics \modif{of coastal regions and estuaries such as the whole Great Barrier Reef \citep{lambrechts2008multi}, the Florida Reef Tract \citep{frys2020fine}, the Scheldt estuary \citep{de2012water}, or the Mahakam delta \citep{sassi2011tidal}}.
+This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics \modif{of coastal regions such as the whole Great Barrier Reef \citep{lambrechts2008multi}, the Florida Reef Tract \citep{frys2020fine}, the Scheldt estuary \citep{de2012water}, or the Mahakam delta \citep{sassi2011tidal}}.
 The 2D barotropic version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities.
@@
 Models are simplifications of the reality and hypotheses must therefore be made on the processes driving the evolution of the system under study.
-Such hypotheses were for instance made by favoring a fine resolution barotropic model over a coarser baroclinic model of the Florida Reef Tract.
+Such hypotheses were for instance made by favoring a fine resolution barotropic model over a coarser resolution baroclinic model of the Florida Reef Tract.
 Furthermore, we assumed that if the modeled waves and hydrodynamics accurately reproduced local observations from buoys, tide gauges and ADCP, then the modeled transport that they drive would be reasonably accurate as well.
@@
 
-\newpage
 \section*{Supporting publications}

### 2022-04-05 chapters/sctld/article.tex
modifs mise en page impression

@@
     \centering
-    \begin{tabular}{|p{3cm}|p{4cm}|p{3.5cm}|}
+    \begin{tabular}{|p{2.5cm}|p{3.5cm}|p{3.5cm}|}
         \hline
@@
 	\centering
-	\includegraphics[width=.6\textwidth]{chapters/sctld/figures/network_sir.png}
+	\includegraphics[width=.7\textwidth]{chapters/sctld/figures/network_sir.png}
 	\caption{\modif{Illustration of the developed connected SIR model.
@@
     \centering
-    \includegraphics[width=.9\textwidth]{chapters/sctld/figures/connect_paper.jpg}
+    \includegraphics[width=.99\textwidth]{chapters/sctld/figures/connect_paper.jpg}
     \caption{Distribution of the indicators derived from the monthly connectivity matrices computed for each type of current during our simulated period. Mean values are indicated by white squares.}
@@
     \centering
-    \includegraphics[width=.95\textwidth]{chapters/sctld/figures/sir_obs.jpg}
+    \includegraphics[width=.99\textwidth]{chapters/sctld/figures/sir_obs.jpg}
     \caption{Disease prevalence averaged over all monitored sites over time as modeled by Eqs.
@@
     \centering
-    \includegraphics[width=.9\textwidth]{chapters/sctld/figures/sctld_validation_corrected.jpg}
+    \includegraphics[width=.99\textwidth]{chapters/sctld/figures/sctld_validation_corrected.jpg}
     \caption{Summary of epidemiological model simulations with calibrated transmission parameters.

### 2022-04-05 include/thanks.tex
modifs mise en page impression

@@
 En outre, il a toujours réussi à accomplir le miracle de dégager du temps dans son agenda bien chargé pour me conseiller ou relire et commenter mon travail avec minutie.
+Je tiens également à remercier tous les membres de mon jury pour le temps qu'ils ont accordé à la lecture de ce manuscrit ainsi que pour leurs judicieux conseils et suggestions lors de la défense privée.
 
@@
 
-Ce travail a été financé par le Fonds de la Recherche Scientifique au moyen d’une bourse FRIA.
+Ce travail a été financé par le Fonds de la Recherche Scientifique - FNRS au moyen d’une bourse FRIA.

### 2022-04-05 include/title.tex
new title
modifs mise en page impression

@@
 \begin{minipage}[b]{0.3\textwidth}
-\hspace*{-0.8cm}\includegraphics[width=5.5cm]{./figures/logo_ucl.jpg}\vspace{0.35cm}
+\hspace*{-0.8cm}\includegraphics[width=5.5cm]{./figures/logo_ucl.jpg}\vspace{0.1cm}
 %\hspace*{3.4cm}\includegraphics[width=4.5cm]{./figures/orb.png}\vspace*{.1cm}
@@
 
-\vspace*{3cm}
+\vspace*{2cm}
 \hspace{-0.8cm}
@@
 \centering
-\LARGE\bfseries Florida's Coral Reef - Impact of ocean transport on management approaches % How modeling tools can inform coral reef management in the Florida Reef Tract
+\LARGE\bfseries Modeling local and regional dispersion of stressors in Florida's Coral Reef
 \end{minipage}
@@
 }
-
+\newline
+\vspace*{4em}
+\newline
 \textcolor{white}{Et pour la tradition, Régis Malpas \& Marc-Henry Comans}
@@
 \centering
-\large Louvain-la-Neuve, February 16, 2022
+\large Louvain-la-Neuve, April 25, 2022
 \end{minipage}

### 2022-04-05 thesis.tex
modifs mise en page impression

@@
 \hypersetup{pdfauthor={Thomas DOBBELAERE}}
-\hypersetup{pdfsubject={Florida's Coral Reef - Impact of ocean transport on management approaches}}
+\hypersetup{pdfsubject={Modeling local and regional dispersion of stressors in Florida's Coral Reef}}
 \hypersetup{pdfkeywords={Biophysical modeling, Florida Coral Reef, SCTLD, SLIM}}
@@
   \input{include/abstract}
+  \clearemptydoublepage:
+  \input{include/thanks}
   \clearemptydoublepage
-  %\input{include/thanks}
-  %\clearemptydoublepage
   \tableofcontents
@@
 \mychapter{Transmission dynamics of the disease from the Marquesas and to the Dry Tortugas}\label{chap:drto}
-\chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
+\chaptermark{Transmission of SCTLD to the DRTO}  
 \input{chapters/drto/drto}
 \mychapter{The onset of the outbreak and the PortMiami Deep Dredge Project} \label{chap:onset}
-\chaptermark{Onset of SCTLD and PortMiami Deep Dredge Project}  
+\chaptermark{The onset of SCTLD and the PMDDP}  
 \input{chapters/onset/onset}

### 2022-04-06 chapters/intro/intro.tex
version impression

@@
     \caption{Goods and services of coral reefs and their linkage with human activities.
-\modif{Despite covering less than 0.5\% of the ocean floor, coral reefs are home for almost of third of the marine biodiversity and are a source of food and livelihood to hundreds of millions of people.} Source: \cite{hughes2017coral}}
+\modif{Despite covering less than 0.5\% of the ocean floor, coral reefs are home for almost a third of the marine biodiversity and are a source of food and livelihood to hundreds of millions of people.} Source: \cite{hughes2017coral}}
     \label{intro:corals}
